---
title: "The `MRMhubExperiment` data object"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  MRMhub: Postprocessing and Quality Control of Small-Molecule Mass Spectrometry Data 
vignette: >
  %\VignetteIndexEntry{The `MRMhubExperiment` data object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

# Overview

The `MRMhubExperiment` object is the main **data container** used in the
MRMhub workflow, see also [Data and Metadata in MRMhub](01_datastructure.html).
It holds all the experimental and processed
data and metadata, as well as other details. `MRMhubExperiment` is an S4
object with following slots:

```         
MRMhubExperiment
  ├─title:  chr "My LCMS Assay"
  ├─analysis_type:  chr NA
  ├─feature_intensity_var:  chr "feature_area"
  ├─dataset_orig: tibble [400 × 26] (S3: tbl_df/tbl/data.frame)
  ├─dataset: tibble [400 × 26] (S3: tbl_df/tbl/data.frame)
  ├─dataset_filtered: tibble [0 × 14] (S3: tbl_df/tbl/data.frame)
  ├─annot_analyses: tibble [25 × 13] (S3: tbl_df/tbl/data.frame)
  ├─annot_features: tibble [16 × 16] (S3: tbl_df/tbl/data.frame)
  ├─annot_istds: tibble [8 × 4] (S3: tbl_df/tbl/data.frame)
  ├─annot_responsecurves: tibble [0 × 3] (S3: tbl_df/tbl/data.frame)
  ├─annot_qcconcentrations: tibble [32 × 5] (S3: tbl_df/tbl/data.frame)
  ├─annot_studysamples: tibble [0 × 0] (S3: tbl_df/tbl/data.frame)
  ├─annot_batches: tibble [1 × 4] (S3: tbl_df/tbl/data.frame)
  ├─metrics_qc: tibble [0 × 0] (S3: tbl_df/tbl/data.frame)
  ├─metrics_calibration: tibble [4 × 15] (S3: tbl_df/tbl/data.frame)
  ├─parameters_processing: tibble [0 × 1] (S3: tbl_df/tbl/data.frame)
  ├─status_processing:  chr "Calibration-quantitated data"
  ├─is_istd_normalized:  logi TRUE
  ├─is_quantitated:  logi TRUE
  ├─is_filtered:  logi FALSE
  ├─has_outliers_tech:  logi FALSE\
  ├─is_isotope_corr:  logi FALSE
  ├─analyses_excluded:  logi NA
  ├─features_excluded:  logi NA
  ├─var_drift_corrected:  Named logi [1:3] FALSE FALSE FALSE
  ├─var_batch_corrected:  Named logi [1:3] FALSE FALSE FALSE
```

### Creating a MRMhubExperiment object

```{r}
library(mrmhub)
myexp <- MRMhubExperiment()
```

### Using MRMhubExperiment objects

Most `MRMhub` functions take an MRMhubExperiment object as input. Data
processing functions return a modified `MRMhubExperiment`, which can be
in subsequent step.

```{r, results = TRUE, message = TRUE}
myexp <- MRMhubExperiment()
myexp <- data_load_example(myexp, 1)
myexp <- normalize_by_istd(myexp)

save_dataset_csv(myexp, "mydata.csv", "norm_intensity", FALSE)

```

R pipes can also be used, allowing to chain multiple functions together.
This can more clearly indicate the processing workflow and make the code
more easier to read.

```{r, results = TRUE, message = TRUE}
  myexp |> 
  MRMhubExperiment() |>
  data_load_example(1) |>
  normalize_by_istd() |>
  save_dataset_csv("mydata.csv", "norm_intensity", FALSE)

```

### Multiple `MRMhubExperiment` objects

Multiple `MRMhubExperiment` objects can be created and processed
independently within the same script.

```{r, results = TRUE}
m_polars <- MRMhubExperiment(title = "Polar metabolites")
m_lipids <- MRMhubExperiment(title = "Non-polar metabolites")

```

### Accessing data and metadata

Functions starting with `data_get_` allow to retrieve data and metadat
from `MRMhubExperiment` object.

```{r, results = TRUE}
myexp <- data_load_example(myexp, 1)
dataset <- get_analyticaldata(myexp, annotated = TRUE)

print(dataset)
```

Alternatively, the `$` syntax can be used access data and metadata
tables in `MRMhubExperiment` objects.

```{r, results = TRUE}
analyses <- myexp$annot_analyses
features <- myexp$annot_features
  print(features)

```

### Saving and Reading MRMhubExperiment objects

```{r, results = TRUE}
myexp <- MRMhubExperiment()
myexp <- data_load_example(myexp, 1)
saveRDS(myexp, file = "myexp-mrmhub.rds", compress = TRUE)
my_saved_exp <- readRDS(file = "myexp-mrmhub.rds")
```

### Status of an `MRMhubExperiment`

A detailed summary of the dataset and processing status can be printed
any time

```{r, results = TRUE}
print(my_saved_exp)
```
