---
title: "Isotopic interference correction"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Reference for MRMhub's LICAR-based isotopic interference correction: the
  mrm_pattern labels, the MRM and MS1 derivation levels, and the co-elution rule.
vignette: >
  %\VignetteIndexEntry{Isotopic interference correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
library(mrmhub)
```

<p class="page-kind page-kind-manual">Manual</p>

In class-based targeted assays, the natural M+2 isotopologue of a lighter species can fall in the
transition window of a species two mass units heavier and inflate its measured area. MRMhub corrects
this with the LICAR method [@gao2021LICARApplicationIsotopic]: for each affected feature it subtracts
a fixed fraction *K* of the interferer's area, where *K* is the interferer's theoretical M+2
abundance. This is a Type II correction (overlap between different species); Type I natural-abundance
(MID) correction of a compound's own isotopes is not performed.

For MRM data the factor is computed at the fragment, not the whole molecule. Whether the heavy
isotope ends up on the product ion or on the neutral loss changes *K*, so a whole-molecule (MS1)
factor does not give the right value for a fragment transition. Class-based LC-MRM therefore uses
`level = "MRM"`; `level = "MS1"` is reserved for genuine full-scan data and is not a fallback when a
product *m/z* is missing.

The [Isotopic interference correction tutorial](tutorial-11-interference-correction.html) gives the
step-by-step workflow; this page is the concept and label reference.

## The `mrm_pattern` annotation

Automatic derivation needs one hand-added column, `mrm_pattern`, in the **Features** sheet of the
metadata workbook. It names the lipid class and the product-ion type, from which
`calc_isotopic_interferences()` builds the fragment formula; the precursor and product *m/z* and the
polarity come from the imported data or the metadata. Feature names are parsed with rgoslin, so the
usual lipid shorthand is accepted; fatty-acyl (FA) and sphingoid-base (LCB) patterns additionally
need a chain-resolved name (`PC 16:0_18:1`, `Cer 18:1;O2/16:0`), while a sum-composition name
(`PC 34:1`) suffices for head-group patterns.

```{r, eval = TRUE, echo = FALSE}
knitr::kable(
  tibble::tribble(
    ~feature_id,        ~feature_class, ~precursor_mz, ~product_mz, ~polarity, ~mrm_pattern,
    "PC 34:1",          "PC",           760.6,         184.1,       "Pos",     "PC (Pos) Pro=184.1",
    "SM 34:1;O2",       "SM",           703.6,         184.1,       "Pos",     "SM (Pos) Pro=184.1",
    "Cer 18:1;O2/16:0", "Cer",          538.5,         264.3,       "Pos",     "Cer (Pos) SphB-2H2O"
  ),
  caption = "The `mrm_pattern` column in the Features metadata sheet (other feature columns omitted; values illustrative)."
)
```

On import the label is validated: an unknown label is an error, and a label whose class disagrees
with the feature name, or a sum-composition name under an FA/LCB pattern, is a warning. The
[`save_metadata_templates()`](../reference/save_metadata_templates.html) workbook offers the labels
as a filtered dropdown.

Most lipidomics assays use a handful of classes: the head-group patterns (`PC (Pos) Pro=184.1`,
`SM (Pos) Pro=184.1`, `LPC (Pos) Pro=184.1`, `PE (Pos) Pre-Pro=141`), the sphingoid-base patterns for
ceramides (`Cer (Pos) SphB-2H2O`, `Hex1Cer (Pos) SphB-2H2O`), and, in negative mode, the fatty-acyl
patterns (`PC (Neg, FA) FA`, `PE (Neg) FA`). Not every valid label is auto-derived (see the note
below the list). The complete list is at the [end of this page](#mrm-pattern-labels).

## Derivation levels

`calc_isotopic_interferences()` discovers the overlaps and stores them in `annot_interferences`; the
`level` argument selects how *K* is computed.

| | `level = "MRM"` | `level = "MS1"` |
|---|---|---|
| Intended data | Class-based LC-MRM (precursor **and** product *m/z*) | Genuine MS1 / full-scan (precursor *m/z* only) |
| Correction basis | Fragment formula (product ion or neutral loss) | Whole-molecule precursor formula |
| Pairing scope | Within an `mrm_pattern` | Within a `feature_class` |
| Applies to | Lipids (needs the class fragment chemistry) | Any compound with a formula |

At the MRM level a head-group transition carries a single overlap. A fatty-acyl or sphingoid-base
transition can carry two at once: one interferer whose extra mass sits on the retained product ion,
and one on the neutral loss. The correction subtracts both. Each overlap is one row of
`annot_interferences`, with `overlap_type` `m2_head`, `m2_front`, `m2_back`, or `ms1_m2`, and
`source` `auto` (derived) or `manual` (declared in the metadata). `calc_isotopic_interferences()`
warns when MS1 derivation is run on data whose product *m/z* differs from the precursor, i.e. real
MRM transitions.

## Co-elution

An M+2 isotopologue elutes at its own species' apex, so an *m/z* match is a real interference only
when the interferer and the victim co-elute. The correction subtracts the interferer's full area and
would otherwise remove genuine signal. The experimental `check_coelution = TRUE` keeps an edge only
when the interferer's apex falls inside the victim's integration window
(`feature_int_start`..`feature_int_end`, or `feature_rt ± FWHM` where borders are absent), and
reports how many resolved pairs it drops. It is off by default while the gate is validated.

## Provenance

The factors are theoretical isotope abundances, computed with `enviPat` 2.8
[@loos2015NontargetEnviPat]. The version is pinned to reproduce the published LICAR values, and
`calc_isotopic_interferences()` warns when a different version is installed. Derivation is
deterministic, and the derived `annot_interferences` table travels with the saved object and the
report workbook, so the correction is reproducible from the metadata alone.

## `mrm_pattern` labels {#mrm-pattern-labels}

The valid labels follow the original LICAR class list [@gao2021LICARApplicationIsotopic], grouped by
the product-ion type the transition monitors: head group, fatty acyl (FA), and sphingoid base (LCB),
which MRMhub extends with neutral-loss and reversed-phase (RPLC) patterns.
[`licar_pattern_choices()`](../reference/licar_pattern_choices.html) returns the same list from R.

```{r, eval = TRUE, echo = FALSE, results = "asis"}
origin_label <- c(
  "Head Group" = "Head group",
  "FA"         = "Fatty acyl (FA)",
  "LCB"        = "Sphingoid base (LCB)",
  "Neutral"    = "Neutral loss",
  "RPLC"       = "RPLC (cross-class)"
)
pat <- do.call(rbind, lapply(names(origin_label), function(o) {
  labs <- mrmhub::licar_pattern_choices(o)
  data.frame(
    `Product-ion type` = unname(origin_label[o]),
    `mrm_pattern` = labs,
    Polarity = ifelse(grepl("(Pos", labs, fixed = TRUE), "Pos",
               ifelse(grepl("(Neg", labs, fixed = TRUE), "Neg", "—")),
    check.names = FALSE, stringsAsFactors = FALSE
  )
}))
DT::datatable(
  pat,
  options = list(pageLength = 12, dom = "ftip"),
  rownames = FALSE,
  filter = "top",
  class = "compact stripe"
)
```

Automatic derivation covers the head-group, fatty-acyl (FA), and sphingoid-base (LCB) patterns.
Neutral-loss classes (`CE`, `DG`, `TG`) and cardiolipin FA (`CL`) can be annotated but are not
auto-derived at the MRM level: `calc_isotopic_interferences()` skips them with a warning, and a
correction for them must be declared manually with `correct_custom_interferences()`. The RPLC labels
and a few placeholder entries (`PC d9`, `MG`) are valid labels but are not offered in the template
dropdown.

## Next steps

- [Isotopic interference correction tutorial](tutorial-11-interference-correction.html): the step-by-step workflow
- [Drift and batch correction](manual-07-corrections.html): applied after interference correction
- [The MRMhubExperiment data object](manual-02-data-object.html#feature-variables): how `_orig` preserves the raw intensities
- [`calc_isotopic_interferences()`](../reference/calc_isotopic_interferences.html) · [`correct_isotopic_interferences()`](../reference/correct_isotopic_interferences.html): function reference

## References
