---
title: "Customising plots"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  How the shared appearance arguments of the MRMhub plotting functions
  control text size, legend placement, and automatic sizing, so a balanced
  figure needs little or no manual ggplot2 theming.
vignette: >
  %\VignetteIndexEntry{Customising plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      fig.width = 7, fig.height = 5, out.width = "100%")
```

<p class="page-kind page-kind-manual">Manual</p>

Every MRMhub `plot_*()` function returns a `ggplot2` object built from a shared house theme:
faint grey gridlines, minor gridlines off, a thin panel border, and dark-navy facet strips with
white labels. The theme is tuned to read well in a rendered report or a printed PDF without
further adjustment. A small, consistent set of arguments exposes the appearance choices that
otherwise required a trailing `+ theme(...)` on almost every call (text size, legend placement,
and legend sizing), so a balanced figure can usually be produced from the plotting call alone.
The functions still return plain `ggplot2` objects, so the full grammar remains available for
anything the arguments do not cover (see [With `+ theme()`](#with-theme)).

## The shared arguments

Every plotting function accepts the three common arguments. The higher-use QC plots (`plot_pca()`,
`plot_pca_loading()`, `plot_runscatter()`, `plot_rla_boxplot()`, `plot_normalization_qc()` and
`plot_qcmetrics_comparison()`) accept further arguments for finer legend and facet control.

| Argument | Applies to | Effect |
|----------|-----------|--------|
| `font_base_size` | all plots | Base font size in points; all plot text scales proportionally |
| `legend_position` | all plots | Legend placement (keyword, corner, or coordinate) |
| `legend_size` | all plots | Scales the whole legend: text, title, key and plotted symbols |
| `show_legend_title` | all plots | `FALSE` hides the legend title |
| `title` | all plots | A string sets the plot title; `NULL`/`NA` show none |
| `strip_text_size` | faceted plots | Facet strip label size |
| `strip_bg_color` | faceted plots | Facet strip background fill (strip text auto-contrasts light/dark) |
| `legend_bg_alpha` | core QC plots | Opacity of a white background box behind an inside legend |
| `aspect_ratio` | `plot_pca()` | Panel aspect ratio; default `1` (square score plot) |

Arguments left at their default (`NULL`, or the per-function default) leave that aspect of the
plot at the house setting. Setting one overrides only that property.

## Setting defaults for a whole notebook

The same appearance choices can be made once, for every subsequent plot, with
`mrmhub_set_plot_defaults()`. This is convenient when a document targets a particular medium (a
smaller base font and point size for a dense multi-panel report, say) and the choice should apply
throughout without repeating it on each call.

```{r}
mrmhub_set_plot_defaults(font_base_size = 8, point_size = 0.8)
```

The unit and resolution used when saving figures with `save_plot()` can be set the same way, so
only the width and height need repeating on each call:

```{r}
mrmhub_set_plot_defaults(units = "mm", dpi = 600)

save_plot(p, "figures/pca.png", width = 180, height = 120)
```

The resolution order for every argument is: a value passed explicitly to a plotting function wins,
then the global default set here, then the function's built-in default (including the automatic
`cols_page` sizing on faceted plots). A one-off call therefore still overrides the global setting:

```{r}
# uses the global font_base_size = 8 set above
plot_pca(mexp, variable = "norm_intensity")

# this single plot overrides it
plot_pca(mexp, variable = "norm_intensity", font_base_size = 11)
```

`mrmhub_get_plot_defaults()` reports the active settings and `mrmhub_reset_plot_defaults()` clears
them. The defaults are ordinary R options (`mrmhub.font_base_size` and so on), so they can also be
scoped to a single section with `withr::local_options()` rather than set for the whole session. The
arguments that can be set globally are `font_base_size`, `point_size`, `legend_position`,
`legend_size`, `show_legend_title`, `strip_bg_color`, and `units` and `dpi` for `save_plot()`.

Figure width and height are deliberately *not* settable globally: the physical size of a saved
figure should be readable from the call that writes it.

## Placing the legend

`legend_position` accepts the standard ggplot2 keywords, four corner shortcuts for a legend drawn
inside the panel, or a raw coordinate.

| Value | Result |
|-------|--------|
| `"right"`, `"left"`, `"top"`, `"bottom"` | Legend outside the panel on that side |
| `"none"` | No legend |
| `"inside-tr"`, `"inside-tl"`, `"inside-br"`, `"inside-bl"` | Inside the panel, anchored to that corner |
| `c(x, y)` | Inside the panel at that coordinate (`0`–`1`) |

All plots default to `"right"`. For a legend drawn inside the panel over the points, a
translucent background box (`legend_bg_alpha`) keeps it readable:

```{r}
plot_normalization_qc(mexp,
  before_norm_var = "intensity",
  after_norm_var  = "norm_intensity",
  plot_type       = "diff",
  legend_position = "inside-br",
  legend_bg_alpha = 0.6)
```

## Sizing text and symbols

`font_base_size` is the single text knob: increasing it scales axis text, titles, strip labels and
the legend together. `legend_size` scales the legend independently: its text, title, key box, and
the plotted symbol (the glyph in the key, which a plain `legend.key.size` does not resize). A value
of `3` or less is read as a multiplier of `font_base_size`; a larger value is an absolute point
size. `strip_text_size` follows the same convention for facet labels.

```{r}
plot_rt_vs_chain(mexp,
  font_base_size  = 6,
  legend_position = "right",
  legend_size     = 0.8)     # legend text and glyphs at 6 * 0.8 pt
```

The equivalent before these arguments existed required a hand-written `theme()` setting
`legend.text`, `legend.title` and `legend.key.size`, plus a `guides(override.aes = ...)` to resize
the glyph. The single `legend_size` argument replaces all four.

## Automatic sizing on paged and faceted plots

Plots that lay panels out in a grid (`plot_runscatter()`, `plot_responsecurves()`,
`plot_qcmetrics_comparison()`, `plot_normalization_qc()` and `plot_rt_vs_chain()`) size their text
and points from the number of facet columns, since column count is the main driver of panel size.
With `autoscale = TRUE` (the default), `font_base_size` and `point_size` left unset are filled from
`cols_page`:

| `cols_page` | `font_base_size` | `point_size` |
|-------------|------------------|--------------|
| 1–2 | 9 | 1.0 |
| 3 | 7 | 0.7 |
| 4–5 | 6 | 0.5 |
| 6 or more | 5 | 0.4 |

A value passed explicitly always wins, so autoscaling only fills what is left unset:

```{r}
# 3-column page: text and points sized automatically
plot_runscatter(mexp, variable = "norm_intensity", cols_page = 3)

# keep the automatic font, but set the point size by hand
plot_runscatter(mexp, variable = "norm_intensity", cols_page = 3,
                point_size = 1.2)
```

Setting `autoscale = FALSE` ignores the grid and uses the single-plot defaults
(`font_base_size = 11`, `point_size = 1.5`) for anything left unset. Single-panel plots such as
`plot_pca()` and `plot_matrixeffects()` use `font_base_size = 11` by default.

## With `+ theme()` {#with-theme}

The arguments cover the common adjustments; anything else is reachable through ordinary ggplot2
layering, appended after the plotting call. A trailing `theme()` takes precedence, so it also
serves as an override for any house-theme element.

```{r}
plot_pca(mexp, variable = "norm_intensity") +
  ggplot2::theme(panel.grid.minor = ggplot2::element_line(linewidth = 0.2)) +
  ggplot2::labs(title = "PCA of normalised intensities")
```

## Next steps

- [Visualisation functions](manual-08-visualization.html): the plotting functions by workflow stage
- [Exploring QC: RunScatter and PCA](tutorial-05-run-scatter.html): the plots in an interpretation walk-through
- [Custom QC report (recipe)](recipe-02-custom-qc-report.html): combining plots into a report
