---
title: "The MRMhubExperiment Data Object"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  How to create, use, and access data in MRMhubExperiment objects.
vignette: >
  %\VignetteIndexEntry{The MRMhubExperiment Data Object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

## Overview

The `MRMhubExperiment` object is the main data container used in the
MRMhub workflow. See also [Data and Metadata in MRMhub](manual-01-data-structure.html).
It holds all experimental and processed data and metadata. 
`MRMhubExperiment` is an S4 object with the following slots:

<details>
<summary><strong>Full slot tree (click to expand)</strong></summary>

```
MRMhubExperiment
  ├─ title:                 chr "My LCMS Assay"
  ├─ analysis_type:         chr NA
  ├─ feature_intensity_var: chr "feature_area"
  ├─ dataset_orig:          tibble [400 × 26]
  ├─ dataset:               tibble [400 × 26]
  ├─ dataset_filtered:      tibble [0 × 14]
  ├─ annot_analyses:        tibble [25 × 13]
  ├─ annot_features:        tibble [16 × 16]
  ├─ annot_istds:           tibble [8 × 4]
  ├─ annot_responsecurves:  tibble [0 × 3]
  ├─ annot_qcconcentrations: tibble [32 × 5]
  ├─ annot_studysamples:    tibble [0 × 0]
  ├─ annot_batches:         tibble [1 × 4]
  ├─ metrics_qc:            tibble [0 × 0]
  ├─ metrics_calibration:   tibble [4 × 15]
  ├─ parameters_processing: tibble [0 × 1]
  ├─ status_processing:     chr "Calibration-quantitated data"
  ├─ is_istd_normalized:    logi TRUE
  ├─ is_quantitated:        logi TRUE
  ├─ is_filtered:           logi FALSE
  ├─ has_outliers_tech:     logi FALSE
  ├─ is_isotope_corr:       logi FALSE
  ├─ analyses_excluded:     logi NA
  ├─ features_excluded:     logi NA
  ├─ var_drift_corrected:   Named logi [1:3]
  └─ var_batch_corrected:   Named logi [1:3]
```

</details>

## Creating a MRMhubExperiment Object

```{r}
library(mrmhub)
myexp <- MRMhubExperiment()
```

## Using MRMhubExperiment Objects

Most MRMhub functions take an `MRMhubExperiment` as input. Data processing 
functions return a modified copy, which can be used in subsequent steps.

```{r, results = TRUE, message = TRUE}
myexp <- MRMhubExperiment()
myexp <- data_load_example(myexp, 1)
myexp <- normalize_by_istd(myexp)

save_dataset_csv(myexp, tempfile(fileext = ".csv"), "norm_intensity", FALSE)
```

R pipes allow chaining multiple functions together, making the processing 
workflow easier to read:

```{r, eval = FALSE}
myexp <- MRMhubExperiment() |>
  data_load_example(1) |>
  normalize_by_istd()
```

## Multiple MRMhubExperiment Objects

You can create and process multiple objects independently within the same script:

```{r, results = TRUE}
m_polars <- MRMhubExperiment(title = "Polar metabolites")
m_lipids <- MRMhubExperiment(title = "Non-polar metabolites")
```

## Accessing Data and Metadata

Functions starting with `get_` retrieve data from the object:

```{r, eval = FALSE}
myexp <- data_load_example(MRMhubExperiment(), 1)
dataset <- get_analyticaldata(myexp, annotated = TRUE)
dataset
```

Alternatively, use `$` syntax to access metadata tables directly:

```{r, eval = FALSE}
analyses <- myexp$annot_analyses
features <- myexp$annot_features
features
```

## Saving and Loading

```{r, eval = FALSE}
saveRDS(myexp, file = "myexp-mrmhub.rds", compress = TRUE)
my_saved_exp <- readRDS(file = "myexp-mrmhub.rds")
```

## Checking Processing Status

View the current dataset and processing status at any time:

```{r, eval = FALSE}
myexp
```

## See Also

- [Data and Metadata](manual-01-data-structure.html) — table overview
- [Data Identifiers](manual-02-data-identifiers.html) — ID columns explained
- [Feature Variables](manual-03-feature-variables.html) — measurement columns
- [Key Concepts](manual-00-key-concepts.html) — terminology
