---
title: "Data file structures"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Reference for the file layout and columns expected by each MRMhub data importer.
vignette: >
  %\VignetteIndexEntry{Data file structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)

# Compact "screenshot" of the first rows of a bundled example file.
show_head <- function(file, delim = ",", n = 3, cols = NULL) {
  path <- system.file("extdata", file, package = "mrmhub")
  df <- readr::read_delim(
    path,
    delim = delim,
    n_max = n,
    show_col_types = FALSE,
    name_repair = "minimal"
  )
  total <- ncol(df)
  if (!is.null(cols)) df <- df[, cols, drop = FALSE]
  cap <- if (!is.null(cols) && length(cols) < total) {
    sprintf("First %d rows of `%s` (showing %d of %d columns)", n, file, length(cols), total)
  } else {
    sprintf("First %d rows of `%s`", n, file)
  }
  knitr::kable(df, caption = cap)
}
```

This page shows the **expected layout** of each data file and what its columns mean, so that
the input can be recognised and prepared before importing. Pick the importer that matches the
software that produced the peak table; see
[Import and prepare data files](manual-05a-which-importer.html) for the decision table.

The previews below are the actual first rows of the example files bundled with the package
(`system.file("extdata", ..., package = "mrmhub")`).

Files fall into two groups:

- **Vendor / tool exports** (MRMhub/INTEGRATOR, MassHunter, Skyline) — the columns are produced
  by the software and are not edited manually. Confirm that the correct report was exported.
- **Generic CSV** (wide / long) — these are prepared manually, so the full column reference is given.

## MRMhub / INTEGRATOR — long CSV/TSV {#str-mrmhub}

INTEGRATOR's long output: one row per analysis × feature. Tab-separated (`.tsv`) or
comma-separated (`.csv`). This is the preferred format; import with
[`import_data_mrmhub()`](../reference/import_data_mrmhub.html).

```{r}
show_head(
  "MRMhub_demo.tsv",
  delim = "\t",
  cols = c("feature_name", "raw_data_filename", "sample_type", "batch", "rt_apex", "area")
)
```

The file also carries `internal_standard`, `time_stamp`, transition columns (`precursor_mz`,
`product_mz`, `collision_energy`, `polarity`), and integration columns (`height`, `FWHM`,
`rt_int_start`, `rt_int_end`). This file is not prepared by hand; it is produced by INTEGRATOR.
See [Data Import (detailed)](manual-05-data-import.html).

## Agilent MassHunter Quant — wide CSV {#str-masshunter}

A MassHunter Quant export is **wide with a two-row header**. The first header row groups
columns by compound (`<Compound> Method` and `<Compound> Results` blocks) after a leading
`Sample` block; the second row holds the actual sub-headers. Import with
[`import_data_masshunter()`](../reference/import_data_masshunter.html).

<table class="table" style="font-size: 0.85em;">
<thead>
<tr>
  <th colspan="4" style="text-align:center; border-bottom:2px solid #5B8FA8;">Sample</th>
  <th colspan="4" style="text-align:center; border-bottom:2px solid #D4914E;">CE 18:1 — Method</th>
  <th colspan="3" style="text-align:center; border-bottom:2px solid #6B9E5E;">CE 18:1 — Results</th>
  <th rowspan="2" style="vertical-align:bottom;">…</th>
</tr>
<tr>
  <th>Data File</th><th>Name</th><th>Type</th><th>Acq. Date-Time</th>
  <th>Precursor Ion</th><th>Product Ion</th><th>Collision Energy</th><th>RT</th>
  <th>RT</th><th>Area</th><th>FWHM</th>
</tr>
</thead>
<tbody>
<tr>
  <td>001_…01.d</td><td>001_…01</td><td>Sample</td><td>4/12/18 18:28</td>
  <td>668.6</td><td>369.3</td><td>10</td><td>7.16</td>
  <td>7.16</td><td>5152996</td><td>0.081</td>
</tr>
<tr>
  <td>002_…02.d</td><td>002_…02</td><td>Sample</td><td>4/12/18 18:39</td>
  <td>668.6</td><td>369.3</td><td>10</td><td>7.16</td>
  <td>7.16</td><td>4789505</td><td>0.080</td>
</tr>
</tbody>
</table>

The `Method`/`Results` blocks repeat for every compound (quantifiers and qualifiers). MassHunter
produces this layout; it is not edited manually. See
[Data Import (detailed)](manual-05-data-import.html) for handling qualifier transitions and
choosing the concentration column.

## Skyline (small molecule) — long CSV {#str-skyline}

Skyline's *Molecule Transition Results* export: one row per transition result. Import with
[`import_data_skyline()`](../reference/import_data_skyline.html).

```{r}
show_head(
  "Skyline_MoleculeTransitionResults.csv",
  cols = c("Molecule Name", "Replicate Name", "Precursor Mz", "Product Mz", "Retention Time", "Area")
)
```

`Replicate Name` becomes `analysis_id`, `Molecule Name` becomes `feature_id`, and transitions are
identified by precursor/product name or *m/z*. Export this report from Skyline; see
[Data Import (detailed)](manual-05-data-import.html).

## Generic spreadsheet (wide CSV) {#str-csv-wide}

One row per sample, one column per feature. Import with
[`import_data_csv_wide()`](../reference/import_data_csv_wide.html), passing `variable_name` to
declare what the values represent.

```{r}
show_head("plain_wide_dataset.csv")
```

| Column | Required | Description |
|---|---|---|
| `analysis_id` | Yes | Unique sample/injection identifier (one row per sample). If absent, the first column is used when its values are unique. |
| *feature columns* | Yes (≥1) | One column per feature; the **column name is the `feature_id`** and the values are of the type given by `variable_name`. |
| `qc_type` | Optional | Sample/QC role (e.g. `SPL`, `TQC`, `BQC`, `SBLK`). Imported when `import_metadata = TRUE`. |
| `batch_id` | Optional | Batch identifier. |
| `is_quantifier` | Optional | Whether the feature is a quantifier. |
| `is_istd` | Optional | Whether the feature is an internal standard. |
| `analysis_order` | Optional | Run order of the analyses. |

**Key arguments:** `variable_name` (required) declares what the values represent — one of
`"area"`, `"height"`, `"intensity"`, `"norm_intensity"`, `"conc"`, or `"response"`. Use
`first_feature_column` to mark where the feature columns begin when extra metadata columns are
present, and `analysis_id_col` to point at a non-standard ID column.

## Generic table (long CSV) {#str-csv-long}

One row per `(sample, feature)` measurement. Import with
[`import_data_csv_long()`](../reference/import_data_csv_long.html). Columns are auto-detected by
name (case-insensitive); use `column_mapping` for non-matching headers.

```{r}
show_head(
  "plain_long_dataset.csv",
  cols = c("feature_id", "raw_data_filename", "sample_type", "batch", "rt_apex", "area")
)
```

(The bundled example uses INTEGRATOR-style headers such as `raw_data_filename` and `rt_apex`, so
it is imported with a `column_mapping`; see the example in
[`import_data_csv_long()`](../reference/import_data_csv_long.html).)

Auto-detected columns:

| Column | Required | Description |
|---|---|---|
| `analysis_id` | Yes | Sample/injection identifier (or supply `raw_data_filename` and map it). |
| `feature_id` | Yes | Feature/compound identifier. |
| `area` / `height` / `intensity` | Yes (≥1) | A feature value column; at least one is needed. |
| `rt` | Optional | Retention time. |
| `fwhm` / `width` | Optional | Peak width metrics. |
| `qc_type` | Optional | Sample/QC role. |
| `sample_id` | Optional | Sample identifier (groups replicates). |
| `batch_id` | Optional | Batch identifier. |
| `istd_feature_id` | Optional | Internal standard used for the feature. |
| `feature_class` | Optional | Feature class/group. |
| `analyte_id` | Optional | Standardised analyte identifier. |
| `precursor_mz` / `product_mz` | Optional | Transition *m/z* values. |

Columns already named with the internal `feature_`/`method_` prefixes (e.g. `feature_area`) are
detected automatically too.

## Next Steps

- [Import and prepare data files](manual-05a-which-importer.html) — the importer decision table
- [Metadata table structures](manual-06a-metadata-table-structures.html) — annotation table columns
- [Data Import (detailed)](manual-05-data-import.html) — full arguments and troubleshooting
