---
title: "Visualisation functions"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  Reference for the MRMhub plotting functions, grouped by workflow stage,
  with the canonical argument forms and customisation guidance.
vignette: >
  %\VignetteIndexEntry{Visualisation functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      fig.width = 7, fig.height = 5, out.width = "100%")
```

<p class="page-kind page-kind-manual">Manual</p>

MRMhub provides plotting functions covering every stage of the workflow. All functions return
`ggplot2` objects (or, for paged outputs, a list of `ggplot2` objects), so the standard ggplot2
grammar can be used to further customise titles, themes, scales, and facets. The functions below are
grouped by the workflow stage at which they are most useful.

## Setup

```{r}
#| results: false
#| message: false
library(mrmhub)
library(ggplot2)

mexp <- readRDS("results/mexp_processed.rds")
```

## Overview

| Function | Stage | Purpose |
|----------|-------|---------|
| `plot_runsequence()` | Acquisition design | Run-order overview of QC types and batches |
| `plot_runscatter()` | Drift / batch QC | Per-feature scatter of values vs analysis order |
| `plot_abundanceprofile()` | Abundance overview | Feature abundance distribution per class |
| `plot_rla_boxplot()` | Normalisation QC | Relative log-abundance boxplots per analysis |
| `plot_normalization_qc()` | Normalisation QC | Before/after ISTD normalisation comparison |
| `plot_pca()` | Multivariate QC | PCA score plot |
| `plot_pca_loading()` | Multivariate QC | PCA loadings |
| `plot_feature_correlations()` | Multivariate QC | Feature-feature correlation heatmap |
| `plot_qcmetrics_comparison()` | QC metrics | CV / bias comparison across QC types |
| `plot_qc_summary_byclass()` | QC filtering | Pass/fail summary by feature class |
| `plot_qc_summary_overall()` | QC filtering | Pass/fail summary across the whole dataset |
| `plot_calibrationcurves()` | External calibration | Calibration curves with fit and residuals |
| `plot_responsecurves()` | Response curves | Linearity check from a dilution series |
| `plot_rt_vs_chain()` | Method check (lipidomics) | RT vs chain length / unsaturation |
| `plot_matrixeffects()` | Method check | Matrix-effect QC |
| `plot_interference_correction()` | Interference correction | QC overview for interference annotations |

# Acquisition design

## `plot_runsequence()`: sequence overview

The run-sequence plot summarises the acquisition design: QC type positions, batch boundaries, and
the analysis timeline. It is an experiment-level plot (no per-feature dimension).

```{r}
plot_runsequence(mexp,
                 qc_types = NA,
                 show_batches = TRUE,
                 batch_zebra_stripe = TRUE,
                 show_timestamp = FALSE)
```

Set `show_timestamp = TRUE` to use the acquisition timestamp on the x-axis; this helps
identify interruptions between or within batches that would not be visible against the analysis
sequence number.

# Drift, batch, and per-feature inspection

## `plot_runscatter()`: values vs analysis order

The primary plot for drift and batch QC. Plots a feature variable (intensity, normalised
intensity, concentration) against the analysis order, optionally with fitted drift trends from the
correction functions. Returns one panel per feature (paged).

```{r}
plot_runscatter(mexp,
                variable = "norm_intensity",
                qc_types = c("BQC", "SPL"),
                rows_page = 1,
                cols_page = 3,
                show_trend = TRUE)
```

Use `variable = "intensity_before"` / `"conc_before"` to inspect pre-correction values, and
`variable = "intensity"` / `"conc"` for post-correction. Setting `output_pdf = TRUE` and a `path`
writes a multi-page PDF, which is practical for large feature lists.

## `plot_abundanceprofile()`: feature abundance distribution

Shows the abundance distribution of each feature, grouped by class. Useful for inspecting class
coverage and identifying features at the limits of the dynamic range.

```{r}
plot_abundanceprofile(mexp,
                      variable = "intensity",
                      qc_types = "SPL",
                      log_scale = TRUE)
```

When `use_qc_metrics = TRUE`, the function reads from the pre-computed `metrics_qc` table (much
faster on large feature lists) and `qc_types` must specify a single QC type.

# Normalisation QC

## `plot_rla_boxplot()`: relative log-abundance per analysis

A boxplot per analysis of `log2(value / median_across_analyses)`. Width and centring of the box
reflect injection-level variability. Useful before and after normalisation to confirm that ISTD
correction removes injection-level bias.

```{r}
plot_rla_boxplot(mexp,
                 variable = "norm_intensity",
                 qc_types = c("BQC", "SPL"))
```

## `plot_normalization_qc()`: before-vs-after comparison

Compares pre- and post-normalisation values for QC samples in three layouts (`plot_type`):
`"scatter"` (point cloud), `"diff"` (signed difference), `"ratio"` (after / before). CV of QC
samples should decrease after normalisation; an increase typically indicates an incorrect ISTD
pairing.

```{r}
plot_normalization_qc(mexp,
                      before_norm_var = "intensity",
                      after_norm_var  = "norm_intensity",
                      plot_type       = "scatter",
                      qc_types        = c("BQC", "SPL"),
                      facet_by_class  = TRUE,
                      cv_threshold_value = 25)
```

# Multivariate QC

## `plot_pca()`: score plot

Two-dimensional PCA score plot with confidence ellipses grouped by `qc_type`, `batch_id`, or
`"none"`. See [QC exploration with PCA](tutorial-05-run-scatter.html) for an interpretation
walk-through.

```{r}
plot_pca(mexp,
         variable          = "norm_intensity",
         qc_types          = c("BQC", "SPL"),
         ellipse_variable  = "qc_type")
```

## `plot_pca_loading()`: loadings

Loadings on the first two PCs. Features at the extremes drive sample separation; a single feature
dominating PC1 should be inspected with `plot_runscatter()` before being attributed to biology.

```{r}
plot_pca_loading(mexp,
                 variable = "norm_intensity",
                 qc_types = c("BQC", "SPL"))
```

## `plot_feature_correlations()`: correlation heatmap

Pairwise correlation matrix across features, useful for identifying redundant transitions and
flagging candidate isobaric interferences.

```{r}
plot_feature_correlations(mexp,
                          variable = "norm_intensity",
                          qc_types = "SPL")
```

# QC metrics and filtering

## `plot_qcmetrics_comparison()`: pairwise QC-metric scatter

Plots one QC metric against another, computed across features. The variable names follow the
pattern `{variable}_cv_{qctype}` (e.g. `intensity_cv_bqc`, `norm_intensity_cv_tqc`) and live in
`mexp@metrics_qc` after `calc_qc_metrics()` is run.

```{r}
mexp <- calc_qc_metrics(mexp)

plot_qcmetrics_comparison(mexp,
                          plot_type        = "scatter",
                          x_variable       = "intensity_cv_bqc",
                          y_variable       = "norm_intensity_cv_bqc",
                          equality_line    = TRUE,
                          threshold_values = 25,
                          log_scale        = FALSE)
```

Points below the equality line indicate features for which CV was reduced by normalisation; points
above indicate features made worse, typically a misassigned ISTD.

## `plot_qc_summary_byclass()` / `plot_qc_summary_overall()`: filter outcome

After `filter_features_qc()`, these two functions summarise how many features passed each filter
rule, broken down by feature class (`_byclass`) or aggregated (`_overall`).

```{r}
plot_qc_summary_overall(mexp)
plot_qc_summary_byclass(mexp)
```

# External calibration and response curves

## `plot_calibrationcurves()`: calibration fits

After `quantify_by_calibration()`, plots calibration response vs concentration with the fitted
model, the calibration points, and points excluded from the fit. Returns a paged grid of features.

```{r}
plot_calibrationcurves(mexp,
                       variable = "norm_intensity",
                       qc_types = NA)
```

The fit model is taken from the calibration setup unless overridden with `fit_overwrite =
"linear"` or `"quadratic"`.

## `plot_responsecurves()`: RQC linearity

Plots feature response across an RQC dilution series. The fitted slope and R² values report on the
linearity of the assay in the range of the QC pool. See `get_response_curve_stats()` for the
numeric summary.

```{r}
plot_responsecurves(mexp,
                    variable = "intensity")
```

# Method-specific checks

## `plot_rt_vs_chain()`: retention time vs chain length (lipidomics)

For lipidomics methods using class-based chromatography, plots RT against carbon number, separated
by class. Outliers from the expected linear relationship within a class point to mis-identified
features.

```{r}
plot_rt_vs_chain(mexp)
```

## `plot_matrixeffects()`: matrix-effect overview

Compares ISTD response in matrix-containing QCs against solvent-only injections to flag
matrix-effect outliers.

```{r}
plot_matrixeffects(mexp)
```

## `plot_interference_correction()`: interference annotations

QC overview for features with interference annotations (see
[Interference correction](tutorial-11-interference-correction.html)).

```{r}
plot_interference_correction(mexp)
```

# Customisation and export

Text size, legend placement and legend sizing are controlled directly through shared arguments
(`font_base_size`, `legend_position`, `legend_size`, and, on the core QC plots, `strip_text_size`,
`show_legend_title`, `legend_bg_alpha`); paged and faceted plots size their text and points
automatically from `cols_page`. See [Customising plots](#customising-plots) for the
full argument reference.

## Customising plots

Every `plot_*()` function returns a `ggplot2` object built from a shared house theme (faint grey
gridlines, a thin panel border, dark-navy facet strips). A small set of arguments exposes the common
appearance choices, so a balanced figure usually comes from the plotting call alone, without a
trailing `+ theme(...)`.

### The shared arguments

Every plotting function accepts the three common arguments. The higher-use QC plots (`plot_pca()`,
`plot_pca_loading()`, `plot_runscatter()`, `plot_rla_boxplot()`, `plot_normalization_qc()` and
`plot_qcmetrics_comparison()`) accept further arguments for finer legend and facet control.

| Argument | Applies to | Effect |
|----------|-----------|--------|
| `font_base_size` | all plots | Base font size in points; all plot text scales proportionally |
| `legend_position` | all plots | Legend placement (keyword, corner, or coordinate) |
| `legend_size` | all plots | Scales the whole legend: text, title, key and plotted symbols |
| `show_legend_title` | all plots | `FALSE` hides the legend title |
| `title` | all plots | A string sets the plot title; `NULL`/`NA` show none |
| `strip_text_size` | faceted plots | Facet strip label size |
| `strip_bg_color` | faceted plots | Facet strip background fill (strip text auto-contrasts light/dark) |
| `legend_bg_alpha` | core QC plots | Opacity of a white background box behind an inside legend |
| `aspect_ratio` | `plot_pca()` | Panel aspect ratio; default `1` (square score plot) |

Arguments left at their default (`NULL`, or the per-function default) leave that aspect of the plot
at the house setting. Setting one overrides only that property.

### Setting defaults for a whole notebook

The same appearance choices can be made once, for every subsequent plot, with
`mrmhub_set_plot_defaults()`, convenient when a document targets a particular medium (a smaller base
font and point size for a dense multi-panel report, say) and the choice should apply throughout
without repeating it on each call.

```{r}
mrmhub_set_plot_defaults(font_base_size = 8, point_size = 0.8)
```

The resolution order for every argument is: a value passed explicitly to a plotting function wins,
then the global default set here, then the function's built-in default (including the automatic
`cols_page` sizing on faceted plots). A one-off call therefore still overrides the global setting:

```{r}
# uses the global font_base_size = 8 set above
plot_pca(mexp, variable = "norm_intensity")

# this single plot overrides it
plot_pca(mexp, variable = "norm_intensity", font_base_size = 11)
```

`mrmhub_get_plot_defaults()` reports the active settings and `mrmhub_reset_plot_defaults()` clears
them. The defaults are ordinary R options (`mrmhub.font_base_size` and so on), so they can also be
scoped to a single section with `withr::local_options()` rather than set for the whole session. The
arguments that can be set globally are `font_base_size`, `point_size`, `legend_position`,
`legend_size`, `show_legend_title`, `strip_bg_color`, and the `units` and `dpi` used by `save_plot()`
(see [Setting the unit and resolution once](#setting-the-unit-and-resolution-once)). Figure width and
height are deliberately not settable globally: the physical size of a saved figure should be readable
from the call that writes it.

### Placing the legend

`legend_position` accepts the standard ggplot2 keywords, four corner shortcuts for a legend drawn
inside the panel, or a raw coordinate.

| Value | Result |
|-------|--------|
| `"right"`, `"left"`, `"top"`, `"bottom"` | Legend outside the panel on that side |
| `"none"` | No legend |
| `"inside-tr"`, `"inside-tl"`, `"inside-br"`, `"inside-bl"` | Inside the panel, anchored to that corner |
| `c(x, y)` | Inside the panel at that coordinate (`0`–`1`) |

All plots default to `"right"`. For a legend drawn inside the panel over the points, a translucent
background box (`legend_bg_alpha`) keeps it readable:

```{r}
plot_normalization_qc(mexp,
  before_norm_var = "intensity",
  after_norm_var  = "norm_intensity",
  plot_type       = "diff",
  legend_position = "inside-br",
  legend_bg_alpha = 0.6)
```

### Sizing text and symbols

`font_base_size` is the single text knob: increasing it scales axis text, titles, strip labels and
the legend together. `legend_size` scales the legend independently: its text, title, key box, and the
plotted symbol (the glyph in the key, which a plain `legend.key.size` does not resize). A value of `3`
or less is read as a multiplier of `font_base_size`; a larger value is an absolute point size.
`strip_text_size` follows the same convention for facet labels.

```{r}
plot_rt_vs_chain(mexp,
  font_base_size  = 6,
  legend_position = "right",
  legend_size     = 0.8)     # legend text and glyphs at 6 * 0.8 pt
```

### Automatic sizing on paged and faceted plots

Plots that lay panels out in a grid (`plot_runscatter()`, `plot_responsecurves()`,
`plot_qcmetrics_comparison()`, `plot_normalization_qc()` and `plot_rt_vs_chain()`) size their text and
points from the number of facet columns, since column count is the main driver of panel size. With
`autoscale = TRUE` (the default), `font_base_size` and `point_size` left unset are filled from
`cols_page`:

| `cols_page` | `font_base_size` | `point_size` |
|-------------|------------------|--------------|
| 1–2 | 9 | 1.0 |
| 3 | 7 | 0.7 |
| 4–5 | 6 | 0.5 |
| 6 or more | 5 | 0.4 |

A value passed explicitly always wins, so autoscaling only fills what is left unset:

```{r}
# 3-column page: text and points sized automatically
plot_runscatter(mexp, variable = "norm_intensity", cols_page = 3)

# keep the automatic font, but set the point size by hand
plot_runscatter(mexp, variable = "norm_intensity", cols_page = 3,
                point_size = 1.2)
```

Setting `autoscale = FALSE` ignores the grid and uses the single-plot defaults (`font_base_size =
11`, `point_size = 1.5`) for anything left unset. Single-panel plots such as `plot_pca()` and
`plot_matrixeffects()` use `font_base_size = 11` by default.

## ggplot2 layering

All functions return `ggplot2` objects, so anything the arguments above do not cover (bespoke
themes, scales, titles) can be appended in the usual way, and a trailing `theme()` overrides the
house theme.

```{r}
plot_runscatter(mexp, variable = "norm_intensity", qc_types = c("BQC", "SPL")) +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::labs(title = "Normalised intensity (BQC vs SPL)")
```

## Saving plots

`save_plot()` writes any plot to a file at a defined physical size and resolution. Sizes are given
in `mm` by default, matching how journals specify figure widths, and `cm`, `in`, `pt` and `px` are
also accepted.

```{r}
p <- plot_pca(mexp,
              variable = "norm_intensity",
              qc_types = c("BQC", "SPL"),
              ellipse_variable = "batch_id")

save_plot(p, "figures/pca_batch.pdf", width = 180, height = 120)
```

The same figure can be written in several formats in one call. Give a path without an extension and
list the formats; each file gets the matching extension.

```{r}
save_plot(p, "figures/pca_batch", format = c("pdf", "png"), width = 180, height = 120)
```

The plot is returned visibly, so piping into `save_plot()` still renders the figure in the notebook
and saving needs no separate line:

```{r}
plot_pca(mexp, variable = "norm_intensity", ellipse_variable = "batch_id") |>
  save_plot("figures/pca_batch.pdf", width = 180, height = 120)
```

In a script or a loop, where re-drawing a dense figure is wasted work, `show_plot = FALSE` skips it
and returns the written paths instead:

```{r}
paths <- save_plot(p, "figures/pca_batch.pdf", width = 180, height = 120,
                   show_plot = FALSE)
```

(With `show_plot = TRUE` the paths are still available, as the `"paths"` attribute of the returned
plot.)

`save_plot()` also accepts what the plot functions return around a plot: the result list of
`plot_rla_boxplot()` (it picks up the `plot` element), a `patchwork` composition, and a list of
plots, which becomes a multi-page PDF.

```{r}
pages <- plot_runscatter(mexp, variable = "norm_intensity", return_plots = TRUE)
save_plot(pages, "figures/runscatter_all.pdf", width = 280, height = 200,
          show_plot = FALSE)
```

### Choosing a format

| Purpose | Format | Device used | Typical `dpi` |
|---|---|---|---|
| Journal figure, vector (default choice) | `"pdf"` | `grDevices::cairo_pdf`, else `grDevices::pdf` | n/a |
| Figure for further editing (Illustrator, Inkscape) | `"svg"` | `svglite::svglite`, else `grDevices::svg` | n/a |
| Slides, Quarto HTML, GitHub | `"png"` | `ragg::agg_png`, else `grDevices::png` | 150--300 |
| Journal requiring raster submission | `"tiff"` | `ragg::agg_tiff`, else `grDevices::tiff` | 300--600 |

Prefer a **vector** format for publication: text stays selectable and searchable, and lines stay
sharp at any magnification.

Prefer a **raster** format when a plot draws very many marks --- a `plot_runscatter()` page covering
several thousand analyses, or a dense `plot_pca()` score plot. Every point becomes a separate object
in a PDF, so such figures produce very large files that are slow to open and to typeset. Saving them
at 300--600 dpi instead keeps the file small with no visible loss.

::: {.callout .callout-tip}
The optional packages `ragg` and `svglite` are used automatically when installed, giving better text
rendering, system-font support and smaller SVG files. Without them the equivalent `grDevices` device
is used and the output is still correct. Installing both is recommended:
`install.packages(c("ragg", "svglite"))`.

PDF output uses the cairo device wherever R was built with cairo support, because plain
`grDevices::pdf()` writes text in a single-byte encoding and silently transliterates anything
outside it --- an en dash becomes `-`, and `≥` becomes `>=`. Unit labels such as `µmol/L` and
statistical annotations depend on those glyphs surviving.
:::

### Setting the unit and resolution once

`units` and `dpi` can be set for a whole notebook through `mrmhub_set_plot_defaults()`, so only
`width` and `height` need repeating. The size itself is always explicit at the call site.

```{r}
mrmhub_set_plot_defaults(units = "mm", dpi = 600)

save_plot(p, "figures/pca_batch.png", width = 180, height = 120)   # 600 dpi
```

### Multi-page PDFs

For paged outputs (`plot_runscatter()`, `plot_calibrationcurves()`, `plot_responsecurves()`,
`plot_feature_correlations()`), the built-in `output_pdf = TRUE` and `path` arguments write a
multi-page PDF directly, rather than iterating in user code. `page_width` and `page_height` set the
page size; without them an A4 page is used, oriented by `page_orientation`.

```{r}
plot_runscatter(mexp,
                variable    = "norm_intensity",
                qc_types    = c("BQC", "SPL"),
                output_pdf  = TRUE,
                path        = "figures/runscatter_all.pdf",
                page_width  = 297,
                page_height = 210)
```

## Combining plots

`patchwork` composes multiple panels into a single figure:

```{r}
library(patchwork)

p_seq  <- plot_runsequence(mexp)
p_pca  <- plot_pca(mexp, variable = "norm_intensity",
                   qc_types = c("BQC", "SPL"), ellipse_variable = "batch_id")
p_rla  <- plot_rla_boxplot(mexp, variable = "norm_intensity",
                           qc_types = c("BQC", "SPL"))

p_seq / (p_pca | p_rla)
```

## Next steps

- [Exploring QC: RunScatter and PCA](tutorial-05-run-scatter.html): `plot_runscatter()`, `plot_pca()` and outlier screening
- [Drift and batch correction (tutorial)](tutorial-04-drift-correction.html): `plot_runscatter()` in context
- [External calibration & QC](tutorial-06-external-calibration.html): calibration plots in workflow
