---
title: "Metadata import"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  MRMhub: postprocessing and quality control of small molecule mass spectrometry data 
vignette: >
  %\VignetteIndexEntry{Metadata import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
library(mrmhub)

if(fs::file_exists("metadata_template.xlsx")) fs::file_delete("metadata_template.xlsx")
if(fs::file_exists("metadata_msorganiser_template.xlsx")) fs::file_delete("metadata_msorganiser_template.xlsx")
```

### Metadata

Metadata in this context refers to analysis metadata, i.e., data that
annotate the analytical data. Metadata can be retrieved from the
imported analysis data file where available, or imported from
separate files or R data frames.

Integrity of metadata and data is essential for correct post-processing.
MRMhub therefore inspects imported data and metadata for completeness and
for consistency of the IDs used across the different metadata tables.
After metadata import, a summary of the identified errors, warnings, and
notes is printed to the console so that possible issues can be identified
and addressed.

### Metadata formats and templates

The structure and the required and optional columns for each metadata type
are described in the Manual and in the help pages of the corresponding
import functions.

To obtain metadata templates, an Excel file containing all metadata table
templates can be saved.

```{r}
mrmhub::save_metadata_templates()
```

### Importing metadata from files/sheets

The analysis data are first imported as outlined in "Importing data". In
this case, metadata present in the analysis data is not imported; only the
peak areas are read.

```{r}
library(mrmhub)
mexp <- mrmhub::MRMhubExperiment()

data_path <- "datasets/sPerfect_MRMhub.tsv"
mexp <- import_data_mrmhub(data = mexp, path = data_path, import_metadata = FALSE)
```

The corresponding metadata can then be added file by file.

```{r}
mexp <- import_metadata_analyses(mexp, 
                                 path = "datasets/analysis_metadata.csv", 
                                 excl_unmatched_analyses = TRUE, 
                                 ignore_warnings = TRUE)

mexp <- import_metadata_features(mexp, 
                                 path = "datasets/feature_metadata.csv",
                                 ignore_warnings= TRUE )
```

Metadata can also be imported from sheets of an Excel workbook, which
allows all metadata to be stored in one file. \
Here, metadata on internal standards and response curves is added to the
MRMhubExperiment object

```{r}
mexp <- import_metadata_istds(mexp, 
                              path = "datasets/metadata_tables.xlsx", 
                              sheet = "ISTDs",
                              ignore_warnings= TRUE)
```

Furthermore, metadata can be imported from R `data.frame` objects, which
allows metadata to be obtained from additional sources, e.g. databases or
a LIMS.

```{r}
df_qcinfo <- readr::read_table(file = "datasets/qc_metadata.txt")
mexp <- import_metadata_qcconcentrations(mexp, table = df_qcinfo)
```

### Importing MSOrganiser metadata file template

Another option to import metadata is via the MSOrganiser template file,
an Excel file (`.xlsx`). This template provides tables for all
metadata types supported by MRMhub, with options to check the validity
and integrity of the metadata.

The template can be obtained from <https://github.com/SLINGhub/mrmhub> or
via a mrmhub function

```{r}
mrmhub::save_metadata_msorganiser_template()
```

Only the metadata tables required by the intended processing workflow
need to be completed. The following import function imports all completed
tables.

```{r}
mexp <- import_metadata_msorganiser(mexp, 
                                    path = "datasets/sPerfect_Metadata.xlsx",
                                    ignore_warnings= TRUE)
```

## Next Steps

- [Import and prepare data files](manual-05a-which-importer.html) — the importer decision table
- [Metadata table structures](manual-06a-metadata-table-structures.html) — per-table columns and templates
- [Data Import](manual-05-data-import.html) — importing the measurement data
