---
title: "Design Decisions: Why This Architecture?"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Explains the key design decisions behind MRMhub QUANT -- for power users 
  and contributors who want to understand or extend the package.
vignette: >
  %\VignetteIndexEntry{Design Decisions: Why This Architecture?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="page-kind page-kind-manual">Manual</p>

## Audience

This article is for:

- Power users who want to understand *why* QUANT works the way it does
- Contributors who want to extend the package
- Developers building tools on top of MRMhub

To start processing data directly, see [Your First Analysis](tutorial-00-first-analysis.html) instead.

## Decision 1: S4 Class (MRMhubExperiment) vs Tidy Tibbles

**Choice:** A single S4 object holds all data, metadata, and processing state.

<details>
<summary><strong>Alternatives considered</strong></summary>

| Alternative | Why not |
|-------------|---------|
| Pure tibble workflow (tidyverse) | No built-in integrity checks; too easy to desync data and metadata |
| SummarizedExperiment (Bioconductor) | Models a feature × sample matrix, not a processing pipeline: no *typed* home for the original data, the processing state, or the non-rectangular annotation tables. Supported as an **export target** instead |
| R6 class | Less formal than S4; no validity checking; mutable-by-reference is error-prone for data analysis |

</details>

**Rationale:**

1. **Traceability** — The S4 object bundles original data, processed data,
   annotations, and processing flags in one place.

2. **Integrity checking** — S4 validity methods ensure data and metadata stay
   in sync. Analyses cannot be accidentally deleted without updating annotations.

3. **Pipeline state** — Flags like `is_istd_normalized`, `var_drift_corrected`
   prevent accidental double-processing.

4. **Not SummarizedExperiment — but exports to it.** SE models a feature × sample
   matrix. It offers no *typed* home for `dataset_orig`, for the processing
   state, or for the annotation tables that are not one-row-per-feature
   (`annot_responsecurves`, `annot_qcconcentrations`, `metrics_calibration`) —
   they can only ride along as an untyped `metadata()` list. And while `colData`
   happily *carries* `qc_type` and `batch_id`, no Bioconductor class *models*
   them, nor calibration curves or internal-standard relationships: nothing
   downstream knows that a `PBLK` column is a blank, or that one feature
   normalizes another. Enforcing that is the work QUANT exists to do.

   The argument is about the *internal* representation only. Several parallel
   intensity variables are no obstacle — SE holds them as parallel assays, which
   is exactly what the exporter relies on. Once an experiment is processed,
   [`save_dataset_summarizedexperiment()`](recipe-04-summarizedexperiment.html)
   hands it to `limma`, `lipidr`, `POMA` and the rest of the ecosystem.

```{r}
# The object structure:
slotNames("MRMhubExperiment")
#>  [1] "title"                "analysis_type"        "feature_intensity_var"
#>  [4] "dataset_orig"         "dataset"              "dataset_filtered"     
#>  [7] "annot_analyses"       "annot_features"       "annot_istds"          
#> [10] "annot_responsecurves" "annot_qcconcentrations" "annot_studysamples"
#> [13] "annot_batches"        "metrics_qc"           "metrics_calibration"  
#> [16] "parameters_processing" "status_processing"   "is_istd_normalized"   
#> [19] "is_quantitated"       "is_filtered"          "has_outliers_tech"    
#> [22] "is_isotope_corr"      "analyses_excluded"    "features_excluded"    
#> [25] "var_drift_corrected"  "var_batch_corrected"
```

## Decision 2: Long-Format Data Exchange

**Choice:** The canonical data exchange format is a long CSV with one row per
feature per sample.

<details>
<summary><strong>Alternatives considered</strong></summary>

| Alternative | Why not |
|-------------|---------|
| Wide format (features as columns) | Scales poorly with many features; ambiguous NA meaning; awkward for filtering |
| HDF5 / Parquet | Adds binary dependency; not human-inspectable at handoff point |
| Database (SQLite) | Overkill for typical study sizes; adds complexity |

</details>

| Long format | Wide format |
|-------------|-------------|
| Handles variable feature counts naturally | Requires NA padding |
| Easy to filter, group, join | Awkward for row-wise operations |
| Natural for ggplot2 | Requires pivot before plotting |
| Explicit about missing vs zero | Ambiguous NA meaning |

```{r}
# Expected long format:
# | analysis_id | feature_id | area   | rt   |
# |-------------|------------|--------|------|
# | Sample_001  | LPC_18:1   | 12345  | 2.31 |
# | Sample_001  | LPC_16:0   | 23456  | 1.98 |
# | Sample_002  | LPC_18:1   | 11234  | 2.30 |
```

## Decision 3: Feature and analyte are distinct identifiers

**Choice:** `feature_id` identifies a distinct signal extracted from the MS data,
while `analyte_id` identifies the compound that signal measures. The two are kept
separate because they are not one-to-one.

<details>
<summary><strong>Alternatives considered</strong></summary>

| Alternative | Why not |
|-------------|---------|
| A single ID equating feature and compound | Cannot express one analyte measured by several features, nor one feature shared by isobaric analytes |
| Transition-level IDs only | Loses the compound-level grouping needed for QC concentrations and reporting |

</details>

**Rationale:** A feature is a distinct signal extracted from the MS data. Feature
and analyte are not one-to-one: one analyte can give several features (isotopes,
adducts, transitions), and one feature can carry several analytes when they are
isobaric or isomeric (e.g. SM/PC). INTEGRATOR selects the quantifier signal for
each feature, so by the time data reaches QUANT each `feature_id` is one value per
sample, while `analyte_id` links the features that measure the same compound.

## Decision 4: Immutable Original Data

**Choice:** `dataset_orig` is never modified after import. All processing
operates on `dataset`.

<details>
<summary><strong>Why this matters</strong></summary>

- **Reproducibility:** re-run any step without re-importing
- **Debugging:** compare original vs processed to verify corrections
- **Auditability:** the raw data is always accessible for review

</details>

```{r}
# You can always reset:
mexp@dataset <- mexp@dataset_orig

# Or compare before/after:
original <- mexp@dataset_orig |> dplyr::filter(feature_id == "LPC_18:1")
processed <- mexp@dataset |> dplyr::filter(feature_id == "LPC_18:1")
```

## Decision 5: QC-Centric Correction

**Choice:** Drift and batch correction are exclusively based on QC samples,
never study samples.

<details>
<summary><strong>Alternatives considered</strong></summary>

| Alternative | Why not |
|-------------|---------|
| Study sample–based correction (e.g., median centering on all) | Removes real biological signal |
| External reference only | Not always available; less responsive to within-run drift |

Reference: Broadhurst et al. (2018). *Metabolomics*, 14, 72.

</details>

**Implication:** Sufficient QC samples (≥ 5 per batch, evenly
distributed) are required for correction to work well.

## Decision 6: Explicit Processing Order

**Choice:** Functions should be called in a specific order. The package communicates
via status flags rather than enforcing with hard errors.

```{r}
# Recommended order:
# 1. import
# 2. add_metadata
# 3. set_analysis_order
# 4. normalize_by_istd
# 5. correct_drift_*
# 6. correct_batch_*
# 7. quantify_by_*
# 8. calc_qc_metrics
# 9. filter_features_qc
```

<details>
<summary><strong>Why not enforce strict order?</strong></summary>

- Some workflows skip steps (no ISTD, no drift correction)
- Researchers need flexibility to experiment
- Hard errors would frustrate exploratory analysis
- Status flags provide guidance without blocking

</details>

## Decision 7: Separation of INTEGRATOR and QUANT

**Choice:** Raw data processing (peak integration) and post-processing 
(normalization, QC) are separate tools in different languages.

| INTEGRATOR | QUANT (R) |
|---------------------|-----------|
| Automated, batch-oriented | Interactive, exploratory |
| Computationally intensive | Statistically intensive |
| Run once per study | Run iteratively |
| No user decisions needed | Many user decisions |

The interface is a CSV file. Either tool can be replaced
independently, and data can be inspected at the handoff point.

## Why QUANT is an R package

QUANT is implemented in R rather than Python because the bulk of its work is
statistical and visual: smoothing, robust regression, multivariate QC, and
ggplot2-driven inspection plots. R provides established, audit-friendly
implementations of these primitives (loess, cubic splines, `prcomp`, kernel
smoothers) and a large neighbouring ecosystem (`tidyverse`, Bioconductor,
`shiny`) familiar to most academic mass spectrometry and bioinformatics groups.
Literate `.Rmd` / `.qmd` notebooks make the final analysis self-documenting: the
same file produces the figures, the report, and the audit trail.

## Extending MRMhub

<details>
<summary><strong>Guidelines for new functions</strong></summary>

All functions should:

1. Accept `MRMhubExperiment` as first argument
2. Return `MRMhubExperiment` (for pipeline functions) or a result object
3. Update relevant status flags
4. Not modify `dataset_orig`

| Extension type | Pattern |
|----------------|---------|
| New correction method | `R/correct-*.R` — takes and returns `MRMhubExperiment` |
| New importer | Add `parse_*` function, register in `import_data_main()` |
| New plot | Follow `R/plots-*.R` — takes mexp, returns ggplot |
| New QC metric | Extend `calc_qc_metrics()` or add a helper |

</details>

## References

- Broadhurst et al. (2018). Guidelines and considerations for the use of
  system suitability and quality control samples in mass spectrometry assays.
  *Metabolomics*, 14, 72.
- Morgan M, Obenchain V, Hester J, Pagès H (2026). SummarizedExperiment: A
  container (S4 class) for matrix-like assays.
  doi:[10.18129/B9.bioc.SummarizedExperiment](https://doi.org/10.18129/B9.bioc.SummarizedExperiment).
  R package version 1.42.0, <https://bioconductor.org/packages/SummarizedExperiment>.
- Mohamed A, Molendijk J, Hill MM (2020). lipidr: A Software Tool for Data Mining
  and Analysis of Lipidomics Datasets. *Journal of Proteome Research*, 19(7),
  2890–2897. doi:[10.1021/acs.jproteome.0c00082](https://doi.org/10.1021/acs.jproteome.0c00082).

## Next steps

- [The MRMhubExperiment Data Object](manual-02-data-object.html) — the object and its slots in detail
- [Key Concepts & Glossary](manual-01-key-concepts.html) — the vocabulary used throughout
- [Drift and Batch Correction](manual-07-corrections.html) — the QC-centric correction in practice
