---
title: "Quarto workflows"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Preserving MRMhub's coloured console feedback and controlling figure size when a workflow notebook is rendered.
vignette: >
  %\VignetteIndexEntry{Quarto workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
```

<p class="page-kind page-kind-manual">Manual</p>

A reproducible MRMhub analysis is recorded as a [Quarto](https://quarto.org)
notebook (`.qmd`): one document that interleaves the processing code, its console
feedback, and the resulting figures and tables. Rendering it re-runs the whole
pipeline from raw import to final report, so the record and the result never drift
apart. [Introduction to R, Quarto, and MRMhub](tutorial-13-intro-to-r-quarto-mrmhub.html)
covers writing such a document from scratch, and the
[Workflow builder](tutorial-12-workflow-builder.html) emits one automatically.

This page documents the two aspects of rendering that are specific to MRMhub:
preserving its coloured, step-by-step console feedback in the output, and
controlling figure size independently of that feedback. Because MRMhub reports
each processing step on the console — a count and a truncated list of the features
or analyses affected — a rendered notebook doubles as an **audit trail** of what
each step did, and keeping that feedback intact is the main concern here.

## Session setup

Two session options in the notebook's setup chunk govern MRMhub's rendered
feedback. Both belong in a hidden chunk (`#| include: false`) so they run without
appearing in the output:

```r
library(mrmhub)

# Render mrmhub's coloured console feedback in HTML output (see below):
mrmhub_enable_cli_color()

# Widen or narrow the truncated feature/analysis lists in console messages:
options(mrmhub.max_report_items = 10)
```

`mrmhub_enable_cli_color()` is the only line specific to notebook rendering; the
`mrmhub.max_report_items` option is an ordinary session preference that also
applies interactively.

## Console feedback in the render

In an interactive R session MRMhub's messages are already coloured: green for a
successful step, yellow for a caution, red for an error. In a non-interactive
render, cli suppresses colour by default, so the same messages arrive as plain
text. `mrmhub_enable_cli_color()` re-enables it: it advertises colour support to
cli, and knitr then converts the emitted ANSI sequences to coloured HTML using the
**fansi** package (installed automatically as a suggested dependency). MRMhub's
step feedback travels on the *message* stream, so a chunk that runs a processing
step must keep `#| message: true` (the default) to preserve it.

::: {.callout .callout-note}
Colour applies to **HTML** output only. PDF and Word have no ANSI concept, so
their console blocks render as plain (but still legible) monospaced text. Severity
is never carried by colour alone: success, caution, and error lines remain
distinguishable by their leading symbol and wording.
:::

**What a message conveys.** A processing message states a **count** and, where a
subset of features or analyses is affected, an **illustrative list** of them. That
list is deliberately truncated (`options(mrmhub.max_report_items = 10)` sets how
many members are shown before the `…`), so a step touching hundreds of features
still prints a single tidy line rather than a wall of identifiers.

**Recovering the full list.** Because the list is truncated for display, the
message names the column in the returned object that carries the *complete*
membership. The truncated names are a preview; the column is the record. To export
the full set, filter the object on that column and write it out:

```{r, eval = FALSE}
# A message reporting features flagged out of calibration range names the
# `feature_conc_out_of_range` column; recover the complete set from the object:
mexp@metrics_qc |>
  dplyr::filter(feature_conc_out_of_range) |>
  dplyr::pull(feature_id) |>
  readr::write_lines("output/features_out_of_range.txt")
```

An **error** aborts the pipeline before it returns an object, so (unlike a warning
or success) there is no result to filter afterwards. An error message is therefore
self-contained: it names the offending values inline and, for the metadata
validation report, prints the full error/warning/note table before stopping. The
traceback is kept collapsed but remains reachable with `rlang::last_trace()`.

## Output formats

The output format is declared in the YAML front matter; several may be combined,
and each is produced by rendering the document once:

```yaml
format:
  html: default
  pdf:
    include-in-header:
      text: |
        \renewcommand{\familydefault}{\sfdefault}
  docx: default
```

- **HTML** is the recommended working format: it preserves the coloured console
  feedback, keeps interactive tables scrollable, and needs no external tooling.
- **PDF** requires a LaTeX installation; the lightweight
  [TinyTeX](https://yihui.org/tinytex/) distribution is sufficient
  (`tinytex::install_tinytex()`). Console blocks render as plain monospaced text.
- **Word** (`.docx`) is convenient for collaborators who annotate in Word, but has
  no styled-console concept: message blocks render as plain text and tables are
  not interactive.

## Figure size: `save_plot()` vs Quarto

Quarto's `fig-width`, `fig-height`, and `fig-dpi` control the figure *embedded in
the rendered document*. `save_plot()` controls a *standalone file* at an exact
physical size. The two are independent: a manuscript figure needs the latter,
because its width must match the journal's column specification regardless of how
the notebook happens to display it.

```{r}
#| eval: false
save_plot(
  p, "figures/fig3",
  format = c("pdf", "png"), width = 180, height = 200)
```

To render the same notebook against different inputs — one report per batch or per
project — declare its input paths as Quarto *parameters* rather than string
literals; [Custom QC report](recipe-02-custom-qc-report.html) is the worked
example.

## Next steps

- [Introduction to R, Quarto, and MRMhub](tutorial-13-intro-to-r-quarto-mrmhub.html): writing a workflow notebook from scratch
- [Using the workflow builder](tutorial-12-workflow-builder.html): generate a ready-to-render `.qmd` from your data
- [Custom QC report](recipe-02-custom-qc-report.html): render one notebook once per batch or project
- [Troubleshooting & FAQ](manual-10-troubleshooting.html): resolving common rendering problems
