---
title: "MRMhub overview"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
description: |
  A technical overview of MRMhub.
vignette: >
  %\VignetteIndexEntry{MRMhub overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", out.width = "100%",
  message = FALSE, warning = FALSE
)
```

<p class="page-kind page-kind-manual">Manual</p>

MRMhub performs postprocessing, quality control, and reporting of targeted mass
spectrometry data. Its core data structure is used by MRMhub's data-exchange,
processing, and plotting functions.

::: {.callout .callout-note}
**MRMhub's data** is structured around two levels: **analyses**, the individual
measurements (often corresponding to MS injections), and **features**, the
distinct signals extracted from the MS data. Each analysis–feature pair is a
**measurement**, which can carry one or more **feature variables** such as peak
area and retention time. A **sample** in this context is a
physical sample that was measured (several analyses can come from one sample) and an **analyte** can be represented by several features (i.e., different transitions, isotopes and
adducts). See also the [Glossary](manual-14-glossary.html).
:::

## The MRMhubExperiment data object

All data (measurements) and detailed metadata describing the analyses and
features, together with the intermediate and final processed data for one
experiment, are stored in a single object of the `MRMhubExperiment` class. The
object also records data processing status. See [The MRMhubExperiment data object](manual-02-data-object.html) for details.

<div style="text-align: center; margin: 2em 0;">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 600 280" style="max-width: 600px; width: 100%; height: auto; font-family: -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;">
  <rect x="5" y="5" width="590" height="270" rx="10" fill="#f8f9fa" stroke="#5B8FA8" stroke-width="2"/>
  <text x="300" y="30" text-anchor="middle" font-weight="700" font-size="14" fill="#2C3E50">MRMhubExperiment</text>
  <line x1="20" y1="40" x2="580" y2="40" stroke="#dee2e6"/>
  <rect x="20" y="55" width="270" height="100" rx="6" fill="#d6e4eb" stroke="#5B8FA8"/>
  <text x="155" y="75" text-anchor="middle" font-weight="600" font-size="12" fill="#2C3E50">DATA</text>
  <text x="35" y="95" font-size="10" fill="#333">dataset_orig: original imported data</text>
  <text x="35" y="112" font-size="10" fill="#333">dataset: annotated and processed data</text>
  <text x="35" y="129" font-size="10" fill="#333">dataset_filtered: QC-filtered output</text>
  <text x="35" y="146" font-size="10" fill="#333">metrics_qc: QC metrics per feature</text>
  <rect x="310" y="55" width="270" height="100" rx="6" fill="#f5e0c8" stroke="#D4914E"/>
  <text x="445" y="75" text-anchor="middle" font-weight="600" font-size="12" fill="#2C3E50">METADATA</text>
  <text x="325" y="95" font-size="10" fill="#333">annot_analyses: sample/run annotations</text>
  <text x="325" y="112" font-size="10" fill="#333">annot_features: feature annotations</text>
  <text x="325" y="129" font-size="10" fill="#333">annot_istds: ISTD concentrations</text>
  <text x="325" y="146" font-size="10" fill="#333">annot_batches / annot_qcconcentrations / ...</text>
  <rect x="20" y="170" width="560" height="50" rx="6" fill="#d8e6d4" stroke="#6B9E5E"/>
  <text x="300" y="192" text-anchor="middle" font-weight="600" font-size="12" fill="#2C3E50">STATUS AND FLAGS</text>
  <text x="300" y="210" text-anchor="middle" font-size="10" fill="#333">is_istd_normalized, is_quantitated, var_drift_corrected, var_batch_corrected, ...</text>
  <text x="300" y="258" text-anchor="middle" font-size="10" fill="#666">Functions take MRMhubExperiment in and return updated MRMhubExperiment</text>
</svg>
</div>



### How a MRMhubExperiment object is used in a workflow

An MRMhubExperiment object is created at the start of a workflow. Data and metadata
are imported into it, and it is then used in all subsequent processing, plotting,
and reporting steps. Each `mrmhub` function takes an MRMhubExperiment object as its
first argument, and processing functions return an updated MRMhubExperiment
object. This returned object is typically assigned back to the same name, which is
then passed to the next function. Plotting functions instead return a plot (typically
`ggplot2` object). MRMhubExperiment objects can also be used with the native R pipe (`|>`), which
forwards the object from one function to the next.

```{r, eval = FALSE}
mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(mexp, path = "data/results.tsv")
mexp <- normalize_by_istd(mexp)

# equivalently, chained with the R pipe:
mexp <- MRMhubExperiment() |>
  import_data_mrmhub(path = "data/results.tsv") |>
  normalize_by_istd()
```

## Core function groups

MRMhub's public functions fall into four groups, and their names follow a
consistent convention:

- **Data exchange** — read measured data and metadata into the object, and write results (tables, plots, reports) back out to disk.
- **Data processing** — the transforms that normalize, correct (drift, batch, isotopic interference), and quantify the measurements, and compute per-feature QC metrics.
- **QC charts** — return plots for visual inspection of the run order, normalization, corrections, calibration, and sample relationships.
- **QC filtering** — apply pass/fail thresholds to features and exclude features or analyses that fail QC.

For these functions arranged by pipeline stage, see the
[Function map](manual-13-function-map.html). The
[Function reference](../reference/index.html) lists every function with full
parameter detail.

## Next steps

- [Glossary](manual-14-glossary.html): definitions of the analytical terms used throughout the documentation
- [The MRMhubExperiment data object](manual-02-data-object.html): the tables, identifiers, and variables in detail
- [Sample types & QC roles](manual-06-sample-types.html): the QC labels used throughout
- [Lipidomics workflow](tutorial-03-lipidomics-workflow.html): these functions in a real script
- [Function reference](../reference/index.html): every function with full parameter detail
