---
title: "Metadata Tables & Matching"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  The metadata tables MRMhub uses — analyses, features, ISTDs, response curves and
  QC concentrations — their structure, and how identifiers are matched to the data.
vignette: >
  %\VignetteIndexEntry{Metadata Tables & Matching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
library(mrmhub)

# First rows of a bundled example CSV.
show_csv <- function(file, n = 4) {
  path <- system.file("extdata", file, package = "mrmhub")
  df <- readr::read_csv(path, n_max = n, show_col_types = FALSE, name_repair = "minimal")
  knitr::kable(df, caption = sprintf("Example file `%s`", file))
}

# Column headers of a sheet in the blank metadata template workbook.
show_template <- function(sheet) {
  path <- system.file("extdata", "mrmhub_metadata_templates.xlsx", package = "mrmhub")
  df <- openxlsx2::read_xlsx(path, sheet = sheet)
  knitr::kable(df, caption = sprintf("Template sheet `%s` (column headers)", sheet))
}
```

<p class="page-kind page-kind-manual">Manual</p>

Metadata in this context refers to analysis metadata, i.e., data that annotate the
analytical data. MRMhub stores it in five tables — analyses, features, internal
standards, response curves, and calibration/QC concentrations — that are attached to an
`MRMhubExperiment` and linked to the measurements by identifier. This page documents the
structure of each table and how identifiers are matched. For the steps to obtain the
templates and import metadata, see
[Import Metadata from Files or a Template](recipe-05-import-metadata.html).

Integrity of metadata and data is essential for correct post-processing: MRMhub inspects
imported data and metadata for completeness and for consistency of the IDs used across the
tables, and prints a summary of the identified errors, warnings, and notes to the console
after import so that possible issues can be identified and addressed.

## Metadata table structures

The previews below show example rows or the blank template headers for each metadata
table, so the tables can be prepared before import. Identifier columns must be consistent
across tables: `analysis_id` / `sample_id` match the analysis metadata and the data,
`analyte_id` links features to QC concentrations, and `istd_feature_id` and `feature_id`
reference the feature metadata. The full column reference for each table is given on its
linked function reference page.

### Analyses (samples) {#str-analyses}

One row per analysis (injection). Import with
[`import_metadata_analyses()`](../reference/import_metadata_analyses.html).

```{r, results = "asis"}
show_csv("MHQuant_demo_metadata_analyses.csv")
```

### Features (analytes) {#str-features}

One row per feature. Import with
[`import_metadata_features()`](../reference/import_metadata_features.html).

```{r, results = "asis"}
show_template("Features")
```

### Internal standards (ISTDs) {#str-istds}

One row per internal standard, giving its known concentration. Import with
[`import_metadata_istds()`](../reference/import_metadata_istds.html).

```{r, results = "asis"}
show_csv("MRMhub_ISTDconc.csv")
```

### Response curves {#str-responsecurves}

Maps response-curve injections to the amount analysed. Import with
[`import_metadata_responsecurves()`](../reference/import_metadata_responsecurves.html).

```{r, results = "asis"}
show_template("ResponseCurves")
```

### Calibration / QC concentrations {#str-qcconcentrations}

Known analyte concentrations in calibration and QC samples. Import with
[`import_metadata_qcconcentrations()`](../reference/import_metadata_qcconcentrations.html).

```{r, results = "asis"}
show_template("QCconcentrations")
```

## Identifier matching

Metadata is linked to the data by identifier: `analysis_id` joins the sample metadata to
the analyses, and `feature_id` joins the feature metadata to the measurements. For the
join to succeed, an identifier in a metadata sheet must match the corresponding identifier
in the data exactly.

To make that robust, identifiers in metadata are normalized on import in the same way as
in the data: leading and trailing spaces are removed, internal runs of whitespace are
collapsed to a single space, and raw-data file extensions (`.mzML`, `.d`, `.raw`,
`.wiff`, `.wiff2`, `.lcd`, `.chrom`) are stripped from `analysis_id`. So `"QC  01"` in a
sample sheet still matches `"QC 01"` in the data.

::: {.callout .callout-caution}
If analyses appear to be missing after `add_metadata()`, a difference in identifier
spelling is the usual cause. Whitespace differences are always squished on both sides and
no warning is issued for them — they are handled silently so a stray space cannot cause a
hard-to-trace mapping failure. Other differences (a raw-data extension the import does not
recognize, a typo, or a different naming scheme) are not normalized, and the affected
analyses are dropped from the join with a warning.
:::

## Next steps

- [Import Metadata from Files or a Template](recipe-05-import-metadata.html) — the import steps
- [Importing Analytical Data](manual-04-data-import.html) — importing the measurement data
- [Sample Types & QC Roles](manual-06-sample-types.html) — the `qc_type` values that drive QC logic
- [The MRMhubExperiment Data Object](manual-02-data-object.html) — where the metadata is stored
