---
title: "Metadata tables & matching"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  The metadata tables MRMhub uses (analyses, features, ISTDs, response curves and
  QC concentrations), their structure, and how identifiers are matched to the data.
vignette: >
  %\VignetteIndexEntry{Metadata tables & matching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
library(mrmhub)

# First rows of a bundled example CSV.
show_csv <- function(file, n = 4) {
  path <- system.file("extdata", file, package = "mrmhub")
  df <- readr::read_csv(path, n_max = n, show_col_types = FALSE, name_repair = "minimal")
  knitr::kable(df, caption = sprintf("Example file `%s`", file))
}

# Column headers of a sheet in the blank metadata template workbook.
show_template <- function(sheet) {
  path <- system.file("extdata", "mrmhub_metadata_templates.xlsx", package = "mrmhub")
  df <- openxlsx2::read_xlsx(path, sheet = sheet)
  knitr::kable(df, caption = sprintf("Template sheet `%s` (column headers)", sheet))
}
```

<p class="page-kind page-kind-manual">Manual</p>

Metadata in this context refers to analysis metadata, i.e., data that annotate the
analytical data. MRMhub stores it in five tables (analyses, features, internal
standards, response curves, and calibration/QC concentrations) that are attached to an
`MRMhubExperiment` and linked to the measurements by identifier. This page documents the
structure of each table and how identifiers are matched. For the steps to obtain the
templates and import metadata, see
[Import metadata from files or a template](recipe-05-import-metadata.html).

Integrity of metadata and data is essential for correct post-processing: MRMhub inspects
imported data and metadata for completeness and for consistency of the IDs used across the
tables, and prints a summary of the identified errors, warnings, and notes to the console
after import so that possible issues can be identified and addressed.

## Metadata table structures

The previews below show example rows or the blank template headers for each metadata
table, so the tables can be prepared before import. Identifier columns must be consistent
across tables: `analysis_id` / `sample_id` match the analysis metadata and the data,
`analyte_id` links features to QC concentrations, and `istd_feature_id` and `feature_id`
reference the feature metadata. The full column reference for each table is given on its
linked function reference page.

### Analyses (samples) {#str-analyses}

One row per analysis (injection). Import with
[`import_metadata_analyses()`](../reference/import_metadata_analyses.html).

```{r, results = "asis"}
show_csv("MHQuant_demo_metadata_analyses.csv")
```

### Features (analytes) {#str-features}

One row per feature. Import with
[`import_metadata_features()`](../reference/import_metadata_features.html).

```{r, results = "asis"}
show_template("Features")
```

### Internal standards (ISTDs) {#str-istds}

One row per internal standard, giving its known concentration. Import with
[`import_metadata_istds()`](../reference/import_metadata_istds.html).

```{r, results = "asis"}
show_csv("MRMhub_ISTDconc.csv")
```

### Response curves {#str-responsecurves}

Maps response-curve injections to the amount analysed. Import with
[`import_metadata_responsecurves()`](../reference/import_metadata_responsecurves.html).

```{r, results = "asis"}
show_template("ResponseCurves")
```

### Calibration / QC concentrations {#str-qcconcentrations}

Known analyte concentrations in calibration and QC samples. Import with
[`import_metadata_qcconcentrations()`](../reference/import_metadata_qcconcentrations.html).

```{r, results = "asis"}
show_template("QCconcentrations")
```

## Identifier matching

Metadata is linked to the data by identifier: `analysis_id` joins the sample metadata to
the analyses, and `feature_id` joins the feature metadata to the measurements. For the
join to succeed, an identifier in a metadata sheet must match the corresponding identifier
in the data exactly.

To make that robust, identifiers in metadata are normalized on import in the same way as
in the data: leading and trailing spaces are removed, internal runs of whitespace are
collapsed to a single space, and raw-data file extensions (`.mzML`, `.d`, `.raw`,
`.wiff`, `.wiff2`, `.lcd`, `.chrom`) are stripped from `analysis_id`. So `"QC  01"` in a
sample sheet still matches `"QC 01"` in the data.

::: {.callout .callout-caution}
If analyses appear to be missing after `add_metadata()`, a difference in identifier
spelling is the usual cause. Whitespace differences are always squished on both sides and
no warning is issued for them: they are handled silently so a stray space cannot cause a
hard-to-trace mapping failure. Other differences (a raw-data extension the import does not
recognize, a typo, or a different naming scheme) are not normalized, and the affected
analyses are dropped from the join with a warning.
:::

## Common defects and fixes

Most processing errors trace back to defects in the annotation tables rather
than to the analytical data. The frequent offenders are a mismatched
`analysis_id` between data and annotation, a missing `qc_type` label (the one
column besides `analysis_id` the importer requires), and `qc_type` values that
fall outside the canonical vocabulary. Each defect below pairs its diagnostic
with the edit that resolves it. After an annotation table is edited in R it is
re-imported, or passed to [`add_metadata()`](../reference/add_metadata.html) as
a named list, so the same validation runs again before the table is linked.

<details>
<summary><strong>Column <code>qc_type</code> missing from <code>annot_analyses</code></strong></summary>

`qc_type` is the only column besides `analysis_id` that the importer strictly
requires. When it is absent it can be derived from a naming convention embedded
in `analysis_id`:

```{r, eval = FALSE}
annot_analyses <- annot_analyses |>
  dplyr::mutate(qc_type = dplyr::case_when(
    stringr::str_detect(analysis_id, "BQC") ~ "BQC",
    stringr::str_detect(analysis_id, "BLK") ~ "SBLK",
    TRUE ~ "SPL"
  ))
```

</details>

<details>
<summary><strong><code>analysis_id</code> in data not found in annotation</strong></summary>

Some injections in the imported dataset have no matching row in the annotation
table. Listing the difference exposes the cause; leading or trailing whitespace
and case mismatches are the most frequent.

```{r, eval = FALSE}
data_ids <- unique(myexp@dataset_orig$analysis_id)
annot_ids <- annot_analyses$analysis_id

# IDs present in data but missing from annotation
setdiff(data_ids, annot_ids)

# Fix trailing whitespace introduced by Excel
annot_analyses <- annot_analyses |>
  dplyr::mutate(analysis_id = trimws(analysis_id))
```

`excl_unmatched_analyses = TRUE` in the importer drops unmatched analyses
silently, useful only when the omission is intentional, such as excluded
conditioning injections.

</details>

<details>
<summary><strong>Non-canonical <code>qc_type</code> values</strong></summary>

MRMhub recognises only its canonical `qc_type` labels, and they are
case-sensitive; see [Sample types & QC roles](manual-06-sample-types.html) for
the full set. Non-standard values are mapped to the canonical set before
validating:

```{r, eval = FALSE}
unique(annot_analyses$qc_type)

annot_analyses <- annot_analyses |>
  dplyr::mutate(qc_type = dplyr::case_when(
    qc_type == "Standard" ~ "CAL",
    qc_type == "Blank" ~ "SBLK",
    qc_type == "Pool" ~ "BQC",
    qc_type == "Sample" ~ "SPL",
    TRUE ~ qc_type
  ))
```

</details>

<details>
<summary><strong>Duplicate identifiers</strong></summary>

Repeated identifiers most often stem from copy-paste errors in Excel, or from
header rows interpreted as data.

```{r, eval = FALSE}
annot_analyses |>
  dplyr::count(analysis_id) |>
  dplyr::filter(n > 1)
```

</details>

<details>
<summary><strong>Feature IDs in annotation do not match data</strong></summary>

Comparing the feature identifiers in both directions separates features present
in the data but unannotated from method-development residue left only in the
annotation. A common cause is differing encoding of special characters
(parentheses, slashes, unicode escapes) between the integration-software output
and a hand-edited Excel file.

```{r, eval = FALSE}
data_features <- unique(myexp@dataset_orig$feature_id)
annot_feat_ids <- annot_features$feature_id

# In data but missing from annotation
setdiff(data_features, annot_feat_ids)

# In annotation but absent from data (often method-development residue)
setdiff(annot_feat_ids, data_features)
```

</details>

## Practical recommendations

- Annotations are re-imported after every edit. A saved Excel file held open in
  a separate process can be silently locked and read only in part.
- CSVs are saved as UTF-8 without BOM. Excel's default "CSV UTF-8" export adds a
  byte-order mark that shifts the first column header.
- Special characters in `analysis_id` and `feature_id` (parentheses, slashes,
  spaces) survive R but complicate downstream column references; restricting IDs
  to letters, digits, underscores, and hyphens avoids this.
- The annotation file is kept under version control alongside the data file.

## Next steps

- [Import metadata from files or a template](recipe-05-import-metadata.html): the import steps
- [Importing analytical data](manual-04-data-import.html): importing the measurement data
- [Sample types & QC roles](manual-06-sample-types.html): the `qc_type` values that drive QC logic
- [The MRMhubExperiment data object](manual-02-data-object.html): where the metadata is stored
