---
title: "Key Concepts & Glossary"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
description: |
  Plain-language overview of MRMhub's data model, terminology, and architecture.
vignette: >
  %\VignetteIndexEntry{Key Concepts & Glossary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", out.width = "100%",
  message = FALSE, warning = FALSE
)
```

## How MRMhub Thinks About Your Data

MRMhub organizes targeted mass spectrometry data around two axes:
 
- **Analyses** (rows) — individual injections/samples in the run
- **Features** (columns) — individual MRM transitions (compound + precursor/product ion pair)

Every measurement is a **peak area** at the intersection of one analysis
and one feature. All of this is stored in a single object: the
`MRMhubExperiment`.

## The MRMhubExperiment Object

<div style="text-align: center; margin: 1.5em 0;">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 480 280" style="max-width: 480px; width: 100%; height: auto; font-family: -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;">
  <rect x="10" y="10" width="460" height="260" rx="12" fill="#f8f9fa" stroke="#4a90a4" stroke-width="2"/>
  <text x="240" y="38" text-anchor="middle" font-weight="700" font-size="15" fill="#1a1a1a">MRMhubExperiment</text>
  <line x1="30" y1="48" x2="450" y2="48" stroke="#dee2e6"/>
  <!-- DATA section -->
  <rect x="30" y="60" width="200" height="90" rx="6" fill="#d6e8f0" stroke="#4a90a4"/>
  <text x="130" y="82" text-anchor="middle" font-weight="600" font-size="12" fill="#2c5f7a">DATA</text>
  <text x="45" y="102" font-size="11" fill="#333">• peak areas (raw)</text>
  <text x="45" y="118" font-size="11" fill="#333">• peak areas (normalized)</text>
  <text x="45" y="134" font-size="11" fill="#333">• calibration curves</text>
  <!-- METADATA section -->
  <rect x="250" y="60" width="200" height="90" rx="6" fill="#f5e0c8" stroke="#c87f3b"/>
  <text x="350" y="82" text-anchor="middle" font-weight="600" font-size="12" fill="#7a4a1a">METADATA</text>
  <text x="265" y="102" font-size="11" fill="#333">• annot_analyses</text>
  <text x="265" y="118" font-size="11" fill="#333">• annot_features</text>
  <text x="265" y="134" font-size="11" fill="#333">• processing log</text>
  <!-- Pipeline concept -->
  <rect x="30" y="170" width="420" height="50" rx="6" fill="#d4e8d4" stroke="#5a9a5a"/>
  <text x="240" y="192" text-anchor="middle" font-weight="600" font-size="12" fill="#2c5f2c">PIPELINE</text>
  <text x="240" y="210" text-anchor="middle" font-size="11" fill="#333">function(mhexp) → updated mhexp → function(mhexp) → ...</text>
  <!-- Footer -->
  <text x="240" y="255" text-anchor="middle" font-size="10" fill="#666">Chain functions together to build your workflow</text>
</svg>
</div>

Functions take an `MRMhubExperiment` and return an updated one; these
calls are chained together to build a workflow.

## Glossary

```{r, echo=FALSE, results='asis'}
glossary <- data.frame(
 Term = c("analysis_id", "feature_id", "ISTD", "RQC", "qualifier",
           "quantifier", "drift correction", "batch correction",
           "long format", "wide format", "MRMhubExperiment",
           "INTEGRATOR", "MRMhub-QUANT"),
 Definition = c(
   "Unique identifier for each injection/sample in the run",
   "Unique identifier for each MRM transition",
   "Internal Standard — a spiked compound used for normalization",
   "Response Quality Control — a serial-dilution (response) curve of the QC pool, used to assess linearity and response",
   "A secondary MRM transition used to confirm compound identity",
   "The primary MRM transition used for quantification",
   "Statistical correction for signal drift across a run",
   "Correction for systematic differences between analytical batches",
   "One row per analysis × feature (tidy format)",
   "One row per analysis, features as columns",
   "The S4 object holding all data + metadata for a project",
   "Standalone tool for peak integration from mzML files",
   "The post-processing module (this R package, mrmhub): QC, normalization, quantitation, reporting"
 ),
 stringsAsFactors = FALSE
)

DT::datatable(
 glossary,
 options = list(
   pageLength = 15,
   dom = "ft",
   ordering = TRUE
 ),
 rownames = FALSE,
 filter = "top",
 class = "compact stripe"
)
```

## Key Identifiers

MRMhub uses consistent column names across all data:

| Column | Where | Purpose |
|--------|-------|---------|
| `analysis_id` | analyses table, data tables | Links a row to a specific injection |
| `feature_id` | features table, data tables | Links a value to a specific transition |
| `qc_type` | analyses table | Sample role — SPL, BQC, RQC, blank, calibration, etc. |
| `analyte_id` | features table | Human-readable compound/analyte name |
| `istd_feature_id` | features table | Maps each analyte to its internal standard |

## Next Steps

- [Data Structures](manual-01-data-structure.html) — technical details of each table
- [Data Identifiers](manual-02-data-identifiers.html) — deep dive on ID columns
- [Feature Variables](manual-03-feature-variables.html) — all feature metadata columns
