---
title: "Key Concepts & Glossary"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
description: |
  Core vocabulary of MRMhub — the data model, terminology, and the function-naming convention.
vignette: >
  %\VignetteIndexEntry{Key Concepts & Glossary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", out.width = "100%",
  message = FALSE, warning = FALSE
)
```

<p class="page-kind page-kind-manual">Manual</p>

MRMhub organizes targeted mass spectrometry data around two axes: **analyses**,
the individual injections or samples in a run, and **features**, the distinct
signals extracted from the MS data. Every measurement is a value — typically a
peak area — at the intersection of one analysis and one feature, and all of these
values are held together with their metadata in a single object, the
`MRMhubExperiment`. Functions take an `MRMhubExperiment` and return an updated
one, and these calls are chained together to build a workflow. See
[The MRMhubExperiment Data Object](manual-02-data-object.html) for how the object
is structured and [Sample Types & QC Roles](manual-06-sample-types.html) for the
sample roles that drive its QC logic.

## Glossary

```{r, echo=FALSE, results='asis'}
glossary <- data.frame(
 Term = c("analysis_id", "feature_id", "analyte_id", "ISTD", "RQC", "qualifier",
           "quantifier", "drift correction", "batch correction",
           "long format", "wide format", "MRMhubExperiment",
           "INTEGRATOR", "MRMhub-QUANT"),
 Definition = c(
   "Unique identifier for each injection/sample in the run",
   "Unique identifier for a feature (a distinct signal extracted from the MS data)",
   "Identifier of the compound/analyte a feature measures; several features can share one analyte",
   "Internal Standard — a spiked compound used for normalization",
   "Response Quality Control — a serial-dilution (response) curve of the QC pool, used to assess linearity and response",
   "A secondary MRM transition used to confirm compound identity",
   "The primary MRM transition used for quantification",
   "Statistical correction for signal drift across a run",
   "Correction for systematic differences between analytical batches",
   "One row per analysis × feature (tidy format)",
   "One row per analysis, features as columns",
   "The S4 object holding all data + metadata for a project",
   "Standalone tool for peak integration from mzML files",
   "The post-processing module (this R package, mrmhub): QC, normalization, quantitation, reporting"
 ),
 stringsAsFactors = FALSE
)

DT::datatable(
 glossary,
 options = list(
   pageLength = 15,
   dom = "ft",
   ordering = TRUE
 ),
 rownames = FALSE,
 filter = "top",
 class = "compact stripe"
)
```

A feature and an analyte are not one-to-one: one analyte can give several features
(isotopes, adducts, transitions), and one feature can carry several analytes when
they are isobaric or isomeric (e.g. SM/PC). The `feature_id` names the signal,
while the `analyte_id` names the compound.

## Reading the function names

A naming convention runs through the public API, so once it is internalised the
function names are largely predictable:

- `import_data_*()` — read raw integrated peak data from a given source
- `import_metadata_*()` — read an annotation table from a given source
- `add_metadata()` / `set_*()` — attach annotation or assign a slot on the experiment
- `normalize_by_*()` — divide one variable by another (e.g. by ISTD)
- `correct_drift_*()` / `correct_batch_*()` — apply a named correction model
- `quantify_by_*()` — compute concentration values via a given strategy
- `calc_*()` — compute per-feature or per-analysis metrics
- `filter_*()` — apply pass/fail thresholds
- `plot_*()` — return a ggplot2 object
- `save_*()` — write to disk
- `get_*()` — read a derived value from the experiment

## Next steps

- [The MRMhubExperiment Data Object](manual-02-data-object.html) — the tables and identifiers in detail
- [Sample Types & QC Roles](manual-06-sample-types.html) — the `qc_type` labels used throughout
- [Basic MRMhub Workflow](tutorial-02-basic-workflow.html) — these functions in a real script
- [Function reference](../reference/index.html) — every function with full parameter detail
