---
title: "The MRMhubExperiment Data Object"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  The primary data container of the MRMhub workflow: its data and metadata
  tables, the identifiers that link them, and the feature variables it stores.
vignette: >
  %\VignetteIndexEntry{The MRMhubExperiment Data Object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
bibliography: references.bib
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, warning = FALSE,
                      message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
library(mrmhub)
```

<p class="page-kind page-kind-manual">Manual</p>

The `MRMhubExperiment` object is the primary data container in the MRMhub
workflow. It holds all the experimental and processed data and metadata, as well
as details of the applied processing steps and the current status of the data.
Most MRMhub functions take the `MRMhubExperiment` object as data input, and
functions that process the data return an updated `MRMhubExperiment` object, which
can then be used in subsequent steps. The data within the object is organized into
**data** and **metadata** categories, each divided into tables (data frames).

<div style="text-align: center; margin: 2em 0;">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 600 280" style="max-width: 600px; width: 100%; height: auto; font-family: -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;">
  <rect x="5" y="5" width="590" height="270" rx="10" fill="#f8f9fa" stroke="#5B8FA8" stroke-width="2"/>
  <text x="300" y="30" text-anchor="middle" font-weight="700" font-size="14" fill="#2C3E50">MRMhubExperiment</text>
  <line x1="20" y1="40" x2="580" y2="40" stroke="#dee2e6"/>
  <rect x="20" y="55" width="270" height="100" rx="6" fill="#d6e4eb" stroke="#5B8FA8"/>
  <text x="155" y="75" text-anchor="middle" font-weight="600" font-size="12" fill="#2C3E50">DATA</text>
  <text x="35" y="95" font-size="10" fill="#333">dataset_orig — original imported data</text>
  <text x="35" y="112" font-size="10" fill="#333">dataset — annotated and processed data</text>
  <text x="35" y="129" font-size="10" fill="#333">dataset_filtered — QC-filtered output</text>
  <text x="35" y="146" font-size="10" fill="#333">metrics_qc — QC metrics per feature</text>
  <rect x="310" y="55" width="270" height="100" rx="6" fill="#f5e0c8" stroke="#D4914E"/>
  <text x="445" y="75" text-anchor="middle" font-weight="600" font-size="12" fill="#2C3E50">METADATA</text>
  <text x="325" y="95" font-size="10" fill="#333">annot_analyses — sample/run annotations</text>
  <text x="325" y="112" font-size="10" fill="#333">annot_features — feature annotations</text>
  <text x="325" y="129" font-size="10" fill="#333">annot_istds — ISTD concentrations</text>
  <text x="325" y="146" font-size="10" fill="#333">annot_batches / annot_qcconcentrations / ...</text>
  <rect x="20" y="170" width="560" height="50" rx="6" fill="#d8e6d4" stroke="#6B9E5E"/>
  <text x="300" y="192" text-anchor="middle" font-weight="600" font-size="12" fill="#2C3E50">STATUS AND FLAGS</text>
  <text x="300" y="210" text-anchor="middle" font-size="10" fill="#333">is_istd_normalized, is_quantitated, var_drift_corrected, var_batch_corrected, ...</text>
  <text x="300" y="258" text-anchor="middle" font-size="10" fill="#666">Functions take MRMhubExperiment in and return updated MRMhubExperiment</text>
</svg>
</div>

## Working with the object

A new object is created with `MRMhubExperiment()`. Most MRMhub functions take an
`MRMhubExperiment` object as input, and data processing functions return a
modified `MRMhubExperiment` that is used in the subsequent step. R pipes can also
be used to chain multiple functions together, which more clearly indicates the
processing workflow and makes the code easier to read:

```{r, eval = FALSE}
mexp <- MRMhubExperiment() |>
  data_load_example(1) |>
  normalize_by_istd()
```

<details>
<summary><strong>New to R?</strong> — what <code>|&gt;</code> and <code>&lt;-</code> do</summary>

`<-` assigns the value on its right to the name on its left, so
`mexp <- MRMhubExperiment()` stores the new object in `mexp`. The native pipe
`|>` passes the object on its left as the first argument of the function on its
right, so `mexp |> normalize_by_istd()` is the same as `normalize_by_istd(mexp)`.
</details>

Multiple `MRMhubExperiment` objects can be created and processed independently
within the same script, which is convenient when polar and non-polar assays, or
several studies, are handled together:

```{r, eval = FALSE}
m_polars <- MRMhubExperiment(title = "Polar metabolites")
m_lipids <- MRMhubExperiment(title = "Non-polar metabolites")
```

Functions starting with `get_` retrieve data and metadata from an
`MRMhubExperiment` object, and the `$` syntax can be used to access the data and
metadata tables directly. The whole object, with all its data, metadata and
processing state, is saved and read back as a single `.rds` file, and a detailed
summary of the dataset and processing status can be printed at any time:

```{r, eval = FALSE}
mexp <- data_load_example(MRMhubExperiment(), 1)

dataset  <- get_analyticaldata(mexp, annotated = TRUE)  # processed data
analyses <- mexp$annot_analyses                         # sample metadata

saveRDS(mexp, "myexp-mrmhub.rds", compress = TRUE)
print(mexp)                                             # dataset + status summary
```

The full list of accessor and processing functions is given in the
[function reference](../reference/index.html).

<details>
<summary><strong>Under the hood</strong> — the full slot structure</summary>

`MRMhubExperiment` is an S4 object with the following slots. The status flags
(`is_*`, `var_*_corrected`) record which steps have been applied and are consulted
by later functions to enforce the recommended processing order:

```
MRMhubExperiment
  ├─ title:                 chr "My LCMS Assay"
  ├─ analysis_type:         chr NA
  ├─ feature_intensity_var: chr "feature_area"
  ├─ dataset_orig:          tibble [400 × 26]
  ├─ dataset:               tibble [400 × 26]
  ├─ dataset_filtered:      tibble [0 × 14]
  ├─ annot_analyses:        tibble [25 × 13]
  ├─ annot_features:        tibble [16 × 16]
  ├─ annot_istds:           tibble [8 × 4]
  ├─ annot_responsecurves:  tibble [0 × 3]
  ├─ annot_qcconcentrations: tibble [32 × 5]
  ├─ annot_batches:         tibble [1 × 4]
  ├─ metrics_qc:            tibble [0 × 0]
  ├─ metrics_calibration:   tibble [4 × 15]
  ├─ status_processing:     chr "Calibration-quantitated data"
  ├─ is_istd_normalized:    logi TRUE
  ├─ is_quantitated:        logi TRUE
  ├─ is_filtered:           logi FALSE
  ├─ var_drift_corrected:   Named logi [1:3]
  └─ var_batch_corrected:   Named logi [1:3]
```

</details>

## Data and metadata tables

The **data** tables hold the raw and processed values, and the **metadata**
(annotation) tables describe the samples, features and standards. The two groups
are linked by the shared identifiers described in the next section.

| Group | Table (slot) | Description |
|-------|-------------|-------------|
| Data | `dataset_orig` | Original imported analysis data; never modified after import. |
| Data | `dataset` | Annotated raw and processed data with the available metadata. |
| Data | `dataset_filtered` | Subset of `dataset` passing the QC criteria. |
| Data | `metrics_qc` | Information and quality-control metrics for features. |
| Metadata | `annot_analyses` | Sample categories, amounts, dilutions, processing batches, run order. |
| Metadata | `annot_features` | Internal standards for normalization, response factors, classification, quantifiers. |
| Metadata | `annot_istds` | Concentrations of internal standards added to samples. |
| Metadata | `annot_batches` | Start and end boundaries for each defined batch. |
| Metadata | `annot_responsecurves` | Response curves: sample amounts across dilution steps. |
| Metadata | `annot_qcconcentrations` | Concentrations of labelled and unlabelled standards in calibration and QC materials. |

## Identifiers {#identifiers}

A small set of key fields organizes the data within the `MRMhubExperiment` object
and is used by many functions in the package. Certain field names differ from
conventional terminology (e.g. `analysis_id` instead of `sample_id`) to allow more
flexible workflows and to reduce confusion with other identifiers: a sample may be
measured multiple times across different methods or processing replicates,
necessitating distinct identifiers, and analytes can be quantified through
multiple transitions or adducts, which is why `feature_id` is designated as the
primary identifier.

| Table | Field | Description |
|-------|-------|-------------|
| Analyses | **`analysis_id`** | Unique identifier of each analysis. |
|  | `qc_type` | QC/sample type (see below). |
|  | `batch_id` | Unique identifier of each batch. |
|  | `sample_id` | Unique identifier of the physical sample that was tested. |
| Features | **`feature_id`** | Unique identifier for each feature. |
|  | `istd_feature_id` | The `feature_id` of the internal standard used to normalize raw intensities. |
|  | `analyte_id` | Unique identifier of the analyte. |

The `qc_type` field categorizes samples by their analytical purpose and is used
throughout the package. It combines standardized terms introduced by
@broadhurst2018GuidelinesConsiderationsUse (SPL, BQC, TQC, LTR, RQC) with
traditional terminology from analytical and clinical chemistry (LQC, MQC, HQC,
CAL, NIST, SST, and the blank types). Each `qc_type` is shown with a consistent
colour scheme and point shape across all MRMhub plots; the full list and their
roles is given in [Sample Types & QC Roles](manual-06-sample-types.html).

## Feature variables {#feature-variables}

Feature variables store the values associated with a feature in a specific sample.
They describe, for example, the absolute or relative abundance, the chromatographic
retention time, the peak shape, and — when processed data are imported —
properties such as measurement accuracy. A feature variable can be referred to by
its internal name, which always starts with `feature_` (e.g. `feature_intensity`),
or by its short name (e.g. `conc`, `intensity`, `norm_intensity`, `rt`), and many
processing and plotting functions take a `variable` argument that selects which one
to use.

The following variables organize the data-processing flow and are stored in the
`dataset` table. The `intensity` variable holds the raw signal (e.g. peak area)
retrieved from one of the original feature variables; all variables downstream are
the result of processing:

```{r, echo = FALSE, results = 'asis'}
key_vars <- data.frame(
  `Short Name` = c("intensity", "norm_intensity", "pmol_total", "conc",
                   "conc_normalized", "conc_bias"),
  `Internal Name` = c("feature_intensity", "feature_norm_intensity",
                      "feature_pmol_total", "feature_conc",
                      "feature_conc_normalized", "feature_conc_ratio"),
  Description = c(
    "Raw signal intensity. A copy of area, height, response, or intensity from the imported raw data.",
    "Internal standard-normalized raw signal intensity.",
    "Total feature (analyte) amount in the measured sample (pmol/sample).",
    "Feature (analyte) concentration.",
    "Normalized concentration (e.g. by a reference sample or other methods).",
    "Ratio of measured to expected concentration; generated by calibrate_by_reference()."
  ),
  stringsAsFactors = FALSE, check.names = FALSE
)

DT::datatable(
  key_vars,
  options = list(dom = "t", pageLength = 10, ordering = FALSE),
  rownames = FALSE,
  class = "compact stripe"
)
```

Some processing steps overwrite feature values with re-calculated ones — for
example feature area after interference correction, or concentrations after
drift/batch correction or reference-sample re-calibration. In these cases the
original values are kept in a backup variable, so the earlier state remains
available; the backup is named after the original variable with a postfix:
`_orig` (the imported intensity, before interference correction), `_raw` (the
uncorrected calculated values, before a correction step), `_before` (the last
value before the most recent correction), `_beforecal` (before
`calibrate_by_reference()`), and `_fit` (model-fit points used by
`plot_runscatter()` to show the trend).

The raw feature variables below are stored in `dataset_orig` and are never
modified by any MRMhub function. One of them is copied to `intensity` at import
(by default `area` if available, then `height`, `response`, or `intensity`), and
the source variable can be set manually with `set_intensity_var()`:

```{r, echo = FALSE, results = 'asis'}
raw_vars <- data.frame(
  `Short Name` = c("area", "height", "response", "intensity", "rt", "fwhm",
                   "width", "int_start", "int_end", "symmetry", "sn_ratio",
                   "accuracy", "ionratio"),
  `Internal Name` = c("feature_area", "feature_height", "feature_response",
                      "feature_intensity", "feature_rt", "feature_fwhm",
                      "feature_width", "feature_int_start", "feature_int_end",
                      "feature_symmetry", "feature_sn_ratio", "feature_accuracy",
                      "feature_ionratio"),
  Description = c("Peak area", "Peak height", "Feature response (vendor-specific)",
                  "Feature intensity (vendor-specific)", "Retention time",
                  "Full Width at Half Maximum", "Peak width",
                  "Peak integration start time", "Peak integration end time",
                  "Peak symmetry", "Signal-to-noise ratio",
                  "Accuracy (relative to target value)", "Ion ratio"),
  stringsAsFactors = FALSE, check.names = FALSE
)

DT::datatable(
  raw_vars,
  options = list(dom = "ft", pageLength = 15, ordering = TRUE),
  rownames = FALSE,
  filter = "top",
  class = "compact stripe"
)
```

## Next steps

- [Importing Analytical Data](manual-04-data-import.html) — load a peak table into a new object
- [Sample Types & QC Roles](manual-06-sample-types.html) — the full list of `qc_type` labels
- [Design Overview](manual-03-design-overview.html) — how data flows through the pipeline
- [Function reference](../reference/index.html) — accessors and processing functions

## References
