---
title: "Importing Analytical Data"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Importing analytical data from different sources into an MRMhubExperiment.
vignette: >
  %\VignetteIndexEntry{Importing Analytical Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
bibliography: references.bib
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE,
                      comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
library(mrmhub)

# Compact "screenshot" of the first rows of a bundled example file.
show_head <- function(file, delim = ",", n = 3, cols = NULL) {
  path <- system.file("extdata", file, package = "mrmhub")
  df <- readr::read_delim(
    path,
    delim = delim,
    n_max = n,
    show_col_types = FALSE,
    name_repair = "minimal"
  )
  total <- ncol(df)
  if (!is.null(cols)) df <- df[, cols, drop = FALSE]
  cap <- if (!is.null(cols) && length(cols) < total) {
    sprintf("First %d rows of `%s` (showing %d of %d columns)", n, file, length(cols), total)
  } else {
    sprintf("First %d rows of `%s`", n, file)
  }
  knitr::kable(df, caption = cap)
}
```

<p class="page-kind page-kind-manual">Manual</p>

Analytical data, i.e. preprocessed data from mass spectrometry experiments, can be
imported from different sources. Data files present in a folder can also be
imported and merged, which is useful when the raw data processing is broken down
into batches that produce separate result files. The following formats are
currently supported; each importer reads a CSV or TSV file and returns an
`MRMhubExperiment`, and the exact file layout and all arguments are documented on
the linked reference pages.

| Source | Function |
|---|---|
| MRMhub / INTEGRATOR *(recommended)* | [`import_data_mrmhub()`](../reference/import_data_mrmhub.html) |
| Agilent MassHunter Quant | [`import_data_masshunter()`](../reference/import_data_masshunter.html) |
| Skyline (small molecule) | [`import_data_skyline()`](../reference/import_data_skyline.html) |
| Generic wide CSV | [`import_data_csv_wide()`](../reference/import_data_csv_wide.html) |
| Generic long CSV | [`import_data_csv_long()`](../reference/import_data_csv_long.html) |

When the analytical results contain metadata, such as sample and feature
annotations, these can be imported into the `MRMhubExperiment` object as well. The
imported metadata is checked for integrity and consistency and then added to the
annotation tables; to include it, set the argument `import_metadata = TRUE` (see
[Importing Metadata](manual-05-metadata.html)).

## MRMhub / INTEGRATOR

Long-format output files produced by MRMhub-INTEGRATOR, the peak-integration
component of the project, can be imported directly. Specific metadata present in
the data file is imported as well when `import_metadata = TRUE`:

```{r}
filepath <- system.file("extdata/MRMhub_demo.tsv", package = "mrmhub")
mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(mexp, filepath, import_metadata = TRUE)
```

## Agilent MassHunter Quant

Peak integration results exported from Agilent MassHunter Quant in CSV format can
be imported, with samples in rows and features in columns. Import of qualifier
results is supported, and sample, method and result metadata present in the file
can also be imported (`import_metadata = TRUE`):

```{r}
filepath <- system.file("extdata/MHQuant_demo.csv", package = "mrmhub")
mexp <- MRMhubExperiment()
mexp <- import_data_masshunter(mexp, filepath, import_metadata = TRUE)
```

## Skyline

Small-molecule peak integration results from [Skyline](https://skyline.ms/) can be
imported from long-format CSV reports. The `analysis_id` is mapped from the
`Replicate Name` column, and unique `feature_id`s are generated from the
`Molecule Name` together with the precursor/product names or *m/z* values, unless
the `Molecule Name` alone already identifies the feature (controlled by
`transition_id_columns`):

```{r}
filepath <- system.file("extdata/Skyline_MoleculeTransitionResults.csv", package = "mrmhub")
mexp <- MRMhubExperiment()
mexp <- import_data_skyline(mexp, filepath,
                            import_metadata = TRUE,
                            transition_id_columns = "mz")
```

<details>
<summary><strong>Exporting the report from Skyline</strong></summary>

Navigate to *File → Export* and select the *Molecule Transition Results* format.
Ensure the export includes `Replicate Name`, `Molecule Name`, and either
`Precursor Mz` and `Product Mz` or `Precursor Name` and `Product Name`, plus at
least one feature variable such as `Area` or `Retention Time`. Further options are
described in [`import_data_skyline()`](../reference/import_data_skyline.html).
</details>

## Generic wide-format CSV

Analysis results, whether raw intensities (e.g. peak areas) or preprocessed data
(e.g. concentrations), can be provided as plain wide-format CSV tables with
analyses in rows and features in columns. The data type held in the table is
declared with the `variable_name` argument:

```{r}
filepath <- system.file("extdata/plain_wide_dataset.csv", package = "mrmhub")
mexp <- MRMhubExperiment()
mexp <- import_data_csv_wide(mexp, path = filepath,
                             variable_name = "area",
                             import_metadata = TRUE)
```

## Generic long-format CSV

Analysis results containing various feature variables can be provided as generic
long-format CSV tables, where each row is a feature–value observation for a sample
and additional columns capture feature variables and sample- or method-related
metadata. By default the file must include `analysis_id`, `feature_id`, and at
least one feature-value column such as `area`; if the file uses different column
names, a `column_mapping` associates the MRMhub column names with those in the
file. The full column list and mapping rules are given in
[`import_data_csv_long()`](../reference/import_data_csv_long.html):

```{r}
filepath <- system.file("extdata/plain_long_dataset.csv", package = "mrmhub")
mexp <- MRMhubExperiment()
mexp <- import_data_csv_long(mexp, path = filepath, import_metadata = TRUE,
  column_mapping = c(
    "analysis_id"     = "raw_data_filename",
    "qc_type"         = "qc_type",
    "feature_id"      = "feature_id",
    "feature_class"   = "feature_class",
    "istd_feature_id" = "istd_feature_id",
    "feature_rt"      = "rt",
    "feature_area"    = "area"))
```

## File formats

The previews below show the first rows of the example files bundled with the
package, so that an input file can be recognised and prepared before importing.
Vendor and tool exports (MRMhub/INTEGRATOR, MassHunter, Skyline) are produced by
the software and are not edited by hand, whereas the generic wide and long CSV
tables are prepared manually. The full column reference for each format is given
on its linked function reference page.

### MRMhub / INTEGRATOR — long CSV/TSV {#str-mrmhub}

The long output has one row per analysis × feature, tab- or comma-separated.
Besides the columns shown it also carries `internal_standard`, `time_stamp`,
transition columns (`precursor_mz`, `product_mz`, `collision_energy`, `polarity`),
and integration columns (`height`, `FWHM`, `rt_int_start`, `rt_int_end`). The file
is produced by INTEGRATOR and is not prepared by hand.

```{r, results = "asis"}
show_head(
  "MRMhub_demo.tsv",
  delim = "\t",
  cols = c("feature_name", "raw_data_filename", "sample_type", "batch", "rt_apex", "area")
)
```

### Agilent MassHunter Quant — wide CSV {#str-masshunter}

A MassHunter Quant export is wide with a two-row header. The first header row
groups columns by compound (`<Compound> Method` and `<Compound> Results` blocks)
after a leading `Sample` block; the second row holds the actual sub-headers. The
`Method`/`Results` blocks repeat for every compound (quantifiers and qualifiers).

<table class="table" style="font-size: 0.85em;">
<thead>
<tr>
  <th colspan="4" style="text-align:center; border-bottom:2px solid #5B8FA8;">Sample</th>
  <th colspan="4" style="text-align:center; border-bottom:2px solid #D4914E;">CE 18:1 — Method</th>
  <th colspan="3" style="text-align:center; border-bottom:2px solid #6B9E5E;">CE 18:1 — Results</th>
  <th rowspan="2" style="vertical-align:bottom;">…</th>
</tr>
<tr>
  <th>Data File</th><th>Name</th><th>Type</th><th>Acq. Date-Time</th>
  <th>Precursor Ion</th><th>Product Ion</th><th>Collision Energy</th><th>RT</th>
  <th>RT</th><th>Area</th><th>FWHM</th>
</tr>
</thead>
<tbody>
<tr>
  <td>001_…01.d</td><td>001_…01</td><td>Sample</td><td>4/12/18 18:28</td>
  <td>668.6</td><td>369.3</td><td>10</td><td>7.16</td>
  <td>7.16</td><td>5152996</td><td>0.081</td>
</tr>
<tr>
  <td>002_…02.d</td><td>002_…02</td><td>Sample</td><td>4/12/18 18:39</td>
  <td>668.6</td><td>369.3</td><td>10</td><td>7.16</td>
  <td>7.16</td><td>4789505</td><td>0.080</td>
</tr>
</tbody>
</table>

### Skyline (small molecule) — long CSV {#str-skyline}

The *Molecule Transition Results* export has one row per transition result;
`Replicate Name` becomes `analysis_id`, `Molecule Name` becomes `feature_id`, and
transitions are identified by precursor/product name or *m/z*.

```{r, results = "asis"}
show_head(
  "Skyline_MoleculeTransitionResults.csv",
  cols = c("Molecule Name", "Replicate Name", "Precursor Mz", "Product Mz", "Retention Time", "Area")
)
```

### Generic wide CSV {#str-csv-wide}

One row per sample, one column per feature; the column name is the `feature_id`
and the values are of the type declared by `variable_name`.

```{r, results = "asis"}
show_head("plain_wide_dataset.csv")
```

### Generic long CSV {#str-csv-long}

One row per `(sample, feature)` measurement. Columns are auto-detected by name
(case-insensitive); the bundled example uses INTEGRATOR-style headers such as
`raw_data_filename` and `rt_apex`, so it is imported with a `column_mapping`.

```{r, results = "asis"}
show_head(
  "plain_long_dataset.csv",
  cols = c("feature_id", "raw_data_filename", "sample_type", "batch", "rt_apex", "area")
)
```

## Multiple files: import and merging

Multiple data files can be imported and merged by passing either a list of file
paths or a folder path, in which case all data files in the directory are
imported. This is useful when raw data processing is divided into batches that
produce separate result files. The merged data is checked for consistency, so that
each `analysis_id`–`feature_id` pair is unique and the same feature cannot be
reported more than once within an analysis.

## Next steps

- [Importing Metadata](manual-05-metadata.html) — attach and validate sample and feature annotations
- [The MRMhubExperiment Data Object](manual-02-data-object.html) — the object the data is loaded into
- [Basic MRMhub Workflow](tutorial-02-basic-workflow.html) — start processing the imported data
- [Function reference](../reference/index.html) — full arguments for every importer
