---
title: "Installation"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Install mrmhub, verify your setup, and fix common install errors.
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

## Requirements

- **R ≥ 4.2** — install from [CRAN](https://cran.r-project.org)
- **An IDE** — [Positron](https://positron.posit.co), [RStudio](https://posit.co/download/rstudio-desktop/), or [VS Code](https://code.visualstudio.com)

## Installation

<div class="callout callout-tip">
<strong>Tip.</strong> Quit and reopen your IDE first — most install failures come from packages already loaded in memory.
</div>

In a fresh R session:

```{r, eval=FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/mrmhub")
```

The same command works on Windows, macOS, and Linux. `pak` pulls prebuilt binaries where available, so compilation tools are usually not needed. If `pak` is unavailable, use `remotes::install_github("SLINGhub/mrmhub")`.

## Verify Your Setup

```{r, eval=FALSE}
library(mrmhub)
check_setup()
```

This checks your R version and dependencies and reports exactly what's missing if anything fails.

## Troubleshooting Common Errors

| Error | Cause | Fix |
|-------|-------|-----|
| `namespace 'rlang' is already loaded` | An old dependency is still loaded in the session | Restart R (`Ctrl+Shift+F10`) and retry the install |
| `cannot open URL` | Firewall or proxy blocking GitHub | `options(download.file.method = "libcurl")`, or clone the repo and use `remotes::install_local("path/to/MRMhub")` |
| `package 'X' was installed under R version …` | Package built for a different R | `install.packages("X")` to rebuild it for your R |
| `there is no package called 'mrmhub'` | Install did not finish | Scroll up for the real error, then retry the install |
| `ERROR: Rtools is required` (Windows) | A source-only dependency needs compilation | Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) matching your R version, restart R, retry |
| `clang: error: ...` (macOS) | Compiler tools missing | Run `xcode-select --install` in Terminal, retry |
| `cannot find -lcurl` (Linux) | System libraries missing | `sudo apt install libcurl4-openssl-dev libxml2-dev libssl-dev libfontconfig1-dev` |

The longer list — proxies, mixed R installations on macOS, INTEGRATOR security warnings — lives in [Troubleshooting & FAQ](manual-09-troubleshooting.html).


## Support

File issues or questions on [GitHub](https://github.com/SLINGhub/MRMhub/issues), or contact the authors directly.


