---
title: "Import and prepare data files"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  A decision guide for choosing the right data and metadata import functions in MRMhub.
vignette: >
  %\VignetteIndexEntry{Import and prepare data files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
```

Loading an experiment takes two steps: import the **peak table** produced by the
integration software, then attach the **metadata** that describes samples, features,
and standards. MRMhub stores everything internally in **long** format (one row per
sample × feature); **wide** files (one row per sample, one column per feature) are
pivoted automatically on import.

::: {.callout .callout-getstarted}

**Start from a blank template.** Save a ready-made metadata workbook, fill it in, and
load it back; there is no need to build the tables from scratch.

```{r, eval = FALSE}
save_metadata_templates()              # Individual metadata tables (XLSX)
save_metadata_msorganiser_template()   # msorganizer metadata template (XLSX)
```

:::

## Data import

Pick the importer that matches the software that produced the peak table. Each reads
a CSV/TSV file and returns an `MRMhubExperiment`.

Click a **Format** to see the file structure and column meanings.

| Source | Format | Function |
|---|---|---|
| MRMhub / INTEGRATOR *(recommended)* | [Long CSV / TSV](manual-05b-data-file-structures.html#str-mrmhub) | [`import_data_mrmhub()`](../reference/import_data_mrmhub.html) |
| Agilent MassHunter Quant | [Wide CSV](manual-05b-data-file-structures.html#str-masshunter) | [`import_data_masshunter()`](../reference/import_data_masshunter.html) |
| Skyline (small molecule) | [Long CSV](manual-05b-data-file-structures.html#str-skyline) | [`import_data_skyline()`](../reference/import_data_skyline.html) |
| Generic spreadsheet (samples × features) | [Wide CSV](manual-05b-data-file-structures.html#str-csv-wide) | [`import_data_csv_wide()`](../reference/import_data_csv_wide.html) |
| Generic table (one row per measurement) | [Long CSV](manual-05b-data-file-structures.html#str-csv-long) | [`import_data_csv_long()`](../reference/import_data_csv_long.html) |
| mzTab-M (Lipid Data Analyzer, MS-DIAL, MZmine) | [mzTab-M 2.0](https://github.com/HUPO-PSI/mzTab-M) | [`import_data_mztab()`](../reference/import_data_mztab.html) |

## Metadata import

Annotate the loaded peak table with metadata. Every importer accepts the same three
input modes — a **CSV file** (`path=`), a **sheet in an Excel workbook** (`path=` +
`sheet=`), or an in-memory **`data.frame`** (`table=`).

Click a **Structure** to see the table columns and what each one is for.

| Metadata table | Structure | Function |
|---|---|---|
| Analyses (samples) | [CSV · Excel · data.frame](manual-06a-metadata-table-structures.html#str-analyses) | [`import_metadata_analyses()`](../reference/import_metadata_analyses.html) |
| Features | [CSV · Excel · data.frame](manual-06a-metadata-table-structures.html#str-features) | [`import_metadata_features()`](../reference/import_metadata_features.html) |
| Internal standards (ISTDs) | [CSV · Excel · data.frame](manual-06a-metadata-table-structures.html#str-istds) | [`import_metadata_istds()`](../reference/import_metadata_istds.html) |
| Response curves | [CSV · Excel · data.frame](manual-06a-metadata-table-structures.html#str-responsecurves) | [`import_metadata_responsecurves()`](../reference/import_metadata_responsecurves.html) |
| Calibration / QC concentrations | [CSV · Excel · data.frame](manual-06a-metadata-table-structures.html#str-qcconcentrations) | [`import_metadata_qcconcentrations()`](../reference/import_metadata_qcconcentrations.html) |
| All tables at once | [MSOrganiser template (.xlsx)](manual-06a-metadata-table-structures.html#str-msorganiser) | [`import_metadata_msorganiser()`](../reference/import_metadata_msorganiser.html) |

A minimal import looks like this:

```{r, eval = FALSE}
exp <- MRMhubExperiment() |> import_data_mrmhub(path = "datasets/integrator_output.tsv")
exp <- import_metadata_msorganiser(exp, path = "datasets/metadata.xlsx")
```

## Next Steps

- [Data file structures](manual-05b-data-file-structures.html) — per-importer file layout and column reference
- [Metadata table structures](manual-06a-metadata-table-structures.html) — per-table columns and what each is for
- [Data Import (detailed)](manual-05-data-import.html) — per-function arguments, column expectations, and troubleshooting
- [Metadata Import](manual-06-metadata-import.html) — required columns and validation
- [Key Concepts](manual-00-key-concepts.html) — terminology reference
