---
title: "Troubleshooting and FAQ"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Solutions to the most common errors and questions when using MRMhub QUANT.
vignette: >
  %\VignetteIndexEntry{Troubleshooting and FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Installation Issues

The single most common cause of failed installations is that one of MRMhub's dependencies is
currently *loaded* in the active R session — typically `rlang`, `cli`, `vctrs`, `dplyr`, `tibble`,
or `ggplot2`. RStudio and Positron load several of these on startup (Environment pane, Plots
pane), so a "fresh session" via *Restart R* is sometimes not enough on Windows. When in doubt,
**quit the IDE entirely** and reopen it before installing.

<details>
<summary><strong>"cannot remove prior installation of package 'X'" or "permission denied"</strong> (Windows, most common)</summary>

A dependency DLL is locked by another running R session, RStudio, or Positron. Restarting R does
not always release the lock on Windows.

```{r}
# Fix:
# 1. Quit RStudio / Positron entirely (close all windows, not just the project)
# 2. Make sure no other R.exe / Rterm.exe is running in Task Manager
# 3. Reopen RStudio / Positron in a new session (NOT loading a project that
#    auto-attaches mrmhub or any tidyverse package via an .Rprofile)
# 4. Immediately, before opening any project, run the install command
remotes::install_github("SLINGhub/mrmhub")
```

If the message names a specific package (e.g. `'rlang'` is locked), install
just that package first into a fresh session, then retry MRMhub.

</details>

<details>
<summary><strong>"namespace 'rlang' is already loaded"</strong></summary>

A package that other packages depend on is locked into the current namespace. Quitting RStudio
entirely (see above) usually resolves it; the lighter touch is:

```{r}
# 1. Restart R: Session > Restart R (or Ctrl+Shift+F10)
# 2. Update the locked package first in the fresh session
install.packages("rlang")
# 3. Then retry mrmhub installation
```

</details>

<details>
<summary><strong>"Rtools is required to build R packages"</strong> (Windows)</summary>

Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) for your R version (Rtools43 for
R 4.3, Rtools44 for R 4.4, ...), then restart R and retry. Rtools is needed for compiling source
packages, which some MRMhub dependencies require when a Windows binary is not yet available.

</details>

<details>
<summary><strong>Use <code>pak</code> instead of <code>remotes</code></strong></summary>

[`pak`](https://pak.r-lib.org/) handles locked packages more cleanly than `remotes` on Windows,
gives clearer error messages on transitive failures, and parallelises downloads. If `remotes`
fails:

```{r}
install.packages("pak")
pak::pak("SLINGhub/mrmhub")
```

`pak` automatically restarts R if needed to release locks, and resolves the dependency graph
in one pass.

</details>

<details>
<summary><strong>"cannot open URL" or timeout during install</strong></summary>

The network is blocking GitHub or a CRAN mirror.

```{r}
# Option 1: change download method
options(download.file.method = "libcurl")

# Option 2: clone locally and install
# In terminal: git clone https://github.com/SLINGhub/MRMhub.git
remotes::install_local("path/to/MRMhub")

# Option 3: if behind a corporate proxy, configure it before installing
Sys.setenv(http_proxy  = "http://proxy.example.com:8080")
Sys.setenv(https_proxy = "http://proxy.example.com:8080")
```

</details>

<details>
<summary><strong>"installation of package 'X' had non-zero exit status"</strong> (Suggests dependency failure)</summary>

By default `remotes::install_github(..., dependencies = TRUE)` installs every Suggests dependency
too. A single broken Suggests package — e.g. `lancer` (a non-CRAN GitHub package required only for
LICAR-style interference correction) when the network blocks GitHub clones — aborts the whole
install. Fix:

```{r}
# Install only the strictly required dependencies
remotes::install_github(
  "SLINGhub/mrmhub",
  dependencies = c("Depends", "Imports", "LinkingTo")
)
```

Optional packages can be added later as needed: `install.packages(c("patchwork", "ggrepel"))`.

</details>

<details>
<summary><strong>Apple Silicon: package compilation fails / "no binary available"</strong></summary>

On Apple Silicon, if a dependency is not yet built as an `arm64` binary on CRAN, R falls back to
building from source, which then requires Xcode Command Line Tools. Two fixes:

```{r}
# Option A: install Xcode Command Line Tools (run in terminal, not R)
# xcode-select --install

# Option B: ask R to skip source builds when only a binary is missing
options(install.packages.compile.from.source = "never")
remotes::install_github("SLINGhub/mrmhub")
```

Option B may leave slightly older versions of those packages but avoids the toolchain
entirely.

</details>

---

## Data Import Issues

<details>
<summary><strong>"Column X not found" or unexpected column names</strong></summary>

The CSV uses different column names than expected. Check with:

```{r}
# Read the first few rows to inspect
readr::read_csv("your_file.csv", n_max = 5)
```

Common causes:

- MassHunter exports have locale-specific headers (e.g., German vs English)
- Extra BOM characters in UTF-8 files (open in a plain text editor to check)
- Delimiter mismatch: file uses `;` but R expects `,`

```{r}
# For semicolon-separated files:
readr::read_csv2("your_file.csv", n_max = 5)
```

</details>

<details>
<summary><strong>"No features found" after import</strong></summary>

The feature names in the data file do not match those in the feature annotation.
Check for:

- Trailing whitespace: `"Compound A "` vs `"Compound A"`
- Case differences: `"compound_a"` vs `"Compound_A"`

```{r}
# Debug: compare feature names
data_features <- unique(mexp@dataset$feature_id)
annot_features <- mexp@annot_features$feature_id

# Find mismatches
setdiff(data_features, annot_features)
setdiff(annot_features, data_features)
```

</details>

<details>
<summary><strong>Which importer should I use?</strong></summary>

See [Import and prepare data files](manual-04-data-import.html) for a visual decision flowchart.

</details>

---

## Processing Issues

<details>
<summary><strong>All values are NA after normalisation</strong></summary>

Most likely cause: the ISTD assignment is missing or incorrect.

```{r}
# Check ISTD assignments
mexp@annot_features |>
  dplyr::select(feature_id, istd_feature_id) |>
  dplyr::filter(is.na(istd_feature_id))
```

If `istd_feature_id` is `NA`, normalization divides by nothing → `NA` result.

</details>

<details>
<summary><strong>Drift correction makes data worse</strong></summary>

This can happen when:

- Too few QC samples (need ≥ 5 per batch for loess)
- QC samples are not evenly spaced across the run
- The drift is not systematic (random noise, not a trend)

```{r}
# Visualise before correcting
plot_runscatter(mexp,
                variable = "norm_intensity",
                include_feature_filter = "your_feature",
                qc_types = c("BQC", "SPL"))
```

If the QC trend is flat, skip drift correction for that batch.

</details>

<details>
<summary><strong>Batch correction flattens real biological signal</strong></summary>

`correct_batch_centering()` assumes QC samples have equal concentration
across batches. If the QC pool differs between batches, centering will
introduce artefacts.

**Solution:** Ensure the same QC pool is used across all batches.

</details>

---

## Object and Data Access

<details>
<summary><strong>How do I extract the final data as a data frame?</strong></summary>

```{r}
# After all processing and filtering:
final_data <- get_analyticaldata(mexp, annotated = TRUE)

# Or the filtered dataset directly:
final_filtered <- mexp@dataset_filtered
```

</details>

<details>
<summary><strong>How do I check what processing has been applied?</strong></summary>

```{r}
# Check processing status flags
mexp@is_istd_normalized
mexp@is_quantitated
mexp@var_drift_corrected
mexp@var_batch_corrected
```

</details>

<details>
<summary><strong>How do I go back to the original data?</strong></summary>

The original data is always preserved:

```{r}
# Reset to original
mexp@dataset <- mexp@dataset_orig
```

</details>

---

## QC and Filtering

<details>
<summary><strong>What do the QC metrics mean?</strong></summary>

| Metric | Meaning | Typical threshold |
|--------|---------|-------------------|
| CV (%) | Coefficient of variation in QC samples | < 20–30% |
| Bias (%) | Systematic deviation from expected value | < 20% |
| n_detected | Number of QC samples with signal | ≥ 50–67% of QCs |

</details>

<details>
<summary><strong>My features are all filtered out</strong></summary>

The thresholds may be too strict. Check:

```{r}
# See QC metrics before filtering
mexp@metrics_qc |>
  dplyr::arrange(dplyr::desc(norm_intensity_cv_bqc))

# Use more lenient thresholds
mexp <- filter_features_qc(
  mexp,
  include_qualifier = FALSE,
  include_istd = FALSE,
  max.cv.normintensity.bqc = 40,
  max.prop.missing.normintensity.spl = 0.5
)
```

</details>

---

## Calibration

<details>
<summary><strong>"No calibration points found"</strong></summary>

The calibration annotation does not match sample labels in the data.
Check that `annot_qcconcentrations` uses the same `analysis_id` values
as the data.

</details>

<details>
<summary><strong>Calibration curve has R² < 0.9</strong></summary>

Consider:

- Removing outlier calibration points
- Using a different weighting scheme (1/x or 1/x²)
- Checking for saturation at high concentrations

```{r}
# Inspect calibration metrics
get_calibration_metrics(mexp) |>
  dplyr::filter(r2 < 0.9)
```

</details>

---

## FAQ

<details>
<summary><strong>Can I use MRMhub with non-MRM data (e.g., PRM, SRM)?</strong></summary>

Yes, provided the data can be formatted as a long CSV with feature IDs
and peak areas. The processing is agnostic to acquisition mode.

</details>

<details>
<summary><strong>Can I process only one batch?</strong></summary>

Yes. Do not apply batch correction. All other functions work with
single-batch data.

</details>

<details>
<summary><strong>Is there a size limit?</strong></summary>

No hard limit, but very large datasets (>100k rows) may be slow for drift
correction. Consider processing batches separately and merging.

</details>

<details>
<summary><strong>Where do I report bugs?</strong></summary>

[GitHub Issues](https://github.com/SLINGhub/MRMhub/issues)

</details>

---

## Next Steps

- [Design Overview](manual-03-design-decisions.html) — understand the full pipeline
- [Installation](manual-00-installation.html) — verify your setup with `check_setup()`
