---
title: "Installation"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Install mrmhub, verify your setup, and fix common install errors.
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

<p class="page-kind page-kind-manual">Manual</p>

## Requirements

- **Operating system:** Windows, macOS, or Linux.
- **R:** version 4.1 or newer, from [CRAN](https://cran.r-project.org). Use the
  current release or the one before it: CRAN builds binary packages only for
  those two, so on an older R some dependencies are compiled from source, which
  needs [Rtools](https://cran.r-project.org/bin/windows/Rtools/) (Windows) or
  the Xcode command line tools (macOS).
- **R Editor (recommended):** [RStudio](https://posit.co/download/rstudio-desktop/)
  or [Positron](https://positron.posit.co).
- **Quarto:** bundled with recent RStudio and Positron releases; otherwise
  install it from [quarto.org](https://quarto.org/docs/get-started/). Required
  only for rendering notebooks.

## Installing and updating MRMhub

::: {.callout .callout-caution}
**Important.** Quit and reopen all RStudio/Positron sessions first. Many installation issues 
are caused by R packages that are loaded in another session while the installer tries to update them.
:::

In a **fresh R session**, we recommend installing with `pak`:

```{r, eval=FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/MRMhub")
```

To update MRMhub later, re-run the same command in a fresh session.

::: {.callout .callout-note}
**No compiler needed.** `mrmhub` contains no compiled code, and CRAN provides
binaries for all its dependencies. If the installer starts compiling anyway, or
asks *"install from sources the package which needs compilation?"*, decline —
or set, before installing:

```{r, eval=FALSE}
options(install.packages.compile.from.source = "never")
```
:::

If `pak` fails, e.g. due to institutional network firewall or
proxy settings, try `remotes` instead:

```{r, eval=FALSE}
if (!require("remotes")) install.packages("remotes")
remotes::install_github("SLINGhub/MRMhub")
```

If installation fails, see the ZIP-file method below or [Troubleshooting](#troubleshooting).

## Installing from a downloaded repository ZIP file

If the methods above fail or R cannot reach GitHub, download a copy of the
repository and install MRMhub from it:

1. Open <https://github.com/SLINGhub/MRMhub> in a browser.
2. Click the green **Code** button and choose **Download ZIP** (or download
   <https://github.com/SLINGhub/MRMhub/archive/refs/heads/main.zip> directly).
3. Unzip the file. It expands to a folder named `MRMhub-main`.
4. In a **fresh R session**, install from that folder with `pak`:

```{r, eval=FALSE}
pak::local_install("path/to/MRMhub-main")
```

Replace `path/to/MRMhub-main` with the actual location of the unzipped folder,
e.g. `"~/Downloads/MRMhub-main"`.

## Confirm `mrmhub` is successfully installed

```{r, eval=FALSE}
library(mrmhub)
```

If this loads without error, you are ready to go. If you plan to use any of the
functions listed under [Optional packages](#optional-packages) below, install
those too. If it loads with errors, see [Troubleshooting](#troubleshooting) below.

## Optional packages

To keep the base installation light, a number of specialised functions in
`mrmhub` rely on additional packages that are not installed upfront. When such a
function is called and its package is not yet installed, `mrmhub` offers to
install it.

| Function | Optional package | Enables |
|----------|------------------|---------|
| `plot_runscatter()` | qpdf; mirai, carrier | only for multi-threaded PDF output |
| `plot_qc_summary_overall(with_venn = TRUE)` | ggvenn, patchwork | Venn diagram of features excluded by QC criteria |
| `plot_matrixeffects()`, `plot_interference_correction()` | ggbeeswarm | beeswarm/quasirandom point layers |
| `correct_isotopic_interferences()`, `calc_average_molweight()` | enviPat | isotope-pattern and molecular-weight calculation |
| `correct_batch_combat()` | sva | ComBat batch correction |
| `correct_batch_serrf()` | ranger | SERRF batch correction |
| `correct_drift_gam()` | mgcv | GAM-based drift correction |
| `build_workflow()` | shiny, bslib | interactive workflow-builder app |
| `save_dataset_summarizedexperiment()` | SummarizedExperiment, S4Vectors, lipidr *(Bioconductor)* | export to a `SummarizedExperiment` |
| Lipid-name parsing (isotope correction, lipid plots) | rgoslin *(Bioconductor)* | parse and normalise lipid shorthand |
| `get_response_curve_stats()` | lancer *(GitHub)* | only for specific response-curve metrics |

To install all of them upfront, run the following in a fresh R session
(again, with all other RStudio/Positron sessions closed):

```{r, eval=FALSE}
pak::pak(c(
  "sva", "ranger", "mgcv", "enviPat",
  "qpdf", "mirai", "carrier", "shiny", "bslib",
  "ggvenn", "patchwork", "ggbeeswarm",
  "rgoslin", "lipidr", "SummarizedExperiment", "S4Vectors",
  "SLINGhub/lancer"
))
```

Alternatively, the CRAN packages can be installed with `install.packages()`. For
the Bioconductor ones (`sva`, `rgoslin`, `lipidr`, `SummarizedExperiment`,
`S4Vectors`), use `pak` as shown above: it selects the Bioconductor release
paired with the installed R version. `BiocManager` may instead resolve to an
older release and report that the current one requires a newer R — that message
is advisory, and upgrading R is not necessary.

## Troubleshooting

If the installation seemed successful but the package does not load, or loads
with errors, run `check_setup()` to report the R version and flag any missing
dependencies:

```{r, eval=FALSE}
mrmhub::check_setup()
```

**Frequent installation errors**

See [Troubleshooting & FAQ](manual-10-troubleshooting.html) for a detailed
list of errors and resolutions.

| Error | Cause | Fix |
|--------------------------|--------------------------|-------------------|
| `namespace 'rlang' is already loaded` | An old dependency is still loaded in the session | Restart R (`Ctrl+Shift+F10`) and retry the install |
| `pak` unavailable or failing | `pak` not installed, or its cache is stale | Install with `remotes::install_github("SLINGhub/MRMhub")` |
| `cannot open URL` | Firewall or proxy blocking GitHub | `options(download.file.method = "libcurl")`, or [install from a downloaded repository ZIP file](#installing-from-a-downloaded-repository-zip-file) |
| `Could not resolve host: api.github.com` | Firewall or proxy blocking the GitHub API | Set `http_proxy`/`https_proxy`, or [install from a downloaded repository ZIP file](#installing-from-a-downloaded-repository-zip-file) |
| `SSL certificate problem: self signed certificate in certificate chain` | Corporate proxy inspecting TLS traffic | Use `remotes` (it uses the system certificate store), or install your organisation's root certificate |
| `package 'X' was installed under R version …` | Package built for a different R | `install.packages("X")` to rebuild it for your R |
| `there is no package called 'mrmhub'` | Install did not finish | Scroll up for the real error, then retry the install |
| `ERROR: Rtools is required`, or packages start compiling (Windows) | A dependency has no binary for your R version | Update R, or set `options(install.packages.compile.from.source = "never")`. If a source build is unavoidable, install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) matching your R version |
| `clang: error: ...` (macOS) | Compiler tools missing | Run `xcode-select --install` in Terminal, retry |
| `cannot find -lcurl` (Linux) | System libraries missing | `sudo apt install libcurl4-openssl-dev libxml2-dev libssl-dev libfontconfig1-dev` |

## Next steps

- [MRMhub overview](manual-01-key-concepts.html): core
  vocabulary and the MRMhubExperiment object
- [Getting started with MRMhub](tutorial-02-getting-started-mrmhub.html): a short
  end-to-end walkthrough
- Questions or bug reports? File an issue on
  [GitHub](https://github.com/SLINGhub/MRMhub/issues), or contact the
  authors directly.