---
title: "Visualisation Functions"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  Reference for the MRMhub plotting functions, grouped by workflow stage,
  with the canonical argument forms and customisation guidance.
vignette: >
  %\VignetteIndexEntry{Visualisation Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      fig.width = 7, fig.height = 5, out.width = "100%")
```

MRMhub provides plotting functions covering every stage of the workflow. All functions return
`ggplot2` objects (or, for paged outputs, a list of `ggplot2` objects), so the standard ggplot2
grammar can be used to further customise titles, themes, scales, and facets. This reference groups
the functions by the workflow stage at which they are most useful.

## Setup

```{r}
#| results: false
#| message: false
library(mrmhub)
library(ggplot2)

mexp <- readRDS("results/mexp_processed.rds")
```

## Overview

| Function | Stage | Purpose |
|----------|-------|---------|
| `plot_runsequence()` | Acquisition design | Run-order overview of QC types and batches |
| `plot_runscatter()` | Drift / batch QC | Per-feature scatter of values vs analysis order |
| `plot_abundanceprofile()` | Abundance overview | Feature abundance distribution per class |
| `plot_rla_boxplot()` | Normalisation QC | Relative log-abundance boxplots per analysis |
| `plot_normalization_qc()` | Normalisation QC | Before/after ISTD normalisation comparison |
| `plot_pca()` | Multivariate QC | PCA score plot |
| `plot_pca_loading()` | Multivariate QC | PCA loadings |
| `plot_feature_correlations()` | Multivariate QC | Feature-feature correlation heatmap |
| `plot_qcmetrics_comparison()` | QC metrics | CV / bias comparison across QC types |
| `plot_qc_summary_byclass()` | QC filtering | Pass/fail summary by feature class |
| `plot_qc_summary_overall()` | QC filtering | Pass/fail summary across the whole dataset |
| `plot_calibrationcurves()` | External calibration | Calibration curves with fit and residuals |
| `plot_responsecurves()` | Response curves | Linearity check from a dilution series |
| `plot_rt_vs_chain()` | Method check (lipidomics) | RT vs chain length / unsaturation |
| `plot_qc_matrixeffects()` | Method check | Matrix-effect QC |
| `plot_qc_interferences()` | Interference correction | QC overview for interference annotations |

# Acquisition design

## `plot_runsequence()` — sequence overview

The run-sequence plot summarises the acquisition design: QC type positions, batch boundaries, and
the analysis timeline. It is an experiment-level plot (no per-feature dimension).

```{r}
plot_runsequence(mexp,
                 qc_types = NA,
                 show_batches = TRUE,
                 batch_zebra_stripe = TRUE,
                 show_timestamp = FALSE)
```

Set `show_timestamp = TRUE` to use the acquisition timestamp on the x-axis — useful for
identifying interruptions between or within batches that would not be visible against the analysis
sequence number.

# Drift, batch, and per-feature inspection

## `plot_runscatter()` — values vs analysis order

The work-horse plot for drift and batch QC. Plots a feature variable (intensity, normalised
intensity, concentration) against the analysis order, optionally with fitted drift trends from the
correction functions. Returns one panel per feature (paged).

```{r}
plot_runscatter(mexp,
                variable = "norm_intensity",
                qc_types = c("BQC", "SPL"),
                rows_page = 1,
                cols_page = 3,
                show_trend = TRUE)
```

Use `variable = "intensity_before"` / `"conc_before"` to inspect pre-correction values, and
`variable = "intensity"` / `"conc"` for post-correction. Setting `output_pdf = TRUE` and a `path`
writes a multi-page PDF — practical for large feature lists.

## `plot_abundanceprofile()` — feature abundance distribution

Shows the abundance distribution of each feature, grouped by class. Useful for inspecting class
coverage and identifying features at the limits of the dynamic range.

```{r}
plot_abundanceprofile(mexp,
                      variable = "intensity",
                      qc_types = "SPL",
                      log_scale = TRUE)
```

When `use_qc_metrics = TRUE`, the function reads from the pre-computed `metrics_qc` table (much
faster on large feature lists) and `qc_types` must specify a single QC type.

# Normalisation QC

## `plot_rla_boxplot()` — relative log-abundance per analysis

A boxplot per analysis of `log2(value / median_across_analyses)`. Width and centring of the box
reflect injection-level variability. Useful before and after normalisation to confirm that ISTD
correction removes injection-level bias.

```{r}
plot_rla_boxplot(mexp,
                 variable = "norm_intensity",
                 qc_types = c("BQC", "SPL"))
```

## `plot_normalization_qc()` — before-vs-after comparison

Compares pre- and post-normalisation values for QC samples in three layouts (`plot_type`):
`"scatter"` (point cloud), `"diff"` (signed difference), `"ratio"` (after / before). CV of QC
samples should decrease after normalisation; an increase typically indicates an incorrect ISTD
pairing.

```{r}
plot_normalization_qc(mexp,
                      before_norm_var = "intensity",
                      after_norm_var  = "norm_intensity",
                      plot_type       = "scatter",
                      qc_types        = c("BQC", "SPL"),
                      facet_by_class  = TRUE,
                      cv_threshold_value = 25)
```

# Multivariate QC

## `plot_pca()` — score plot

Two-dimensional PCA score plot with confidence ellipses grouped by `qc_type`, `batch_id`, or
`"none"`. See [QC Exploration with PCA](tutorial-09-pca-exploration.html) for an interpretation
walk-through.

```{r}
plot_pca(mexp,
         variable          = "norm_intensity",
         qc_types          = c("BQC", "SPL"),
         ellipse_variable  = "qc_type")
```

## `plot_pca_loading()` — loadings

Loadings on the first two PCs. Features at the extremes drive sample separation; a single feature
dominating PC1 should be inspected with `plot_runscatter()` before being attributed to biology.

```{r}
plot_pca_loading(mexp,
                 variable = "norm_intensity",
                 qc_types = c("BQC", "SPL"))
```

## `plot_feature_correlations()` — correlation heatmap

Pairwise correlation matrix across features, useful for identifying redundant transitions and
flagging candidate isobaric interferences.

```{r}
plot_feature_correlations(mexp,
                          variable = "norm_intensity",
                          qc_types = "SPL")
```

# QC metrics and filtering

## `plot_qcmetrics_comparison()` — pairwise QC-metric scatter

Plots one QC metric against another, computed across features. The variable names follow the
pattern `{variable}_cv_{qctype}` (e.g. `intensity_cv_bqc`, `norm_intensity_cv_tqc`) and live in
`mexp@metrics_qc` after `calc_qc_metrics()` is run.

```{r}
mexp <- calc_qc_metrics(mexp)

plot_qcmetrics_comparison(mexp,
                          plot_type        = "scatter",
                          x_variable       = "intensity_cv_bqc",
                          y_variable       = "norm_intensity_cv_bqc",
                          equality_line    = TRUE,
                          threshold_values = 25,
                          log_scale        = FALSE)
```

Points below the equality line indicate features for which CV was reduced by normalisation; points
above indicate features made worse — typically a misassigned ISTD.

## `plot_qc_summary_byclass()` / `plot_qc_summary_overall()` — filter outcome

After `filter_features_qc()`, these two functions summarise how many features passed each filter
rule, broken down by feature class (`_byclass`) or aggregated (`_overall`).

```{r}
plot_qc_summary_overall(mexp)
plot_qc_summary_byclass(mexp)
```

# External calibration and response curves

## `plot_calibrationcurves()` — calibration fits

After `quantify_by_calibration()`, plots calibration response vs concentration with the fitted
model, the calibration points, and points excluded from the fit. Returns a paged grid of features.

```{r}
plot_calibrationcurves(mexp,
                       variable = "norm_intensity",
                       qc_types = NA)
```

The fit model is taken from the calibration setup unless overridden with `fit_overwrite =
"linear"` or `"quadratic"`.

## `plot_responsecurves()` — RQC linearity

Plots feature response across an RQC dilution series. The fitted slope and R² values report on the
linearity of the assay in the range of the QC pool. See `get_response_curve_stats()` for the
numeric summary.

```{r}
plot_responsecurves(mexp,
                    variable = "intensity")
```

# Method-specific checks

## `plot_rt_vs_chain()` — retention time vs chain length (lipidomics)

For lipidomics methods using class-based chromatography, plots RT against carbon number, separated
by class. Outliers from the expected linear relationship within a class point to mis-identified
features.

```{r}
plot_rt_vs_chain(mexp)
```

## `plot_qc_matrixeffects()` — matrix-effect overview

Compares ISTD response in matrix-containing QCs against solvent-only injections to flag
matrix-effect outliers.

```{r}
plot_qc_matrixeffects(mexp)
```

## `plot_qc_interferences()` — interference annotations

QC overview for features with interference annotations (see
[Interference Correction](tutorial-11-interference-correction.html)).

```{r}
plot_qc_interferences(mexp)
```

# Customisation and export

## ggplot2 layering

All functions return `ggplot2` objects, so themes, scales, and titles can be appended in the usual
way.

```{r}
plot_runscatter(mexp, variable = "norm_intensity", qc_types = c("BQC", "SPL")) +
  ggplot2::theme_minimal(base_size = 9) +
  ggplot2::labs(title = "Normalised intensity (BQC vs SPL)")
```

## Saving plots

For a single plot:

```{r}
p <- plot_pca(mexp,
              variable = "norm_intensity",
              qc_types = c("BQC", "SPL"),
              ellipse_variable = "batch_id")
ggplot2::ggsave("figures/pca_batch.png", p, width = 8, height = 6, dpi = 300)
```

For paged outputs (`plot_runscatter()`, `plot_calibrationcurves()`), use the function's built-in
`output_pdf = TRUE` and `path` arguments to write a multi-page PDF directly, rather than iterating
in user code.

```{r}
plot_runscatter(mexp,
                variable   = "norm_intensity",
                qc_types   = c("BQC", "SPL"),
                output_pdf = TRUE,
                path       = "figures/runscatter_all.pdf")
```

## Combining plots

`patchwork` composes multiple panels into a single figure:

```{r}
library(patchwork)

p_seq  <- plot_runsequence(mexp)
p_pca  <- plot_pca(mexp, variable = "norm_intensity",
                   qc_types = c("BQC", "SPL"), ellipse_variable = "batch_id")
p_rla  <- plot_rla_boxplot(mexp, variable = "norm_intensity",
                           qc_types = c("BQC", "SPL"))

p_seq / (p_pca | p_rla)
```

## See Also

- [Run-Scatter Plot (tutorial)](tutorial-05-run-scatter.html) — detailed `plot_runscatter()` usage
- [QC Exploration with PCA](tutorial-09-pca-exploration.html) — PCA interpretation
- [Drift Correction (tutorial)](tutorial-04-drift-correction.html) — `plot_runscatter()` in context
- [External Calibration & QC (recipe)](recipe-01-ext-calibration-qc.html) — calibration plots in workflow
- [Custom QC Report (recipe)](recipe-02-custom-qc-report.html) — combine plots in a report
