---
title: "Glossary"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
description: |
  Definitions of the analytical terms used throughout the MRMhub documentation.
vignette: >
  %\VignetteIndexEntry{Glossary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", out.width = "100%",
  message = FALSE, warning = FALSE
)
```

<p class="page-kind page-kind-manual">Manual</p>

Definitions of the analytical terms used throughout the documentation. Identifiers
and variable names within the data object are described in
[The MRMhubExperiment data object](manual-02-data-object.html).

```{r, echo=FALSE, results='asis'}
glossary <- data.frame(
  Term = c(
    "Analysis", "Feature", "Analyte", "Quantifier", "Qualifier",
    "Internal standard (ISTD)", "Study sample", "QC sample",
    "MRM / SRM", "Transition", "Precursor / MS1", "Product / MS2",
    "Peak area", "Normalization", "Isotopic interference", "Drift",
    "Batch effect", "Calibration curve", "Response curve", "%CV",
    "Long / wide format"
  ),
  Definition = c(
    "A single injection or sample in a run; the unit that carries a QC role.",
    "A distinct signal extracted from the MS data for one compound (e.g. a chromatographic peak for one transition); what MRMhub measures.",
    "The compound a feature quantifies. Not one-to-one with features: one analyte can give several (isotopes, adducts, transitions), and isobaric/isomeric compounds can share a feature.",
    "The MRM transition used to measure a compound's concentration.",
    "A secondary MRM transition used to confirm a compound's identity.",
    "A compound (usually stable-isotope-labelled) spiked at a known amount into every sample, used as a reference for normalization and quantitation.",
    "An unknown — the sample whose concentrations are reported.",
    "A quality-control injection (pooled QC, dilution series, reference material, or blank) used to anchor corrections and metrics rather than reported.",
    "Multiple- (Selected-) Reaction Monitoring: a targeted MS acquisition that follows a defined precursor→product ion pair for each compound.",
    "One precursor→product ion pair monitored for a compound; a compound may have a quantifier and one or more qualifier transitions.",
    "The intact analyte ion selected in the first quadrupole (MS1) before fragmentation.",
    "A fragment ion of the precursor, measured in the second stage (MS2) to give the transition its specificity.",
    "The integrated area under a chromatographic peak, the raw measure of a feature's abundance.",
    "Dividing a feature's signal by a reference (typically its ISTD) to remove technical variation between injections.",
    "Signal a labelled or isobaric compound contributes to another's transition through overlapping isotope masses; corrected before quantitation.",
    "A gradual change in signal across a run (e.g. from instrument fouling), modelled from QC samples and removed.",
    "A systematic difference in signal between analytical batches, aligned using QC samples.",
    "The relationship between known concentration and measured response, fitted from calibration standards to convert signal to concentration.",
    "A serial-dilution series of a QC pool used to assess linearity and dynamic range (not to assign concentrations).",
    "Coefficient of variation (SD / mean, as a percentage): the precision metric used to judge QC and feature quality.",
    "Long format has one row per analysis × feature (tidy); wide format has one row per analysis with features as columns."
  ),
  stringsAsFactors = FALSE
)

DT::datatable(
  glossary,
  options = list(pageLength = 25, dom = "ft", ordering = TRUE),
  rownames = FALSE,
  filter = "top",
  class = "compact stripe"
)
```

## Next steps

- [MRMhub overview](manual-01-key-concepts.html): the data object and function groups
- [The MRMhubExperiment data object](manual-02-data-object.html): the tables, identifiers, and variables in detail
- [Sample types & QC roles](manual-06-sample-types.html): the QC labels used throughout
