---
title: "Importing Metadata"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Attaching and validating sample, feature, ISTD and QC metadata on an MRMhubExperiment.
vignette: >
  %\VignetteIndexEntry{Importing Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
library(mrmhub)

if(fs::file_exists("metadata_template.xlsx")) fs::file_delete("metadata_template.xlsx")
if(fs::file_exists("metadata_msorganiser_template.xlsx")) fs::file_delete("metadata_msorganiser_template.xlsx")

# First rows of a bundled example CSV.
show_csv <- function(file, n = 4) {
  path <- system.file("extdata", file, package = "mrmhub")
  df <- readr::read_csv(path, n_max = n, show_col_types = FALSE, name_repair = "minimal")
  knitr::kable(df, caption = sprintf("Example file `%s`", file))
}

# Column headers of a sheet in the blank metadata template workbook.
show_template <- function(sheet) {
  path <- system.file("extdata", "mrmhub_metadata_templates.xlsx", package = "mrmhub")
  df <- openxlsx2::read_xlsx(path, sheet = sheet)
  knitr::kable(df, caption = sprintf("Template sheet `%s` (column headers)", sheet))
}
```

<p class="page-kind page-kind-manual">Manual</p>

Metadata in this context refers to analysis metadata, i.e., data that
annotate the analytical data. Metadata can be retrieved from the
imported analysis data file where available, or imported from separate
files or R data frames. Integrity of metadata and data is essential for
correct post-processing: MRMhub inspects imported data and metadata for
completeness and for consistency of the IDs used across the different
metadata tables, and after import a summary of the identified errors,
warnings, and notes is printed to the console so that possible issues can
be identified and addressed.

## Metadata formats and templates

The structure and the required and optional columns for each metadata type
are described below under [Metadata table structures](#metadata-table-structures)
and in the help pages of the corresponding import functions. To obtain metadata
templates, an Excel file containing all metadata table templates can be saved.

```{r}
mrmhub::save_metadata_templates()
```

## Importing metadata from files and sheets

The analysis data are first imported as outlined in
[Importing Analytical Data](manual-04-data-import.html). In this case,
metadata present in the analysis data is not imported; only the peak areas
are read.

```{r}
library(mrmhub)
mexp <- mrmhub::MRMhubExperiment()

data_path <- "datasets/sPerfect_MRMhub.tsv"
mexp <- import_data_mrmhub(data = mexp, path = data_path, import_metadata = FALSE)
```

The corresponding metadata can then be added file by file.

```{r}
mexp <- import_metadata_analyses(mexp,
                                 path = "datasets/analysis_metadata.csv",
                                 excl_unmatched_analyses = TRUE,
                                 ignore_warnings = TRUE)

mexp <- import_metadata_features(mexp,
                                 path = "datasets/feature_metadata.csv",
                                 ignore_warnings= TRUE )
```

Metadata can also be imported from sheets of an Excel workbook, which
allows all metadata to be stored in one file. Here, metadata on internal
standards and response curves is added to the MRMhubExperiment object.

```{r}
mexp <- import_metadata_istds(mexp,
                              path = "datasets/metadata_tables.xlsx",
                              sheet = "ISTDs",
                              ignore_warnings= TRUE)
```

Furthermore, metadata can be imported from R `data.frame` objects, which
allows metadata to be obtained from additional sources, e.g. databases or
a LIMS.

```{r}
df_qcinfo <- readr::read_table(file = "datasets/qc_metadata.txt")
mexp <- import_metadata_qcconcentrations(mexp, table = df_qcinfo)
```

## Importing an MSOrganiser metadata template

Another option to import metadata is via the MSOrganiser template file, an
Excel file (`.xlsx`). This template provides tables for all metadata types
supported by MRMhub, with options to check the validity and integrity of the
metadata. The template can be obtained from
<https://github.com/SLINGhub/mrmhub> or via a mrmhub function.

```{r}
mrmhub::save_metadata_msorganiser_template()
```

Only the metadata tables required by the intended processing workflow need
to be completed. The following import function imports all completed tables.

```{r}
mexp <- import_metadata_msorganiser(mexp,
                                    path = "datasets/sPerfect_Metadata.xlsx",
                                    ignore_warnings= TRUE)
```

## Metadata table structures

The previews below show example rows or the blank template headers for each
metadata table, so the tables can be prepared before import. Identifier columns
must be consistent across tables: `analysis_id` / `sample_id` match the analysis
metadata and the data, `analyte_id` links features to QC concentrations, and
`istd_feature_id` and `feature_id` reference the feature metadata. The full
column reference for each table is given on its linked function reference page.

### Analyses (samples) {#str-analyses}

One row per analysis (injection). Import with
[`import_metadata_analyses()`](../reference/import_metadata_analyses.html).

```{r, results = "asis"}
show_csv("MHQuant_demo_metadata_analyses.csv")
```

### Features (analytes) {#str-features}

One row per feature. Import with
[`import_metadata_features()`](../reference/import_metadata_features.html).

```{r, results = "asis"}
show_template("Features")
```

### Internal standards (ISTDs) {#str-istds}

One row per internal standard, giving its known concentration. Import with
[`import_metadata_istds()`](../reference/import_metadata_istds.html).

```{r, results = "asis"}
show_csv("MRMhub_ISTDconc.csv")
```

### Response curves {#str-responsecurves}

Maps response-curve injections to the amount analysed. Import with
[`import_metadata_responsecurves()`](../reference/import_metadata_responsecurves.html).

```{r, results = "asis"}
show_template("ResponseCurves")
```

### Calibration / QC concentrations {#str-qcconcentrations}

Known analyte concentrations in calibration and QC samples. Import with
[`import_metadata_qcconcentrations()`](../reference/import_metadata_qcconcentrations.html).

```{r, results = "asis"}
show_template("QCconcentrations")
```

## Identifier matching

Metadata is linked to the data by identifier: `analysis_id` joins the sample
metadata to the analyses, and `feature_id` joins the feature metadata to the
measurements. For the join to succeed, an identifier in a metadata sheet must
match the corresponding identifier in the data exactly.

To make that robust, identifiers in metadata are normalized on import in the
same way as in the data: leading and trailing spaces are removed, internal runs
of whitespace are collapsed to a single space, and raw-data file extensions
(`.mzML`, `.d`, `.raw`, `.wiff`, `.wiff2`, `.lcd`, `.chrom`) are stripped from
`analysis_id`. So `"QC  01"` in a sample sheet still matches `"QC 01"` in the
data.

::: {.callout .callout-caution}
If analyses appear to be missing after `add_metadata()`, a difference in
identifier spelling is the usual cause. Whitespace differences are always
squished on both sides and no warning is issued for them — they are handled
silently so a stray space cannot cause a hard-to-trace mapping failure. Other
differences (a raw-data extension the import does not recognize, a typo, or a
different naming scheme) are not normalized, and the affected analyses are
dropped from the join with a warning.
:::

## Next steps

- [Importing Analytical Data](manual-04-data-import.html) — importing the measurement data
- [Sample Types & QC Roles](manual-06-sample-types.html) — the `qc_type` values that drive QC logic
- [The MRMhubExperiment Data Object](manual-02-data-object.html) — where the metadata is stored
- [Function reference](../reference/index.html) — full arguments for every metadata importer
