---
title: "Data Identifiers in MRMhub"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Explanation of the key identifier columns used to organize data in MRMhub.
vignette: >
  %\VignetteIndexEntry{Data Identifiers in MRMhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
bibliography: references.bib
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, warning = FALSE,
                      message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

## Key Data Identifiers

The following fields are essential for organizing data in MRMhub. They link
the data tables within the `MRMhubExperiment` object and are used by processing
functions throughout the package.

| Table | Field | Description |
|-------|-------|-------------|
| Analyses | **`analysis_id`** | Unique identifier for each injection/analysis. |
|  | `qc_type` | Sample type label (e.g., SPL, BQC, CAL). See below. |
|  | `batch_id` | Identifies the analytical batch. |
|  | `sample_id` | Identifies the physical sample that was tested. |
| Features | **`feature_id`** | Unique identifier for each MRM transition/feature. |
|  | `istd_feature_id` | The `feature_id` of the internal standard used for normalization. |
|  | `analyte_id` | Identifies the analyte (compound). |

<div class="alert alert-info" role="alert">
<strong>Why <code>analysis_id</code> instead of <code>sample_id</code>?</strong>
A sample may be measured multiple times (replicates, different methods). Each 
injection needs a unique identifier. Similarly, one analyte can be measured 
through multiple transitions, so <code>feature_id</code> (not <code>analyte_id</code>) 
is the primary key.
</div>

## QC Types (Sample Types)

QC types categorize samples by their analytical purpose. This classification
combines nomenclature from @broadhurst2018GuidelinesConsiderationsUse (SPL, BQC,
TQC, LTR, RQC) with traditional terms from analytical and clinical chemistry
(LQC, MQC, HQC, CAL, NIST, SST, blanks).

```{r, echo = FALSE, results = 'asis'}
qc_types <- data.frame(
  `QC Type` = c("SPL", "BQC", "TQC", "LQC", "MQC", "HQC", "CAL", "RQC",
                "LTR", "NIST", "SST", "PBLK", "UBLK", "SBLK", "IBLK", "MBLK"),
  Group = c("Unknowns", "QC", "QC", "QC", "QC", "QC", "Calibration", "QC",
            "Reference", "Reference", "QA", "Blank", "Blank", "Blank", "Blank", "Blank"),
  Description = c(
    "Study sample (unknowns)",
    "Batch/Process QC (pooled sample)",
    "Technical QC",
    "Low-concentration QC",
    "Mid-concentration QC",
    "High-concentration QC",
    "Calibration standard",
    "Response QC (dilution series)",
    "Long-Term Reference",
    "NIST Standard Reference Material",
    "System Suitability Test",
    "Process Blank (no matrix, processed with ISTD)",
    "Unextracted Blank (no matrix, ISTD, no processing)",
    "Solvent Blank (no matrix, no ISTD, solvents only)",
    "Instrument Blank (with/without injection)",
    "Matrix Blank (matrix processed without ISTDs)"
  ),
  stringsAsFactors = FALSE, check.names = FALSE
)

DT::datatable(
  qc_types,
  options = list(dom = "ft", pageLength = 20, ordering = TRUE),
  rownames = FALSE,
  filter = "top",
  class = "compact stripe"
)
```

QC types are shown with consistent colors and point shapes in all MRMhub plots,
allowing quick visual identification across different figures.

## See Also

- [Sample Types & QC Roles](manual-00-sample-types.html) — detailed definitions and usage rules
- [Data Structures](manual-01-data-structure.html) — overview of all tables
- [Feature Variables](manual-03-feature-variables.html) — measurement variable columns

## References
