---
title: "Design Overview: How Data Flows Through MRMhub"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 3
description: |
  A bird's-eye view of the MRMhub data pipeline from raw instrument files 
  to final quantitative results.
vignette: >
  %\VignetteIndexEntry{Design Overview: How Data Flows Through MRMhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## The Big Picture

MRMhub processes targeted mass spectrometry data in two stages:

<div style="text-align: center; margin: 2em 0;">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 860 100" style="max-width: 860px; width: 100%; height: auto; font-family: -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;">
  <style>
    .arch-box { rx: 8; ry: 8; stroke-width: 1.5; }
    .arch-label { font-size: 12px; font-weight: 600; fill: #1a1a1a; }
    .arch-sub { font-size: 10px; fill: #555; }
    .arch-arrow { fill: #666; }
  </style>
  <rect class="arch-box" x="0" y="15" width="140" height="70" fill="#e8e8e8" stroke="#888"/>
  <text class="arch-label" x="70" y="45" text-anchor="middle">Raw Files</text>
  <text class="arch-sub" x="70" y="62" text-anchor="middle">(.wiff / .d)</text>
  <polygon class="arch-arrow" points="150,50 162,43 162,57"/>
  <rect class="arch-box" x="172" y="15" width="140" height="70" fill="#d6e8f0" stroke="#4a90a4"/>
  <text class="arch-label" x="242" y="45" text-anchor="middle">msconvert</text>
  <text class="arch-sub" x="242" y="62" text-anchor="middle">→ .mzML</text>
  <polygon class="arch-arrow" points="322,50 334,43 334,57"/>
  <rect class="arch-box" x="344" y="15" width="140" height="70" fill="#f5e0c8" stroke="#c87f3b"/>
  <text class="arch-label" x="414" y="45" text-anchor="middle">INTEGRATOR</text>
  <text class="arch-sub" x="414" y="62" text-anchor="middle">Peak integration</text>
  <polygon class="arch-arrow" points="494,50 506,43 506,57"/>
  <rect class="arch-box" x="516" y="15" width="140" height="70" fill="#f0f0f0" stroke="#888"/>
  <text class="arch-label" x="586" y="45" text-anchor="middle">long.csv</text>
  <text class="arch-sub" x="586" y="62" text-anchor="middle">Tidy exchange</text>
  <polygon class="arch-arrow" points="666,50 678,43 678,57"/>
  <a href="manual-05-data-import.html">
    <rect class="arch-box" x="688" y="15" width="160" height="70" fill="#d4e8d4" stroke="#5a9a5a"/>
    <text class="arch-label" x="768" y="45" text-anchor="middle">QUANT (R)</text>
    <text class="arch-sub" x="768" y="62" text-anchor="middle">Post-processing</text>
  </a>
</svg>
</div>

**Stage 1 — INTEGRATOR** converts vendor raw files into a standardized long-format CSV
containing peak areas per feature per sample. This handles:

- File format conversion (msconvert)
- Peak detection, picking, and integration
- Batch-aware file organization

**Stage 2 — QUANT (this R package)** takes the long CSV and provides:

- Data import and validation
- Internal standard (ISTD) normalization
- Drift and batch correction
- External calibration and quantification
- Quality control metrics and filtering
- Visualization and reporting

## Data Flow Inside QUANT

<div style="text-align: center; margin: 2em 0;">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 700 220" style="max-width: 700px; width: 100%; height: auto; font-family: -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;">
  <rect x="5" y="5" width="690" height="210" rx="10" fill="#f8f9fa" stroke="#4a90a4" stroke-width="1.5"/>
  <text x="350" y="28" text-anchor="middle" font-weight="700" font-size="13" fill="#2c5f7a">MRMhubExperiment Object</text>
  <line x1="20" y1="36" x2="680" y2="36" stroke="#dee2e6"/>
  <!-- Input arrows -->
  <text x="20" y="70" font-size="11" fill="#555">long.csv →</text>
  <text x="20" y="130" font-size="11" fill="#555">metadata →</text>
  <!-- Data slots -->
  <rect x="100" y="50" width="180" height="70" rx="6" fill="#d6e8f0" stroke="#4a90a4"/>
  <text x="190" y="70" text-anchor="middle" font-size="11" font-weight="600" fill="#2c5f7a">Datasets</text>
  <text x="115" y="88" font-size="10" fill="#333">dataset_orig (immutable)</text>
  <text x="115" y="103" font-size="10" fill="#333">dataset (working)</text>
  <text x="115" y="118" font-size="10" fill="#333">dataset_filtered (final)</text>
  <!-- Metadata slots -->
  <rect x="300" y="50" width="180" height="70" rx="6" fill="#f5e0c8" stroke="#c87f3b"/>
  <text x="390" y="70" text-anchor="middle" font-size="11" font-weight="600" fill="#7a4a1a">Metadata</text>
  <text x="315" y="88" font-size="10" fill="#333">annot_analyses</text>
  <text x="315" y="103" font-size="10" fill="#333">annot_features</text>
  <text x="315" y="118" font-size="10" fill="#333">annot_istds</text>
  <!-- QC/Metrics slots -->
  <rect x="500" y="50" width="180" height="70" rx="6" fill="#d4e8d4" stroke="#5a9a5a"/>
  <text x="590" y="70" text-anchor="middle" font-size="11" font-weight="600" fill="#2c5f2c">Metrics</text>
  <text x="515" y="88" font-size="10" fill="#333">metrics_qc</text>
  <text x="515" y="103" font-size="10" fill="#333">metrics_calibration</text>
  <text x="515" y="118" font-size="10" fill="#333">status_processing</text>
  <!-- Pipeline -->
  <rect x="100" y="140" width="580" height="55" rx="6" fill="#e8d4e0" stroke="#a45a8a"/>
  <text x="390" y="162" text-anchor="middle" font-size="11" font-weight="600" fill="#5a2a4a">Processing Pipeline</text>
  <text x="390" y="182" text-anchor="middle" font-size="10" fill="#333">import → normalize → correct_drift → correct_batch → quantify → QC → export</text>
</svg>
</div>

### Step-by-step inside QUANT:

```{r}
library(mrmhub)

# 1. Import data into MRMhubExperiment
mexp <- import_data_csv_long("path/to/long.csv",
                             analysis_type = "lipidomics")

# 2. Add metadata (sample and feature annotations)
mexp <- add_metadata(mexp,
                     annot_analyses = annot_analyses,
                     annot_features = annot_features)

# 3. Set analysis order (critical for drift correction)
mexp <- set_analysis_order(mexp)

# 4. Normalize by internal standards
mexp <- normalize_by_istd(mexp)

# 5. Correct drift (within-batch signal drift)
mexp <- correct_drift_loess(mexp)

# 6. Correct batch effects (between-batch)
mexp <- correct_batch_centering(mexp)

# 7. Quantify (using ISTD or external calibration)
mexp <- quantify_by_istd(mexp)
# -- OR --
mexp <- quantify_by_calibration(mexp)

# 8. QC filtering
mexp <- calc_qc_metrics(mexp)
mexp <- filter_features_qc(mexp)

# 9. Export
save_report_xlsx(mexp, "results.xlsx")
```

## The MRMhubExperiment Object

<details>
<summary><strong>All slots (click to expand)</strong></summary>

| Slot | Purpose |
|------|---------|
| `dataset_orig` | Original imported data (never modified) |
| `dataset` | Working data (modified by corrections/normalisation) |
| `dataset_filtered` | Final filtered data after QC |
| `annot_analyses` | Sample/run metadata (run order, batch, sample type) |
| `annot_features` | Feature metadata (compound name, ISTD assignment, RT) |
| `annot_istds` | ISTD concentrations and assignments |
| `metrics_qc` | QC metrics per feature (CV, bias, n) |
| `metrics_calibration` | Calibration curve statistics |
| `status_processing` | Current pipeline status |

</details>

## Key Accessor Functions

```{r}
# Get the current working dataset as a tibble
get_analyticaldata(mexp)

# Counts
get_feature_count(mexp)
get_analysis_count(mexp)

# Feature list
get_featurelist(mexp)

# Batch boundaries (for plotting)
get_batch_boundaries(mexp)

# Run start/end
get_analysis_start(mexp)
get_analysis_end(mexp)
```

## Decision Points

| Question | Answer |
|----------|--------|
| Do I need INTEGRATOR? | Only if starting from raw vendor files (.wiff, .d). If you already have peak areas in CSV, skip directly to QUANT. |
| Which importer? | See [Import and prepare data files](manual-05a-which-importer.html) |
| ISTD or external calibration? | ISTD: relative quantification. External calibration: absolute concentrations with calibration curve. |
| Drift correction needed? | Check with `plot_runscatter()` — if QC samples show systematic trend, apply correction. |

## Next Steps

- [Your First Analysis](tutorial-00-first-analysis.html) — try the pipeline hands-on
- [Key Concepts & Glossary](manual-00-key-concepts.html) — terminology reference
- [Import and prepare data files](manual-05a-which-importer.html) — choose the right entry point
