---
title: "Getting help from an AI assistant"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  How a lab scientist can use Claude, ChatGPT, or a local model to help plan,
  write, and troubleshoot MRMhub QUANT workflows -- accurately, and without
  exposing study data.
vignette: >
  %\VignetteIndexEntry{Getting help from an AI assistant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="page-kind page-kind-manual">Manual</p>

## Who this is for

You run targeted MS assays and have a chat assistant open -- Claude Desktop,
ChatGPT, or a model running on your own machine -- and you would like it to help
with the analysis in MRMhub QUANT. You may write a little R, or none at all.

An assistant can genuinely help here, but only if you set it up correctly. Out
of the box it has read very little about a specialised, young package like
MRMhub, so it tends to guess. This page shows the everyday tasks it is good at,
the one step that makes its answers reliable, and the two rules that keep your
results and your data safe. If you have never run the package by hand, start
with [Your first analysis](tutorial-00-first-analysis.html) so you have a
baseline to compare against.

## What it can -- and cannot -- do for you

An assistant is a fast, patient explainer and a first-draft writer. It is not a
substitute for running the analysis or for your own judgement.

**Good at**

- Explaining a step or a concept in plain language ("why is drift correction
  fitted on QC samples only?").
- Drafting R code for your workflow, which you then run.
- Reading an R error message and suggesting a fix.
- Helping you plan an analysis from a description of your samples and standards.

**Not able to**

- Run your data or check that the numbers are right -- only you can do that.
- Guarantee the code is correct; it can produce plausible code that fails.
- Know MRMhub's functions reliably unless you show it the package's own
  documentation first.

That last point is the one worth acting on, and it is easy.

## Make its answers accurate: show it the package's own docs

Left to memory, an assistant invents function names that sound right but do not
exist -- `normalize()` instead of `normalize_by_istd()`, or a single
`run_pipeline()` that was never written. The fix is to give it the package's real
reference and tell it to use only what it finds there.

MRMhub publishes a compact map of its entire interface as a single file, built
automatically with the documentation site:

<https://slinghub.github.io/MRMhub/quant/llms.txt>

Open that link, copy the contents into your chat, and start with an instruction
like this:

```text
I am analysing targeted MS data in R with the package mrmhub. Below is the
package's function reference. Use ONLY functions that appear in it. If you are
unsure a function exists, say so rather than inventing one.

<paste the contents of https://slinghub.github.io/MRMhub/quant/llms.txt here>
```

If your assistant can browse the web -- Claude with web access, ChatGPT with
browsing -- you can skip the copy-and-paste and simply give it the link, asking
it to read that page and follow the links to individual function pages as needed.

## Common tasks

Each task below is a short prompt you can adapt. Paste the reference first (or set
it up once, further down) so the assistant is working from the real interface.

### Understand a step

```text
In one short paragraph, explain what correct_drift_loess() does and why drift
and batch corrections in mrmhub are fitted on QC samples, not study samples.
```

### Draft a script for your workflow

Ask for a complete script, and insist it keep MRMhub's core pattern: every
processing step takes an `MRMhubExperiment` and returns an updated one.

```text
Write an R script using mrmhub that imports my data, normalises by internal
standard, corrects signal drift, quantifies against calibration curves,
computes QC metrics, filters failing features, and writes an Excel report.
Keep the mexp <- f(mexp, ...) pattern for every step, in recommended order.
```

A grounded assistant should return real function names in the right order. The
runnable core looks like the block below; it imports the bundled demo so it works
as is, and you would swap in your own file:

```{r}
library(mrmhub)

# Bundled demo file, so this runs as written; use your own file instead
demo <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")

mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(mexp, path = demo, import_metadata = TRUE)

# Normalise each feature by its internal standard -> feature_norm_intensity
mexp <- normalize_by_istd(mexp)

# Write the multi-sheet Excel report
save_report_xlsx(mexp, path = "results.xlsx")
```

From that base, ask the assistant to add drift correction, calibration, QC
metrics, and filtering for your own data; the full pipeline and its order are laid
out in [Basic MRMhub workflow](tutorial-02-basic-workflow.html).

### Explain and fix an error

Paste the error together with the line that produced it -- the assistant reads
both far better than a description.

```text
Running this line: mexp <- normalize_by_istd(mexp)
I get this error: <paste the exact error message>
What does it mean, and how do I fix it?
```

### Choose or understand a parameter

Point it at that function's reference page, published alongside the map above (for
example `.../quant/reference/filter_features_qc.md`), and ask what an argument
controls and how to set it for your assay.

### Plan the analysis

```text
My samples are plasma extracts with stable-isotope-labelled internal standards
and a 7-point external calibration curve. Which mrmhub steps apply, in what
order, and where do QC samples come in?
```

## Set it up once

If you use the assistant regularly, load the reference once instead of pasting it
into every chat. In **Claude**, create a Project and add the `llms.txt` file (and
a manual page or two) to its knowledge. In **ChatGPT**, do the same with a Project
or a custom GPT, attaching the same files as knowledge. Every conversation in that
Project is then grounded in the real interface from the start.

## The no-code shortcut: `build_workflow()`

If you would rather not write R at all, the package gives the assistant a head
start. `build_workflow()` opens an interactive application: you point it at your
data and metadata, tick the processing steps you want, and it validates your
inputs and generates a correct, runnable Quarto (`.qmd`) workflow to download (see
[Build a workflow without code](tutorial-12-workflow-builder.html)).

Handing that generated script to an assistant -- to explain a step, or to adjust
one part -- is far safer than asking it to write a pipeline from nothing, because
it starts from something already valid.

## Two rules you cannot skip

::: {.callout .callout-caution}
**Verify every AI-generated pipeline before trusting its results.**

- **Run it on data you understand** first -- the bundled demo
  (`system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")`) is ideal. A
  made-up function fails immediately as "could not find function".
- **Cross-check every function name** against the
  [function reference](../reference/index.html). If a call is not listed there, it
  does not exist, however convincing it looks.
- **Confirm the step order** and that drift and batch corrections are fitted on QC
  samples, not study samples. See
  [Design decisions](manual-03-design-decisions.html) for why this matters.
- **Inspect the numbers**, not just that the script ran. A pipeline can complete
  and still be wrong. Sanity-check against QC plots and metrics.
:::

::: {.callout .callout-caution}
**Do not paste study data into a cloud assistant.** Sample intensities, subject
metadata, and unpublished results sent to a hosted service leave your control and
may be retained. Share *code and the function map* -- `llms.txt`, reference pages,
and the function calls themselves -- never the measurements. When the data must
stay on the machine, use a locally hosted model, which keeps everything
on-device.
:::

## Going further

More capable tooling exists for those who want it. Agentic coding assistants such
as Claude Code and Cursor operate directly on a project folder and can read the
package source as they write. Claude Desktop can connect to your files through an
MCP connector, though their contents then reach the provider like any other
message. Locally hosted models (through Ollama or LM Studio) keep everything
on-device, but a smaller local model has memorised even less of a niche package,
so it depends *more* on being shown the reference, not less. Whichever you use,
the two rules above still apply.

## Next steps

- [Design decisions](manual-03-design-decisions.html): the conventions an assistant must respect -- the `mexp -> mexp` pattern and the pipeline order
- [Key concepts and glossary](manual-01-key-concepts.html): the data model and terms to share with the assistant
- [Your first analysis](tutorial-00-first-analysis.html): a hand-run baseline to compare generated code against
