---
title: "Troubleshooting and FAQ"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Solutions to the most common errors and questions when using MRMhub QUANT.
vignette: >
  %\VignetteIndexEntry{Troubleshooting and FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="page-kind page-kind-manual">Manual</p>

Common errors and questions, grouped by workflow stage. For installation problems,
`check_setup()` reports what the R environment is missing; unresolved bugs belong
on [GitHub Issues](https://github.com/SLINGhub/MRMhub/issues).

## Installation and setup

**Q: How do I confirm my R environment is ready for MRMhub?**

**A:** `check_setup()` checks the R version (≥ 4.1.0 is required), the packages
MRMhub needs, and the optional packages that unlock extra features. Missing
required packages are listed with a copy-pasteable install command.

```{r}
check_setup()
```

**Q: `build_workflow()` reports that shiny or bslib is not installed.**

**A:** The Workflow Builder is an optional feature, so `shiny` and `bslib` are
not installed with MRMhub. Run interactively, `build_workflow()` offers to
install them; otherwise add them yourself.

```{r}
install.packages(c("shiny", "bslib"))
```

If instead it reports *"Could not find the workflow builder app. Try
reinstalling mrmhub."*, the app files are missing from the installed package —
reinstall MRMhub.

**Q: On Windows, installation fails with *"cannot remove prior installation of
package 'X'"* or *"permission denied"*.**

**A:** A dependency's DLL is locked by another running R session, RStudio, or
Positron, and *Restart R* does not always release the lock on Windows. Quit the
IDE entirely (close all windows), confirm no `R.exe`/`Rterm.exe` is left in Task
Manager, reopen, and run the install command before opening any project that
auto-attaches a tidyverse package.

**Q: *"namespace 'rlang' is already loaded"* (or another core package).**

**A:** A package that others depend on is pinned in the current namespace.
Restart R, update the named package first in the fresh session, then retry.

```{r}
install.packages("rlang")
```

**Q: After updating MRMhub, a function fails with *"lazy-load database
'.../mrmhub.rdb' is corrupt"*.**

**A:** The package was reinstalled while it was loaded in the running session,
so R still holds the index of the previous installation and looks up functions
at offsets that no longer match the new file on disk. Nothing is actually
damaged. Restart R and attach the package again.

```{r}
library(mrmhub)
```

Restarting R after every install or update of MRMhub avoids this. In RStudio and
Positron, *Session ▸ Restart R* (<kbd>Ctrl/Cmd</kbd>+<kbd>Shift</kbd>+<kbd>F10</kbd>).

**Q: *"Rtools is required to build R packages"* (Windows).**

**A:** Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) matching
your R version (Rtools43 for R 4.3, Rtools44 for R 4.4, …), restart R, and
retry. Rtools compiles the source packages a few dependencies still need when no
Windows binary is available yet.

**Q: `remotes` keeps failing on a locked or transitive dependency.**

**A:** [`pak`](https://pak.r-lib.org/) resolves the dependency graph in one pass,
handles locked packages more cleanly on Windows, and gives clearer messages on
transitive failures.

```{r}
install.packages("pak")
pak::pak("SLINGhub/mrmhub")
```

**Q: *"installation of package 'X' had non-zero exit status"*.**

**A:** By default every optional (Suggests) dependency is installed too, so one
broken optional package aborts the whole install. Install only the strictly
required dependencies and add optional ones later.

```{r}
remotes::install_github(
  "SLINGhub/mrmhub",
  dependencies = c("Depends", "Imports", "LinkingTo")
)
```

## Importing data

**Q: Which importer should I use?**

**A:** See [Import and prepare data files](manual-04-data-import.html) for a
decision flowchart matching each vendor/file layout to its importer.

**Q: `import_data_masshunter()` aborts with *"Data file is in an unsupported or
corrupted format…"*.**

**A:** Almost always the wrong importer for the file — a plain, Skyline, or
INTEGRATOR CSV passed to the MassHunter parser. Use the importer that matches the
file's format. If the file really is a MassHunter export, re-export it from MH
Quant with compounds as columns.

**Q: An importer aborts with *"Required `analysis_id` column is missing"* (or
`feature_id`).**

**A:** The column holding the analysis or feature identifier was not recognised.
Map it explicitly with `column_mapping` (names are matched case-insensitively).

```{r}
import_data_csv_long(
  mexp,
  path = "your_file.csv",
  column_mapping = c(analysis_id = "Sample", feature_id = "Compound")
)
```

**Q: *"The file was read as a single column using the ',' delimiter."***

**A:** The file is semicolon- or tab-delimited (common in European Excel
locales), so a comma reader sees one column. Re-export as *CSV UTF-8 (Comma
delimited)* and re-import.

**Q: *"Imported data measures the same feature more than once in the same
analysis."***

**A:** Read the values in the message. **Identical** values mean the same data
was read twice (the same result file, or an overlapping part of it). **Differing**
values mean two real measurements carry the same IDs: two injections sharing a
`raw_data_filename` (the default `analysis_id`), or the wrong column mapped to
`analysis_id`/`feature_id`.

**Q: A wide CSV aborts with *"All columns with feature values must be numeric."***

**A:** Metadata columns are mixed into the feature block. Point
`first_feature_column` at the first real feature column, and pass `na_strings`
for any text missing-value tokens.

```{r}
import_data_csv_wide(
  mexp,
  path = "your_file.csv",
  first_feature_column = 5,
  na_strings = c("NA", "n.a.", "")
)
```

**Q: Skyline import aborts with *"`Molecule Name` is not unique identifier for
each transition."***

**A:** Several transitions share a molecule name. Build unique IDs from the
precursor/product names or masses.

```{r}
import_data_skyline(
  mexp,
  path = "your_file.csv",
  transition_id_columns = "name"
)
```

**Q: Feature names in the data don't match the feature annotation.**

**A:** Trailing whitespace or case differences (`"Compound A "` vs
`"Compound A"`) break the join. Compare both directions:

```{r}
data_ids  <- unique(mexp@dataset$feature_id)
annot_ids <- mexp@annot_features$feature_id
setdiff(data_ids, annot_ids)  # in data, not annotated
setdiff(annot_ids, data_ids)  # annotated, not in data
```

## Adding metadata

**Q: *"None of the analyses in the data match the imported analysis metadata."***

**A:** Either the wrong metadata file, or the `analysis_id`s are formatted
differently in the two sources. Import strips raw-file extensions and squishes
whitespace, so a residual mismatch means the IDs genuinely differ — compare them
with `setdiff()` as above.

**Q: *"Not all analyses in the data have corresponding metadata."***

**A:** Some analyses have no row in the analysis metadata. Add the missing rows,
or import with `excl_unmatched_analyses = TRUE` to drop the unannotated
analyses.

**Q: The metadata import aborts on validation warnings.**

**A:** A validation report is printed, then *"Please verify warnings in
corresponding metadata. Use `ignore_warnings = TRUE` to ignore warnings."* Fix
the flagged issues, or re-import with `ignore_warnings = TRUE`. A hard
*"Metadata validation failed — see the report above."* (duplicate IDs, missing
mandatory values) cannot be bypassed and must be corrected.

## Processing

**Q: Normalised intensities come back `NA`.**

**A:** `normalize_by_istd()` aborts by default when a feature has no internal
standard assigned (*"…no ISTD was defined…"*), or when no feature carries an
`istd_feature_id` at all (*"No ISTDs defined in feature metadata."*). Find the
unassigned features, then assign an `istd_feature_id` to each — or pass
`ignore_missing_annotation = TRUE` to set them to `NA` and continue.

```{r}
mexp@annot_features |>
  dplyr::select(feature_id, istd_feature_id) |>
  dplyr::filter(is.na(istd_feature_id))
```

**Q: *"Data needs to be ISTD normalized, please run 'normalize_by_istd' first."***

**A:** Quantitation and drift/batch correction on normalised or concentration
variables run after normalisation. Run `normalize_by_istd()` before
`quantify_by_istd()` or `quantify_by_calibration()`.

**Q: *"Acquisition timestamps are not present in analysis results."***

**A:** The vendor export carries no acquisition time, so `set_analysis_order()`
cannot order by `"timestamp"`. Order by the result-file name or by a column in
the metadata instead.

```{r}
set_analysis_order(mexp, order_by = "resultfile")
```

**Q: Drift or batch correction aborts with *"One or more specified `qc_types` are
not present in the dataset."***

**A:** The reference QC label has no matching rows — a misspelled QC type, no
such QC samples in the run, or `add_metadata()` was not run so `qc_type` is
still empty. Drift and batch correction fit on QC samples only, so a valid QC
type must be present.

**Q: Correction fails with *"Normalized intensities not available"* or
*"Concentration data are not available"*.**

**A:** You asked to correct a variable that has not been produced yet. Run the
upstream step first (`normalize_by_istd()`, then a `quantify_*()`), or correct
`variable = "intensity"`.

**Q: *"Batch correction was not applied as there is only one batch."***

**A:** Between-batch correction needs more than one `batch_id`. With a single
batch there is nothing to centre — skip it and run the rest of the pipeline
normally.

**Q: Drift correction makes the data worse.**

**A:** Loess fitting needs enough evenly-spaced QC samples per batch; with too
few, or when the QC trend is flat (noise, not drift), the fit can add variance.
Inspect the QC trend before correcting, and skip drift correction for batches
where it is flat.

```{r}
plot_runscatter(
  mexp,
  variable = "norm_intensity",
  qc_types = c("BQC", "SPL"),
  include_feature_filter = "your_feature"
)
```

## Quantitation and calibration

**Q: `quantify_by_istd()` aborts about missing ISTD concentrations or amounts.**

**A:** ISTD quantitation needs internal-standard concentrations plus per-analysis
amounts. Import ISTD metadata (`annot_istds`) with concentrations in nmol/L or
ng/mL, and make sure `sample_amount` and `istd_volume` are present in the
analysis metadata. Missing amounts can be tolerated with
`ignore_missing_annotation = TRUE`.

**Q: *"Calibration curve data missing…"***

**A:** No calibrator samples and no concentrations are annotated. Flag the
calibrator analyses with `qc_type = "CAL"` in the analysis metadata, and supply
their known concentrations in the QC-concentration metadata
(`annot_qcconcentrations`).

**Q: *"Calibration curve annotations for N features are missing."***

**A:** Those features have no concentrations in `annot_qcconcentrations`, which
is keyed on `sample_id` and `analyte_id`. Add the missing entries, or skip those
features with `ignore_missing_annotation = TRUE`.

**Q: A calibration curve has R² < 0.9.**

**A:** Inspect the fits, then try removing outlier calibration points, switching
`fit_model` (`"linear"`/`"quadratic"`) or `fit_weighting` (`"none"`, `"1/x"`,
`"1/x^2"`) — globally or per feature via the `curve_fit_model` /
`curve_fit_weighting` metadata columns — and check for saturation at high
concentrations.

```{r}
get_calibration_metrics(mexp) |>
  dplyr::filter(r2 < 0.9)
```

## QC and filtering

**Q: What do the QC metrics mean?**

| Metric | Meaning | Typical threshold |
|--------|---------|-------------------|
| CV (%) | Coefficient of variation in QC samples | < 20–30% |
| Bias (%) | Systematic deviation from expected value | < 20% |
| n_detected | Number of QC samples with signal | ≥ 50–67% of QCs |

**Q: All my features are filtered out.**

**A:** `filter_features_qc()` requires you to state `include_qualifier` and
`include_istd` explicitly. If every feature then fails the thresholds it raises
no error — `dataset_filtered` simply comes back empty, and the next step aborts
with *"Data has not been QC-filtered."* Inspect the metrics and loosen the
thresholds.

```{r}
mexp <- filter_features_qc(
  mexp,
  include_qualifier = FALSE,
  include_istd = FALSE,
  max.cv.normintensity.bqc = 40,
  max.prop.missing.normintensity.spl = 0.5
)
```

## Working with the object

**Q: The pipeline does not run through, or the results look wrong.**

**A:** A common cause is an inconsistent object name across chunks — assigning a step to one name
but running the next step on another (a typo, or mixing `mexp` and `myexp`). Because every function
takes and returns the `MRMhubExperiment`, a mismatched name quietly processes an earlier or wrong
object instead of raising an error. Use a single name throughout, assign each step back to it, and
confirm with `mrmhub_status()` that the steps you expect have run.

**Q: How do I extract the final data as a data frame?**

**A:** `get_analyticaldata()` returns the metadata-joined working table; the
filtered table is a slot.

```{r}
final <- get_analyticaldata(mexp, annotated = TRUE)
final_filtered <- mexp@dataset_filtered
```

**Q: How do I check what processing has been applied?**

**A:** The status flags record it.

```{r}
mexp@is_istd_normalized
mexp@is_quantitated
mexp@var_drift_corrected
mexp@var_batch_corrected
```

**Q: How do I go back to the original data?**

**A:** The raw import is always preserved in `dataset_orig`. Restoring it resets
the working table (which drops the processing columns added since import).

```{r}
mexp@dataset <- mexp@dataset_orig
```

## Workflow Builder

**Q: Some processing steps are greyed out with a yellow badge.**

**A:** Steps are gated by the metadata you imported; the badge names what is
missing — *No feature metadata*, *No ISTDs defined*, *No ISTD concentrations*,
*No QC concentrations*, *No sample_id metadata*, or *Only one batch*. Import the
metadata that step needs (or, for the batch step, a run with more than one
batch) to enable it.

**Q: A banner says *"N things to check before this will run."***

**A:** The builder pre-checks each selected step. Amber means warnings only — it
still generates the workflow; red means errors to fix first. Open *Show details*
to see each message, then import the missing metadata or select the upstream
step it names.

**Q: The banner says *"Configure your columns."***

**A:** A generic long/wide CSV cannot be read until its columns are mapped.
Click *Configure columns…* at the top of the sidebar, map the identifier and
value columns, and *Apply* — the sidebar then shows *columns configured*.

**Q: Which metadata source do I choose, and why did *"Embedded in the data
file"* disappear?**

**A:** There are four sources: embedded in the data file, an MSOrganiser
workbook, separate metadata tables (one `.xlsx`), or none for now. The app probes
your uploaded file and only offers *Embedded* when the data actually carries
in-file metadata columns; when it does not, that option is removed — use one of
the workbook routes.

**Q: A red banner says *"Import failed:"***

**A:** The importer threw while reading your files. The most common causes are
the wrong importer for the file's format, or a non-MSOrganiser workbook (or an
unsupported template version) fed to the MSOrganiser route. The message after
*"Import failed:"* is the underlying error — see the import and metadata
sections above.

**Q: I clicked *Browse…* and nothing happened.**

**A:** The native file and folder pickers only work when the app runs locally
with a system dialog available. On a server or headless session there is no
picker — type or paste the path into the text field instead.

**Q: Will my large INTEGRATOR file upload?**

**A:** Yes, up to 500 MB (the app raises Shiny's default 5 MB cap). The data
input accepts `.csv`, `.tsv`, and `.txt`; Excel files are only for the metadata
routes.

**Q: Did the app run my analysis?**

**A:** No — the Workflow Builder validates your inputs and generates a Quarto
workflow (`mrmhub_workflow.qmd`); it never runs the pipeline. Download or copy
the `.qmd`, then render it yourself.

```bash
quarto render mrmhub_workflow.qmd
```

**Q: How do I try it without my own data?**

**A:** *Try the example* loads the bundled demo dataset (with embedded metadata)
so you can reach a green *Ready to render* banner end to end.

## FAQ

**Q: Can I change the default text size, point size, or legend in plots?**

**A:** `mrmhub_set_plot_defaults()` sets session-wide defaults for the shared
appearance arguments of the `plot_*()` functions, so you can set them once at the
top of a notebook instead of on every call.

```{r}
mrmhub_set_plot_defaults(font_base_size = 8, point_size = 0.8)
```

It also accepts `legend_position`, `legend_size`, `show_legend_title`, and
`strip_bg_color`; a value passed directly to a plotting function always wins.
`mrmhub_get_plot_defaults()` shows the active settings and
`mrmhub_reset_plot_defaults()` clears them. See
[Customising plots](manual-08-visualization.html#customising-plots) for the full
reference.

**Q: Can I use MRMhub with non-MRM data (PRM, SRM)?**

**A:** Yes, provided the data can be formatted as a long CSV with feature IDs and
peak areas. The processing is agnostic to acquisition mode.

**Q: Can I process only one batch?**

**A:** Yes. Skip batch correction; every other function works with single-batch
data.

**Q: Is there a size limit?**

**A:** No hard limit, but very large datasets are slower in the drift and batch
steps. If needed, process batches separately and merge.

**Q: Where do I report bugs?**

**A:** [GitHub Issues](https://github.com/SLINGhub/MRMhub/issues).

## Next steps

- [Import and prepare data files](manual-04-data-import.html): pick the right importer for your export
- [Metadata](manual-05-metadata.html): the annotation tables the pipeline joins against
- [Installation](manual-00-installation.html): diagnose setup problems with `check_setup()`
- [Design decisions behind MRMhub QUANT](manual-03-design-decisions.html): why the pipeline is built this way
