---
title: "Installation"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Install mrmhub, verify your setup, and fix common install errors.
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

<p class="page-kind page-kind-manual">

Manual

</p>

## Requirements

The `mrmhub` R package requires:

- **Operating system:** Windows, macOS, or Linux.
- **R:** version 4.1 or newer, from [CRAN](https://cran.r-project.org).
- **IDE (recommended):** [RStudio](https://posit.co/download/rstudio-desktop/)
  or [Positron](https://positron.posit.co), which combine a code editor,
  console, and package management in a single window.
- **Quarto:** bundled with recent RStudio and Positron releases; otherwise
  install it from [quarto.org](https://quarto.org/docs/get-started/). Required
  only for rendering reports and workflow documents.

All package dependencies are resolved and installed automatically, so no
further configuration is needed beyond a working R installation.

## Installing MRMhub

::: {.callout .callout-caution}
**Important.** Quit and reopen all RStudio/Positron sessions first; many
install failures come from R packages already loaded in another session
:::

In a fresh R session:

```{r, eval=FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/MRMhub")
```

We recommend `pak`, but if it fails behind a proxy or on an unusual
platform, `remotes` works as well:

```{r, eval=FALSE}
if (!require("remotes")) install.packages("remotes")
remotes::install_github("SLINGhub/MRMhub")
```

## Confirm `mrmhub` is sucessfully installed

```{r, eval=FALSE}
library(mrmhub)
```

If this loads without error, you are ready to go. Head to [Your first
analysis](tutorial-00-first-analysis.html).

## Optional packages

To keep the base installation light, `mrmhub` installs only the packages
needed for the common workflow. A number of specialised functions rely on
additional packages that are not pulled in upfront. When such a function is
called and its package is missing, `mrmhub` reports what is needed and, in
an interactive session, offers to install it on the spot — so most users
never install these manually.

Ordered from the most commonly used functions to the more specialised ones:

| Function | Optional package | Enables |
|----------|------------------|---------|
| QC and EDA plots | ggvenn, patchwork, ggforce, ggbeeswarm, ggsignif | QC-summary and exploratory plots |
| `plot_pca()` | ggnewscale | PCA score plots |
| `plot_runscatter()` | qpdf; mirai, carrier | multi-page PDF output; parallel rendering |
| `correct_batch_combat()` | sva | ComBat batch correction |
| `correct_batch_serrf()` | ranger | SERRF batch correction |
| `correct_drift_gam()` | mgcv | GAM-based drift correction |
| `plot_responsecurves()` | ggpmisc | response-curve plots with fit annotations |
| `build_workflow()` | shiny, bslib | interactive workflow-builder app |
| `save_dataset_summarizedexperiment()` | SummarizedExperiment, S4Vectors, lipidr *(Bioconductor)* | export to a `SummarizedExperiment` |
| Lipid-name parsing (isotope correction, lipid plots) | rgoslin *(Bioconductor)* | parse and normalise lipid shorthand |
| `correct_isotopic_interferences()`, `calc_average_molweight()` | enviPat | isotope-pattern and molecular-weight calculation |
| `get_response_curve_stats()` | lancer *(GitHub)* | response-curve quality metrics |

To install all of them upfront, run the following in a fresh R session
(again, with all other RStudio/Positron sessions closed):

```{r, eval=FALSE}
pak::pak(c(
  "sva", "ranger", "mgcv", "enviPat",
  "ggnewscale", "ggpmisc", "ggvenn", "patchwork",
  "ggforce", "ggbeeswarm", "ggsignif",
  "qpdf", "mirai", "carrier", "shiny", "bslib",
  "rgoslin", "lipidr", "SummarizedExperiment", "S4Vectors",
  "SLINGhub/lancer"
))
```

`pak` resolves CRAN, Bioconductor, and GitHub sources in one call. If it
fails behind a proxy or on an unusual platform, install the CRAN packages
with `install.packages()` and the Bioconductor ones
(`rgoslin`, `lipidr`, `SummarizedExperiment`, `S4Vectors`) with
`BiocManager::install()`.

## Troubleshooting

If the installation seemed succesfull but but the package doesnt load or
with errors, run `check_setup()` to report the R version and flag any
missing dependencies:

```{r, eval=FALSE}
check_setup()
```

**Most frequent installation errors**

See [Troubleshooting & FAQ](manual-10-troubleshooting.html) for detailed
error and resolution list.

| Error | Cause | Fix |
|--------------------------|--------------------------|-------------------|
| `namespace 'rlang' is already loaded` | An old dependency is still loaded in the session | Restart R (`Ctrl+Shift+F10`) and retry the install |
| `pak` unavailable or failing | `pak` not installed, or its cache is stale | Install with `remotes::install_github("SLINGhub/MRMhub")` |
| `cannot open URL` | Firewall or proxy blocking GitHub | `options(download.file.method = "libcurl")`, or clone the repo and use `remotes::install_local("path/to/MRMhub")` |
| `package 'X' was installed under R version …` | Package built for a different R | `install.packages("X")` to rebuild it for your R |
| `there is no package called 'mrmhub'` | Install did not finish | Scroll up for the real error, then retry the install |
| `ERROR: Rtools is required` (Windows) | A source-only dependency needs compilation | Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) matching your R version, restart R, retry |
| `clang: error: ...` (macOS) | Compiler tools missing | Run `xcode-select --install` in Terminal, retry |
| `cannot find -lcurl` (Linux) | System libraries missing | `sudo apt install libcurl4-openssl-dev libxml2-dev libssl-dev libfontconfig1-dev` |

## Next steps

- [Key concepts & glossary](manual-01-key-concepts.html): core
  vocabulary and the MRMhubExperiment object
- [Your first analysis](tutorial-00-first-analysis.html): a short
  end-to-end walkthrough
- Questions or bug reports? File an issue on
  [GitHub](https://github.com/SLINGhub/MRMhub/issues), or contact the
  authors directly.