---
title: "Installation"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Install mrmhub, verify your setup, and fix common install errors.
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

<p class="page-kind page-kind-manual">Manual</p>

## Requirements

- **Operating system:** Windows, macOS, or Linux.
- **R:** version 4.1 or newer, from [CRAN](https://cran.r-project.org).
- **R Editor (recommended):** [RStudio](https://posit.co/download/rstudio-desktop/)
  or [Positron](https://positron.posit.co).
- **Quarto:** bundled with recent RStudio and Positron releases; otherwise
  install it from [quarto.org](https://quarto.org/docs/get-started/). Required
  only for rendering notebooks.

## Installing MRMhub

::: {.callout .callout-caution}
**Important.** Quit and reopen all RStudio/Positron sessions first. Many installation issues 
are caused by R packages that are loaded in another session while the installer tries to update them.
:::

In a **fresh R session**, we recommend installing with `pak`:

```{r, eval=FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/MRMhub")
```

If `pak` installation fails, e.g. due to institutional network firewall or
proxy settings, try installing
with `remotes` instead:

```{r, eval=FALSE}
if (!require("remotes")) install.packages("remotes")
remotes::install_github("SLINGhub/MRMhub")
```

If installation fails, see [Troubleshooting](#troubleshooting) below.

## Confirm `mrmhub` is successfully installed

```{r, eval=FALSE}
library(mrmhub)
```

If this loads without error, you are ready to go. If you plan to use any of the
functions listed under [Optional packages](#optional-packages) below, install
those too. If it loads with errors, see [Troubleshooting](#troubleshooting) below.

## Optional packages

To keep the base installation light, a number of specialised functions in
`mrmhub` rely on additional packages that are not installed upfront. When such a
function is called and its package is not yet installed, `mrmhub` offers to
install it.

| Function | Optional package | Enables |
|----------|------------------|---------|
| `plot_runscatter()` | qpdf; mirai, carrier | only for multi-threaded PDF output |
| `correct_isotopic_interferences()`, `calc_average_molweight()` | enviPat | isotope-pattern and molecular-weight calculation |
| `correct_batch_combat()` | sva | ComBat batch correction |
| `correct_batch_serrf()` | ranger | SERRF batch correction |
| `correct_drift_gam()` | mgcv | GAM-based drift correction |
| `build_workflow()` | shiny, bslib | interactive workflow-builder app |
| `save_dataset_summarizedexperiment()` | SummarizedExperiment, S4Vectors, lipidr *(Bioconductor)* | export to a `SummarizedExperiment` |
| Lipid-name parsing (isotope correction, lipid plots) | rgoslin *(Bioconductor)* | parse and normalise lipid shorthand |
| `get_response_curve_stats()` | lancer *(GitHub)* | only for specific response-curve metrics |

To install all of them upfront, run the following in a fresh R session
(again, with all other RStudio/Positron sessions closed):

```{r, eval=FALSE}
pak::pak(c(
  "sva", "ranger", "mgcv", "enviPat",
  "qpdf", "mirai", "carrier", "shiny", "bslib",
  "rgoslin", "lipidr", "SummarizedExperiment", "S4Vectors",
  "SLINGhub/lancer"
))
```

Alternatively, install the CRAN packages
with `install.packages()` and the Bioconductor ones
(`rgoslin`, `lipidr`, `SummarizedExperiment`, `S4Vectors`) with
`BiocManager::install()`.

## Troubleshooting

If the installation seemed successful but the package does not load, or loads
with errors, run `check_setup()` to report the R version and flag any missing
dependencies:

```{r, eval=FALSE}
mrmhub::check_setup()
```

**Frequent installation errors**

See [Troubleshooting & FAQ](manual-10-troubleshooting.html) for a detailed
list of errors and resolutions.

| Error | Cause | Fix |
|--------------------------|--------------------------|-------------------|
| `namespace 'rlang' is already loaded` | An old dependency is still loaded in the session | Restart R (`Ctrl+Shift+F10`) and retry the install |
| `pak` unavailable or failing | `pak` not installed, or its cache is stale | Install with `remotes::install_github("SLINGhub/MRMhub")` |
| `cannot open URL` | Firewall or proxy blocking GitHub | `options(download.file.method = "libcurl")`, or clone the repo and use `remotes::install_local("path/to/MRMhub")` |
| `Could not resolve host: api.github.com` | Firewall or proxy blocking the GitHub API | Set `http_proxy`/`https_proxy`, or download and unzip the repo and use `remotes::install_local("path/to/MRMhub")` (or `pak::local_install("path/to/MRMhub")`) |
| `SSL certificate problem: self signed certificate in certificate chain` | Corporate proxy inspecting TLS traffic | Use `remotes` (it uses the system certificate store), or install your organisation's root certificate |
| `package 'X' was installed under R version …` | Package built for a different R | `install.packages("X")` to rebuild it for your R |
| `there is no package called 'mrmhub'` | Install did not finish | Scroll up for the real error, then retry the install |
| `ERROR: Rtools is required` (Windows) | A source-only dependency needs compilation | Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/) matching your R version, restart R, retry |
| `clang: error: ...` (macOS) | Compiler tools missing | Run `xcode-select --install` in Terminal, retry |
| `cannot find -lcurl` (Linux) | System libraries missing | `sudo apt install libcurl4-openssl-dev libxml2-dev libssl-dev libfontconfig1-dev` |

## Next steps

- [MRMhub overview](manual-01-key-concepts.html): core
  vocabulary and the MRMhubExperiment object
- [Introduction to R, Quarto & MRMhub](tutorial-13-intro-to-r-quarto-mrmhub.html): a short
  end-to-end walkthrough
- Questions or bug reports? File an issue on
  [GitHub](https://github.com/SLINGhub/MRMhub/issues), or contact the
  authors directly.