---
title: "Feature Variables in MRMhub"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Reference for all feature variables (measurement columns) available in MRMhub.
vignette: >
  %\VignetteIndexEntry{Feature Variables in MRMhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
bibliography: references.bib
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, warning = FALSE,
                      message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

Feature variables represent measurement values associated with a feature in a 
specific sample. They describe properties such as peak area, concentration, 
retention time, or peak shape.

Feature variables can be accessed in MRMhub functions by their internal name 
(starting with `feature_`, e.g., `feature_intensity`) or by their short name 
(e.g., `conc`, `intensity`, `norm_intensity`, `rt`).

## Key Feature Variables

These variables are essential for the data processing flow. The `intensity` 
variable holds the raw signal (e.g., peak area) and is the starting point for 
all downstream processing. All key feature variables are stored in the `dataset` 
table of the `MRMhubExperiment` object.

```{r, echo = FALSE, results = 'asis'}
key_vars <- data.frame(
  `Short Name` = c("intensity", "norm_intensity", "pmol_total", "conc", 
                   "conc_normalized", "conc_bias"),
  `Internal Name` = c("feature_intensity", "feature_norm_intensity", 
                      "feature_pmol_total", "feature_conc", 
                      "feature_conc_normalized", "feature_conc_ratio"),
  Description = c(
    "Raw signal intensity. Copied from area, height, response, or intensity at import.",
    "Internal standard-normalized raw signal intensity.",
    "Total analyte amount in the measured sample (pmol/sample).",
    "Analyte concentration.",
    "Normalized concentration (e.g., by reference sample or other methods).",
    "Ratio of measured to expected concentration. Generated by calibrate_by_reference()."
  ),
  stringsAsFactors = FALSE, check.names = FALSE
)

DT::datatable(
  key_vars,
  options = list(dom = "t", pageLength = 10, ordering = FALSE),
  rownames = FALSE,
  class = "compact stripe"
)
```

::: {.callout .callout-note}
Many processing and plotting functions accept a `variable` argument
that selects which feature variable to use as input.
:::

## Backup Feature Variables

Some processing steps (drift correction, batch correction, recalibration) 
overwrite existing values. In these cases, the original values are saved in a 
backup variable with a postfix appended to the variable name.

| Postfix | Example | Created when |
|---------|---------|-------------|
| `_orig` | `feature_intensity_orig` | Before interference correction. Always holds the original imported intensity. |
| `_raw` | `feature_conc_raw` | Before any correction step. Holds the uncorrected calculated values. |
| `_before` | `feature_conc_before` | Before the most recent correction step (e.g., after drift correction, before batch correction). |
| `_beforecal` | `feature_conc_beforecal` | Before `calibrate_by_reference()` was applied. Only for concentration variables. |
| `_fit` | `conc_before_fit` | Model fit data points from drift/batch correction. Used by `plot_runscatter()` to show trends. |

## Raw Feature Variables

These variables are stored in the `dataset_orig` table and are never modified by 
any MRMhub function. One of them is copied to `intensity` at import (by default 
`area` if available, then `height`, `response`, or `intensity` in that order).
The source variable can be set manually with `set_intensity_var()`.

```{r, echo = FALSE, results = 'asis'}
raw_vars <- data.frame(
  `Short Name` = c("area", "height", "response", "intensity", "rt", "fwhm", 
                   "width", "int_start", "int_end", "symmetry", "sn_ratio", 
                   "accuracy", "ionratio"),
  `Internal Name` = c("feature_area", "feature_height", "feature_response", 
                      "feature_intensity", "feature_rt", "feature_fwhm", 
                      "feature_width", "feature_int_start", "feature_int_end", 
                      "feature_symmetry", "feature_sn_ratio", "feature_accuracy", 
                      "feature_ionratio"),
  Description = c("Peak area", "Peak height", "Feature response (vendor-specific)", 
                  "Feature intensity (vendor-specific)", "Retention time", 
                  "Full Width at Half Maximum", "Peak width", 
                  "Peak integration start time", "Peak integration end time", 
                  "Peak symmetry", "Signal-to-noise ratio", 
                  "Accuracy (relative to target value)", "Ion ratio"),
  stringsAsFactors = FALSE, check.names = FALSE
)

DT::datatable(
  raw_vars,
  options = list(dom = "ft", pageLength = 15, ordering = TRUE),
  rownames = FALSE,
  filter = "top",
  class = "compact stripe"
)
```

## Next Steps

- [Data Structures](manual-01-data-structure.html) — overview of all data tables
- [Data Identifiers](manual-02-data-identifiers.html) — ID columns
- [Key Concepts & Glossary](manual-00-key-concepts.html) — terminology reference