---
title: "Data and Metadata in MRMhub"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Overview of how MRMhub organizes analytical data and metadata within
  the MRMhubExperiment object.
vignette: >
  %\VignetteIndexEntry{Data and Metadata in MRMhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, warning = FALSE,
                      message = FALSE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

## Overview

The **`MRMhubExperiment`** object is the primary data container in the MRMhub 
workflow. It holds all experimental and processed data and metadata, as well as 
details of applied processing steps and the current status of the data. Most 
MRMhub functions take the `MRMhubExperiment` object as input and return an 
updated copy.

Data within the `MRMhubExperiment` is organized into **data** and **metadata** 
categories, each divided into tables (data frames).

<div style="text-align: center; margin: 2em 0;">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 600 280" style="max-width: 600px; width: 100%; height: auto; font-family: -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;">
  <rect x="5" y="5" width="590" height="270" rx="10" fill="#f8f9fa" stroke="#4a90a4" stroke-width="2"/>
  <text x="300" y="30" text-anchor="middle" font-weight="700" font-size="14" fill="#1a1a1a">MRMhubExperiment</text>
  <line x1="20" y1="40" x2="580" y2="40" stroke="#dee2e6"/>
  <rect x="20" y="55" width="270" height="100" rx="6" fill="#d6e8f0" stroke="#4a90a4"/>
  <text x="155" y="75" text-anchor="middle" font-weight="600" font-size="12" fill="#2c5f7a">DATA</text>
  <text x="35" y="95" font-size="10" fill="#333">dataset_orig — original imported data</text>
  <text x="35" y="112" font-size="10" fill="#333">dataset — annotated and processed data</text>
  <text x="35" y="129" font-size="10" fill="#333">dataset_filtered — QC-filtered output</text>
  <text x="35" y="146" font-size="10" fill="#333">feature_metrics — QC metrics per feature</text>
  <rect x="310" y="55" width="270" height="100" rx="6" fill="#f5e0c8" stroke="#c87f3b"/>
  <text x="445" y="75" text-anchor="middle" font-weight="600" font-size="12" fill="#7a4a1a">METADATA</text>
  <text x="325" y="95" font-size="10" fill="#333">annot_analyses — sample/run annotations</text>
  <text x="325" y="112" font-size="10" fill="#333">annot_features — feature annotations</text>
  <text x="325" y="129" font-size="10" fill="#333">annot_istds — ISTD concentrations</text>
  <text x="325" y="146" font-size="10" fill="#333">annot_batches / annot_standards / ...</text>
  <rect x="20" y="170" width="560" height="50" rx="6" fill="#d4e8d4" stroke="#5a9a5a"/>
  <text x="300" y="192" text-anchor="middle" font-weight="600" font-size="12" fill="#2c5f2c">STATUS AND FLAGS</text>
  <text x="300" y="210" text-anchor="middle" font-size="10" fill="#333">is_istd_normalized, is_quantitated, var_drift_corrected, var_batch_corrected, ...</text>
  <text x="300" y="258" text-anchor="middle" font-size="10" fill="#666">Functions take MRMhubExperiment in and return updated MRMhubExperiment</text>
</svg>
</div>

## Data Tables

| Category | Table (Slot) | Description |
|----------|-------------|-------------|
| Raw Data | `dataset_orig` | Original imported data — never modified after import. |
| Processed Data | `dataset` | Working copy with all annotations and processing results. |
| QC-filtered Data | `dataset_filtered` | Subset of `dataset` passing QC criteria. |
| Feature Metrics | `feature_metrics` | Quality control metrics calculated per feature. |

## Metadata Tables

| Category | Table (Slot) | Description |
|----------|-------------|-------------|
| Analyses | `annot_analyses` | Sample categories, amounts, dilutions, batch assignments, and run order. |
| Features | `annot_features` | ISTD assignments, response factors, feature classification, quantifier/qualifier flags. |
| Internal Standards | `annot_istds` | ISTD concentrations added to each sample. |
| Batches | `annot_batches` | Start and end boundaries for each analytical batch. |
| Response Curves | `annot_responsecurves` | Dilution series definitions for response QC (RQC). |
| Calibration | `annot_standards` | Known concentrations of standards in calibration and QC samples. |

## How They Connect

All tables are linked by shared identifiers:

- **`analysis_id`** — links rows in `dataset` to rows in `annot_analyses`
- **`feature_id`** — links measurements to feature properties in `annot_features`
- **`batch_id`** — connects analyses to batch-level metadata in `annot_batches`

```{r, eval = FALSE}
library(mrmhub)
mexp <- MRMhubExperiment()
mexp <- data_load_example(mexp, 1)

# Access data
get_analyticaldata(mexp)

# Access metadata
mexp$annot_analyses
mexp$annot_features
```

## See Also

- [Data Identifiers](manual-02-data-identifiers.html) — detailed explanation of ID columns
- [Feature Variables](manual-03-feature-variables.html) — all measurement variables
- [MRMhubExperiment Object](manual-04-mrmhub-experiment.html) — creating and using the object
- [Key Concepts & Glossary](manual-00-key-concepts.html) — plain-language overview
