---
title: "Import & Export mzTab-M"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  Exchange data with the HUPO-PSI mzTab-M standard format: export a processed
  MRMhubExperiment for sharing or MetaboLights submission, and import results
  from tools such as Lipid Data Analyzer, MS-DIAL or MZmine.
vignette: >
  %\VignetteIndexEntry{Import & Export mzTab-M}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

::: {.callout .callout-note}
**Level:** Intermediate &nbsp;|&nbsp;
**Output:** `.mzTab` file (mzTab-M 2.0.0-M) &nbsp;|&nbsp;
**Requires:** a processed `MRMhubExperiment`
:::

## Goal

Export the quantitative results of a processed experiment to
[mzTab-M](https://github.com/HUPO-PSI/mzTab-M), the HUPO-PSI community standard
for reporting metabolomics / lipidomics quantification. mzTab-M is a plain,
tab-delimited text format that opens in Excel yet is fully machine-readable,
and is the format expected by repositories such as
[MetaboLights](https://www.ebi.ac.uk/metabolights/).

## Prerequisites

```{r}
library(mrmhub)

# A fully processed MRMhubExperiment. Here we build one from the bundled
# `lipidomics_dataset`; in practice this would be your own processed object.
mexp <- lipidomics_dataset |>
  normalize_by_istd() |>
  quantify_by_istd()
```

## Basic export

```{r}
# write to a temporary directory for this example; use your own path in practice
out_dir <- tempdir()

save_dataset_mztab(mexp, file.path(out_dir, "experiment.mzTab"))
```

By default the **final concentrations** (`feature_conc`) are written as the
per-sample abundances, with the concentration unit declared in the file
header. If the experiment has not been quantified, the exporter automatically
falls back to the raw `feature_intensity` and declares an
"Arbitrary quantification unit".

Choose a different abundance variable with `variable`:

```{r}
save_dataset_mztab(mexp, file.path(out_dir, "raw_areas.mzTab"), variable = "area")
save_dataset_mztab(
  mexp,
  file.path(out_dir, "intensities.mzTab"),
  variable = "intensity"
)
```

## What gets written

The full dataset is exported &mdash; every analysis (including QC, blank and
calibration samples) and every feature:

| mzTab-M section | mrmhub source |
|---|---|
| `MTD` metadata | experiment title, units, one `ms_run`/`assay` per analysis, one `study_variable` per `qc_type` |
| `SMF` (feature) | one row per `feature_id` (quantifiers, qualifiers **and** ISTDs); `abundance_assay[n]` = chosen variable; ISTDs flagged via `opt_global_is_internal_standard` |
| `SML` (summary) | one row per analyte, grouping its features; the quantifier drives the summary abundance and per-group mean / %CV |
| `SME` (evidence) | a minimal identification stub per feature |

The metadata header can be enriched with optional arguments:

```{r}
save_dataset_mztab(
  mexp, file.path(out_dir, "experiment.mzTab"),
  instrument = "Agilent 6495C QqQ",
  contact = "Jane Doe",
  publication = "doi:10.1234/example"
)
```

::: {.callout .callout-caution}
**mzTab-M is a quantification report, not a full processing record.**
Internal-standard relationships, QC and calibration metrics,
drift/batch-correction state, and the QC-type / batch structure are not part of
the mzTab-M model and are therefore *not* reproduced on round-trip. The
file captures identities, the chosen abundance matrix, and the study-variable
grouping. Keep the `MRMhubExperiment` (or the Excel report from
`save_report_xlsx()`) as the authoritative processing record.
:::

## Validating the file

The output targets mzTab-M **2.0.0-M**. To confirm conformance, upload the file
to the official HUPO-PSI / LIFS web validator at
<https://apps.lifs-tools.org/mztabvalidator/>, or, if the reference R
package [`rmzTabM`](https://lifs-tools.github.io/rmzTabM/) is installed, parse it
back:

```{r}
# optional, GitHub-only reference implementation
m <- rmzTabM::readMzTab(file.path(out_dir, "experiment.mzTab"))
rmzTabM::extractSmallMoleculeFeatures(m)
```

`mrmhub` itself has **no runtime dependency** on `rmzTabM` &mdash; the writer is
self-contained.

## Importing mzTab-M

`import_data_mztab()` ingests mzTab-M produced by other tools &mdash; for
example [Lipid Data Analyzer](http://genome.tugraz.at/lda2/), MS-DIAL or
MZmine &mdash; into an `MRMhubExperiment`:

```{r}
mexp <- MRMhubExperiment(title = "Imported lipidomics")
mexp <- import_data_mztab(mexp, "LDA_export.mzTab")
```

Each Small Molecule Feature (`SMF`) becomes an mrmhub feature and each assay an
analysis. The per-assay abundances are imported as `feature_intensity`, and
feature identities (name, formula, neutral mass, m/z, retention time) are taken
from the `SMF`/`SML` sections. Where one analyte is reported as several
features (e.g. different adducts), the adduct is appended to keep `feature_id`
unique (`Cer d18:1/16:0 | [M-H]-`).

::: {.callout .callout-caution}
**Import is partial by nature.** mzTab-M carries a single
abundance per feature, so internal-standard relationships, QC-type assignments,
and calibration metadata are *not* present and must be supplied with
`add_metadata()`. `study_variable` groups are imported
best-effort as `batch_id` (mzTab-M has no analytical-batch concept).
:::

## Next Steps

- [Custom QC Report](recipe-02-custom-qc-report.html) &mdash; a richer
  human-readable report.
- [Feature Variables](manual-03-feature-variables.html) &mdash; what `conc`,
  `intensity`, `area` and friends mean.
- [The MRMhubExperiment Object](manual-04-mrmhub-experiment.html) &mdash; the
  slots behind the export.
