---
title: "Validating and fixing metadata"
description: >
  Validate analysis, feature, and ISTD annotations against imported data,
  diagnose the common defects, and generate a metadata template.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Validating and fixing metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="page-meta"><span class="page-kind page-kind-recipe">Recipe</span> <span class="page-prereq">Prerequisites: <a href="recipe-05-import-metadata.html">Importing metadata</a></span></p>

Most processing errors in MRMhub trace back to defects in the annotation tables rather than the
analytical data itself. The frequent offenders are a mismatched `analysis_id` between data and
annotation, a missing `qc_type` label (the one column besides `analysis_id` the importer requires),
absent `batch_id` or `analysis_order` columns that later steps depend on, and `qc_type` values that
fall outside the vocabulary MRMhub uses for QC selection. This recipe covers generating a template,
validating annotations against the imported data, and resolving the typical defects. For the
structure of each annotation table and how identifiers are matched, see
[Metadata tables and matching](manual-05-metadata.html); for the canonical `qc_type` labels, see
[Sample types and QC roles](manual-06-sample-types.html).

## Generate a metadata template

`save_metadata_templates()` writes an XLSX template with one sheet per annotation table
(`annot_analyses`, `annot_features`, `annot_istds`, `annot_responsecurves`,
`annot_qcconcentrations`), each carrying the expected column names and example rows.

```{r}
library(mrmhub)

save_metadata_templates(path = "metadata/sPerfect_metadata.xlsx")
```

Fill in the sheets in Excel, save, and proceed to import. The MRMhub Metadata Organizer template
(`save_metadata_msorganiser_template()`) is a richer alternative that bundles every metadata table
into a single workbook, for labs that prefer an Excel-driven layout.

## Extract metadata from imported data

If the imported data file already carries embedded metadata — for example an INTEGRATOR result with
`qc_type`, `batch_id`, and `analysis_order` columns — `import_metadata_from_data()` lifts those
columns into the annotation slots and links them to the dataset in a single call, returning the
updated `MRMhubExperiment`.

```{r}
myexp <- MRMhubExperiment()
myexp <- import_data_mrmhub(
  myexp,
  path = "datasets/sPerfect_MRMhub.tsv",
  import_metadata = FALSE
)

myexp <- import_metadata_from_data(myexp)
```

This is the fast path for data that already contains analysis metadata in the import file. For
manual edits or additional annotation — ISTDs, calibration concentrations, biological covariates —
fill in the XLSX template above and use the per-table importers below.

## Import and validate annotation tables

The per-table importers (`import_metadata_analyses()`, `import_metadata_features()`,
`import_metadata_istds()`, `import_metadata_responsecurves()`, `import_metadata_qcconcentrations()`)
read a CSV or Excel sheet, validate it against the imported data, and link the validated table to
the experiment.

```{r}
myexp <- import_metadata_analyses(
  myexp,
  path = "metadata/sPerfect_metadata.xlsx",
  sheet = "annot_analyses",
  excl_unmatched_analyses = FALSE
)

myexp <- import_metadata_features(
  myexp,
  path = "metadata/sPerfect_metadata.xlsx",
  sheet = "annot_features"
)

myexp <- import_metadata_istds(
  myexp,
  path = "metadata/sPerfect_metadata.xlsx",
  sheet = "annot_istds"
)
```

The importer checks the table against the data before attaching it — required columns, unique and
consistent identifiers, and canonical `qc_type` labels — and reports any defect as a `cli` message
naming the offending table, column, and IDs. The [metadata tables reference](manual-05-metadata.html)
lists the required columns and matching rules for each table. When annotation tables are assembled in
R rather than read from a file, pass them to [`add_metadata()`](../reference/add_metadata.html) as a
named list; the same validation is applied before the tables are linked.

## Common defects and fixes

<details>
<summary><strong>Column <code>qc_type</code> missing from <code>annot_analyses</code></strong></summary>

`qc_type` is the only column besides `analysis_id` that the importer strictly requires. Add it
before re-importing:

```{r}
annot_analyses <- annot_analyses |>
  dplyr::mutate(qc_type = dplyr::case_when(
    stringr::str_detect(analysis_id, "BQC") ~ "BQC",
    stringr::str_detect(analysis_id, "BLK") ~ "SBLK",
    TRUE ~ "SPL"
  ))
```

</details>

<details>
<summary><strong><code>analysis_id</code> in data not found in annotation</strong></summary>

Some injections in the imported dataset have no matching row in the annotation table. List them and
inspect the difference — leading or trailing whitespace and case mismatches are the most frequent
causes.

```{r}
data_ids <- unique(myexp@dataset_orig$analysis_id)
annot_ids <- annot_analyses$analysis_id

# IDs present in data but missing from annotation
setdiff(data_ids, annot_ids)

# Fix trailing whitespace introduced by Excel
annot_analyses <- annot_analyses |>
  dplyr::mutate(analysis_id = trimws(analysis_id))
```

`excl_unmatched_analyses = TRUE` in the importer drops unmatched analyses silently — useful only
when the omission is intentional, such as excluded conditioning injections.

</details>

<details>
<summary><strong>Non-canonical <code>qc_type</code> values</strong></summary>

MRMhub only recognises its canonical `qc_type` labels, and they are case-sensitive; see
[Sample types and QC roles](manual-06-sample-types.html) for the full set. Map any non-standard
values before validating:

```{r}
unique(annot_analyses$qc_type)

annot_analyses <- annot_analyses |>
  dplyr::mutate(qc_type = dplyr::case_when(
    qc_type == "Standard" ~ "CAL",
    qc_type == "Blank" ~ "SBLK",
    qc_type == "Pool" ~ "BQC",
    qc_type == "Sample" ~ "SPL",
    TRUE ~ qc_type
  ))
```

</details>

<details>
<summary><strong>Duplicate identifiers</strong></summary>

```{r}
annot_analyses |>
  dplyr::count(analysis_id) |>
  dplyr::filter(n > 1)
```

Common cause: copy-paste errors in Excel, or repeated header rows interpreted as data.

</details>

<details>
<summary><strong>Feature IDs in annotation do not match data</strong></summary>

```{r}
data_features <- unique(myexp@dataset_orig$feature_id)
annot_feat_ids <- annot_features$feature_id

# In data but missing from annotation
setdiff(data_features, annot_feat_ids)

# In annotation but absent from data (often method-development residue)
setdiff(annot_feat_ids, data_features)
```

A common cause is differing encoding of special characters — parentheses, slashes, unicode escapes —
between the integration software output and a hand-edited Excel file.

</details>

## Practical recommendations

- Re-import annotations after every edit. Saved Excel files held open in a separate process can be
  silently locked and read only in part.
- Save CSVs as UTF-8 without BOM. Excel's default "CSV UTF-8" export adds a byte-order mark that
  shifts the first column header.
- Avoid special characters in `analysis_id` and `feature_id` — parentheses, slashes, spaces. They
  survive R but complicate downstream column references; keep to letters, digits, underscores, and
  hyphens.
- Keep the annotation file under version control alongside the data file.

## Next steps

- [Sample types and QC roles](manual-06-sample-types.html) — canonical `qc_type` vocabulary
- [Metadata tables and matching](manual-05-metadata.html) — required columns and identifier rules
- [Importing metadata](recipe-05-import-metadata.html) — attaching metadata from files or a template
- [Importing analytical data](manual-04-data-import.html) — importing peak-area / concentration data
