---
title: "Import metadata from files or a template"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  Attach sample, feature, ISTD and QC metadata to an MRMhubExperiment, file by file,
  from Excel sheets, from a data.frame, or in one step from an MSOrganiser template.
vignette: >
  %\VignetteIndexEntry{Import metadata from files or a template}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = FALSE, message = FALSE, comment = "#>")
set.seed(1041)
options(dplyr.print_max = 10)
library(mrmhub)

if (fs::file_exists("metadata_template.xlsx")) fs::file_delete("metadata_template.xlsx")
if (fs::file_exists("metadata_msorganiser_template.xlsx")) {
  fs::file_delete("metadata_msorganiser_template.xlsx")
}
```

<p class="page-kind page-kind-recipe">Recipe</p>

MRMhub attaches sample, feature, ISTD, response-curve and QC metadata to an
`MRMhubExperiment` after the analytical data has been imported. Metadata can be added
table by table from individual files, from the sheets of a single Excel workbook, from R
`data.frame` objects, or in one step from an MSOrganiser template. For the structure of
each table, see [Metadata tables & matching](manual-05-metadata.html).

## Obtain the templates

Blank Excel templates for all metadata tables can be saved to the working directory:

```{r}
mrmhub::save_metadata_templates()
```

## Import from files and sheets

The analytical data is imported first, as outlined in
[Importing analytical data](manual-04-data-import.html), reading only the peak areas
(`import_metadata = FALSE`).

```{r}
mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(
  data = mexp,
  path = "datasets/sPerfect_MRMhub.tsv",
  import_metadata = FALSE)
```

The metadata is then added table by table. Sample and feature metadata are read here from
individual CSV files:

```{r}
mexp <- import_metadata_analyses(
  mexp,
  path = "datasets/analysis_metadata.csv",
  excl_unmatched_analyses = TRUE,
  ignore_warnings = TRUE)
mexp <- import_metadata_features(
  mexp,
  path = "datasets/feature_metadata.csv",
  ignore_warnings = TRUE)
```

Metadata can also be read from the sheets of a single Excel workbook, keeping all tables
in one file, here the internal-standard metadata:

```{r}
mexp <- import_metadata_istds(
  mexp,
  path = "datasets/metadata_tables.xlsx",
  sheet = "ISTDs",
  ignore_warnings = TRUE)
```

Metadata can equally be supplied from an R `data.frame`, allowing it to come from other
sources such as a database or a LIMS:

```{r}
df_qcinfo <- readr::read_table("datasets/qc_metadata.txt")
mexp <- import_metadata_qcconcentrations(mexp, table = df_qcinfo)
```

## Import an MSOrganiser template

Alternatively, all metadata can be imported in one step from an MSOrganiser template, an
Excel file providing tables for every metadata type, with built-in validity and integrity
checks. Obtain it from <https://github.com/SLINGhub/mrmhub> or save it directly:

```{r}
mrmhub::save_metadata_msorganiser_template()
```

Only the tables required by the intended workflow need to be completed; the import reads
all completed tables:

```{r}
mexp <- import_metadata_msorganiser(
  mexp,
  path = "datasets/sPerfect_Metadata.xlsx",
  ignore_warnings = TRUE)
```

## Next steps

- [Metadata tables & matching](manual-05-metadata.html): the structure of each metadata table
- [Validating and fixing metadata](recipe-04-validate-metadata.html): checking metadata before processing
- [Importing analytical data](manual-04-data-import.html): importing the measurement data
