---
title: "Custom QC Report"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  Create a detailed HTML QC report from a processed MRMhubExperiment
  using a parameterized Quarto template.
vignette: >
  %\VignetteIndexEntry{Custom QC Report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="page-kind page-kind-recipe">Recipe</p>

::: {.callout .callout-note}
**Level** Intermediate &nbsp;·&nbsp;
**Output** Self-contained HTML report &nbsp;·&nbsp;
**Requires** `gt`, `ggplot2`, `patchwork` (optional), `quarto` (for template rendering)
:::

## Goal

Generate a self-contained HTML QC report that includes:

- Summary statistics (feature counts, sample counts, batches)
- CV distribution plots
- Run-order scatter plots for key features
- Calibration curve summaries (if applicable)
- Feature pass/fail table
- Flagged samples

## Prerequisites

```{r}
library(mrmhub)
library(dplyr)
library(ggplot2)
library(gt)

# A fully processed MRMhubExperiment
mexp <- readRDS("results/mexp_processed.rds")
```

## Part 1: Summary Table

```{r}
summary_tbl <- tibble::tibble(
  Metric = c("Total analyses", "Study samples (SPL)", "QC samples",
             "Blanks", "Features (total)", "Features (passed QC)",
             "Batches", "Drift corrected?", "Batch corrected?"),
  Value = c(
    get_analysis_count(mexp),
    mexp@annot_analyses |> filter(qc_type == "SPL") |> nrow(),
    mexp@annot_analyses |> filter(grepl("QC", qc_type)) |> nrow(),
    mexp@annot_analyses |> filter(grepl("BLK", qc_type)) |> nrow(),
    nrow(mexp@annot_features),
    get_feature_count(mexp),
    length(unique(mexp@annot_analyses$batch_id)),
    ifelse(length(mexp@var_drift_corrected) > 0, "Yes", "No"),
    ifelse(length(mexp@var_batch_corrected) > 0, "Yes", "No")
  )
)

summary_tbl |> gt::gt() |> gt::tab_header(title = "Study Summary")
```

## Part 2: CV Distribution

```{r, fig.height=4}
cv_data <- mexp@metrics_qc |>
  select(feature_id, cv = norm_intensity_cv_bqc) |>
  filter(!is.na(cv))

ggplot(cv_data, aes(x = cv)) +
  geom_histogram(binwidth = 5, fill = "#5B8FA8", colour = "white") +
  geom_vline(xintercept = 30, linetype = "dashed", colour = "red") +
  labs(
    title = "QC CV Distribution",
    subtitle = paste0(sum(cv_data$cv <= 30), "/", nrow(cv_data),
                      " features with CV \u2264 30%"),
    x = "CV (%)", y = "Count"
  ) +
  theme_minimal()
```

## Part 3: Run-Order Scatter (Worst Features)

```{r, fig.height=6}
# Plot run scatter for the 4 features with highest CV
top_cv_features <- cv_data |>
  arrange(desc(cv)) |>
  head(4) |>
  pull(feature_id)

plots <- lapply(top_cv_features, function(feat) {
  plot_runscatter(mexp,
                  variable = "norm_intensity",
                  include_feature_filter = feat,
                  qc_types = c("BQC", "SPL")) +
    ggtitle(feat)
})

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(plots, ncol = 2)
}
```

## Part 4: Feature Pass/Fail Table

```{r}
qc_table <- mexp@metrics_qc |>
  select(feature_id, norm_intensity_cv_bqc, normint_dratio_sd_bqc) |>
  mutate(
    cv_pass = norm_intensity_cv_bqc <= 30,
    dratio_pass = normint_dratio_sd_bqc <= 0.5,
    status = case_when(
      cv_pass & dratio_pass ~ "PASS",
      !cv_pass & !dratio_pass ~ "FAIL (CV + D-ratio)",
      !cv_pass ~ "FAIL (CV)",
      !dratio_pass ~ "FAIL (D-ratio)"
    )
  ) |>
  arrange(desc(norm_intensity_cv_bqc))

qc_table |>
  gt::gt() |>
  gt::fmt_number(columns = c(norm_intensity_cv_bqc, normint_dratio_sd_bqc), decimals = 1) |>
  gt::data_color(
    columns = status,
    fn = function(x) ifelse(x == "PASS", "#d4e8d4", "#f8d4d4")
  ) |>
  gt::tab_header(title = "Feature QC Summary")
```

## Part 5: Calibration Summary (if applicable)

```{r}
if (nrow(mexp@metrics_calibration) > 0) {
  cal_summary <- get_calibration_metrics(mexp) |>
    select(feature_id, r2, fit_model, fit_weighting, lowest_cal, highest_cal) |>
    arrange(r2)

  cal_summary |>
    gt::gt() |>
    gt::fmt_number(columns = c(r2, lowest_cal, highest_cal), decimals = 4) |>
    gt::tab_header(title = "Calibration Curve Metrics")
}
```

## Part 6: Flagged Samples

```{r}
if (length(mexp@analyses_excluded) > 0) {
  tibble::tibble(
    analysis_id = mexp@analyses_excluded,
    reason = "Excluded during processing"
  ) |> gt::gt() |> gt::tab_header(title = "Excluded Analyses")
}
```

## Parameterized Quarto Template

<details>
<summary><strong>Full template (click to expand)</strong></summary>

Save as `qc-report-template.qmd`:

````
---
title: "QC Report"
date: today
format:
  html:
    self-contained: true
    toc: true
params:
  rds_path: "results/mexp_processed.rds"
---
````

````
```{r setup}
#| include: false
library(mrmhub)
library(dplyr)
library(ggplot2)
library(gt)
mexp <- readRDS(params$rds_path)
```
````

Then include each Part above as a code chunk in the template.

</details>

Render the template:

```{r}
quarto::quarto_render(
  "qc-report-template.qmd",
  execute_params = list(rds_path = "results/mexp_processed.rds")
)
```

## Tips

- **Self-contained HTML** — a single file that can be emailed or archived.
- **Parameterize** the RDS path so the same template works for any study.
- **Date stamp** (`date: today`) for an audit trail.
- **Session info** at the end for reproducibility.

## Next steps

- [Basic MRMhub Workflow](tutorial-02-basic-workflow.html) — full processing before reporting
- [External Calibration & QC](recipe-01-ext-calibration-qc.html) — calibration workflow
- [Troubleshooting & FAQ](manual-10-troubleshooting.html) — common issues
