---
title: "Custom QC report"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  Create a detailed HTML QC report from a processed MRMhubExperiment
  using a parameterized Quarto template.
vignette: >
  %\VignetteIndexEntry{Custom QC report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="page-meta"><span class="page-kind page-kind-recipe">Recipe</span> <span class="page-prereq">Prerequisites: <a href="tutorial-02-basic-workflow.html">Basic workflow</a></span></p>

A QC report gathers the quality indicators for a processed experiment into a single
self-contained HTML file that can be shared, emailed or archived. This recipe assembles one
from a fully processed `MRMhubExperiment`: a summary of sample and feature counts, the QC CV
distribution, run-order scatter plots for the least reproducible features, a feature pass/fail
table, calibration metrics where a calibration was run, and any excluded analyses. Each piece
can be dropped into the parameterized Quarto template given at the end, so the same report
regenerates for any study.

The report reads an experiment that has already been through import, metadata annotation,
normalization and QC-metric calculation:

```{r}
library(mrmhub)
library(dplyr)
library(ggplot2)
library(gt)

mexp <- readRDS("results/mexp_processed.rds")
```

## Summary table

A compact overview: counts of analyses, samples, blanks and features, the number of batches,
and whether drift and batch correction were applied.

```{r}
summary_tbl <- tibble::tibble(
  Metric = c("Total analyses", "Study samples (SPL)", "QC samples",
             "Blanks", "Features (total)", "Features (measured)",
             "Batches", "Drift corrected?", "Batch corrected?"),
  Value = c(
    get_analysis_count(mexp),
    mexp@annot_analyses |> filter(qc_type == "SPL") |> nrow(),
    mexp@annot_analyses |> filter(grepl("QC", qc_type)) |> nrow(),
    mexp@annot_analyses |> filter(grepl("BLK", qc_type)) |> nrow(),
    nrow(mexp@annot_features),
    get_feature_count(mexp),
    length(unique(mexp@annot_analyses$batch_id)),
    ifelse(any(mexp@var_drift_corrected), "Yes", "No"),
    ifelse(any(mexp@var_batch_corrected), "Yes", "No")
  )
)

summary_tbl |> gt::gt() |> gt::tab_header(title = "Study Summary")
```

## CV distribution

The coefficient of variation across the batch QC (BQC) samples is the primary precision
metric; the histogram shows how many features fall within a chosen CV limit. The 30 % line
here is illustrative. See [Sample types and QC roles](manual-06-sample-types.html) for which
QC types drive this metric.

```{r, fig.height=4}
cv_data <- mexp@metrics_qc |>
  select(feature_id, cv = norm_intensity_cv_bqc) |>
  filter(!is.na(cv))

ggplot(cv_data, aes(x = cv)) +
  geom_histogram(binwidth = 5, fill = "#5B8FA8", colour = "white") +
  geom_vline(xintercept = 30, linetype = "dashed", colour = "red") +
  labs(
    title = "QC CV Distribution",
    subtitle = paste0(sum(cv_data$cv <= 30), "/", nrow(cv_data),
                      " features with CV \u2264 30%"),
    x = "CV (%)", y = "Count"
  ) +
  theme_minimal()
```

## Run-order scatter for the worst features

Plotting the least reproducible features against acquisition order shows whether the high CV
comes from drift, a step change between batches, or scattered outliers.

```{r, fig.height=6}
top_cv_features <- cv_data |>
  arrange(desc(cv)) |>
  head(4) |>
  pull(feature_id)

plots <- lapply(top_cv_features, function(feat) {
  plot_runscatter(mexp,
                  variable = "norm_intensity",
                  include_feature_filter = feat,
                  qc_types = c("BQC", "SPL")) +
    ggtitle(feat)
})

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(plots, ncol = 2)
}
```

## Feature pass/fail table

A per-feature verdict from the BQC CV and D-ratio. The thresholds below (CV ≤ 30 %,
D-ratio ≤ 0.5) are illustrative defaults; MRMhub's own feature filtering is applied with
`filter_features_qc()`.

```{r}
qc_table <- mexp@metrics_qc |>
  select(feature_id, norm_intensity_cv_bqc, normint_dratio_sd_bqc) |>
  mutate(
    cv_pass = norm_intensity_cv_bqc <= 30,
    dratio_pass = normint_dratio_sd_bqc <= 0.5,
    status = case_when(
      cv_pass & dratio_pass ~ "PASS",
      !cv_pass & !dratio_pass ~ "FAIL (CV + D-ratio)",
      !cv_pass ~ "FAIL (CV)",
      !dratio_pass ~ "FAIL (D-ratio)"
    )
  ) |>
  arrange(desc(norm_intensity_cv_bqc))

qc_table |>
  gt::gt() |>
  gt::fmt_number(
    columns = c(norm_intensity_cv_bqc, normint_dratio_sd_bqc),
    decimals = 1) |>
  gt::data_color(
    columns = status,
    fn = function(x) ifelse(x == "PASS", "#d4e8d4", "#f8d4d4")
  ) |>
  gt::tab_header(title = "Feature QC Summary")
```

## Calibration summary

Where the experiment was quantified against external calibration curves, the fit metrics
summarise each curve: model, weighting, R² and the calibrated range.

```{r}
if (nrow(mexp@metrics_calibration) > 0) {
  cal_summary <- get_calibration_metrics(mexp) |>
    select(
      feature_id, r2, fit_model, fit_weighting,
      lowest_cal, highest_cal) |>
    arrange(r2)

  cal_summary |>
    gt::gt() |>
    gt::fmt_number(columns = c(r2, lowest_cal, highest_cal), decimals = 4) |>
    gt::tab_header(title = "Calibration Curve Metrics")
}
```

## Excluded analyses

Any analyses removed from processing and reporting, kept in the report for the audit trail.

```{r}
if (!all(is.na(mexp@analyses_excluded))) {
  tibble::tibble(
    analysis_id = mexp@analyses_excluded,
    reason = "Excluded during processing"
  ) |> gt::gt() |> gt::tab_header(title = "Excluded Analyses")
}
```

## Parameterized Quarto template

<details>
<summary><strong>Full template (click to expand)</strong></summary>

Save as `qc-report-template.qmd`:

````
---
title: "QC Report"
date: today
format:
  html:
    self-contained: true
    toc: true
params:
  rds_path: "results/mexp_processed.rds"
---
````

````
```{r setup}
#| include: false
library(mrmhub)
library(dplyr)
library(ggplot2)
library(gt)
mexp <- readRDS(params$rds_path)
```
````

Then include each section above as a code chunk in the template.

</details>

Render the template:

```{r}
quarto::quarto_render(
  "qc-report-template.qmd",
  execute_params = list(rds_path = "results/mexp_processed.rds"))
```

## Tips

- **Self-contained HTML**: a single file that can be emailed or archived.
- **Parameterize** the RDS path so the same template works for any study.
- **Date stamp** (`date: today`) for an audit trail.
- **Session info** at the end for reproducibility.
- **Export the figures too**: `save_plot(p, "figures/cv", format = c("pdf", "png"), width = 180,
  height = 120)` writes a print-ready vector file and a web-ready raster one in a single call. See
  [Saving plots](manual-08-visualization.html#saving-plots).

## Next steps

- [Basic MRMhub workflow](tutorial-02-basic-workflow.html): full processing before reporting
- [External calibration & QC](recipe-01-ext-calibration-qc.html): calibration workflow
- [Troubleshooting & FAQ](manual-10-troubleshooting.html): common issues
