---
title: "Build a workflow without code"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
description: >
  Use the point-and-click builder to validate your data and metadata and
  generate a runnable Quarto workflow.
vignette: >
  %\VignetteIndexEntry{Build a workflow without code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
source("_common.R")
knitr::opts_chunk$set(eval = FALSE)
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Beginner</span> <span class="page-prereq">Prerequisites: <a href="manual-00-installation.html">MRMhub installed</a></span></p>

`build_workflow()` opens an interactive application that turns a data file and
its metadata into a runnable Quarto (`.qmd`) workflow. It validates the inputs
against the real importers, disables processing steps the metadata cannot
support, and previews the generated document for download, while the code-first
workflow it produces remains the reproducible artefact.

## 1. Launching the builder

```{r}
library(mrmhub)

build_workflow()
```

The application runs locally in the default browser. It depends on the `shiny`
and `bslib` packages; if either is missing, the first call prompts to install
them (they are not pulled in when `mrmhub` itself is installed). To try the
builder without any files of your own, use the **Load bundled demo data** button,
which loads the INTEGRATOR demo dataset shipped with the package.

## 2. The sidebar, step by step

The sidebar is worked top to bottom in four numbered sections.

**1 · Data source.** Choose the tool or format that produced the data (this
selects the matching importer: `import_data_mrmhub()`,
`import_data_masshunter()`, `import_data_skyline()`, or the generic
`import_data_csv_long()` / `import_data_csv_wide()`), then upload the file. For
a generic CSV the column names are not fixed, so a dialog opens to map them:

- **Long CSV**: pick the columns holding the *analysis id*, the *feature id*,
  and the *value*, plus what the value represents (area, height, intensity,
  concentration, or normalised intensity). These become a `column_mapping`.
- **Wide CSV**: state what the matrix values represent and, optionally, the
  analysis-id column and the index of the first feature column.

::: {.callout .callout-caution}
Until a generic CSV's columns are mapped the data cannot be validated: this is
the same requirement `import_data_csv_long()` enforces when it reports a missing
`feature_id` column. The other importers read fixed column names and need no
mapping.
:::

**2 · Metadata.** Select where sample, feature, and internal-standard
annotations come from: embedded in the data file, an MSOrganiser workbook, a
multi-sheet metadata workbook, or **individual files**: separate CSVs for
analyses, features, and ISTDs imported with `import_metadata_analyses()`,
`import_metadata_features()`, and `import_metadata_istds()`.

**3 · Project folder.** Optionally point the builder at a project directory
(the **Browse…** button opens a native folder picker). When set, the uploaded
files are copied into `<folder>/data/`, the report is written to
`<folder>/output/`, and **Create project & write .qmd** lays down the directory
structure with the workflow document beside it. Left empty, the generated code
references the data file by the path shown in the *Data file path* field, which
can be edited to an absolute path.

**4 · Processing steps.** Tick the steps to include. Steps whose metadata is
absent are disabled and annotated with a short reason and an information icon
carrying the full explanation: for example, *Quantify by internal standard* is
disabled with "No ISTD concentrations" until an ISTD concentration table is
imported. See [Reading the validation panel](#reading-the-validation-panel)
below.

Drift and batch correction share a **method** selector (Gaussian kernel, cubic
spline, or LOESS) that also sets the reference QC type: the Gaussian kernel uses
study samples (`SPL`); spline and LOESS use pooled QCs (the first of `BQC`,
`TQC`, or `QC` present in the data). Batch centring reuses the same reference.
The corrected feature variable is chosen automatically as the highest processing
level the selected steps reach: `conc` once a quantification step is included,
otherwise `norm_intensity`, otherwise raw `intensity`.

## 3. Reading the validation panel

The **Validation** panel runs the selected steps' preconditions against the
imported experiment and reports one line per finding:

- **✗ error**: the import failed, or a selected step would abort (for example,
  `normalize_by_istd()` selected while no feature carries an `istd_feature_id`).
- **⚠ warning**: a step needs an upstream step or an optional field (for
  example, a drift reference QC type absent from the data).
- **✓ ok**: the data imported and the selected steps have no outstanding
  issues.

Because the panel calls the same importer and precondition checks as a scripted
run, a workflow that validates cleanly here runs the same way from the console.

## 4. What it generates

The **Generated workflow** pane updates live as the sidebar changes. The YAML
front matter records the title and the requested output formats:

```yaml
---
title: "MRMhub Workflow"
toc: true
execute:
  warning: true
format:
  html: default
---
```

The document then opens with a hidden setup chunk (`#| include: false`) that
loads the package and calls `mrmhub_enable_cli_color()`, so mrmhub's coloured
console feedback renders in the HTML output. See
[Quarto workflows](manual-11-quarto-workflows.html) for the reporting options and
output-format details.

Each selected step then becomes its own `##` section with an `{r}` chunk. For
the demo dataset with normalisation, ISTD quantification, drift and batch
correction, and QC filtering selected, and a project folder set, the emitted
code is, in order:

```{r}
mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(
  mexp, path = "data/MRMhub_demo.tsv", import_metadata = TRUE)

mexp <- normalize_by_istd(mexp)
mexp <- quantify_by_istd(mexp)

mexp <- correct_drift_gaussiankernel(
  mexp, variable = "conc", ref_qc_types = "SPL")
mexp <- correct_batch_centering(
  mexp, variable = "conc", ref_qc_types = "SPL")

mexp <- calc_qc_metrics(mexp)
mexp <- filter_features_qc(
  mexp, include_qualifier = FALSE, include_istd = FALSE)

save_report_xlsx(mexp, path = "output/results.xlsx")
```

The emitted calls and their argument defaults mirror the
[Basic MRMhub workflow](tutorial-02-basic-workflow.html); the builder is a way to
assemble that script, not a separate engine. **Copy code** copies the document to
the clipboard, and **Download .qmd** (or the project button) writes it to disk.

## 5. Rendering the workflow

The **Output format(s)** control writes the requested Quarto formats into the
document's YAML: HTML, Word (`.docx`), and PDF may be combined. The PDF format
is configured for a sans-serif typeface using the default LaTeX engine. Rendering
is performed with Quarto once the document is saved:

```bash
quarto render mrmhub_workflow.qmd
```

::: {.callout .callout-tip}
The generated file is an ordinary Quarto document. Open it in RStudio, Positron,
or VS Code, adjust paths or parameters, and run the chunks interactively; the
builder is a starting point, not a black box.
:::

For scripted use, the same document can be produced without the application by
passing a specification list to `generate_workflow_qmd()`.

## Next steps

- [Your first analysis](tutorial-00-first-analysis.html): the equivalent
  five-minute workflow written by hand
- [Basic MRMhub workflow](tutorial-02-basic-workflow.html): the full pipeline
  the builder mirrors
- [Importing analytical data](manual-04-data-import.html): file formats and the
  importer decision tree
- [Importing metadata](manual-05-metadata.html): metadata tables and their
  required columns
- [Quarto workflows](manual-11-quarto-workflows.html): the setup chunk, coloured
  console output, and rendering to HTML, PDF, and Word
