---
title: "Drift and batch correction"
description: >
  Correct run-order signal drift and between-batch effects from QC samples, and
  combine the two in the recommended order.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Drift and batch correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Intermediate</span> <span class="page-prereq">Prerequisites: <a href="tutorial-02-basic-workflow.html">Basic workflow</a></span></p>

Signal intensities in a mass-spectrometry run drift with injection order and shift between
analytical batches. MRMhub corrects run-order drift by smoothing a trend through reference
samples, and batch effects by median centering. Both operate on a data variable of an
`MRMhubExperiment` (raw intensities, normalized intensities, or concentrations) and are
fitted on QC or study samples only, never mixed. After this tutorial you can select an appropriate correction method for a given
QC design and apply drift and batch correction in the correct order.

## 1. Import data

We import pre-calculated raw concentration values from a CSV file. Batch-wise correction requires a
`batch_id` column; see `import_data_csv_wide()` for the expected format.

```{r import}
library(mrmhub)

myexp <- MRMhubExperiment()
myexp <- import_data_csv_wide(
  myexp,
  path = "smooth-testdata.csv",
  variable_name = "conc",
  import_metadata = TRUE
)
```

## 2. QC-based drift correction

Apply a QC-based drift correction with a cubic spline fitted through the batch QC samples (`BQC`).
The reported change in median CV summarises whether the correction improved analytical precision.

```{r drift-cubicspline}
#| message: true
#| code-overflow: wrap
mexp_drift <- correct_drift_cubicspline(
  myexp,
  variable = "conc",
  batch_wise = FALSE,
  ref_qc_types = "BQC",
  recalc_trend_after = TRUE
)
```

```{r drift-cubicspline-before}
#| fig.cap: !expr fig_cap("Concentrations before QC-based drift correction; the fitted trend (green) follows the BQC points (red).")
#| fig.alt: RunScatter of concentrations before QC-based drift correction
plot_runscatter(
  mexp_drift,
  variable = "conc_before",
  qc_types = c("BQC", "SPL"),
  rows_page = 1, cols_page = 3, show_trend = TRUE
)
```

```{r drift-cubicspline-after}
#| fig.cap: !expr fig_cap("Concentrations after QC-based drift correction; the BQC trend is flat.")
#| fig.alt: RunScatter of concentrations after QC-based drift correction
plot_runscatter(
  mexp_drift,
  variable = "conc",
  qc_types = c("BQC", "SPL"),
  rows_page = 1, cols_page = 3, show_trend = TRUE
)
```

The fitted trend (green) tracks the `BQC` points (red) before correction and is flat afterwards.
Flattening the `BQC` trend does not, however, necessarily straighten the study-sample (`SPL`) trend,
since pooled QCs can differ from the study samples in handling and matrix.

## 3. Sample-based drift correction

When QC samples do not represent the study-sample trend, fit the drift on the study samples instead.
Gaussian kernel smoothing is suited to this because study samples are numerous but individually
noisy; `kernel_size` sets the smoothing window. The study-sample trend is then well corrected,
though for this dataset the difference from the QC-based result above is modest.

```{r drift-gaussian}
#| code-overflow: wrap
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  batch_wise = FALSE,
  ref_qc_types = "SPL",
  kernel_size = 10,
  recalc_trend_after = TRUE
)
```

```{r drift-gaussian-after}
#| fig.cap: !expr fig_cap("Concentrations after sample-based Gaussian-kernel drift correction; the study-sample trend is flat.")
#| fig.alt: RunScatter after sample-based Gaussian-kernel drift correction
plot_runscatter(
  mexp_drift,
  variable = "conc",
  qc_types = c("BQC", "SPL"),
  rows_page = 1, cols_page = 3, show_trend = TRUE
)
```

## 4. Within-batch drift correction

The corrections above span all batches. When batch effects interrupt the drift, fit the trend within
each batch by setting `batch_wise = TRUE`.

```{r drift-batchwise}
#| code-overflow: wrap
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  batch_wise = TRUE,
  ref_qc_types = "SPL",
  kernel_size = 10,
  recalc_trend_after = TRUE
)
```

```{r drift-batchwise-after}
#| fig.cap: !expr fig_cap("Concentrations after within-batch drift correction; residual trend differences remain between batches.")
#| fig.alt: RunScatter after within-batch drift correction, residual between-batch differences
plot_runscatter(
  mexp_drift,
  variable = "conc",
  qc_types = c("BQC", "SPL"),
  rows_page = 1, cols_page = 3, show_trend = TRUE
)
```

Clear trend differences remain between batches because each batch is fitted independently, so
differing sample sizes and batch effects are not reconciled. A batch correction is therefore
usually applied after batch-wise drift correction. Apply a subsequent batch correction by median
centering to align the batches.

```{r drift-then-batch}
#| fig.cap: !expr fig_cap("Concentrations after within-batch drift correction followed by median-centering batch correction; batches are aligned.")
#| fig.alt: RunScatter after within-batch drift correction and batch centering
mexp_drift <- correct_batch_centering(
  mexp_drift,
  variable = "conc",
  ref_qc_types = "SPL",
  correct_scale = TRUE
)

plot_runscatter(
  mexp_drift,
  variable = "conc",
  qc_types = c("BQC", "SPL"),
  rows_page = 1, cols_page = 3, show_trend = TRUE
)
```

## 5. Batch-effect correction

Batch effects (systematic differences between analytical batches) are corrected on their own with
`correct_batch_centering()`. The following uses a seven-batch dataset to show the effect clearly.
Each batch is centered on a reference QC type, here the study samples (`ref_qc_types = "SPL"`).

```{r batch-import}
myexp_batch <- MRMhubExperiment()
myexp_batch <- import_data_csv_wide(
  myexp_batch,
  path = "simdata-u1000-sd100_7batches.csv",
  variable_name = "conc",
  import_metadata = TRUE
)
```

```{r batch-center}
#| message: true
#| code-overflow: wrap
myexp_batch <- correct_batch_centering(
  myexp_batch,
  variable = "conc",
  ref_qc_types = "SPL",
  correct_scale = FALSE
)
```

```{r batch-before}
#| fig.cap: !expr fig_cap("Concentrations before batch correction; batch medians are offset from one another.")
#| fig.alt: RunScatter before batch correction, batches offset
plot_runscatter(
  myexp_batch,
  variable = "conc_before",
  rows_page = 1, cols_page = 1
)
```

```{r batch-after}
#| fig.cap: !expr fig_cap("Concentrations after median-centering batch correction; batch medians are aligned.")
#| fig.alt: RunScatter after median-centering batch correction, batches aligned
plot_runscatter(
  myexp_batch,
  variable = "conc",
  rows_page = 1, cols_page = 1
)
```

The batches are aligned in location, but their spread still differs. Setting `correct_scale = TRUE`
also equalises the variance between batches.

```{r batch-scale}
#| fig.cap: !expr fig_cap("Concentrations after batch correction with variance scaling; both location and spread are consistent across batches.")
#| fig.alt: RunScatter after batch correction with variance scaling
myexp_batch <- correct_batch_centering(
  myexp_batch,
  variable = "conc",
  ref_qc_types = "SPL",
  correct_scale = TRUE
)

plot_runscatter(
  myexp_batch,
  variable = "conc",
  rows_page = 1, cols_page = 1
)
```

## 6. Choosing a drift-correction method

MRMhub provides four drift-correction methods. Loess and cubic spline are typically fitted on QC
samples, Gaussian kernel on study samples; a fourth, GAM smoothing via `correct_drift_gam()`, is also
available. See [Drift and batch correction (reference)](manual-07-corrections.html) for the full
parameter description of all four.

| Method | Function | Reference | Typical use |
|--------|----------|-----------|-------------|
| Loess | `correct_drift_loess()` | QC | Frequent QC injections; robust to single outlier QCs |
| Cubic spline | `correct_drift_cubicspline()` | QC | Frequent QC injections; flexible, sensitive to outlier QCs |
| Gaussian kernel | `correct_drift_gaussiankernel()` | Study | Sparse QCs; large, well-randomised sample sets only |

Loess uses locally weighted regression and is less sensitive to individual outlier QCs than a cubic
spline; its `span` controls smoothness.

```{r drift-loess}
#| code-overflow: wrap
mexp_drift_loess <- correct_drift_loess(
  myexp,
  variable = "conc",
  batch_wise = TRUE,
  ref_qc_types = "BQC",
  recalc_trend_after = TRUE
)
```

## 7. Export corrected data

Continue processing the corrected object with MRMhub functions, or export the corrected variable
to CSV.

```{r export}
#| message: true
save_dataset_csv(
  mexp_drift,
  path = "drift-batch-corrected-conc-data.csv",
  variable = "conc",
  filter_data = FALSE
)
```

## Next steps

- [RunScatter and PCA QC exploration](tutorial-05-run-scatter.html): visualise run-order and batch effects
- [Drift and batch correction (reference)](manual-07-corrections.html): full method documentation
- [Calibration by a reference sample](tutorial-07-calibration-reference.html): normalise to a reference material
