---
title: "Exporting to standard and community formats"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  Move a processed MRMhubExperiment into interchange formats: mzTab-M for
  repositories, or a Bioconductor SummarizedExperiment for downstream analysis.
vignette: >
  %\VignetteIndexEntry{Exporting to standard and community formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("_common.R")
knitr::opts_chunk$set(eval = FALSE)
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-prereq">Prerequisites: <a href="tutorial-03-lipidomics-workflow.html">Full workflow</a></span></p>

Sharing results or running downstream statistics usually means handing the data
to another tool. MRMhub offers two export routes: **mzTab-M**, the HUPO-PSI
community standard that repositories expect, and a Bioconductor
**SummarizedExperiment** for downstream analysis with `limma` (differential
abundance) or `lipidr` (lipid-specific analysis).

Both examples start from a processed object. Here we build one from the bundled
`lipidomics_dataset`; in practice this would be your own.

```{r}
library(mrmhub)

mexp <- lipidomics_dataset |>
  normalize_by_istd() |>
  quantify_by_istd()
```

## 1. mzTab-M

[mzTab-M](https://github.com/HUPO-PSI/mzTab-M) is the HUPO-PSI community standard
for reporting metabolomics and lipidomics quantification. It is a plain,
tab-delimited text format that opens in Excel yet is fully machine-readable, and
is what repositories such as [MetaboLights](https://www.ebi.ac.uk/metabolights/)
expect on submission.

`save_dataset_mztab()` writes a processed experiment to an mzTab-M 2.0.0-M file:

```{r}
out_dir <- tempdir()

save_dataset_mztab(mexp, file.path(out_dir, "experiment.mzTab"))
```

By default the final concentrations (`feature_conc`) are written as the
per-sample abundances, with the concentration unit declared in the header. If the
experiment has not been quantified, the exporter falls back to the raw
`feature_intensity` and declares an "Arbitrary quantification unit". Choose a
different abundance variable with `variable`:

```{r}
save_dataset_mztab(
  mexp, file.path(out_dir, "raw_areas.mzTab"),
  variable = "area")
```

The full dataset is exported, every analysis (including QC, blank and
calibration samples) and every feature:

| mzTab-M section | mrmhub source |
|---|---|
| `MTD` metadata | title, units, one `ms_run`/`assay` per analysis, one `study_variable` per `qc_type` |
| `SMF` (feature) | one row per `feature_id` (quantifiers, qualifiers **and** ISTDs); `abundance_assay[n]` = chosen variable; ISTDs flagged via `opt_global_is_internal_standard` |
| `SML` (summary) | one row per analyte, grouping its features; the quantifier drives the summary abundance and per-group mean / %CV |
| `SME` (evidence) | a minimal identification stub per feature |

Enrich the metadata header with optional arguments:

```{r}
save_dataset_mztab(
  mexp, file.path(out_dir, "experiment.mzTab"),
  instrument = "Agilent 6495C QqQ",
  contact = "Jane Doe",
  publication = "doi:10.1234/example")
```

mzTab-M is a quantification report, not a full processing record: internal-standard
relationships, QC and calibration metrics, and drift/batch state are not part of the
model and are not reproduced on round-trip. Keep the `MRMhubExperiment` (or the Excel
report from `save_report_xlsx()`) as the definitive record.

### Validating the file

The output targets mzTab-M **2.0.0-M**. To confirm conformance, upload the file
to the HUPO-PSI / LIFS web validator at
<https://apps.lifs-tools.org/mztabvalidator/>, or parse it back with the
reference R package [`rmzTabM`](https://lifs-tools.github.io/rmzTabM/):

```{r}
m <- rmzTabM::readMzTab(file.path(out_dir, "experiment.mzTab"))
rmzTabM::extractSmallMoleculeFeatures(m)
```

MRMhub has **no runtime dependency** on `rmzTabM`; the writer is
self-contained.

### Importing mzTab-M

`import_data_mztab()` ingests mzTab-M produced by other tools (for example
[Lipid Data Analyzer](http://genome.tugraz.at/lda2/), MS-DIAL or MZmine)
into an `MRMhubExperiment`:

```{r}
mexp_in <- MRMhubExperiment(title = "Imported lipidomics")
mexp_in <- import_data_mztab(mexp_in, "LDA_export.mzTab")
```

Each Small Molecule Feature (`SMF`) becomes a feature and each assay an analysis.
The per-assay abundances are imported as `feature_intensity`, and feature
identities (name, formula, neutral mass, m/z, retention time) come from the
`SMF`/`SML` sections. Where one analyte is reported as several features (e.g.
different adducts), the adduct is appended to keep `feature_id` unique
(`Cer d18:1/16:0 | [M-H]-`).

Import is partial by nature: mzTab-M carries a single abundance per feature, so
internal-standard relationships, QC-type assignments, and calibration metadata
are absent and must be supplied with `add_metadata()`. `study_variable` groups
are imported best-effort as `batch_id`.

## 2. SummarizedExperiment

[SummarizedExperiment](https://bioconductor.org/packages/SummarizedExperiment/)
is the Bioconductor container many statistical and lipidomics tools consume.
`save_dataset_summarizedexperiment()` produces one, with features as **rows** and
analyses as **columns**:

```{r}
library(SummarizedExperiment)

se <- save_dataset_summarizedexperiment(mexp)
se
#> class: SummarizedExperiment
#> dim: 29 499
#> metadata(11): title analysis_type ... var_batch_corrected mrmhub_version
#> assays(9): rt area ... pmol_total conc
#> rownames(29): CE 18:1 CE 18:1 d7 (ISTD) ... TG 48:2 [-18:1] TG 48:2 [SIM]
#> rowData names(18): feature_id feature_class ... remarks feature_label
#> colnames(499): Longit_BLANK-01 (Eluent A) ... Longit_BLANK-07 (Eluent A)
#> colData names(13): analysis_order analysis_id ... annot_order_num remarks
```

Install the package once with `BiocManager::install("SummarizedExperiment")`.
Pass a `path` to also write the object to disk as an `.rds`; without one it is
only returned:

```{r}
save_dataset_summarizedexperiment(mexp, "experiment.rds")
```

The mapping from experiment slots is:

| Component | mrmhub source |
|---|---|
| `assays()` | one matrix per feature variable, named without the `feature_` prefix |
| `rowData()` | `annot_features`: one row per `feature_id` |
| `colData()` | `annot_analyses`: one row per `analysis_id` |
| `metadata()` | title, analysis type, processing status, `is_*` flags, concentration unit, MRMhub version |

The parallel feature variables become **parallel assays**, so raw, normalized and
quantified values live side-by-side in one object:

```{r}
assayNames(se)
#> [1] "rt"     "area"   "height" "fwhm"   "width"  "intensity"
#> [7] "norm_intensity" "pmol_total" "conc"

assay(se, "conc")[1:3, 1:2]
```

Export a subset of variables with `variable`, or QC-filtered data by filtering
first and passing `filter_data = TRUE`:

```{r}
se_conc <- save_dataset_summarizedexperiment(mexp, variable = "conc")

mexp_filt <- mexp |>
  calc_qc_metrics() |>
  filter_features_qc(
    include_qualifier = FALSE,
    include_istd = FALSE,
    max.cv.conc.bqc = 25)

se_filt <- save_dataset_summarizedexperiment(mexp_filt, filter_data = TRUE)
```

### Subsetting to study samples

Everything is exported and flagged rather than dropped, since downstream tools need
it (`lipidr`, for example, requires the internal-standard annotation). But nothing
downstream reads `qc_type`, so subset to study samples before a PCA, normalization,
or differential test; otherwise blanks and QCs are folded in with the samples:

```{r}
se_spl <- se[!rowData(se)$is_istd, se$qc_type == "SPL"]
dim(se_spl)
#> [1]  20 374
```

QC metrics are deliberately *not* written to `rowData()`: nothing downstream reads
them, and QC filtering belongs in MRMhub where it is tested. Use
`filter_data = TRUE` as above, or `save_feature_qc_metrics()` for the metrics.

### Differential abundance with limma

[limma](https://bioconductor.org/packages/limma/) works directly on continuous
data such as concentrations. `voom()` is **not** used: it models count data and
is invalid here.

::: {.callout .callout-note}
The bundled `lipidomics_dataset` carries no phenotype: every study sample is
plasma from the same longitudinal series. The grouping below is **simulated to
demonstrate the interface only**; the p-values are meaningless by construction.
With your own data the group would come from your sample metadata and already be
in `colData()`.
:::

```{r}
library(limma)

set.seed(1)
se_spl$group <- factor(
  sample(c("ctrl", "trt"), ncol(se_spl), replace = TRUE))

# concentrations are right-skewed; limma assumes normality
y <- log2(assay(se_spl, "conc"))

design <- model.matrix(~ 0 + group, data = colData(se_spl))
colnames(design) <- levels(se_spl$group)

fit <- lmFit(y, design)
fit <- contrasts.fit(fit, makeContrasts(trt - ctrl, levels = design))
fit <- eBayes(fit)

topTable(fit, number = 5)
```

Because `colData()` carries the full analysis annotation, covariates are
available without extra joins, for example blocking on batch with
`~ 0 + group + batch_id`.

### Lipid-specific analysis with lipidr

[lipidr](https://www.lipidr.org) works on a `LipidomicsExperiment`, a subclass of
`SummarizedExperiment`. Produce one directly:

```{r}
le <- save_dataset_summarizedexperiment(
  mexp,
  variable = "intensity",
  as = "LipidomicsExperiment")
```

MRMhub fills in what lipidr requires: `Molecule` from `feature_id`, `Class` from
`feature_class`, `istd` from `is_istd`, plus the `summarized` / `logged` /
`normalized` flags lipidr reads but does **not** validate; an object
missing them constructs cleanly and then misbehaves.

::: {.callout .callout-caution}
**Use a peak-area scale variable with lipidr, not concentrations.** lipidr
log-transforms with `log = TRUE` by default and **clamps values below 1 to 1**
first, which assumes Skyline peak-area magnitudes. Concentrations in µmol/L are
mostly below 1, so they are silently flattened to `log2(1) = 0`; `eBayes()` then
reports zero residual variances and every fold change collapses. MRMhub warns
when you export a mostly-sub-1 assay this way.

The trigger is absolute scale, not concentration as such: the clamp is a fixed
threshold at 1. If you need concentrations, pass `log = FALSE` to lipidr, or
log-transform them yourself beforehand.
:::

```{r}
library(lipidr)

le_spl <- le[!rowData(le)$istd, le$qc_type == "SPL"]
set.seed(1)
le_spl$group <- factor(
  sample(c("ctrl", "trt"), ncol(le_spl), replace = TRUE))

le_norm <- normalize_pqn(le_spl, measure = "intensity", log = TRUE)

de_results <- de_analysis(
  le_norm,
  trt - ctrl,
  measure = "intensity",
  group_col = "group")
plot_results_volcano(de_results, show.labels = FALSE)
```

## Next steps

- [The MRMhubExperiment data object](manual-02-data-object.html#feature-variables):
  what `conc`, `intensity` and `area` mean, and the slots behind these exports.
- [Design decisions](manual-03-design-decisions.html): why `MRMhubExperiment` is
  not itself a Bioconductor class.
- [Lipidomics workflow](tutorial-03-lipidomics-workflow.html): QC-filter features
  before exporting.
