---
title: "Basic MRMhub Workflow"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  A realistic end-to-end workflow covering project setup, metadata import, 
  drift/batch correction, QC filtering, and export.
vignette: >
  %\VignetteIndexEntry{Basic MRMhub Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = TRUE, message = TRUE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

<p class="page-kind page-kind-tutorial">Tutorial</p>

::: {.callout .callout-note}
**Time** ~15 min &nbsp;·&nbsp; **Level** Beginner–Intermediate &nbsp;·&nbsp; **Prerequisites** [Your First Analysis](tutorial-00-first-analysis.html)
:::

This tutorial outlines the key steps in a MRMhub workflow, based on a lipidomics
dataset; see the [Lipidomics Data Processing](tutorial-03-lipidomics-workflow.html)
tutorial for a more detailed example. Where [Your First Analysis](tutorial-00-first-analysis.html)
uses bundled demo data and covers only import, normalize and export, this tutorial
uses real file paths, imports metadata via an MSOrganiser template, applies drift
and batch correction, runs QC filtering, and exports a full report. The examples
are simplified and may not apply to every dataset and experimental setup, so
consult the other tutorials and recipes for additional workflows and data types.

## Step 1: Set Up a Project

To start a new MRMhub data analysis, create an RStudio or Positron
project. (See [Using RStudio Projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects)
or the [Positron User Guide](https://positron.posit.co/).)

The project should contain the following subfolders:

```
my_study/
├── data/           # raw data and metadata files
├── output/         # exported results
└── analysis.Rmd    # your processing notebook
```

R/Notebook (`.Rmd`) or [Quarto](https://docs.posit.co/ide/user/ide/guide/documents/quarto-project.html) 
(`.qmd`) files are good choices for documented processing workflows. They
combine code with formatted text to document every processing decision.

Start with a new notebook or R script and load the `mrmhub` package:

```{r}
library(mrmhub)
```

## Step 2: Create a MRMhubExperiment

The `MRMhubExperiment` object is the main **data container** in the
MRMhub workflow. See [The MRMhubExperiment Object](manual-02-data-object.html) 
for details.

Create a new empty object:

```{r}
myexp <- MRMhubExperiment()
```

## Step 3: Import Analysis Data

Import peak area data from an INTEGRATOR output file. This file also
contains some metadata (e.g., `qc_type`, `batch_id`) which we import
at the same time.

```{r load-rawdata}
myexp <- import_data_mrmhub(
  data = myexp, 
  path = "datasets/sPerfect_MRMhub.tsv", 
  import_metadata = TRUE
)
```

<details>
<summary><strong>Using a different data source?</strong></summary>

See [Import and prepare data files](manual-04-data-import.html) for a
decision flowchart, or [Data Import](manual-04-data-import.html) for
the full API reference.

</details>

## Step 4: Add Metadata

Processing steps require additional information not available in the
data file: internal standard assignments, concentrations, sample
amounts, etc.

Metadata can be imported from separate CSV/Excel files or R data frames
(see [Metadata Import](manual-05-metadata.html)), or from an
MSOrganiser template:

```{r import-metadata}
myexp <- import_metadata_msorganiser(
  myexp, 
  path = "datasets/sPerfect_Metadata.xlsx", 
  ignore_warnings = TRUE
)
```

The validation checks may produce warnings, which by default cause the import to
fail. Setting `ignore_warnings = TRUE` allows it to proceed; the warnings are still
shown in the console table, marked with an asterisk (`*`) in the `status` column,
and should be reviewed to ensure nothing critical is missed. See
[Validating and Fixing Metadata](tutorial-10-metadata-validation.html) for how to
inspect and resolve them.

## Step 5: Process the Data

Apply normalization, quantification, drift correction, batch correction,
and QC filtering:

```{r processing}
# Normalize by internal standards
myexp <- normalize_by_istd(myexp)

# Quantify using ISTD-based approach
myexp <- quantify_by_istd(myexp)

# Correct run-order drift (Gaussian kernel, sample-based)
myexp <- correct_drift_gaussiankernel(
  data = myexp,
  variable = "conc",
  ref_qc_types = c("SPL")
)

# Correct batch effects (median centering)
myexp <- correct_batch_centering(
  myexp, 
  variable = "conc",
  ref_qc_types = "SPL"
)

# Apply QC-based feature filtering
myexp <- filter_features_qc(
  data = myexp,
  include_qualifier = FALSE,
  include_istd = FALSE, 
  min.signalblank.median.spl.sblk = 10,
  max.cv.conc.bqc = 25
)
```

Note that the drift correction above uses the study samples
(`ref_qc_types = "SPL"`) as the reference. Study-sample smoothing is only
appropriate for large, well-randomised sample sets; the QC-based convention
(Broadhurst et al. 2018) instead fits the drift trend on dedicated QC injections.

<details>
<summary><strong>Processing step order</strong></summary>

The recommended order is:

1. Normalize by ISTD
2. Quantify (ISTD or external calibration)
3. Drift correction
4. Batch correction
5. QC filtering

See [Design Decisions](manual-03-design-decisions.html) for the rationale behind this order.

</details>

## Step 6: Visualize Results

Create a run scatter plot to check signal trends and correction quality:

```{r}
plot_runscatter(
  data = myexp,
  variable = "conc",
  include_feature_filter = "PC 4",
  include_istd = FALSE,
  cap_outliers = TRUE,
  log_scale = FALSE,
  output_pdf = FALSE,
  path = "./output/runscatter_PC408_beforecorr.pdf",
  cols_page = 3, rows_page = 2
)
```

See [Visualisation Functions](manual-08-visualization.html) for the full
list of plotting functions grouped by workflow stage.


## Step 7: Export and Share

```{r save-report}
# Detailed Excel report with multiple sheets
save_report_xlsx(myexp, path = tempfile(fileext = ".xlsx"))

# Flat CSV with concentration values that passed QC
save_dataset_csv(
  data = myexp, 
  path = tempfile(fileext = ".csv"),
  variable = "conc", 
  qc_types = "SPL", 
  include_qualifier = FALSE,
  filter_data = TRUE
)

# Save the complete object for reproducibility or sharing
saveRDS(myexp, file = tempfile(fileext = ".rds"), compress = TRUE)
```

The `.rds` file preserves the entire `MRMhubExperiment` object, so a colleague can
open it in R and run their own processing, plots and QC checks — no code is needed
to reproduce the data state.

## Next steps

- [Lipidomics Data Processing](tutorial-03-lipidomics-workflow.html) — a more detailed lipidomics example
- [Drift and Batch Correction](tutorial-04-drift-correction.html) — correction methods and diagnostics
- [Exploring QC: RunScatter and PCA](tutorial-05-run-scatter.html) — QC visualisation and outlier screening
- [External Calibration & QC](recipe-01-ext-calibration-qc.html) — quantitation with calibration curves

