---
title: "Export to Bioconductor (SummarizedExperiment)"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
description: |
  Export a processed MRMhubExperiment as a Bioconductor SummarizedExperiment, and
  take it downstream: differential abundance with limma, or lipid-specific
  analysis with lipidr.
vignette: >
  %\VignetteIndexEntry{Export to Bioconductor (SummarizedExperiment)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Intermediate</span> <span class="page-prereq">Prerequisites: <a href="tutorial-02-basic-workflow.html">Basic workflow</a></span></p>

Export a processed `MRMhubExperiment` to a Bioconductor
[SummarizedExperiment](https://bioconductor.org/packages/SummarizedExperiment/),
then take it downstream: differential abundance with `limma`, lipid-specific
analysis with `lipidr`.

## 1. Setup

```{r}
library(mrmhub)
library(SummarizedExperiment)

# A processed MRMhubExperiment. Here built from the bundled
# `lipidomics_dataset`; in practice this would be your own processed object.
mexp <- lipidomics_dataset |>
  normalize_by_istd() |>
  quantify_by_istd()
```

`SummarizedExperiment` is a Bioconductor package. Install it once with:

```{r}
# install.packages("BiocManager")
BiocManager::install("SummarizedExperiment")
```

## 2. Basic export

```{r}
se <- save_dataset_summarizedexperiment(mexp)
se
#> class: SummarizedExperiment
#> dim: 29 499
#> metadata(11): title analysis_type ... var_batch_corrected mrmhub_version
#> assays(9): rt area ... pmol_total conc
#> rownames(29): CE 18:1 CE 18:1 d7 (ISTD) ... TG 48:2 [-18:1] TG 48:2 [SIM]
#> rowData names(17): feature_id feature_class ... remarks feature_label
#> colnames(499): Longit_BLANK-01 (Eluent A) ... Longit_BLANK-07 (Eluent A)
#> colData names(13): analysis_order analysis_id ... annot_order_num remarks
```

Pass a `path` to also write the object to disk as an `.rds`; without one, the
object is returned:

```{r}
save_dataset_summarizedexperiment(mexp, "experiment.rds")
```

## 3. Anatomy of the exported object

Features are **rows** and analyses are **columns**, per the SummarizedExperiment
convention.

| Component | mrmhub source |
|---|---|
| `assays()` | one matrix per feature variable, named without the `feature_` prefix |
| `rowData()` | `annot_features`: one row per `feature_id` |
| `colData()` | `annot_analyses`: one row per `analysis_id` |
| `metadata()` | title, analysis type, processing status, `is_*` flags, concentration unit, MRMhub version |

The parallel feature variables become **parallel assays**, so raw, normalized
and quantified values live side-by-side in one object:

```{r}
assayNames(se)
#> [1] "rt"     "area"   "height" "fwhm"   "width"  "intensity"
#> [7] "norm_intensity" "pmol_total" "conc"

assay(se, "conc")[1:3, 1:2]
```

Export a subset with `variable`:

```{r}
se_conc <- save_dataset_summarizedexperiment(mexp, variable = "conc")
se_two <- save_dataset_summarizedexperiment(
  mexp,
  variable = c("intensity", "conc"))
```

To export QC-filtered data instead of the full dataset, filter first and pass
`filter_data = TRUE`:

```{r}
mexp_filt <- mexp |>
  calc_qc_metrics() |>
  filter_features_qc(
    include_qualifier = FALSE,
    include_istd = FALSE,
    max.cv.conc.bqc = 25)

se_filt <- save_dataset_summarizedexperiment(mexp_filt, filter_data = TRUE)
```

## 4. Subsetting: study samples only

**Everything is exported**: internal standards, QC samples, blanks and
calibrants are all present and flagged rather than dropped, because downstream
tools need them: `lipidr`, for one, requires the internal-standard annotation.
But nothing downstream reads `qc_type`, so a PCA, normalization or differential
test will fold blanks, calibrants and QC replicates in with the study samples.
Subset to study samples before any such analysis, here in one line:

```{r}
se_spl <- se[!rowData(se)$is_istd, se$qc_type == "SPL"]
dim(se_spl)
#> [1]  20 374
```

QC metrics are deliberately *not* written to `rowData()`: nothing downstream
reads them, and QC filtering belongs in MRMhub where it is tested. Use
`filter_data = TRUE` as above, or `save_feature_qc_metrics()` for the metrics
themselves. To attach them to the features anyway:

```{r}
rowData(se) <- cbind(
  rowData(se),
  mexp_filt@metrics_qc[
    match(rownames(se), mexp_filt@metrics_qc$feature_id),
  ])
```

## 5. Differential abundance with limma

[limma](https://bioconductor.org/packages/limma/) works directly on continuous
data such as concentrations. Note that `voom()` is **not** used: it models count
data and is invalid here.

::: {.callout .callout-note}
The bundled `lipidomics_dataset` carries no phenotype: every study sample is
plasma from the same longitudinal series. The grouping below is **simulated to
demonstrate the interface only**; the p-values are meaningless by construction.
With your own data, the group would come from your sample metadata and already
be present in `colData()`.
:::

```{r}
library(limma)

set.seed(1)
se_spl$group <- factor(
  sample(c("ctrl", "trt"), ncol(se_spl), replace = TRUE))

# log-transform: concentrations are right-skewed and limma assumes normality
y <- log2(assay(se_spl, "conc"))

design <- model.matrix(~ 0 + group, data = colData(se_spl))
colnames(design) <- levels(se_spl$group)

fit <- lmFit(y, design)
fit <- contrasts.fit(fit, makeContrasts(trt - ctrl, levels = design))
fit <- eBayes(fit)

topTable(fit, number = 5)
#>                      logFC   AveExpr        t    P.Value adj.P.Val         B
#> LPC 18:1 (a)    0.09562082 -7.678714 1.672959 0.09516504 0.4750705 -4.212455
#> TG 48:2 [SIM]   0.22391165 -3.283165 1.660831 0.09757883 0.4750705 -4.225380
#> TG 48:1 [SIM]   0.22768141 -3.357736 1.616869 0.10674305 0.4750705 -4.271461
#> TG 48:1 [-18:1] 0.21339199 -4.809001 1.533404 0.12601528 0.4750705 -4.355608
#> TG 48:2 [-16:0] 0.20881625 -4.975439 1.480413 0.13959852 0.4750705 -4.406757
```

Because `colData()` carries the full analysis annotation, covariates are
available without extra joins, for example blocking on batch:

```{r}
design <- model.matrix(~ 0 + group + batch_id, data = colData(se_spl))
```

## 6. Lipid-specific analysis with lipidr

[lipidr](https://www.lipidr.org) works on a `LipidomicsExperiment`, a subclass of
`SummarizedExperiment`. Produce one directly:

```{r}
le <- save_dataset_summarizedexperiment(
  mexp,
  variable = "intensity",
  as = "LipidomicsExperiment")
```

MRMhub fills in what lipidr requires: `Molecule` from `feature_id`, `Class` from
`feature_class`, `istd` from `is_istd`, plus the `summarized` / `logged` /
`normalized` flags lipidr reads but does **not** validate: an object missing
them constructs cleanly and then misbehaves.

::: {.callout .callout-caution}
**Use a peak-area scale variable with lipidr, not concentrations.** lipidr
log-transforms with `log = TRUE` by default and **clamps values below 1 to 1**
first, which assumes Skyline peak-area magnitudes. Concentrations in µmol/L are
mostly below 1, so they are silently flattened to `log2(1) = 0`; `eBayes()` then
reports zero residual variances and every fold change collapses. MRMhub warns
when you export a mostly-sub-1 assay this way.

The trigger is absolute scale, not concentration as such: the clamp is a fixed
threshold at 1, so any assay whose values sit mostly below it is affected. If you
need concentrations, pass `log = FALSE` to lipidr, or log-transform them
yourself beforehand.
:::

```{r}
library(lipidr)

le_spl <- le[!rowData(le)$istd, le$qc_type == "SPL"]
set.seed(1)
le_spl$group <- factor(
  sample(c("ctrl", "trt"), ncol(le_spl), replace = TRUE))

plot_samples(le_spl, type = "boxplot", measure = "intensity")

le_norm <- normalize_pqn(le_spl, measure = "intensity", log = TRUE)

de_results <- de_analysis(
  le_norm,
  trt - ctrl,
  measure = "intensity",
  group_col = "group")
plot_results_volcano(de_results, show.labels = FALSE)
```

## References

- Morgan M, Obenchain V, Hester J, Pagès H (2026). SummarizedExperiment: A
  container (S4 class) for matrix-like assays.
  doi:[10.18129/B9.bioc.SummarizedExperiment](https://doi.org/10.18129/B9.bioc.SummarizedExperiment).
  R package version 1.42.0, <https://bioconductor.org/packages/SummarizedExperiment>.
- Mohamed A, Molendijk J, Hill MM (2020). lipidr: A Software Tool for Data Mining
  and Analysis of Lipidomics Datasets. *Journal of Proteome Research*, 19(7),
  2890–2897. doi:[10.1021/acs.jproteome.0c00082](https://doi.org/10.1021/acs.jproteome.0c00082).
- Ritchie ME, Phipson B, Wu D, Hu Y, Law CW, Shi W, Smyth GK (2015). limma powers
  differential expression analyses for RNA-sequencing and microarray studies.
  *Nucleic Acids Research*, 43(7), e47.
  doi:[10.1093/nar/gkv007](https://doi.org/10.1093/nar/gkv007).

## Next steps

- [Import & export mzTab-M](recipe-03-mztab-export.html): the standard exchange format for repositories such as MetaboLights.
- [Design decisions](manual-03-design-decisions.html): why `MRMhubExperiment` is not itself a Bioconductor class.
- [Lipidomics workflow](tutorial-03-lipidomics-workflow.html): QC-filter features before exporting.
- [Custom QC report](recipe-02-custom-qc-report.html): reporting straight from MRMhub.
