---
title: "Introduction to R, Quarto, and MRMhub"
description: >
  Install R and an IDE, create a Quarto project, and run a complete MRMhub
  analysis in it — written for readers new to R and Quarto.
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Introduction to R, Quarto, and MRMhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Beginner</span></p>

A MRMhub analysis is written as a document that mixes prose with the R code that produces the
results, so the report and the analysis are one file. This tutorial assumes no prior experience with
R or Quarto. It starts from an empty computer, installs the software, sets up a project, and then
runs a complete analysis of the bundled demonstration data — importing, normalizing, plotting, and
exporting — using MRMhub functions alone, without any custom R code. After working through it you
can open a fresh Quarto document and reproduce the same workflow on your own data.

## 1. Install R and an editor

Two pieces of software are needed. **R** is the language the analysis runs in; an **integrated
development environment** (IDE) is the editor you write and run it in.

- Install R from [CRAN](https://cran.r-project.org/) (newest version). Pick the download for your OS
  and accept the defaults.
- Install an IDE. [RStudio](https://posit.co/download/rstudio-desktop/) is the most common choice and
  the one this tutorial follows. [Positron](https://positron.posit.co/) is a great alternative built
  on VS Code.

With R and RStudio installed, open RStudio and install MRMhub from the console (the bottom-left
pane). The first line installs `pak`, a package installer; the second uses it to fetch MRMhub and its
dependencies:

```{r install, eval = FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/MRMhub")
```

If the installation reports an error, the [Installation guide](manual-00-installation.html) lists the
common causes (a missing compiler on Windows or macOS, a firewall blocking the download) and their
fixes.

## 2. Create a Quarto project

A **project** keeps everything for one analysis — data, code, and results — in a single folder that
the IDE treats as a unit. In RStudio, choose *File → New Project → New Directory → Quarto Project*,
give it a name, and create it. Quarto's [project documentation](https://quarto.org/docs/tools/rstudio/)
covers the dialog in detail.

Inside the new project, add two folders to keep raw inputs separate from generated results — the
convention MRMhub's own workflows follow:

```
my_study/
├── data/            # raw INTEGRATOR output and metadata files
├── output/          # exported CSVs, PDFs, and reports
└── analysis.qmd     # this document
```

A Quarto document (`.qmd`) is the home for the workflow: it interleaves explanatory text with code,
documenting each decision alongside the result it produces.

## 3. Add text and code

A `.qmd` file is plain text with two kinds of content. **Prose** is written in Markdown — the same
ordinary syntax used across the web, where `#` starts a heading, `**bold**` sets bold text, and `-`
begins a list item. **Code** lives in *chunks*: R code fenced between ```` ```{r} ```` and
```` ``` ````. In RStudio, *Insert → Code Chunk* (or Ctrl/Cmd + Alt + I) adds an empty one.

```` markdown
Some explanatory text describing the next step.

```{r first-step}
1 + 1
```

More text explaining what comes next, then a second chunk that builds on the first.

```{r second-step}
2 + 2
```
````

The word after `r` — `first-step`, `second-step` above — is an optional label naming the chunk.

RStudio offers two ways to edit a `.qmd`: the **Source** editor shows the raw Markdown, and the
**Visual** editor a formatted, word-processor-like view. Both edit the same file; Quarto's
[authoring guide](https://quarto.org/docs/get-started/hello/rstudio.html) introduces both.

A chunk's behaviour is set by **cell options** — lines at the top of the chunk, each prefixed with
`#|`, written as `#| option: value`. A handful cover almost every need in an analysis document:

- `#| echo: false` — run the chunk but hide its source code, showing only the result.
- `#| eval: false` — show the code without running it (useful for an export step you don't want
  re-run on every render).
- `#| message: true` — keep the chunk's messages, including MRMhub's coloured step-by-step console
  feedback; `#| warning: false` hides warnings.
- `#| fig-width:` and `#| fig-height:` — the width and height of a figure, in inches. Set them per
  chunk here, or once for the whole document in the YAML header; [Quarto
  workflows](manual-11-quarto-workflows.html) explains how these interact with MRMhub's plot sizing.

The [Quarto documentation](https://quarto.org/docs/computations/execution-options.html) is the
complete reference.

## 4. Preview and run code

There are two distinct actions. **Running a chunk** executes its code immediately and shows the
result inline, right beneath the chunk. Each chunk carries a small toolbar at its top-right: the
green ▶ arrow runs that chunk, the ⬇ beside it first runs every earlier chunk, and the ⚙ opens the
chunk's options. Pressing Ctrl/Cmd + Shift + Enter with the cursor inside the chunk does the same as
▶. **Rendering** turns the whole document into a finished HTML (or PDF) report: click the
📄 **Render** button in the editor toolbar to run every chunk in order and assemble the output. The
**Preview** that opens updates on each render, so the report stays in step with the code.

While developing an analysis, run chunks one at a time to check each step; render at the end to
produce the shareable report.

## 5. The setup chunk

The first chunk of an analysis loads the packages and sets defaults that apply throughout. Loading
MRMhub makes its functions available; `mrmhub_enable_cli_color()` colours the console messages in the
rendered report; and `mrmhub_set_plot_defaults()` sets a house font size and resolution for every
figure, so individual plots need not repeat them.

```{r first-chunk}
library(mrmhub)
mrmhub_enable_cli_color()
mrmhub_set_plot_defaults(font_base_size = 8)
```

## 6. Load the data

Every analysis begins with a `MRMhubExperiment` — the container that holds the data and every result
computed from it (see [The MRMhubExperiment object](manual-02-data-object.html)). We create an empty
one and import the demonstration dataset bundled with the package. Using `system.file()` locates that
file wherever the package is installed, so the code runs unchanged on any machine; in your own
project you would instead give the path to a file in `data/`.

```{r load-data}
#| message: true
demo_file <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")

mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(
  mexp,
  path = demo_file,
  import_metadata = TRUE)
```

Setting `import_metadata = TRUE` reads the sample and internal-standard annotations embedded in the
INTEGRATOR file at the same time. MRMhub reads several other formats; [Importing analytical
data](manual-04-data-import.html) has a decision flowchart and the full importer reference.

## 7. Process the data

The core processing step normalizes each feature to its assigned internal standard, which corrects
for extraction and injection variation. The result is stored in the `feature_norm_intensity`
variable, leaving the raw intensities untouched.

```{r process}
#| message: true
mexp <- normalize_by_istd(mexp)
```

This is the shortest useful pipeline. A real study usually adds drift and batch correction and QC
filtering in a set order; [A basic MRMhub workflow](tutorial-02-basic-workflow.html) and [Drift and
batch correction](tutorial-04-drift-correction.html) build on the object from here. If you would
rather not write this code by hand, `build_workflow()` opens an interactive application that
validates your data and metadata and generates an equivalent Quarto document to download (see
[Build a workflow without code](tutorial-12-workflow-builder.html)).

## 8. Check the processing status

`mrmhub_status()` prints a full report of the object: what was imported, which steps have run, and
which stages still lie ahead. It is the quickest way to confirm the experiment is in the state you
expect before moving on to plots and export.

```{r status}
#| message: true
mrmhub_status(mexp)
```

## 9. Plot signals across the run

A run-scatter plot shows each feature's signal against injection order — a quick check for trends or
outliers. We plot the raw peak areas for two lipid classes, selected by a regular expression on their
names.

```{r runscatter}
#| fig.cap: !expr fig_cap("Raw peak areas of Cer and PC features across injection order.")
#| fig.alt: Run-scatter of raw peak areas for Cer and PC features by injection order
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "^(Cer|PC)",
  rows_page = 2, cols_page = 3)
```

To save the plot to a file instead of showing it, `plot_runscatter()` can write a multi-page PDF —
one page per grid of features — directly to `output/`:

```{r runscatter-pdf, eval = FALSE}
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "^(Cer|PC)",
  rows_page = 2, cols_page = 3,
  output_pdf = TRUE,
  path = "output/runscatter.pdf")
```

## 10. Run a PCA

A principal-component analysis (PCA) summarises all features at once, placing each sample in a plane
so that similar samples fall close together — useful for spotting QC clustering or outliers.

```{r pca}
#| fig.cap: !expr fig_cap("PCA of the normalized intensities, coloured by QC type.")
#| fig.alt: PCA score plot of samples coloured by QC type
plot_pca(mexp, variable = "norm_intensity")
```

To keep a copy of the figure, pipe it into `save_plot()`. Because `save_plot()` returns the plot
visibly, one line both writes the PDF and still renders the figure in the report; `width` and
`height` are in millimetres:

```{r pca-save, eval = FALSE}
plot_pca(mexp, variable = "norm_intensity") |>
  save_plot("output/pca.pdf", width = 180, height = 120)
```

Giving `save_plot()` a path with no extension and `format = c("pdf", "png")` writes both formats at
once; [Visualisation functions](manual-08-visualization.html) lists the plots available at each stage.

## 11. Export results

Finally, export the results for downstream use. A single-variable CSV gives a plain table of
normalized intensities, and `save_report_xlsx()` writes a multi-sheet Excel workbook holding every
data table and QC metric in one file.

```{r export, eval = FALSE}
save_dataset_csv(
  mexp,
  path = "output/norm_intensities.csv",
  variable = "norm_intensity")

save_report_xlsx(
  mexp,
  path = "output/report.xlsx",
  filtered_variable = "norm_intensity")
```

With these chunks written, click **Render** once more: RStudio runs the whole analysis top to bottom
and produces a self-contained report, with the CSV, PDFs, and Excel workbook waiting in `output/`.

## Next steps

- [A basic MRMhub workflow](tutorial-02-basic-workflow.html): the full pipeline with quantification and QC filtering
- [Drift and batch correction](tutorial-04-drift-correction.html): correcting run-order and batch effects
- [Exploring QC: RunScatter and PCA](tutorial-05-run-scatter.html): QC visualisation in depth
- [Quarto workflows](manual-11-quarto-workflows.html): rendering, figure sizing, and report options
