---
title: "Getting started with R and Quarto notebooks"
description: >
  Install R and an IDE, create a Quarto project, learn to read and run notebook
  code, and render a report — for readers new to R and Quarto.
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Getting started with R and Quarto notebooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span></p>

A MRMhub analysis is written as a document that mixes prose with the R code that produces the
results, so the report and the analysis are one file. This tutorial assumes no prior experience with
R or Quarto: it starts from an empty computer, installs the software, sets up the project and notebook
to work in, and shows how to read and run the code. Once your machine is set up, [Getting started with
MRMhub](tutorial-02-getting-started-mrmhub.html) runs a complete analysis in the project you create
here.

## 1. Install R and an editor

Two pieces of software are needed. **R** is the language the analysis runs in; an **integrated
development environment** (IDE) is the editor you write and run it in. Install R first — the IDE
looks for it on startup.

- Install R from [CRAN](https://cran.r-project.org/) (version 4.1 or newer). Pick the download for
  your operating system and accept the defaults.
- Install an IDE. [RStudio](https://posit.co/download/rstudio-desktop/) is the most common choice and
  the one this tutorial follows. If you already work in VS Code, [Positron](https://positron.posit.co/)
  is an RStudio-like alternative built on the same editor.

Both are installed once per machine and shared by every project you create afterwards.

## 2. RStudio at a glance

When RStudio opens it shows four panes, each used at some point in an analysis:

- **Source** (top-left) — where you edit the notebook and run its code. Its toolbar carries the
  **Render** button, and each code chunk gets a green ▶ arrow that runs it.
- **Console** (bottom-left) — runs a single line of R immediately, without saving it in the
  notebook. This is where software is installed and where MRMhub's step messages appear.
- **Environment** (top-right) — lists the objects that currently exist in the session, such as the
  `mexp` data object once the notebook has created it.
- **Files / Plots / Help** (bottom-right) — browses the project folder, shows figures, and displays
  function help pages.

<div style="text-align: center; margin: 2em 0;">
<svg xmlns="http://www.w3.org/2000/svg" viewBox="0 0 640 470" style="max-width: 700px; width: 100%; height: auto; font-family: -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;" role="img" aria-label="Schematic of the RStudio window: the Source pane top-left holds the notebook with an R code chunk, a Render button and a green run arrow; the Console bottom-left runs single commands; the Environment pane top-right lists objects; Files, Plots and Help are bottom-right.">
  <rect x="8" y="8" width="624" height="428" rx="10" fill="#f8f9fa" stroke="#2C3E50" stroke-width="2"/>
  <circle cx="30" cy="29" r="5" fill="#dee2e6"/>
  <circle cx="48" cy="29" r="5" fill="#dee2e6"/>
  <circle cx="66" cy="29" r="5" fill="#dee2e6"/>
  <text x="320" y="34" text-anchor="middle" font-weight="700" font-size="15" fill="#2C3E50">my_study — RStudio</text>
  <line x1="8" y1="46" x2="632" y2="46" stroke="#dee2e6"/>
  <rect x="20" y="58" width="290" height="190" rx="6" fill="#d6e4eb" stroke="#5B8FA8"/>
  <rect x="20" y="58" width="290" height="24" rx="6" fill="#ffffff" fill-opacity="0.55"/>
  <text x="32" y="75" font-size="10" fill="#2C3E50">analysis.qmd</text>
  <rect x="240" y="62" width="62" height="16" rx="3" fill="#6B9E5E"/>
  <text x="271" y="74" text-anchor="middle" font-size="10" font-weight="600" fill="#ffffff">Render</text>
  <text x="32" y="104" font-weight="600" font-size="14" fill="#2C3E50">Source</text>
  <text x="292" y="105" text-anchor="middle" font-size="13" fill="#6B9E5E">▶</text>
  <text x="32" y="124" font-size="11" fill="#333">edit prose and code chunks here</text>
  <rect x="32" y="136" width="252" height="66" rx="4" fill="#ffffff" fill-opacity="0.8" stroke="#a9c6d4"/>
  <text x="42" y="154" font-size="10.5" font-family="ui-monospace, Menlo, Consolas, monospace" fill="#7a8b94">&#96;&#96;&#96;{r}</text>
  <text x="42" y="174" font-size="10.5" font-family="ui-monospace, Menlo, Consolas, monospace" fill="#31586b">mexp &lt;- normalize_by_istd(mexp)</text>
  <text x="42" y="194" font-size="10.5" font-family="ui-monospace, Menlo, Consolas, monospace" fill="#7a8b94">&#96;&#96;&#96;</text>
  <text x="32" y="228" font-size="11" fill="#333">▶ runs a chunk · Render builds the report</text>
  <rect x="326" y="58" width="294" height="190" rx="6" fill="#d8e6d4" stroke="#6B9E5E"/>
  <text x="338" y="86" font-weight="600" font-size="14" fill="#2C3E50">Environment</text>
  <text x="338" y="114" font-size="10.5" font-family="ui-monospace, Menlo, Consolas, monospace" fill="#3a5f30">mexp    MRMhubExperiment</text>
  <text x="338" y="148" font-size="11" fill="#333">the objects that exist right now —</text>
  <text x="338" y="168" font-size="11" fill="#333">empty again in a fresh session</text>
  <text x="338" y="196" font-size="11" fill="#333">click a name to inspect it</text>
  <rect x="20" y="258" width="290" height="170" rx="6" fill="#f5e0c8" stroke="#D4914E"/>
  <text x="32" y="286" font-weight="600" font-size="14" fill="#2C3E50">Console</text>
  <text x="32" y="312" font-size="10.5" font-family="ui-monospace, Menlo, Consolas, monospace" fill="#8a5a25">&gt; pak::pak(&quot;SLINGhub/MRMhub&quot;)</text>
  <text x="32" y="332" font-size="10.5" font-family="ui-monospace, Menlo, Consolas, monospace" fill="#8a5a25">&gt; mrmhub::build_workflow()</text>
  <text x="32" y="352" font-size="10.5" font-family="ui-monospace, Menlo, Consolas, monospace" fill="#3a5f30">✔ Normalized 84 features by ISTD</text>
  <text x="32" y="384" font-size="11" fill="#333">runs a single line of R immediately;</text>
  <text x="32" y="404" font-size="11" fill="#333">MRMhub's step messages arrive here</text>
  <rect x="326" y="258" width="294" height="170" rx="6" fill="#f2dcdc" stroke="#C27171"/>
  <text x="338" y="286" font-weight="600" font-size="14" fill="#2C3E50">Files · Plots · Help</text>
  <text x="338" y="316" font-size="11" fill="#333">browse data/ and output/</text>
  <text x="338" y="342" font-size="11" fill="#333">view figures in Plots</text>
  <text x="338" y="368" font-size="11" fill="#333">read ?function pages in Help</text>
  <text x="320" y="458" text-anchor="middle" font-size="11" fill="#666">RStudio's default four-pane layout, with the project folder open as my_study</text>
</svg>
</div>

The panes can be resized or rearranged, but the defaults are fine to start; nothing here needs
configuring.

## 3. Install MRMhub

With R and RStudio installed, open RStudio and install MRMhub from the Console (the bottom-left
pane). The first line installs `pak`, a package installer; the second uses it to fetch MRMhub and its
dependencies:

```{r install, eval = FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/MRMhub")
```

Installing is a once-per-machine step, and running the same two lines again later updates MRMhub
to the current version. Making its functions available is a separate, per-session step —
`library(mrmhub)` — which belongs in the notebook rather than the Console, so that every render
loads the package itself.

If the installation reports an error, the [Installation guide](manual-00-installation.html) lists the
common causes (a missing compiler on Windows or macOS, a firewall blocking the download) and their
fixes.

## 4. Create a Quarto project

A **project** keeps everything for one analysis — data, code, and results — in a single folder that
the IDE treats as a unit. In RStudio, choose *File → New Project → New Directory → Quarto Project*,
give it a name, and create it. Quarto itself ships with RStudio, so nothing further needs
installing; if the *Quarto Project* entry is missing, update RStudio or install Quarto from
[quarto.org](https://quarto.org/docs/download/). Quarto's
[project documentation](https://quarto.org/docs/tools/rstudio/) covers the dialog in detail.

The dialog creates the folder, a first `.qmd` document, and an `.Rproj` file. That `.Rproj` file is
how the project is reopened later — by double-clicking it, or through the project menu at the
top-right of the RStudio window. Opening the project, rather than the `.qmd` on its own, is what
points R at the right folder.

Inside the new project, add two folders to keep raw inputs separate from generated results — the
convention MRMhub's own workflows follow:

```
my_study/
├── my_study.Rproj   # opens the project
├── data/            # raw INTEGRATOR output and metadata files
├── output/          # exported CSVs, PDFs, and reports
└── analysis.qmd     # this document
```

Paths in your code are written relative to this project folder: RStudio treats it as the working
directory, so `data/my_file.csv` points to the same place on any machine, without a full path. This
is why the `data/` and `output/` folders above are all the analysis needs to find its files, and why
the whole folder can be zipped and shared as a self-contained unit.

## 5. Write text and code in a `.qmd`

A `.qmd` file is plain text with three kinds of content: a header, prose, and code chunks.

At the top, a short **header** fenced by `---` sets the title and the output format — `html`, `pdf`,
or `docx`. A new project starts with a sensible HTML header, and
[Quarto workflows](manual-11-quarto-workflows.html) covers the options that matter for a MRMhub
report.

Below the header, **prose** is written in Markdown — plain paragraphs, with `#` for headings,
`**bold**`, and `-` for lists. Quarto's
[Markdown basics](https://quarto.org/docs/authoring/markdown-basics.html) is the full reference for
tables, figures, cross-references, and citations.

**Code** lives in *chunks*: R code fenced between ```` ```{r} ```` and ```` ``` ````. In RStudio,
*Insert → Code Chunk* (or Ctrl/Cmd + Alt + I) adds an empty one. A chunk that loads the MRMhub
package, with a sentence of prose introducing it, looks like this:

````{verbatim, lang = "markdown"}
Load the package before calling any of its functions.

```{r}
library(mrmhub)
```
````

### Chunk options

Each chunk can carry **options** that control whether its code runs, whether the code itself is
shown, and which parts of its output reach the report. They are written as `#|` comment lines
directly below the chunk's opening fence, one per line, as `option: value`:

````{verbatim, lang = "markdown"}
```{r}
#| label: normalize
#| message: true
mexp <- normalize_by_istd(mexp)
```
````

The options a processing notebook needs are few:

| Option | What it controls |
|---|---|
| <code>#&#124; label:</code> | A unique name for the chunk; also names any figure it produces. |
| <code>#&#124; eval:</code> | Whether the code runs. `false` shows the code without executing it. |
| <code>#&#124; echo:</code> | Whether the code itself appears in the report. |
| <code>#&#124; output:</code> | Whether the results appear — printed values, tables, figures. |
| <code>#&#124; message:</code> | Whether messages appear. MRMhub's step feedback travels here. |
| <code>#&#124; warning:</code> | Whether warnings appear. |
| <code>#&#124; include:</code> | `false` runs the chunk but hides both code and output — used for setup chunks. |
| <code>#&#124; cache:</code> | Stores the result, so an unchanged chunk is not re-run on the next render. |
| <code>#&#124; fig-width:</code>, <code>#&#124; fig-height:</code> | Size, in inches, of the figures the chunk produces. |

MRMhub reports what each processing step did as a console message — a count, and a truncated list
of the features or analyses affected — so keeping `#| message: true` on the chunks that run a step
turns the rendered report into a record of the processing. Defaults for the whole document are set
in the header instead of chunk by chunk:

```yaml
execute:
  warning: true
  message: true
```

Quarto's [execution options](https://quarto.org/docs/computations/execution-options.html) page is the
complete reference for both forms.

RStudio offers two ways to edit a `.qmd`: the **Source** editor shows the raw Markdown, while the
**Visual** editor (the *Visual* toggle, top-left of the document) shows a formatted,
word-processor-like view. Both edit the same file, and Quarto's
[authoring guide](https://quarto.org/docs/get-started/hello/rstudio.html) introduces both. You may
also meet the older R Markdown format (`.Rmd`), which Quarto renders unchanged; it writes the same
options inside the fence, as `{r, message=TRUE}`.

## 6. Run code and render the report

There are two distinct actions. **Running a chunk** executes its code immediately and shows the
result inline: click the green ▶ arrow at the chunk's top-right, or press Ctrl/Cmd + Shift + Enter
with the cursor inside it. Because each chunk builds on the objects the earlier ones created,
*Run → Run All Chunks Above* (Ctrl/Cmd + Alt + P) is the way back to a working state after reopening
the project, when the session starts with an empty Environment.

**Rendering** turns the whole document into a finished report: click **Render** (Ctrl/Cmd + Shift +
K), and Quarto runs every chunk in order and assembles the output — `analysis.html` alongside the
document — which opens in the Preview pane and refreshes on each render.

Rendering runs in a **fresh R session**, top to bottom, so the report is built from what the
document itself contains and nothing else. A chunk that works interactively because a needed object
is still in the Environment pane will fail on render if no chunk creates it. This is the point of
rendering: it proves the analysis reproduces from the raw data. While developing, run chunks one at
a time to check each step; render at the end to produce the shareable report.

Try it now: insert a chunk into your document, put `library(mrmhub)` in it, and run it. Loading
MRMhub without error confirms the installation worked.

```{r load-mrmhub}
library(mrmhub)
```

If a chunk stops with an error, the Console prints the message under the offending call; the last
line is usually the informative one. [Troubleshooting & FAQ](manual-10-troubleshooting.html)
collects the errors that come up most often, and what each one means.

## 7. Reading the code you'll run

The next tutorial uses MRMhub functions and no custom R, but three pieces of R syntax recur
throughout. Recognising them is enough to follow along:

- **Assignment** (`<-`) stores a result under a name. `mexp <- MRMhubExperiment()` creates a data
  object and names it `mexp`; writing `mexp` again later refers back to it.
- **Function calls with named arguments.** In `import_data_mrmhub(mexp, path = demo_file)`, the
  function acts on `mexp`, and `path =` names which input the file is. Most MRMhub functions take the
  data object first and options as named arguments.
- **The pipe** (`|>`) passes a result straight into the next function, so
  `plot_pca(mexp) |> save_plot("pca.pdf")` plots and then saves in one line, with no temporary name.

Each processing step follows the same shape — take the data object, return an updated one, and store
it back under the same name: `mexp <- normalize_by_istd(mexp)`. The functions do not modify the
object in place, so the assignment is what carries the result forward; running
`normalize_by_istd(mexp)` on its own leaves `mexp` as it was. All the documentation names the object
`mexp`, and keeping that one name throughout a notebook avoids running a later step on an earlier
version of the data.

## 8. Getting help

To see what a function does and which arguments it takes, type `?` followed by its name in the
Console:

```{r help, eval = FALSE}
?plot_pca
```

Its help page opens in the **Help** pane, listing every argument with an explanation and, usually,
examples that can be copied and run. The same pages for every function are in the
[function reference](../reference/index.html), the [MRMhub overview](manual-01-key-concepts.html)
introduces the concepts they build on, and problems that look like a bug belong in the
[issue tracker](https://github.com/SLINGhub/MRMhub/issues).

With the software installed, the project set up, and the syntax demystified, you are ready to run a
real analysis. Continue with [Getting started with MRMhub](tutorial-02-getting-started-mrmhub.html),
which imports the demo data and takes it through to an exported report.

## Next steps

- [Getting started with MRMhub](tutorial-02-getting-started-mrmhub.html): run a complete analysis in the project you just created
- [Quarto workflows](manual-11-quarto-workflows.html): rendering, figure sizing, and report options
- [Installation guide](manual-00-installation.html): detailed setup and troubleshooting
