---
title: "Quick tour"
description: >
  A short walkthrough on the bundled demo data: import a result file, normalize
  by internal standard, inspect the run, and export: no external files needed.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
vignette: >
  %\VignetteIndexEntry{Quick tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Beginner</span> <span class="page-prereq">Prerequisites: <a href="manual-00-installation.html">MRMhub installed</a></span></p>

This walkthrough takes an MRMhub result from import to a normalized, exportable
dataset in a handful of lines, using demo data bundled with the package so
nothing external is needed. It is the shortest path through the pipeline; the
workflow tutorials that follow unpack each step in full.

## 1. Run the complete workflow

The block below is the whole analysis: import the demo result and normalize each
feature by its internal standard. Run it and you have a processed dataset.

```{r pipeline}
#| message: true
library(mrmhub)

demo_file <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")

mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(mexp, path = demo_file, import_metadata = TRUE)
mexp <- normalize_by_istd(mexp)
```

We then look at the raw peak areas across the analytical run, the first thing to
inspect for run-order drift. To keep the figure to a single page, we restrict it
to the PC species.

```{r runscatter}
#| fig.height: 4
#| fig.cap: !expr fig_cap("Raw peak areas of the PC features plotted against injection order.")
#| fig.alt: RunScatter of raw peak areas against run order for the PC features
plot_runscatter(mexp, variable = "area", include_feature_filter = "^PC ")
```

A principal-component plot of the internal-standard–normalized signal gives a
first look at how the samples group and whether any injection separates from the
rest.

```{r pca}
#| fig.height: 4
#| fig.cap: !expr fig_cap("PCA of the internal-standard–normalized feature intensities.")
#| fig.alt: PCA score plot of ISTD-normalized feature intensities
plot_pca(mexp, variable = "norm_intensity")
```

If you prefer not to write code, `build_workflow()` opens an
interactive application that validates your data and metadata, flags pipeline
mismatches, and generates an equivalent Quarto workflow to download (see
[Build a workflow without code](tutorial-12-workflow-builder.html)); the script
above remains the reproducible path.

## 2. Export the results

Write the processed data to a multi-sheet Excel report, or to a single tidy CSV:

```{r export}
#| eval: false
save_report_xlsx(mexp, path = "my_first_results.xlsx")
save_dataset_csv(mexp, path = "my_first_results.csv")
```

## 3. Check the processing status

`mrmhub_status()` prints the full processing and metadata report: what was
imported, which steps have run, and which stages still lie ahead. It is the
quickest way to confirm the object is in the state you expect before moving on.

```{r status}
#| message: true
mrmhub_status(mexp)
```

## Next steps

- [Key concepts and glossary](manual-01-key-concepts.html): understand the data model
- [Preparing and importing data](tutorial-01-prep-data.html): import your own data and metadata
- [Basic MRMhub workflow](tutorial-02-basic-workflow.html): the full end-to-end pipeline
- [Lipidomics data processing](tutorial-03-lipidomics-workflow.html): a detailed real-world workflow
- [Visualisation functions](manual-08-visualization.html): the QC plots, and saving figures at a defined size
- [Customising plots](manual-13-plot-customization.html): size text, legends and points for reports
- [Getting help from an AI assistant](manual-09-ai-assistants.html): use Claude or ChatGPT to help write and troubleshoot your workflow, accurately and safely
