---
title: "Getting started with MRMhub"
description: >
  Create a MRMhubExperiment, import the bundled demo data, then normalize,
  plot, and export — a first end-to-end MRMhub analysis in a Quarto notebook.
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Getting started with MRMhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span></p>

A MRMhub analysis is written as a Quarto document that mixes prose with the R code that produces the
results, so the report and the analysis are one file. This tutorial runs a complete analysis of the
bundled demonstration data — importing, normalizing, plotting, and exporting — using MRMhub functions
alone, without any custom R code. After working through it you can open a fresh Quarto document and
reproduce the same workflow on your own data.

New to R or Quarto notebooks? [Getting started with R and Quarto
notebooks](tutorial-13-getting-started-r-quarto.html) installs the software and sets up the project
this analysis runs in; the [Installation guide](manual-00-installation.html) covers installing MRMhub
itself. For a first orientation to the terminology and data structure, see the [MRMhub
overview](manual-01-key-concepts.html).

If you would rather not write code, `build_workflow()` opens an interactive app that validates your
data and metadata and generates an equivalent Quarto workflow to download (see [Build a workflow
without code](tutorial-12-workflow-builder.html)); the notebook below remains the reproducible path.

## 1. The setup chunk

The first chunk of an analysis loads the packages and sets defaults that apply throughout. Loading
MRMhub makes its functions available; `mrmhub_enable_cli_color()` colours the console messages in the
rendered report; and `mrmhub_set_plot_defaults()` sets a house font size and resolution for every
figure, so individual plots need not repeat them.

```{r first-chunk}
library(mrmhub)
mrmhub_enable_cli_color()
mrmhub_set_plot_defaults(font_base_size = 8)
```

Figure dimensions are controlled by Quarto itself, through the document's `fig-width` and
`fig-height` options; [Quarto workflows](manual-11-quarto-workflows.html) explains how those interact
with MRMhub's plot sizing.

## 2. Load the data

Every analysis begins with a `MRMhubExperiment` — the container that holds the data and every result
computed from it (see [The MRMhubExperiment object](manual-02-data-object.html)). We create an empty
one and import the demonstration dataset bundled with the package. Using `system.file()` locates that
file wherever the package is installed, so the code runs unchanged on any machine; in your own
project you would instead give the path to a file in `data/`.

```{r load-data}
#| message: true
demo_file <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")

mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(
  mexp,
  path = demo_file,
  import_metadata = TRUE)
```

Setting `import_metadata = TRUE` reads the sample and internal-standard annotations embedded in the
INTEGRATOR file at the same time. MRMhub reads several other formats; [Importing analytical
data](manual-04-data-import.html) has a decision flowchart and the full importer reference.

## 3. Process the data

The core processing step normalizes each feature to its assigned internal standard, which corrects
for extraction and injection variation. The result is stored in the `feature_norm_intensity`
variable, leaving the raw intensities untouched.

```{r process}
#| message: true
mexp <- normalize_by_istd(mexp)
```

This is the shortest useful pipeline. A real study usually adds drift and batch correction and QC
filtering in a set order; [Lipidomics data processing](tutorial-03-lipidomics-workflow.html) and
[Drift and batch correction](tutorial-04-drift-correction.html) build on the object from here.

## 4. Check the processing status

`mrmhub_status()` prints a report of what was imported, which processing steps have run, and which
stages still lie ahead — the quickest way to confirm the object is in the state you expect before
moving on.

```{r status}
#| message: true
mrmhub_status(mexp)
```

## 5. Plot signals across the run

A run-scatter plot shows each feature's signal against injection order — a quick check for trends or
outliers. We plot the normalized intensities for two lipid classes, selected by a regular expression
on their names.

```{r runscatter}
#| fig.cap: !expr fig_cap("Normalized intensities of Cer and PC features across injection order.")
#| fig.alt: Run-scatter of normalized intensities for Cer and PC features by injection order
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "^(Cer|PC)",
  rows_page = 2, cols_page = 3)
```

To save the plot to a file instead of showing it, `plot_runscatter()` can write a multi-page PDF —
one page per grid of features — directly to `output/`:

```{r runscatter-pdf, eval = FALSE}
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "^(Cer|PC)",
  rows_page = 2, cols_page = 3,
  output_pdf = TRUE,
  path = "output/runscatter.pdf")
```

## 6. Run a PCA

A principal-component analysis (PCA) summarises all features at once, placing each sample in a plane
so that similar samples fall close together — useful for spotting QC clustering or outliers.

```{r pca}
#| fig.cap: !expr fig_cap("PCA of the normalized intensities, coloured by QC type.")
#| fig.alt: PCA score plot of samples coloured by QC type
plot_pca(mexp, variable = "norm_intensity")
```

To keep a copy of the figure, pipe it into `save_plot()`. Because `save_plot()` returns the plot
visibly, one line both writes the PDF and still renders the figure in the report; `width` and
`height` are in millimetres:

```{r pca-save, eval = FALSE}
plot_pca(mexp, variable = "norm_intensity") |>
  save_plot("output/pca.pdf", width = 180, height = 120)
```

Giving `save_plot()` a path with no extension and `format = c("pdf", "png")` writes both formats at
once; [Visualisation functions](manual-08-visualization.html) lists the plots available at each stage.

## 7. Export results

Finally, export the results for downstream use. A single-variable CSV gives a plain table of
normalized intensities, and `save_report_xlsx()` writes a multi-sheet Excel workbook holding every
data table and QC metric in one file. In this short pipeline those QC-metric and
filtered-data sheets are still empty; they populate once `calc_qc_metrics()` and
`filter_features_qc()` have run, as in the [lipidomics workflow](tutorial-03-lipidomics-workflow.html).

```{r export, eval = FALSE}
save_dataset_csv(
  mexp,
  path = "output/norm_intensities.csv",
  variable = "norm_intensity")

save_report_xlsx(
  mexp,
  path = "output/report.xlsx",
  filtered_variable = "norm_intensity")
```

The complete processing state can also be serialized to a single `.rds` file:

```{r save-rds, eval = FALSE}
saveRDS(mexp, file = "output/experiment.rds")
```

The `.rds` preserves the entire `MRMhubExperiment`, so a colleague can open it in R and reproduce the
exact data state — running their own plots and QC checks — without re-running the pipeline.

With these chunks written, click **Render** once more: RStudio runs the whole analysis top to bottom
and produces a self-contained report, with the CSV, PDFs, Excel workbook, and saved object waiting in
`output/`.

## Next steps

- [Lipidomics data processing](tutorial-03-lipidomics-workflow.html): the full pipeline with quantification and QC filtering
- [Drift and batch correction](tutorial-04-drift-correction.html): correcting run-order and batch effects
- [Exploring QC: RunScatter and PCA](tutorial-05-run-scatter.html): QC visualisation in depth
- [Quarto workflows](manual-11-quarto-workflows.html): rendering, figure sizing, and report options
