---
title: "Drift and Batch Correction"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Drift and Batch Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

<p class="page-kind page-kind-tutorial">Tutorial</p>

MRMhub provides functions for run-order drift correction, differing by the employed
smoothing algorithm and suited to either QC- or study-sample-based correction, and
for batch-effect correction by median centering. The data must be provided via an
MRMhubExperiment object; raw data that was imported or processed data can be
corrected, such as `intensity` or `conc` values. These functions have various
options for customization; please refer to the manual page on
[Drift and Batch Correction](manual-07-corrections.html) for more details. This
tutorial first demonstrates the drift-correction methods, then batch-effect
correction and how the two are combined.

::: {.callout .callout-note}
**Time** ~20 min &nbsp;·&nbsp; **Level** Intermediate &nbsp;·&nbsp; **Prerequisites** [Basic workflow](tutorial-02-basic-workflow.html)
:::

## Import data

In this tutorial, we import pre-calculated raw concentration values from a CSV file. This file must
contain a column with batch information (`batch_id`) if batch-wise correction should be applied, see
`import_data_csv_wide()` for more information.

```{r setup}
#| results: false  
#| message: false
library(mrmhub)

myexp <- mrmhub::MRMhubExperiment()

myexp <- import_data_csv_wide(
  myexp,
  path = "smooth-testdata.csv",
  variable_name = "conc",
  import_metadata = TRUE)
```

## QC-based smoothing

We apply a QC-based drift correction using cubic spline. See `correct_drift_cubicspline()` for
more information.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_cubicspline(
  myexp,
  batch_wise = FALSE,
  variable = "conc",
  ref_qc_types = "BQC",
  recalc_trend_after = TRUE)
```

Next, we inspect the data before and after the correction. We observe that the green trendline
follows the QC samples (in red). After the correction, the trend has been fully smoothed out,
resulting in a straight line.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

## Sample trend-based smoothing

Above, we observe that the QC samples do not fully represent the trends of the study samples. This
discrepancy can occur when the QC samples, which often based on pooled samples, differ in handling
or properties from the study samples.

We therefore apply a sample-based drift correction using gaussian kernel smoothing. See
`correct_drift_gaussiankernel()` for more information.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  kernel_size = 10,
  batch_wise = FALSE,
  ref_qc_types = "SPL",
  recalc_trend_after = TRUE)

```

Again, we inspect the data before and after the correction. We observe that the green trendline
follows the samples (in grey). After the correction, the trend has been fully smoothed out,
resulting in a straight line.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)

```

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)    
```
Now, the trends of the study samples appear to be fairly well corrected. However, in this example,
the differences compared to the previous QC-based smoothing are not very pronounced.

## Within-batch drift correction

In the examples before we applied drift correction across all batches. However, if batch-effects are
present that break the drifts, it is recommended to apply drift correction within each batch.

We explore a within-batch drift correction using a sample-based gaussian kernel smoothing from above
in this next example, by setting `batch_wise = TRUE`.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  kernel_size = 10,
  batch_wise = TRUE,
  ref_qc_types = "SPL",
  recalc_trend_after = TRUE)
```

Now, each batch has its own trendline corresponding the trend in each batch of each feature.
Contrary to the previous correction across batches, we now observe clear differences in the trends
still present after the correction. This is an effect of the correction applied independently to each
batch, due to different sample sizes and present batch effects. Therefore, it is often necessary to
apply batch correction after batch-wise drift correction

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

We therefore apply a subsequent batch correction using median-centering, resulting in an alignment
of the batches.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
mexp_drift <- mrmhub::correct_batch_centering(
  mexp_drift, 
  ref_qc_types = "SPL", 
  variable = "conc",
  correct_scale = TRUE
)

plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

## Batch-effect correction

Batch effects — systematic differences between analytical batches — are corrected
by median centering with `correct_batch_centering()`. The following example uses a
separate dataset of seven batches to show the effect clearly.

```{r}
#| results: false
#| message: false
myexp_batch <- mrmhub::MRMhubExperiment()
myexp_batch <- import_data_csv_wide(
  myexp_batch,
  path = "simdata-u1000-sd100_7batches.csv",
  variable_name = "conc",
  import_metadata = TRUE)
```

The concentration values of each batch are centered on a reference QC type, here the
study samples (`ref_qc_types = "SPL"`). The `correct_scale` parameter controls
whether the batch-to-batch differences in variance (scale) are also corrected; with
`correct_scale = FALSE` only the median (location) of each batch is aligned.

```{r}
#| results: false
#| code-overflow: 'wrap'
myexp_batch <- correct_batch_centering(
  data = myexp_batch,
  variable = "conc",
  ref_qc_types = "SPL",
  correct_scale = FALSE
)
```

The data before and after batch correction are compared below; the batches are now
aligned. If the study samples or other QC types do not follow the reference samples,
they may not be corrected appropriately.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false
#| message: false
plot_runscatter(myexp_batch, variable = "conc_before", rows_page = 1, cols_page = 1)
plot_runscatter(myexp_batch, variable = "conc", rows_page = 1, cols_page = 1)
```

### Correcting variance as well as location

Although the batches are aligned, the spread (variance) of the points still varies
considerably between batches. This is corrected by scaling the variance with
`correct_scale = TRUE`, after which the spread is fairly consistent across batches.

```{r}
#| fig.alt: >
#|   RunScatter plot after batch correction with variance scaling
#| results: false
#| message: false
myexp_batch <- mrmhub::correct_batch_centering(
  myexp_batch,
  ref_qc_types = "SPL",
  variable = "conc",
  correct_scale = TRUE
)

plot_runscatter(myexp_batch, variable = "conc", rows_page = 1, cols_page = 1)
```

## Method comparison

`MRMhub` provides four drift correction methods. Loess and cubic spline are typically used with
QC samples as the reference, gaussian kernel smoothing with study samples; a fourth method,
generalized additive model (GAM) smoothing via `correct_drift_gam()`, is also available. The method
comparison below summarises the typical use cases of the three most common — see the
[Drift and Batch Correction (reference)](manual-07-corrections.html) for the full
parameter description of all four.

| Method | Function | Reference samples | Typical use |
|--------|----------|-------------------|-------------|
| Loess | `correct_drift_loess()` | QC samples | Frequent QC injections; smooth trends; robust to single-point outliers |
| Cubic Spline | `correct_drift_cubicspline()` | QC samples | Frequent QC injections; flexible curves; sensitive to outlier QC points |
| Gaussian Kernel | `correct_drift_gaussiankernel()` | Study samples | Sparse QCs; only suitable for large, well-randomised sample sets |

## Loess drift correction

Loess uses locally weighted polynomial regression and is less sensitive to individual outlier QC
points than cubic spline. The `span` parameter controls smoothness (higher = smoother).

```{r}
#| results: false
#| code-overflow: 'wrap'
mexp_drift_loess <- correct_drift_loess(
  myexp,
  variable = "conc",
  batch_wise = TRUE,
  ref_qc_types = "BQC",
  recalc_trend_after = TRUE)
```

Inspect the result with `plot_runscatter()` before exporting.

## Export corrected data

Next, we can either continue to work with the corrected data using `MRMhub` functions or export the
data.

```{r}
save_dataset_csv(
  mexp_drift, 
  path = "drift-batch-corrected-conc-data.csv", 
  variable = "conc", 
  filter_data = FALSE
  )
```

## Next steps

- [RunScatter and PCA QC exploration](tutorial-05-run-scatter.html) — visualise run-order and batch effects
- [Drift and Batch Correction (reference)](manual-07-corrections.html) — full method documentation
- [Calibration by Reference Sample](tutorial-07-calibration-reference.html) — normalise to a reference material
