---
title: "Validating and Fixing Metadata"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
description: |
  How to validate analysis, feature, and ISTD annotations before processing,
  diagnose common defects, and generate a metadata template from your data.
vignette: >
  %\VignetteIndexEntry{Validating and Fixing Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      out.width = "100%")
```

<p class="page-kind page-kind-tutorial">Tutorial</p>

Most processing errors in MRMhub trace back to defects in the annotation tables rather than the
analytical data itself. The most frequent offenders are mismatched `analysis_id` between data and
annotation, missing required columns (`qc_type`, `batch_id`, `analysis_order`), and `qc_type`
labels that do not match the canonical vocabulary used internally for QC selection. This tutorial
covers generating a template, validating annotations against imported data, and resolving
the typical defects.

See [Sample Types and QC Roles](manual-06-sample-types.html) for the canonical `qc_type` vocabulary
and [Metadata Import](manual-05-metadata.html) for the per-table importer reference.

::: {.callout .callout-note}
**Time** ~10 min &nbsp;·&nbsp; **Level** Beginner &nbsp;·&nbsp; **Prerequisites** [Your first analysis](tutorial-00-first-analysis.html)
:::

## 1. Generate a metadata template

`save_metadata_templates()` writes an XLSX template containing one sheet per annotation table
(`annot_analyses`, `annot_features`, `annot_istds`, `annot_responsecurves`,
`annot_qcconcentrations`). The template carries the expected column names and example rows.

```{r}
library(mrmhub)

save_metadata_templates(path = "metadata/sPerfect_metadata.xlsx")
```

Fill in the sheets in Excel, save, and proceed to import. The `MRMhub Metadata Organizer` template
(`save_metadata_msorganiser_template()`) is a richer alternative that bundles every metadata table
into a single `.xlsx` workbook, for labs that prefer Excel-driven workflows.

## 2. Extract a template from imported data

If the imported data file already carries embedded metadata (e.g. an INTEGRATOR `long.tsv` with
`qc_type`, `batch_id`, and `analysis_order` columns), `import_metadata_from_data()` lifts those
columns into the annotation slots and links them to the dataset in a single call. The function
returns the updated `MRMhubExperiment`.

```{r}
myexp <- MRMhubExperiment()
myexp <- import_data_mrmhub(myexp,
                            path = "datasets/sPerfect_MRMhub.tsv",
                            import_metadata = FALSE)

# Pull embedded metadata, validate, and link in one step
myexp <- import_metadata_from_data(myexp)
```

This is a fast path for data that already contains analysis metadata in the import file. For
manual edits or additional annotation (ISTDs, calibration concentrations, biological covariates),
fill in the XLSX template from Step 1 and use the per-table importers in Step 3.

## 3. Import and validate annotation tables

The per-table importers (`import_metadata_analyses()`, `import_metadata_features()`,
`import_metadata_istds()`, `import_metadata_responsecurves()`, `import_metadata_qcconcentrations()`)
read a CSV or Excel file, validate it against the imported data, and link the validated table to the
experiment. Defects are reported via `cli` warnings and aborts; on success the table is attached.

```{r}
myexp <- import_metadata_analyses(myexp,
                                  path = "metadata/sPerfect_metadata.xlsx",
                                  sheet = "annot_analyses",
                                  excl_unmatched_analyses = FALSE,
                                  ignore_warnings = FALSE)

myexp <- import_metadata_features(myexp,
                                  path = "metadata/sPerfect_metadata.xlsx",
                                  sheet = "annot_features")

myexp <- import_metadata_istds(myexp,
                               path = "metadata/sPerfect_metadata.xlsx",
                               sheet = "annot_istds")
```

The import validates:

- presence of required columns (`analysis_id`, `qc_type` for analyses; `feature_id` for features;
  `istd_feature_id` and an `istd_conc_*` column for ISTDs);
- uniqueness of identifiers (`analysis_id`, `feature_id`);
- consistency between annotation IDs and IDs present in `dataset_orig`;
- canonical `qc_type` labels (see [Sample Types](manual-06-sample-types.html)).

Defects are reported using `cli` messages that name the offending table, column, and IDs. When
annotation tables are assembled programmatically rather than read from a file, pass them to
[`add_metadata()`](../reference/add_metadata.html) as a named list; the same validation is applied
before the tables are linked.

## 4. Common defects and resolutions

<details>
<summary><strong>Column <code>qc_type</code> missing from <code>annot_analyses</code></strong></summary>

`qc_type` is the only column besides `analysis_id` that is strictly required. Add it before
re-importing:

```{r}
annot_analyses <- annot_analyses |>
  dplyr::mutate(qc_type = dplyr::case_when(
    stringr::str_detect(analysis_id, "BQC")  ~ "BQC",
    stringr::str_detect(analysis_id, "BLK")  ~ "SBLK",
    TRUE                                     ~ "SPL"
  ))
```

</details>

<details>
<summary><strong><code>analysis_id</code> in data not found in annotation</strong></summary>

Some injections in the imported dataset have no matching row in the annotation table. List them and
inspect the difference — leading/trailing whitespace and case mismatches are the most frequent
causes.

```{r}
data_ids  <- unique(myexp@dataset_orig$analysis_id)
annot_ids <- annot_analyses$analysis_id

# IDs present in data but missing from annotation
setdiff(data_ids, annot_ids)

# Fix trailing whitespace introduced by Excel
annot_analyses <- annot_analyses |>
  dplyr::mutate(analysis_id = trimws(analysis_id))
```

`excl_unmatched_analyses = TRUE` in the importer drops unmatched analyses silently — useful only
when the omission is intentional (e.g., excluded conditioning injections).

</details>

<details>
<summary><strong>Non-canonical <code>qc_type</code> values</strong></summary>

The canonical vocabulary is `BQC`, `TQC`, `PQC`, `HQC`, `MQC`, `LQC`, `QC`, `CAL`, `RQC`, `EQC`,
`NIST`, `LTR`, `EQA`, `SPL`, `SST`, `SBLK`, `TBLK`, `PBLK`, `MBLK`, `UBLK`. Labels are
case-sensitive. Replace non-standard values before validation:

```{r}
unique(annot_analyses$qc_type)

annot_analyses <- annot_analyses |>
  dplyr::mutate(qc_type = dplyr::case_when(
    qc_type == "Standard" ~ "CAL",
    qc_type == "Blank"    ~ "SBLK",
    qc_type == "Pool"     ~ "BQC",
    qc_type == "Sample"   ~ "SPL",
    TRUE                  ~ qc_type
  ))
```

</details>

<details>
<summary><strong>Duplicate identifiers</strong></summary>

```{r}
annot_analyses |>
  dplyr::count(analysis_id) |>
  dplyr::filter(n > 1)
```

Common cause: copy-paste errors in Excel, or repeated header rows interpreted as data.

</details>

<details>
<summary><strong>Feature IDs in annotation do not match data</strong></summary>

```{r}
data_features  <- unique(myexp@dataset_orig$feature_id)
annot_feat_ids <- annot_features$feature_id

# In data but missing from annotation
setdiff(data_features, annot_feat_ids)

# In annotation but absent from data (often method-development residue)
setdiff(annot_feat_ids, data_features)
```

A common cause is differing encoding of special characters (parentheses, slashes, unicode escapes)
between the integration software output and a hand-edited Excel file.

</details>

## 5. Practical recommendations

- Re-import annotations after every edit. Saved Excel files held open in a separate process can be
  silently locked and read partially.
- Save CSVs as UTF-8 without BOM. Excel's default "CSV UTF-8" export adds a byte-order mark that
  shifts the first column header.
- Avoid special characters in `analysis_id` and `feature_id` (parentheses, slashes, spaces) — they
  survive R but complicate downstream column references. Letters, digits, underscores, hyphens.
- Keep the annotation file under version control alongside the data file.

## Next steps

- [Sample Types & QC Roles](manual-06-sample-types.html) — canonical `qc_type` vocabulary
- [Importing Metadata](manual-05-metadata.html) — per-table importer reference
- [Importing Analytical Data](manual-04-data-import.html) — importing peak area / concentration data
- [Your First Analysis](tutorial-00-first-analysis.html) — minimal end-to-end workflow
