---
title: "Isotopic interference correction"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4
description: >
  Correct fragment-level isotopic (M+2) overlap in class-based targeted MRM
  assays: annotate mrm_pattern, auto-derive the overlaps, inspect, and subtract.
vignette: >
  %\VignetteIndexEntry{Isotopic interference correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
bibliography: references.bib
---

```{r setup, include = FALSE}
source("_common.R")
knitr::opts_chunk$set(eval = FALSE)
```

<p class="page-meta">

[Tutorial]{.page-kind .page-kind-tutorial} [Advanced]{.page-level} [Prerequisites:
<a href="tutorial-02-basic-workflow.html">Basic workflow</a>]{.page-prereq}

</p>

Class-based chromatography — HILIC, normal-phase LC, or SFC — co-elutes same-class lipid species,
whose isotopic envelopes overlap and bias quantification [@gao2021LICARApplicationIsotopic]. In these
assays the overlap is the rule and has to be corrected — a *Type II* isotopic correction (overlap
between distinct species, as opposed to *Type I* correction of a compound's own isotopologue pattern).

The correction is precursor-based for shotgun (MS1) data but must be done at the fragment level for
MRM, where LICAR [@gao2021LICARApplicationIsotopic] is the sole method. MRMhub implements LICAR: it
runs on the raw feature intensities (`feature_intensity`) and applies to *every* sample, unlike drift
and batch correction, which are fitted on QC samples only. The
[Isotopic interference correction](manual-12-interference-correction.html) manual page covers the
concepts.

## 1. Annotate the MRM pattern

Automatic derivation needs one manual annotation: an `mrm_pattern` column on the `Features` metadata
sheet, encoding the lipid class and the product-ion origin so `calc_isotopic_interferences()` can
build the fragment formula. Precursor/product m/z and the optional `polarity` come from the import or
the metadata; only `mrm_pattern` is added by hand:

```{r, eval = TRUE, echo = FALSE}
tibble::tribble(
  ~feature_id, ~feature_class, ~precursor_mz, ~product_mz, ~polarity, ~mrm_pattern,
  "PC 34:2", "PC", 758.6, 184.1, "Pos", "PC (Pos) Pro=184.1",
  "PC 34:1", "PC", 760.6, 184.1, "Pos", "PC (Pos) Pro=184.1",
  "PC 34:0", "PC", 762.6, 184.1, "Pos", "PC (Pos) Pro=184.1",
  "Cer 18:1;O2/16:0", "Cer", 538.5, 264.3, "Pos", "Cer (Pos) SphB-2H2O",
  "PC 34:1 [FA 16:0]", "PC", 804.6, 255.2, "Neg", "PC (Neg, FA) FA"
) |>
  knitr::kable(caption = "Example `Features` metadata (`mrm_pattern` is the added annotation).")
```

The valid labels are listed by [`licar_pattern_choices()`](../reference/licar_pattern_choices.html):

```{r}
library(mrmhub)
licar_pattern_choices("Head Group")
```

The `save_metadata_templates()` template offers `mrm_pattern` as a filtered dropdown. On import,
MRMhub validates it: an unknown label is an error; a name/pattern class mismatch, or a sum-composition
name under a chain-resolved (FA/LCB) pattern, is a warning.

```{r}
mexp <- readRDS("results/mexp_processed.rds")
```

## 2. Derive the interference relationships

[`calc_isotopic_interferences()`](../reference/calc_isotopic_interferences.html) discovers the M+2
overlaps and stores them in `annot_interferences`. Set `level` to the acquisition: `"MRM"` for
class-based LC-MRM (fragment-level) or `"MS1"` for genuine full-scan data. Always use `"MRM"` for MRM
data — `"MS1"` is not a valid fallback when a product m/z is missing.

```{r}
# Class-based LC-MRM: fragment-level M+2 (needs precursor + product m/z).
mexp <- calc_isotopic_interferences(mexp, level = "MRM")

# Genuine MS1 / full-scan: whole-molecule M+2 (needs precursor m/z only).
# mexp <- calc_isotopic_interferences(mexp, level = "MS1")
```

Factors are computed with `enviPat` 2.8 [@loos2015NontargetEnviPat] and are deterministic. Raise
`min_contribution` to drop negligible pairs; see
[co-elution filtering](#co-elution-filtering-experimental) to keep only co-eluting pairs.

## 3. Inspect the derived relationships

The two-stage API lets you review the edges before subtracting. Each row of `annot_interferences` is
one overlap, with its contribution factor *K* in `interference_contribution` and an `overlap_type`
(`m2_front`, `m2_back`, or `ms1_m2`):

```{r}
mexp@annot_interferences
```

[`summarize_interferences()`](../reference/summarize_interferences.html) rolls this up per feature —
how many are affected, split by source and overlap type, with the contribution-factor range:

```{r}
summarize_interferences(mexp)
```

A feature carrying both a `m2_front` and a `m2_back` row is corrected against two interferers; the
`source` column marks each edge as auto-derived (`"auto"`) or declared (`"manual"`).

## 4. Apply the correction

[`correct_isotopic_interferences()`](../reference/correct_isotopic_interferences.html) subtracts all
derived edges from the raw intensities in one pass (its reference page gives the formula and
zero-clamping).

```{r}
mexp <- correct_isotopic_interferences(mexp)
```

Apply it **before** ISTD normalisation, drift, and batch correction — running it later resets those
steps (with a warning). Uncorrected values are kept in `feature_intensity_orig`. Overlaps you
annotate yourself (in-source fragments, co-eluting isobars) are applied with
[`correct_custom_interferences()`](../reference/correct_custom_interferences.html).

## 5. Correct a single pair manually

For a one-off correction, or to check a factor before trusting the automatic derivation, use
[`correct_interference_manual()`](../reference/correct_interference_manual.html), where `variable` is
the `dataset` column (`feature_intensity`):

```{r}
mexp <- correct_interference_manual(
  mexp,
  variable = "feature_intensity",
  feature = "PC 32:0",
  interfering_feature = "SM 36:1 M+3",
  interference_contribution = 0.0107,
  neg_to_na = FALSE,
  updated_feature_id = NA)
```

`updated_feature_id` renames the corrected feature so raw and corrected channels can coexist. A
manual factor can come from a theoretical calculation (`enviPat`), a pure-standard injection (which
also captures Q1/Q3 transmission and in-source effects), or a published class-level table.

## 6. Verify the correction

```{r, eval = TRUE, include = FALSE, message = FALSE, warning = FALSE}
# Build a corrected object for the QC plots below from the bundled multi-class
# LICAR HILIC demo (PC, SM, Cer, dhCer in positive mode plus PC/PE via the
# negative FA fragment), mirroring sections 2-4: import the long analysis data
# (carrying precursor/product m/z and polarity) together with the mrm_pattern
# feature sheet, auto-derive the M+2 edges, and subtract them. The derivation
# needs enviPat 2.8 and rgoslin.
library(mrmhub)
demo_csv <- system.file("extdata", "licar_hilic_demo.csv", package = "mrmhub")
demo_features <- system.file(
  "extdata",
  "licar_hilic_demo_features.csv",
  package = "mrmhub"
)

mexp <- MRMhubExperiment() |>
  import_data_csv_long(path = demo_csv, import_metadata = TRUE, silent = TRUE) |>
  import_metadata_features(path = demo_features, ignore_warnings = TRUE) |>
  calc_isotopic_interferences(level = "MRM") |>
  correct_isotopic_interferences()
```

```{r}
d <- get_analyticaldata(mexp, annotated = TRUE) |>
  dplyr::filter(feature_id == "PC 32:0") |>
  dplyr::select(
    analysis_id, qc_type,
    intensity_before = feature_intensity_orig,
    intensity_after = feature_intensity) |>
  dplyr::mutate(
    pct_change =
      100 * (intensity_after - intensity_before) / intensity_before)

summary(d$pct_change)
```

In blanks (`SBLK`/`PBLK`) the residual signal should approach zero; a non-zero blank median points to
an underestimated factor. [`plot_qc_interference_impact()`](../reference/plot_qc_interference_impact.html)
shows how many features were corrected at each magnitude (percent of signal removed):

```{r qc-impact}
#| eval: true
#| fig.cap: !expr fig_cap("How many features were corrected at each magnitude (percent of signal removed) in the study samples.")
#| fig.alt: Distribution of per-feature interference-correction magnitude in study samples
plot_qc_interference_impact(mexp, qc_types = "SPL")
```

For a per-feature view,
[`plot_interference_correction()`](../reference/plot_interference_correction.html) plots each
feature's residual signal as a percent of its uncorrected value, split by QC type. `min_correction_pct`
restricts it to strongly affected features (here ≥ 40 % removed), `sort_by_effect` ranks by magnitude,
and `top_n` keeps the largest:

```{r per-feature}
#| eval: true
#| fig.cap: !expr fig_cap("Per-feature residual signal after correction, as a percent of the uncorrected value, for features with at least 40% removed, split by QC type.")
#| fig.alt: Per-feature residual signal after interference correction, split by QC type
plot_interference_correction(
  mexp,
  qc_types = "SPL",
  min_correction_pct = 40,
  sort_by_effect = "desc",
  point_size = 1.5, point_alpha = 0.8)
```

Each point is one study sample; `SPL` points have transparent fill, so raise `point_size` /
`point_alpha` when only a few are shown.

## Co-elution filtering (experimental) {#co-elution-filtering-experimental}

An interferer's M+2 lands inside a victim's integrated area only if the two co-elute. The experimental
`check_coelution` gate keeps an m/z-matched edge only when the interferer's apex falls within the
victim's integration window, dropping chromatographically resolved pairs:

```{r}
mexp <- calc_isotopic_interferences(
  mexp, level = "MRM", check_coelution = TRUE)
```

It is **off by default** while the gate is validated. When enabled, the pairs it drops — and, where
retention data are missing, those it cannot verify — are reported in the console.

## Next steps

- [Isotopic interference correction](manual-12-interference-correction.html): the conceptual
  reference
- [Drift and batch correction](tutorial-04-drift-correction.html): apply after interference
  correction
- [Basic MRMhub workflow](tutorial-02-basic-workflow.html): full processing pipeline
- [The MRMhubExperiment data object](manual-02-data-object.html#feature-variables): how `_orig`
  postfixes preserve raw values

## References