---
title: "Interference Correction"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4
description: |
  Annotate mrm_pattern, derive isotopic interference relationships automatically
  with derive_interferences(), inspect them, and apply the subtraction with
  correct_interferences(). Manual single-pair correction is also covered.
vignette: >
  %\VignetteIndexEntry{Interference Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      fig.width = 7, fig.height = 4, out.width = "100%")
```

<p class="page-kind page-kind-tutorial">Tutorial</p>

In class-based targeted assays, the natural-abundance heavy isotopologues of one species can overlap
the transition of a species two mass units heavier, inflating its measured area. MRMhub derives these
isotopic (M+2) interference relationships automatically from a per-feature `mrm_pattern` annotation
and removes them with a contribution-based subtraction ported from the LICAR method
[@gao2021LICARApplicationIsotopic]. This tutorial walks through the end-to-end workflow; the concepts
(front/back overlaps, fragment-based-for-MRM, the co-elution requirement) are documented in the
[Isotopic interference correction](manual-12-interference-correction.html) manual page.

::: {.callout .callout-note}
**Time** ~15 min &nbsp;·&nbsp; **Level** Advanced &nbsp;·&nbsp; **Prerequisites** [Basic workflow](tutorial-02-basic-workflow.html)
:::

## 1. When is correction needed?

Isotopic interference correction should be considered when:

1. blank injections show a non-zero signal in a feature that should be zero;
2. a species and an M+2 "shoulder" two mass units heavier show strong injection-to-injection
   correlation, suggestive of isotopologue overlap;
3. the theoretical M+2 contribution from an adjacent same-class species is large enough to bias the
   measurement (more than a few percent of the target signal).

Correction operates on raw feature intensities (`feature_intensity`) and should be applied *before*
ISTD normalisation, drift, and batch correction, so those downstream steps use corrected raw signals.

## 2. Annotate `mrm_pattern`

Automatic derivation needs one input: the `mrm_pattern` column on the `Features` metadata sheet,
giving each feature its class + MRM pattern (e.g. `PC (Pos) Pro=184.1`, `Cer (Pos) SphB-2H2O`,
`PC (Neg, FA) FA`). The label encodes both the lipid class and the product-ion origin. The valid
labels are listed by [`licar_pattern_choices()`](../reference/licar_pattern_choices.html):

```{r}
library(mrmhub)
licar_pattern_choices("Head Group")
```

The metadata template (`save_metadata_templates()`) provides `mrm_pattern` as a filtered dropdown on
the `Features` sheet, with an optional `polarity` column that narrows the choices and colour warnings
when a species name does not match the chosen pattern's class. On import, MRMhub validates the column:
an unknown label is an error, and a name/pattern class mismatch or a sum-only name under a
chain-resolved (FA/LCB) pattern is a warning.

```{r}
mexp <- readRDS("results/mexp_processed.rds")
```

## 3. Derive interference relationships

[`derive_interferences()`](../reference/derive_interferences.html) discovers the M+2 overlaps and
stores them in the `annot_interferences` slot. Choose the derivation `level` to match the acquisition:

```{r}
# Class-based LC-MRM: fragment-level front/back correction (needs precursor + product m/z).
mexp <- derive_interferences(mexp, level = "MRM")

# Genuine MS1 / full-scan: whole-molecule M+2 (needs precursor m/z only).
# mexp <- derive_interferences(mexp, level = "MS1")
```

::: {.callout .callout-caution}
**MS1 is not a fallback for MRM.** MRM correction must be fragment-based, because a heavy isotope's
contribution depends on whether it lands on the retained product ion or the neutral loss
[@gao2021LICARApplicationIsotopic]. Use `level = "MS1"` only for genuine full-scan data — never as a
substitute for MRM data missing a product m/z. See the [manual page](manual-12-interference-correction.html)
for the full rationale.
:::

By default (`check_coelution = TRUE`) an m/z-matched edge is kept only when the interferer's peak apex
falls inside the victim's integration window; chromatographically resolved pairs are dropped and
reported. Derived factors are pinned to `enviPat` 2.8 [@loos2015NontargetEnviPat] and are
deterministic, so re-importing the annotation and re-deriving reproduces identical edges.

## 4. Inspect the derived relationships

The two-step API lets you review the derived edges before subtracting anything. Each row is one
overlap, with its contribution factor `K` and an `overlap_type` (`m2_front`, `m2_back`, or `ms1_m2`):

```{r}
mexp@annot_interferences
```

A victim carrying both a `m2_front` and a `m2_back` row is corrected against two interferers. The
`source` column distinguishes auto-derived (`"auto"`) from manually annotated (`"manual"`) edges;
both are applied together.

## 5. Apply the correction

[`correct_interferences()`](../reference/correct_interferences.html) reads `annot_interferences`
(unioned with any legacy manual annotation columns) and subtracts every edge in one pass:

$$
\text{Value}_\text{corrected} = \text{Value}_\text{target} -
   \sum_i K_i \cdot \text{Value}_{\text{interfering}_i}
$$

```{r}
mexp <- correct_interferences(mexp, variable = "feature_intensity")
```

Upstream features in a chain (A ← B ← C) are corrected first via a topological ordering; circular
dependencies are detected and aborted with an informative error. For auto-derived edges, a
subtraction is skipped when the running-corrected interferer has already fallen to zero, and corrected
values are clamped at zero (LICAR parity). The original raw values are preserved in
`feature_intensity_orig`.

## 6. Manual correction of a single pair

For a one-off correction, or to validate a factor before trusting the automatic derivation, use
[`correct_interference_manual()`](../reference/correct_interference_manual.html). `variable` here is the
actual `dataset` column name (`feature_intensity`).

```{r}
mexp <- correct_interference_manual(
  mexp,
  variable                  = "feature_intensity",
  feature                   = "PC 32:0",
  interfering_feature       = "SM 36:1 M+3",
  interference_contribution = 0.0107,
  neg_to_na                 = FALSE,
  updated_feature_id        = NA
)
```

Setting `updated_feature_id` renames the corrected feature so raw and corrected channels can coexist
under different IDs. Manual factors can be sourced from a theoretical isotopologue calculation
(`enviPat`), an empirical pure-standard injection (which also captures Q1/Q3 transmission and
in-source effects), or published class-level interference tables — verified empirically when moving a
method between instruments.

## 7. Verifying the correction

```{r}
d <- get_analyticaldata(mexp, annotated = TRUE) |>
  dplyr::filter(feature_id == "PC 32:0") |>
  dplyr::select(analysis_id, qc_type,
                intensity_before = feature_intensity_orig,
                intensity_after  = feature_intensity) |>
  dplyr::mutate(pct_change = 100 * (intensity_after - intensity_before) / intensity_before)

summary(d$pct_change)
```

For blanks (`SBLK`/`PBLK`), residual signal after correction should approach zero. A non-zero blank
median often indicates an underestimated contribution factor.

## Recommendations

- Apply interference correction on raw `feature_intensity` *before* normalisation, drift, and batch
  correction.
- Prefer `level = "MRM"` for MRM data and supply the product m/z; reserve `level = "MS1"` for genuine
  full-scan measurements.
- Inspect `annot_interferences` before applying, especially any pairs the co-elution gate dropped.
- Keep the `mrm_pattern` annotation with the project: it reproduces the derived factors under the
  `enviPat` 2.8 pin.
- After correction, verify that blank residual signal approaches zero.

## Next steps

- [Isotopic interference correction](manual-12-interference-correction.html) — the conceptual reference
- [Drift and Batch Correction](tutorial-04-drift-correction.html) — apply after interference correction
- [Basic MRMhub Workflow](tutorial-02-basic-workflow.html) — full processing pipeline
- [The MRMhubExperiment Data Object](manual-02-data-object.html#feature-variables) — how `_orig` postfixes preserve raw values

## References
