---
title: "A Real-World MRMhub Workflow"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
description: |
  A realistic end-to-end workflow covering project setup, metadata import, 
  drift/batch correction, QC filtering, and export.
vignette: >
  %\VignetteIndexEntry{A Real-World MRMhub Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, results = TRUE, message = TRUE, comment = "#>", out.width = "100%")
set.seed(1041)
options(dplyr.print_max = 10)
```

::: {.callout .callout-note}
**Time:** ~15 min &nbsp;|&nbsp; **Level:** Beginner–Intermediate &nbsp;|&nbsp; **Prerequisites:** [Your First Analysis](tutorial-00-first-analysis.html)
:::

::: {.callout .callout-caution}
**How this differs from "Your First Analysis":** That tutorial uses
bundled demo data and covers only import, normalize, and export in 5 minutes.
This tutorial uses real file paths, imports metadata via MSOrganiser, applies
drift and batch correction, runs QC filtering, and exports a full report —
a realistic end-to-end workflow.
:::

This tutorial outlines the key steps in a MRMhub workflow, based on a
lipidomics dataset. See the [Lipidomics Data Processing](tutorial-03-lipidomics-workflow.html) 
tutorial for an even more detailed example.

These examples are simplified and may not apply to every dataset and
experimental setup. See other tutorials and recipes for additional
workflows and data types.

# Step 1: Set Up a Project

To start a new MRMhub data analysis, create an RStudio or Positron
project. (See [Using RStudio Projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects)
or the [Positron User Guide](https://positron.posit.co/).)

The project should contain the following subfolders:

```
my_study/
├── data/           # raw data and metadata files
├── output/         # exported results
└── analysis.Rmd    # your processing notebook
```

R/Notebook (`.Rmd`) or [Quarto](https://docs.posit.co/ide/user/ide/guide/documents/quarto-project.html) 
(`.qmd`) files are good choices for documented processing workflows. They
combine code with formatted text to document every processing decision.

Start with a new notebook or R script and load the `mrmhub` package:

```{r}
library(mrmhub)
```

# Step 2: Create a MRMhubExperiment

The `MRMhubExperiment` object is the main **data container** in the
MRMhub workflow. See [The MRMhubExperiment Object](manual-04-mrmhub-experiment.html) 
for details.

Create a new empty object:

```{r}
myexp <- MRMhubExperiment()
```

# Step 3: Import Analysis Data

Import peak area data from an INTEGRATOR output file. This file also
contains some metadata (e.g., `qc_type`, `batch_id`) which we import
at the same time.

```{r load-rawdata}
myexp <- import_data_mrmhub(
  data = myexp, 
  path = "datasets/sPerfect_MRMhub.tsv", 
  import_metadata = TRUE
)
```

<details>
<summary><strong>Using a different data source?</strong></summary>

See [Import and prepare data files](manual-05a-which-importer.html) for a
decision flowchart, or [Data Import](manual-05-data-import.html) for
the full API reference.

</details>

# Step 4: Add Metadata

Processing steps require additional information not available in the
data file: internal standard assignments, concentrations, sample
amounts, etc.

Metadata can be imported from separate CSV/Excel files or R data frames
(see [Metadata Import](manual-06-metadata-import.html)), or from an
MSOrganiser template:

```{r import-metadata}
myexp <- import_metadata_msorganiser(
  myexp, 
  path = "datasets/sPerfect_Metadata.xlsx", 
  ignore_warnings = TRUE
)
```

::: {.callout .callout-note}
The validation checks may produce warnings. Setting
`ignore_warnings = TRUE` allows the import to proceed, but
warnings are still displayed in the console (marked with `*`).
Review them to ensure nothing critical is missed.
:::

<details>
<summary><strong>Validating metadata before import</strong></summary>

Use `assert_metadata()` to check annotation tables before linking. The function takes the metadata
as a named list and returns the validated list, which is then passed to `add_metadata()`:

```{r, eval = FALSE}
validated <- assert_metadata(
  myexp,
  metadata = list(annot_analyses = annot_analyses,
                  annot_features = annot_features),
  ignore_warnings = FALSE,
  excl_unmatched_analyses = FALSE
)

myexp <- add_metadata(myexp, metadata = validated)
```

See [Validating and Fixing Metadata](tutorial-10-metadata-validation.html) for a full walkthrough.

</details>

# Step 5: Process the Data

Apply normalization, quantification, drift correction, batch correction,
and QC filtering:

```{r processing}
# Normalize by internal standards
myexp <- normalize_by_istd(myexp)

# Quantify using ISTD-based approach
myexp <- quantify_by_istd(myexp)

# Correct run-order drift (Gaussian kernel, sample-based)
myexp <- correct_drift_gaussiankernel(
  data = myexp,
  variable = "conc",
  ref_qc_types = c("SPL")
)

# Correct batch effects (median centering)
myexp <- correct_batch_centering(
  myexp, 
  variable = "conc",
  ref_qc_types = "SPL"
)

# Apply QC-based feature filtering
myexp <- filter_features_qc(
  data = myexp,
  include_qualifier = FALSE,
  include_istd = FALSE, 
  min.signalblank.median.spl.sblk = 10,
  max.cv.conc.bqc = 25
)
```

Note that the drift correction above uses the study samples
(`ref_qc_types = "SPL"`) as the reference. Study-sample smoothing is only
appropriate for large, well-randomised sample sets; the QC-based convention
(Broadhurst et al. 2018) instead fits the drift trend on dedicated QC injections.

<details>
<summary><strong>Processing step order</strong></summary>

The recommended order is:

1. Normalize by ISTD
2. Quantify (ISTD or external calibration)
3. Drift correction
4. Batch correction
5. QC filtering

See [Design Overview](manual-00-architecture.html) for the full pipeline diagram.

</details>

# Step 6: Visualize Results

Create a run scatter plot to check signal trends and correction quality:

```{r}
plot_runscatter(
  data = myexp,
  variable = "conc",
  include_feature_filter = "PC 4",
  include_istd = FALSE,
  cap_outliers = TRUE,
  log_scale = FALSE,
  output_pdf = FALSE,
  path = "./output/runscatter_PC408_beforecorr.pdf",
  cols_page = 3, rows_page = 2
)
```

See [Visualisation Functions](manual-08-visualization.html) for the full
list of plotting functions grouped by workflow stage.


# Step 7: Export and Share

```{r save-report}
# Detailed Excel report with multiple sheets
save_report_xlsx(myexp, path = tempfile(fileext = ".xlsx"))

# Flat CSV with concentration values that passed QC
save_dataset_csv(
  data = myexp, 
  path = tempfile(fileext = ".csv"),
  variable = "conc", 
  qc_types = "SPL", 
  include_qualifier = FALSE,
  filter_data = TRUE
)

# Save the complete object for reproducibility or sharing
saveRDS(myexp, file = tempfile(fileext = ".rds"), compress = TRUE)
```

::: {.callout .callout-tip}
The `.rds` file preserves the entire `MRMhubExperiment`
object. A colleague can open it in R and run their own processing, plots,
and QC checks; no code is needed to reproduce the data state.
:::

## Next Steps

- [Lipidomics Workflow](tutorial-03-lipidomics-workflow.html) — more detailed lipidomics example
- [Drift Correction](tutorial-04-drift-correction.html) — drift correction methods
- [Batch Correction](tutorial-06-batch-correction.html) — batch correction options
- [PCA Exploration](tutorial-09-pca-exploration.html) — QC exploration with PCA
- [External Calibration & QC](recipe-01-ext-calibration-qc.html) — external calibration

