---
title: "A basic MRMhub workflow"
description: >
  A realistic end-to-end workflow, from project setup through drift and
  batch correction, QC filtering, and export.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{A basic MRMhub workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Beginner</span> <span class="page-prereq">Prerequisites: <a href="tutorial-00-first-analysis.html">Your First Analysis</a></span></p>

MRMhub processes targeted mass-spectrometry data through an ordered sequence: import, metadata
annotation, normalization, quantification, drift and batch correction, QC filtering, and export.
This tutorial walks through that full sequence on a lipidomics dataset, using real file paths, an
MSOrganiser metadata template, and a complete exported report. The examples are deliberately
simplified; the [Lipidomics data processing](tutorial-03-lipidomics-workflow.html) tutorial and the
recipes cover dataset- and instrument-specific variations. After this tutorial you can set up a
project, run the standard pipeline end to end, and export a QC-filtered report.

## 1. Set up a project

To start a new analysis, create an RStudio or Positron project (see
[Using RStudio Projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects)
or the [Positron User Guide](https://positron.posit.co/)). A predictable folder layout keeps raw
data, results, and the processing notebook separate:

```
my_study/
├── data/           # raw data and metadata files
├── output/         # exported results
└── analysis.Rmd    # your processing notebook
```

An R Notebook (`.Rmd`) or
[Quarto](https://docs.posit.co/ide/user/ide/guide/documents/quarto-project.html) (`.qmd`) document
is a good home for a processing workflow: it interleaves code with prose, documenting every
decision. Load the package to begin:

```{r}
library(mrmhub)
```

## 2. Create the experiment and import data

The `MRMhubExperiment` object is the central data container in a MRMhub workflow (see
[The MRMhubExperiment object](manual-02-data-object.html)). We create an empty object, then import
peak-area data from an INTEGRATOR output file; the file also carries some metadata (`qc_type`,
`batch_id`), which we import at the same time.

```{r}
myexp <- MRMhubExperiment()
```

```{r load-rawdata}
#| message: true
myexp <- import_data_mrmhub(
  myexp,
  path = "datasets/sPerfect_MRMhub.tsv",
  import_metadata = TRUE)
```

MRMhub reads several other formats; [Import and prepare data files](manual-04-data-import.html) has a
decision flowchart and the full importer reference.

## 3. Add metadata

Most processing steps need information absent from the data file: internal-standard assignments,
calibrant concentrations, sample amounts. Metadata can come from separate CSV or Excel files, from R
data frames (see [Metadata import](manual-05-metadata.html)), or from an MSOrganiser template:

```{r import-metadata}
#| message: true
myexp <- import_metadata_msorganiser(
  myexp,
  path = "datasets/sPerfect_Metadata.xlsx",
  ignore_warnings = TRUE)
```

Validation checks may raise warnings, which by default halt the import. Setting
`ignore_warnings = TRUE` lets it proceed; the warnings still appear in the console table, marked with
an asterisk (`*`) in the `status` column, and should be reviewed so nothing critical is missed.
[Validating and fixing metadata](recipe-04-validate-metadata.html) shows how to inspect and
resolve them.

## 4. Process the data

With data and metadata in place, we run the core pipeline: normalize to internal standards, quantify,
correct run-order drift and between-batch effects, and filter features on QC criteria.

```{r processing}
myexp <- normalize_by_istd(myexp)
myexp <- quantify_by_istd(myexp)

myexp <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  ref_qc_types = c("SPL"))

myexp <- correct_batch_centering(
  myexp,
  variable = "conc",
  ref_qc_types = "SPL")

myexp <- filter_features_qc(
  myexp,
  include_qualifier = FALSE,
  include_istd = FALSE,
  min.signalblank.median.spl.sblk = 10,
  max.cv.conc.bqc = 25)
```

The drift and batch corrections here are fitted on the study samples (`ref_qc_types = "SPL"`), which
suits large, well-randomised sample sets. [Drift and batch correction](tutorial-04-drift-correction.html)
covers fitting on dedicated QC samples instead and how to choose a method.

## 5. Visualize results

A run-scatter plot shows signal against injection order, a quick visual check on trends and
correction quality. We plot a single lipid class here;
[Visualisation functions](manual-08-visualization.html) lists the full set of plots grouped by
workflow stage.

```{r runscatter}
#| fig.cap: !expr fig_cap("Run-scatter of PC concentrations by injection order after drift and batch correction.")
#| fig.alt: Run-scatter of PC feature concentrations across injection order
plot_runscatter(
  myexp,
  variable = "conc",
  include_feature_filter = "PC 4",
  include_istd = FALSE,
  cap_outliers = TRUE,
  log_scale = FALSE,
  output_pdf = FALSE,
  path = "./output/runscatter_PC408_beforecorr.pdf",
  cols_page = 3, rows_page = 2)
```

## 6. Export and share

Finally, we export results for downstream analysis and archiving: a detailed Excel report capturing
every table and QC metric, a flat CSV of just the QC-passing concentrations, and the serialized
object holding the complete processing state.

```{r save-report}
save_report_xlsx(myexp, path = tempfile(fileext = ".xlsx"))

save_dataset_csv(
  myexp,
  path = tempfile(fileext = ".csv"),
  variable = "conc",
  qc_types = "SPL",
  include_qualifier = FALSE,
  filter_data = TRUE)

saveRDS(myexp, file = tempfile(fileext = ".rds"), compress = TRUE)
```

The `.rds` file preserves the entire `MRMhubExperiment`, so a colleague can open it in R and run their
own plots and QC checks, reproducing the exact data state without re-running the pipeline.

## Next steps

- [Lipidomics data processing](tutorial-03-lipidomics-workflow.html): a more detailed lipidomics example
- [Drift and batch correction](tutorial-04-drift-correction.html): correction methods and diagnostics
- [Exploring QC: RunScatter and PCA](tutorial-05-run-scatter.html): QC visualisation and outlier screening
- [External calibration and QC](recipe-01-ext-calibration-qc.html): quantitation with calibration curves
