---
title: "Drift Correction (Smoothing)"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Drift Correction (Smoothing)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

MRMhub provides different functions for run-order drift correction, differing by the employed
smoothing algorithm. These functions can be categorized into functions that are either more suited
for QC or study sample-based drift correction.

The data must be provided via a MRMhubExperiment object, whereby raw data that was imported or
processed data can be corrected, such as `intensity` or `conc` values.

These drift correction functions have various options for customization, please refer the manual
page on [Drift and Batch Correction](manual-07-drift-batch-correction.html) for more details.

::: {.callout .callout-note}
**Time:** ~15 min &nbsp;|&nbsp; **Level:** Intermediate &nbsp;|&nbsp; **Prerequisites:** [Basic workflow](tutorial-02-basic-workflow.html)
:::

## Import data

In this tutorial, we import pre-calculated raw concentration values from a CSV file. This file must
contain a column with batch information (`batch_id`) if batch-wise correction should be applied, see
`import_data_csv_wide()` for more information.

```{r setup}
#| results: false  
#| message: false
library(mrmhub)

myexp <- mrmhub::MRMhubExperiment()

myexp <- import_data_csv_wide(
  myexp,
  path = "smooth-testdata.csv",
  variable_name = "conc",
  import_metadata = TRUE)
```

## QC-based smoothing

We apply a QC-based drift correction using cubic spline. See `correct_drift_cubicspline()` for
more information.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_cubicspline(
  myexp,
  batch_wise = FALSE,
  variable = "conc",
  ref_qc_types = "BQC",
  recalc_trend_after = TRUE)
```

Next, we inspect the data before and after the correction. We observe that the green trendline
follows the QC samples (in red). After the correction, the trend has been fully smoothed out,
resulting in a straight line.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

## Sample trend-based smoothing

Above, we observe that the QC samples do not fully represent the trends of the study samples. This
discrepancy can occur when the QC samples, which often based on pooled samples, differ in handling
or properties from the study samples.

We therefore apply a sample-based drift correction using gaussian kernel smoothing. See
`correct_drift_gaussiankernel()` for more information.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  kernel_size = 10,
  batch_wise = FALSE,
  ref_qc_types = "SPL",
  recalc_trend_after = TRUE)

```

Again, we inspect the data before and after the correction. We observe that the green trendline
follows the samples (in grey). After the correction, the trend has been fully smoothed out,
resulting in a straight line.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)

```

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)    
```
Now, the trends of the study samples appear to be fairly well corrected. However, in this example,
the differences compared to the previous QC-based smoothing are not very pronounced.

## Within-batch drift correction

In the examples before we applied drift correction across all batches. However, if batch-effects are
present that break the drifts, it is recommended to apply drift correction within each batch.

We explore a within-batch drift correction using a sample-based gaussian kernel smoothing from above
in this next example, by setting `batch_wise = TRUE`.

```{r}
#| results: false  
#| code-overflow: 'wrap'
mexp_drift <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  kernel_size = 10,
  batch_wise = TRUE,
  ref_qc_types = "SPL",
  recalc_trend_after = TRUE)
```

Now, each batch has its own trendline corresponding the trend in each batch of each feature.
Contrary to the previous correction across batches, we now observe clear differences in the trends
still present after the correction. This is an effect of the correction applied independently to each
batch, due to different sample sizes and present batch effects. Therefore, it is often necessary to
apply batch correction after batch-wise drift correction

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(mexp_drift, variable = "conc_before", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

We therefore apply a subsequent batch correction using median-centering, resulting in an alignment
of the batches.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
mexp_drift <- mrmhub::correct_batch_centering(
  mexp_drift, 
  ref_qc_types = "SPL", 
  variable = "conc",
  correct_scale = TRUE
)

plot_runscatter(mexp_drift, variable = "conc", qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 3, show_trend = TRUE)
```

## Method comparison

`MRMhub` provides four drift correction methods. Loess and cubic spline are typically used with
QC samples as the reference, gaussian kernel smoothing with study samples; a fourth method,
generalized additive model (GAM) smoothing via `correct_drift_gam()`, is also available. The method
comparison below summarises the typical use cases of the three most common — see the
[Drift and Batch Correction (reference)](manual-07-drift-batch-correction.html) for the full
parameter description of all four.

| Method | Function | Reference samples | Typical use |
|--------|----------|-------------------|-------------|
| Loess | `correct_drift_loess()` | QC samples | Frequent QC injections; smooth trends; robust to single-point outliers |
| Cubic Spline | `correct_drift_cubicspline()` | QC samples | Frequent QC injections; flexible curves; sensitive to outlier QC points |
| Gaussian Kernel | `correct_drift_gaussiankernel()` | Study samples | Sparse QCs; only suitable for large, well-randomised sample sets |

## Loess drift correction

Loess uses locally weighted polynomial regression and is less sensitive to individual outlier QC
points than cubic spline. The `span` parameter controls smoothness (higher = smoother).

```{r}
#| results: false
#| code-overflow: 'wrap'
mexp_drift_loess <- correct_drift_loess(
  myexp,
  variable = "conc",
  batch_wise = TRUE,
  ref_qc_types = "BQC",
  recalc_trend_after = TRUE)
```

Inspect the result with `plot_runscatter()` before exporting.

## Export corrected data

Next, we can either continue to work with the corrected data using `MRMhub` functions or export the
data.

```{r}
save_dataset_csv(
  mexp_drift, 
  path = "drift-batch-corrected-conc-data.csv", 
  variable = "conc", 
  filter_data = FALSE
  )
```

## Next Steps

- [RunScatter Plot](tutorial-05-run-scatter.html) — visualise run order effects
- [Batch Correction](tutorial-06-batch-correction.html) — correct inter-batch variability
- [Drift & Batch Correction (reference)](manual-07-drift-batch-correction.html) — full method documentation
