---
title: "Correcting isotopic interference"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4
description: >
  Annotate the mrm_pattern, derive the M+2 overlaps, inspect them, and subtract
  them — the step-by-step isotopic interference-correction workflow.
vignette: >
  %\VignetteIndexEntry{Correcting isotopic interference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
bibliography: references.bib
---

```{r setup, include = FALSE}
source("_common.R")
knitr::opts_chunk$set(eval = FALSE)
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-prereq">Prerequisites: <a href="tutorial-03-lipidomics-workflow.html">Full workflow</a></span></p>

MRMhub corrects isotopic (M+2) overlap between co-eluting same-class lipids with LICAR
[@gao2021LICARApplicationIsotopic]. It runs on the raw `feature_intensity` and, unlike drift and
batch correction, applies to every sample. This tutorial annotates the pattern, derives the overlaps,
inspects them, and subtracts them; the
[Isotopic interference correction](manual-12-interference-correction.html) manual covers the
concepts, the derivation levels, and the full `mrm_pattern` list.

## 1. Annotate the MRM pattern

Automatic derivation needs one hand-added column, `mrm_pattern`, on the `Features` metadata sheet; it
names the lipid class and the product-ion type (the manual lists every label and shows the column in
context). The valid labels for a product-ion type are returned by `licar_pattern_choices()`:

```{r}
library(mrmhub)
licar_pattern_choices("Head Group")
```

MRMhub validates the label on import. Load a processed object to continue:

```{r}
mexp <- readRDS("results/mexp_processed.rds")
```

## 2. Derive the interference relationships

[`calc_isotopic_interferences()`](../reference/calc_isotopic_interferences.html) discovers the M+2
overlaps and stores them in `annot_interferences`. Set `level = "MRM"` for class-based LC-MRM;
`"MS1"` is only for genuine full-scan data, not a fallback when a product *m/z* is missing (see the
manual).

```{r}
# Class-based LC-MRM: fragment-level M+2 (needs precursor + product m/z).
mexp <- calc_isotopic_interferences(mexp, level = "MRM")

# Genuine MS1 / full-scan: whole-molecule M+2 (needs precursor m/z only).
# mexp <- calc_isotopic_interferences(mexp, level = "MS1")
```

Derivation is deterministic; raise `min_contribution` to drop negligible pairs.

## 3. Inspect the derived relationships

Before subtracting, review the derived overlaps. Each row of `annot_interferences` is one overlap,
described by its `overlap_type` and `source` columns (defined in the manual):

```{r}
mexp@annot_interferences
```

[`summarize_interferences()`](../reference/summarize_interferences.html) rolls this up per feature:
how many are affected, split by source and overlap type, with the contribution-factor range.

```{r}
summarize_interferences(mexp)
```

## 4. Apply the correction

[`correct_isotopic_interferences()`](../reference/correct_isotopic_interferences.html) subtracts all
derived overlaps from the raw intensities in one pass (its reference page gives the formula and
zero-clamping).

```{r}
mexp <- correct_isotopic_interferences(mexp)
```

Apply it before ISTD normalisation, drift, and batch correction; running it later resets those steps
(with a warning). Uncorrected values are kept in `feature_intensity_orig`. Overlaps you annotate
yourself (in-source fragments, co-eluting isobars) are applied with
[`correct_custom_interferences()`](../reference/correct_custom_interferences.html).

## 5. Correct a single pair manually

For a one-off correction, or to check a factor before trusting the automatic derivation, use
[`correct_interference_manual()`](../reference/correct_interference_manual.html), where `variable` is
the `dataset` column:

```{r}
mexp <- correct_interference_manual(
  mexp,
  variable = "feature_intensity",
  feature = "PC 32:0",
  interfering_feature = "SM 36:1 M+3",
  interference_contribution = 0.0107,
  neg_to_na = FALSE,
  updated_feature_id = NA)
```

`updated_feature_id` renames the corrected feature so the raw and corrected channels can coexist.

## 6. Verify the correction

```{r, eval = TRUE, include = FALSE, message = FALSE, warning = FALSE}
# Build a corrected object for the QC plots from the bundled multi-class LICAR
# HILIC demo, mirroring sections 2-4 (import long data + mrm_pattern sheet,
# auto-derive M+2 edges, subtract). Needs enviPat 2.8 and rgoslin.
library(mrmhub)
demo_csv <- system.file("extdata", "licar_hilic_demo.csv", package = "mrmhub")
demo_features <- system.file(
  "extdata",
  "licar_hilic_demo_features.csv",
  package = "mrmhub"
)

mexp <- MRMhubExperiment() |>
  import_data_csv_long(path = demo_csv, import_metadata = TRUE, silent = TRUE) |>
  import_metadata_features(path = demo_features, ignore_warnings = TRUE) |>
  calc_isotopic_interferences(level = "MRM") |>
  correct_isotopic_interferences()
```

Compare a corrected feature's intensities before and after; in blanks (`SBLK`/`PBLK`) the residual
signal should approach zero, and a non-zero blank median points to an underestimated factor.

```{r}
d <- get_analyticaldata(mexp, annotated = TRUE) |>
  dplyr::filter(feature_id == "PC 32:0") |>
  dplyr::select(
    analysis_id, qc_type,
    intensity_before = feature_intensity_orig,
    intensity_after = feature_intensity) |>
  dplyr::mutate(
    pct_change =
      100 * (intensity_after - intensity_before) / intensity_before)

summary(d$pct_change)
```

[`plot_qc_interference_impact()`](../reference/plot_qc_interference_impact.html) shows how many
features were corrected at each magnitude (percent of signal removed):

```{r qc-impact}
#| eval: true
#| fig.cap: !expr fig_cap("How many features were corrected at each magnitude (percent of signal removed) in the study samples.")
#| fig.alt: Distribution of per-feature interference-correction magnitude in study samples
plot_qc_interference_impact(mexp, qc_types = "SPL")
```

[`plot_interference_correction()`](../reference/plot_interference_correction.html) plots each
feature's residual signal as a percent of its uncorrected value, split by QC type. `min_correction_pct`
restricts it to strongly affected features (here ≥ 40 % removed), `sort_by_effect` ranks by magnitude,
and `top_n` keeps the largest:

```{r per-feature}
#| eval: true
#| fig.cap: !expr fig_cap("Per-feature residual signal after correction, as a percent of the uncorrected value, for features with at least 40% removed, split by QC type.")
#| fig.alt: Per-feature residual signal after interference correction, split by QC type
plot_interference_correction(
  mexp,
  qc_types = "SPL",
  min_correction_pct = 40,
  sort_by_effect = "desc",
  point_size = 1.5, point_alpha = 0.8)
```

## Co-elution filtering (experimental) {#co-elution-filtering-experimental}

::: {.callout .callout-caution}
`check_coelution` is **experimental** and off by default. Verify the pairs it keeps and drops on your
own data before relying on it.
:::

The correction subtracts the full source area, so it is valid only where the source and target peaks
co-elute and/or are co-integrated. The experimental `check_coelution` gate keeps only the
*m/z*-matched edges that pass this test and drops chromatographically resolved pairs (see the
manual):

```{r}
mexp <- calc_isotopic_interferences(
  mexp, level = "MRM", check_coelution = TRUE)
```

It is off by default while the gate is validated, and reports how many pairs it drops.

## Next steps

- [Isotopic interference correction (manual)](manual-12-interference-correction.html): concepts, derivation levels, and the full label list
- [Drift and batch correction](tutorial-04-drift-correction.html): apply after interference correction
- [Lipidomics workflow](tutorial-03-lipidomics-workflow.html): the full processing pipeline
- [The MRMhubExperiment data object](manual-02-data-object.html#feature-variables): how `_orig` preserves the raw intensities

## References
