---
title: "Lipidomics data processing"
description: >
  Post-process a targeted lipidomics run from peak areas to a curated, quantified
  dataset: quality assessment, ISTD normalisation, drift and batch correction, and
  QC filtering.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Lipidomics data processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
source("_common.R")
set.seed(1041)
options(dplyr.print_max = 10)
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Advanced</span> <span class="page-prereq">Prerequisites: <a href="tutorial-01-prep-data.html">Preparing and importing data</a></span></p>

This tutorial follows a complete postprocessing and quality control workflow for
a targeted lipidomics analysis. Starting from integrated peak areas, we work
towards a curated table of lipid species concentrations ready for statistical
analysis. Along the way we assess the analytical and data quality of the run,
normalise and quantify against internal standards, correct for run-order drift
and between-batch effects, filter features on QC criteria, and export the result.
The dataset (a longitudinal plasma study (`sPerfect`) measured across six
batches) is realistic enough to show the decisions a real analysis demands.

## 1. Set up a project

A new analysis is easiest to manage inside an RStudio or Positron project
(see [Using RStudio Projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects)
or the [Positron User Guide](https://positron.posit.co/)), with a
predictable folder layout that keeps the raw data, the exported results,
and the processing notebook apart:

```
my_study/
├── data/           # raw data and metadata files
├── output/         # exported results
└── analysis.Rmd    # your processing notebook
```

An R Notebook (`.Rmd`) or
[Quarto](https://docs.posit.co/ide/user/ide/guide/documents/quarto-project.html)
(`.qmd`) document is a natural home for a workflow like this one: it
interleaves code with prose, keeping a record of every decision beside the
result it produced. With the project in place, load the package to begin:

```{r library}
library(mrmhub)
```

## 2. Importing analysis results

We begin by importing the MRMhub result file, which holds the areas of the
integrated peaks (features) for every processed raw data file. The same file also
carries the peak retention times and widths and metadata read from the mzML
files, such as acquisition time stamps and precursor/product m/z values; setting
`import_metadata = TRUE` brings those across as well. The imported data lands in
the `dataset_orig` slot and is copied to the working `dataset`. Printing the
object at any stage with `print(myexp)` reports its processing status, and in
RStudio you can expand `myexp` in the Environment pane.

```{r load-rawdata}
#| message: true
myexp <- MRMhubExperiment(title = "sPerfect")
myexp <- import_data_mrmhub(
  myexp,
  path = "./datasets/sPerfect_MRMhub.tsv",
  import_metadata = TRUE
)
```

## 3. A glimpse on the imported data

The data is stored in long format (one row per analysis–feature pair) so each
measurement carries its own area, retention time, and metadata side by side.
Print the working table below, or enter `View(myexp@dataset_orig)` in the console
to browse and filter it in the RStudio viewer. For routine access it is cleaner
to use `get_analyticaldata(myexp, annotated = TRUE)`, which returns the same data
already joined to its sample and feature annotations.

```{r view-rawdata}
print(myexp@dataset)
```

## 4. Analytical design and timeline

Before touching the signals, it is worth understanding how the run was
structured. The plot below lays out the batch boundaries, the positions of the
quality control (QC) samples, and the date, duration, and run time of the
analysis: context that informs every later choice about drift, batches, and
which QC type to trust. Setting `show_timestamp = TRUE` overlays acquisition
times and reveals any long interruptions within or between batches.

```{r runsequence}
#| fig.cap: !expr fig_cap("Run sequence of the analysis (batch structure, QC-sample positions, and acquisition timeline).")
#| fig.alt: RunSequence plot of the analysis showing batches and QC positions
plot_runsequence(
  myexp,
  qc_types = NA,
  show_batches = TRUE,
  batch_zebra_stripe = TRUE,
  batch_fill_color = "#fffbdb",
  segment_linewidth = 0.5,
  show_timestamp = FALSE
)
```

## 5. Overview of chromatographic separation

A first look at the chromatography confirms that species elute where expected.
The plot shows the retention time distribution of all detected lipid species
across the study samples, arranged by class.

```{r rt-overview}
#| fig.cap: !expr fig_cap("Retention-time distribution of all detected lipid species across the study samples.")
#| fig.alt: Retention-time distribution of all detected lipid species
plot_abundanceprofile(
  myexp,
  variable = "rt",
  qc_types = "SPL",
  log_scale = FALSE,
  density_strip = TRUE,
  show_sum = FALSE,
  x_label = NA,
  feature_map = "lipidomics"
)
```

## 6. Peak picking QC

Within a lipid class, retention time increases smoothly with chain length and
decreases with the number of double bonds, so plotting RT against total carbon
count (coloured by double-bond number) turns peak-picking errors into visible
outliers. Species whose residual from the class trend exceeds
`outlier_residual_min` are labelled, flagging likely misannotations or
mis-integrated peaks. The check is only as strong as the coverage behind it: here
just a few species per class were measured, so the trend is poorly constrained
for some classes and double-bond counts.

```{r peak-picking-qc}
#| fig-height: 6.5
#| fig.cap: !expr fig_cap("Retention time versus total carbon number per lipid class; labelled points flag possible misannotations.")
#| fig.alt: Retention time versus carbon number per lipid class, outliers labelled
plot_rt_vs_chain(
  myexp,
  qc_types = "SPL",
  x_var = "total_c",
  outlier_residual_min = 0.3,
  font_base_size = 8,
  point_size = 1
)
```

## 7. Signal trends of internal standards

The internal standards (ISTDs) are our clearest window on instrument stability.
The same ISTD amount was spiked into every sample except the solvent blanks
(`SBLK`), so in the absence of technical variation each ISTD should read a
constant intensity across all samples and sample types. Plotting them in
injection order across the six batches makes any drift, batch offset, or blockage
immediately visible. Setting `output_pdf = TRUE` saves the plots to the `output`
subfolder for closer inspection.

```{r runscatter-istd}
#| message: true
#| fig.cap: !expr fig_cap("Internal-standard intensities across all six batches; the spiked amount is constant, so a flat trend is expected.")
#| fig.alt: RunScatter of internal-standard intensities across batches
plot_runscatter(
  myexp,
  variable = "intensity",
  qc_types = c("BQC", "TQC", "SPL", "PBLK", "SBLK"),
  include_feature_filter = "ISTD",
  exclude_feature_filter = "Hex|282",
  cap_outliers = TRUE,
  log_scale = FALSE,
  show_batches = TRUE,
  font_base_size = 5,
  output_pdf = FALSE,
  cols_page = 4, rows_page = 3
)
```

## 8. Adding detailed metadata

Everything from here on (sample roles, ISTD assignments, spike amounts,
concentrations) comes from the analysis metadata. The MRMhub Excel template
collects, organises, and pre-validates it in one place; open the `.xlsx` file in
the `datasets` folder to see its structure. On import the metadata is validated:
a hard error (a duplicate or missing ID) stops the import with a summary of the
problems, while softer issues are reported as warnings you can work through, or
bypass with `ignore_warnings = TRUE`. We then fix the processing order to the
injection sequence with `set_analysis_order()`, and tell MRMhub which of the
imported signal variables to treat as the working feature intensity
(`set_intensity_var("area")`), the value all subsequent calculations start from.

```{r import-metadata}
#| message: true
myexp <- import_metadata_msorganiser(
  myexp,
  path = "datasets/sPerfect_Metadata.xlsx",
  ignore_warnings = TRUE
)
myexp <- set_analysis_order(myexp, order_by = "timestamp")
myexp <- set_intensity_var(myexp, variable_name = "area")
```

## 9. Overall trends and possible outliers

To judge technical trends that affect most analytes at once, the RLA (Relative
Log Abundance) plot is the tool of choice (De Livera et al., Analytical
Chemistry, 2015). Each feature is expressed relative to its across- or
within-batch median and summarised as one boxplot per sample; a well-behaved
sample sits tight around zero, so boxes that shift or spread flag pipetting
errors, sample spillage, injection-volume changes, or drops in instrument
sensitivity. In this run one sample stands out from its batch as a candidate
technical outlier.

```{r rla-plot}
#| fig.cap: !expr fig_cap("Within-batch relative log abundance per sample; deviating boxes flag pipetting, injection, or sensitivity issues.")
#| fig.alt: RLA boxplot per sample within batches
plot_rla_boxplot(
  myexp,
  variable = "intensity",
  rla_type_batch = "within",
  qc_types = c("BQC", "SPL", "RQC", "TQC", "PBLK"),
  filter_data = FALSE,
  show_timestamp = FALSE,
  outlier_exclude = FALSE,
  x_gridlines = FALSE,
  batch_zebra_stripe = FALSE,
  linewidth = 0.1
)
```

## 10. PCA of all QC types

Principal component analysis condenses all features into a few axes and gives a
complementary overview of how the study and QC samples relate, useful for
spotting batch effects, technical outliers, and systematic differences between
sample types. Here it offers a second, independent view of the candidate outlier
flagged by the RLA plot.

```{r pca-before}
#| fig.cap: !expr fig_cap("PCA of study and QC samples before outlier removal.")
#| fig.alt: PCA of study and QC samples before outlier removal
plot_pca(
  myexp,
  variable = "feature_intensity",
  qc_types = c("SPL", "BQC", "TQC"),
  filter_data = FALSE,
  pca_dims = c(1, 2),
  labels_threshold_mad = 3,
  log_transform = TRUE,
  include_istd = FALSE,
  point_size = 2, point_alpha = 0.7, ellipse_alpha = 0.3, font_base_size = 8
)
```

## 11. Excluding technical outliers

With the outlier corroborated by both plots, we remove it from all downstream
processing using `exclude_analyses()`. The function does not delete any data: it
sets the sample's `valid_analysis` flag to `FALSE`, so it is simply skipped in
later steps and can be reinstated at any time. Replotting the PCA lets us confirm
the effect of the exclusion on the sample structure.

```{r outlier-removal}
#| fig.cap: !expr fig_cap("PCA after excluding the flagged technical outlier.")
#| fig.alt: PCA of study and QC samples after excluding the outlier
myexp <- exclude_analyses(
  myexp,
  analyses = "Longit_batch6_51",
  clear_existing = TRUE
)

plot_pca(
  myexp,
  variable = "intensity",
  qc_types = c("SPL", "BQC", "TQC"),
  filter_data = FALSE,
  pca_dims = c(1, 2),
  labels_threshold_mad = 3,
  log_transform = TRUE,
  include_istd = FALSE,
  shared_labeltext_hide = NA,
  point_size = 2, point_alpha = 0.7, ellipse_alpha = 0.3, font_base_size = 8
)
```

## 12. Response curves

A linear response is a prerequisite for comparing concentrations between samples.
Plasma lipid abundances span a wide dynamic range, and because each
class-specific ISTD is spiked at a single concentration, the response may be
linear near the ISTD level yet flatten for far more or less abundant species.
Verifying this is a worthwhile part of the analytical quality assessment, so we
measured injection-volume series at the start and end of the run as a dedicated
QC. We first drop the very lowest-abundance features, then plot the curves for a
subset of PC species; `include_feature_filter` accepts regular expressions, so
any class or chain-length pattern can be selected.

```{r responsecurves}
#| fig.cap: !expr fig_cap("Response curves for selected PC species over the injection-volume series.")
#| fig.alt: Response curves for selected PC species
myexp <- filter_features_qc(
  myexp,
  include_qualifier = FALSE,
  include_istd = TRUE,
  min.intensity.median.spl = 200
)
plot_responsecurves(
  myexp,
  variable = "intensity",
  filter_data = TRUE,
  include_feature_filter = "^PC 3[0-5]",
  output_pdf = FALSE,
  cols_page = 5, rows_page = 4
)
```

## 13. Isotope interference correction

Some peaks of interest are co-integrated with the isotopic peaks of other lipid
species, for example the M+2 isotope of a species two mass units lighter falling
in the same transition. `correct_custom_interferences()` subtracts these declared
contributions from the raw feature intensities, reading which feature interferes
with which, and by how much, from the metadata; the original values are preserved
in `feature_intensity_before`. The interference factors here were derived with
LICAR (<https://github.com/SLINGhub/LICAR>). Inspect the "Features (Analytes)"
sheet of the metadata file to see which species are affected.

```{r isotope-correction}
#| fig.cap: !expr fig_cap("Selected species before and after isotope-interference correction.")
#| fig.alt: QC plot of species before and after interference correction
myexp <- correct_custom_interferences(myexp)
plot_interference_correction(
  myexp,
  qc_types = c("BQC", "SPL", "TQC", "LTR")
)
```

## 14. Normalization and quantification based on ISTDs

Quantification proceeds in two steps. `normalize_by_istd()` divides each lipid's
peak area by the area of its assigned class-specific internal standard *within
the same injection*, writing the ratio to `feature_norm_intensity`; this cancels
most of the injection-to-injection and matrix variation shared by an analyte and
its ISTD. `quantify_by_istd()` then converts those ratios into absolute
concentrations from three quantities held in the metadata (the ISTD
concentration in the spike solution, the spiked volume, and the sample amount)
and stores the result in `feature_conc`. Because both new variables sit beside
the raw areas, the RLA and PCA plots above can be rerun on the normalised data
with `variable = "norm_intensity"` or `variable = "conc"`.

```{r normalization}
#| message: true
myexp <- normalize_by_istd(myexp)
myexp <- quantify_by_istd(myexp)
```

## 15. Effects of class-wide ISTD normalization

Class-specific ISTDs are common practice in lipidomics, but they are rarely
authentic standards for every species they normalise. A non-authentic ISTD may
elute at a different time, experience different matrix effects, and fragment
differently, so its response can diverge from that of the analyte. Normalising
against such an ISTD can then *add* variability rather than remove it. Because a
good ISTD should reduce the coefficient of variation (CV), comparing the CV of QC
and study samples before and after normalisation is a direct way to catch a
misbehaving standard: a class whose CV rises after normalisation is a warning
sign worth investigating before trusting its concentrations.

```{r norm-effects}
#| fig.cap: !expr fig_cap("Change in CV per feature after ISTD normalisation, faceted by lipid class.")
#| fig.alt: Normalization QC plot of CV change per feature by class
myexp <- filter_features_qc(
  myexp,
  include_qualifier = FALSE,
  include_istd = TRUE,
  min.intensity.median.spl = 1000
)
plot_normalization_qc(
  myexp,
  before_norm_var = "intensity",
  after_norm_var = "norm_intensity",
  plot_type = "diff",
  qc_types = c("TQC", "BQC", "SPL"),
  facet_by_class = TRUE,
  point_size = 2,
  y_lim = c(-5, 15)
)
```

## 16. Drift correction

Even after ISTD normalisation, concentrations can drift gradually within a batch
as the instrument response changes. We correct this with a Gaussian-kernel
smoother fitted through the study samples in each batch (`batch_wise = TRUE`),
which suits large, well-randomised sample sets like this longitudinal study;
`kernel_size` sets the width of the smoothing window. The corrected values
replace `conc`, the pre-correction values are kept in `conc_before`, and the
fitted trend in `conc_before_fit`. The summary the function prints is not a formal
diagnostic of the fit but a quick check (reported as a change in median CV)
that the correction improved precision without introducing artefacts. A scale
correction along the fit is available via `scale_smooth = TRUE`.

```{r drift-corr}
#| message: true
myexp <- correct_drift_gaussiankernel(
  myexp,
  variable = "conc",
  ref_qc_types = "SPL",
  ignore_istd = TRUE,
  batch_wise = TRUE,
  replace_previous = TRUE,
  recalc_trend_after = TRUE,
  kernel_size = 10,
  outlier_filter = FALSE,
  outlier_ksd = 5,
  location_smooth = TRUE,
  scale_smooth = FALSE,
  show_progress = FALSE
)
```

To see the effect, we plot one example species (PC 40:8) before and after
correction. Since we reuse the same plot several times, we wrap it in a small
function with the recurring parameters preset; `variable` selects which stage to
show (`conc_before` or `conc`) and `feature` which species.

```{r trendplot-fun}
my_trend_plot <- function(variable, feature) {
  plot_runscatter(
    myexp,
    variable = variable,
    qc_types = c("BQC", "TQC", "SPL"),
    include_feature_filter = feature,
    exclude_feature_filter = "ISTD",
    cap_outliers = TRUE,
    log_scale = FALSE,
    show_trend = TRUE,
    output_pdf = FALSE,
    cols_page = 1, rows_page = 1
  )
}
```

```{r plot-trend-before}
#| fig.cap: !expr fig_cap("PC 40:8 raw concentrations before drift correction, with the fitted within-batch trend.")
#| fig.alt: RunScatter of PC 40:8 concentrations before drift correction
my_trend_plot("conc_before", "PC 40:8")
```

```{r plot-trend-after}
#| fig.cap: !expr fig_cap("PC 40:8 concentrations after within-batch drift correction.")
#| fig.alt: RunScatter of PC 40:8 concentrations after drift correction
my_trend_plot("conc", "PC 40:8")
```

## 17. Batch-effect correction

Drift correction flattens the trend inside each batch but leaves the batches at
different levels, since each is fitted independently. `correct_batch_centering()`
aligns them by shifting every batch to a common median, computed from the study
samples (`ref_qc_types = "SPL"`); `correct_scale = TRUE` additionally equalises
the between-batch spread. As with drift correction, the aligned values replace
`conc` and the previous ones are retained in `conc_before`. We replot the example
species to confirm the batches now line up.

```{r batch-effect-corr}
#| fig.cap: !expr fig_cap("PC 40:8 after drift correction followed by batch centering; batch trends are aligned.")
#| fig.alt: RunScatter of PC 40:8 after batch centering
myexp <- correct_batch_centering(
  myexp,
  variable = "conc",
  ref_qc_types = "SPL",
  replace_previous = TRUE,
  correct_location = TRUE,
  correct_scale = TRUE,
  log_transform_internal = TRUE
)

my_trend_plot("conc", "PC 40:8")
```

## 18. Saving runscatter plots of all features as PDF

For a full record it helps to save runscatter plots for every species, or a
chosen subset, to a multi-page PDF in the `output` subfolder. Blanks are usually
worth excluding, as they can show erratic concentrations when feature and ISTD
signals sit close together or below the limit of detection. The
`include_feature_filter` and `exclude_feature_filter` arguments select analytes
and both accept regular expressions, which a language model can help you
construct for more intricate patterns.

```{r saving-runscatter}
#| eval: false
plot_runscatter(
  myexp,
  variable = "conc",
  qc_types = c("BQC", "TQC", "SPL"),
  include_feature_filter = NA,
  exclude_feature_filter = "ISTD",
  cap_outliers = TRUE,
  log_scale = FALSE,
  show_trend = TRUE,
  output_pdf = TRUE,
  path = "./output/runscatter_after-drift-batch-correction.pdf",
  page_width = 297, page_height = 210,
  cols_page = 2, rows_page = 2,
  show_progress = TRUE
)
```

`page_width` and `page_height` set the page size; without them an A4 page is used, oriented by
`page_orientation`.

::: {.callout .callout-tip}
The single figures made earlier in this workflow are saved with `save_plot()`, which writes any plot
at a defined size and resolution and can produce several formats in one call:

```{r}
#| eval: false
plot_pca(myexp, variable = "conc", qc_types = c("BQC", "SPL")) |>
  save_plot(path = "./output/pca", format = c("pdf", "png"),
            width = 180, height = 120)
```

The plot is returned visibly, so it still appears in the notebook while being written to file.

See [Saving plots](manual-08-visualization.html#saving-plots) for the available formats and when to
prefer a vector or a raster one.
:::

## 19. QC-based feature filtering

The final curation step removes features that fail defined QC criteria.
`filter_features_qc()` combines several: the quality of the response curves
(minimum r², slope, and maximum y-intercept), the signal-to-blank ratio against
process blanks, the absolute signal level as a proxy for the limit of detection,
and the precision in the batch QCs (`max.cv.conc.bqc`). The criteria are applied
*hierarchically* (a feature counts as failing CV only once it has passed the S/B
and signal filters) and `features.to.keep` retains named species regardless. The
filter can be run repeatedly, overwriting or, with `clear_existing = FALSE`,
amending the previous set.

```{r qc-filter}
#| message: true
myexp <- filter_features_qc(
  myexp,
  clear_existing = TRUE,
  use_batch_medians = TRUE,
  include_qualifier = FALSE,
  include_istd = FALSE,
  response.curves.selection = c(1, 2),
  response.curves.summary = "mean",
  min.rsquare.response = 0.8,
  min.slope.response = 0.75,
  max.yintercept.response = 0.5,
  min.signalblank.median.spl.pblk = 10,
  min.intensity.median.spl = 100,
  max.cv.conc.bqc = 25,
  features.to.keep = c(
    "CE 20:4", "CE 22:5", "CE 22:6", "CE 16:0", "CE 18:0"
  )
)
```

## 20. Summary of the QC filtering

The plot below summarises the filtering per lipid class. Green segments count the
species that passed every criterion; the remaining segments count those that
failed each one. Because the criteria are hierarchical, a species is attributed
to the *first* filter it fails (for example `CV`), having already cleared the
lower ones (`S/B`, `LOD`), so the counts partition the features rather than
double-counting them.

```{r qc-summary-byclass}
#| fig.cap: !expr fig_cap("Feature QC filtering outcome per lipid class; green marks species passing all criteria.")
#| fig.alt: Feature QC filter summary by lipid class
plot_qc_summary_byclass(myexp)
```

The next plot gives the same picture across all features, with the total retained,
and adds a Venn diagram showing how many features each individual criterion
excluded, and where those exclusions overlap.

```{r qc-summary-overall}
#| fig.cap: !expr fig_cap("Overall feature filtering summary with a Venn diagram of exclusion criteria.")
#| fig.alt: Overall feature filter summary with Venn diagram
plot_qc_summary_overall(myexp)
```

## 21. Lipidome profile

As a final sanity check, we plot the concentration profile of the filtered
dataset. Comparing these values (the most abundant species, the summed
concentration per class, or the ratios between classes) against in-house
reference ranges or the literature is the quickest way to confirm that
quantification landed in a plausible range and that no gross errors slipped
through.

```{r lipidprofile}
#| fig.cap: !expr fig_cap("Concentration profile of the filtered study-sample lipidome.")
#| fig.alt: Concentration profile of the filtered lipidome
plot_abundanceprofile(
  myexp,
  variable = "conc",
  qc_types = "SPL",
  filter_data = TRUE,
  log_scale = TRUE,
  x_label = NA,
  feature_map = "lipidomics"
)
```

## 22. Saving a report with data, metadata and processing details

The full post-processing can be written to a formatted Excel workbook, with
separate sheets for the raw and processed datasets, the associated metadata, the
per-feature QC metrics, and a record of the processing steps applied: a
self-contained account of how the numbers were produced.

```{r save-report}
#| message: true
save_report_xlsx(myexp, path = tempfile(fileext = ".xlsx"))
```

For downstream statistics it is often easier to export a single flat, wide CSV of
a chosen data subset. This is the format used to share the data for the
statistical analysis in the next part of the workshop.

```{r save-csv}
save_dataset_csv(
  myexp,
  path = tempfile(fileext = ".csv"),
  variable = "conc",
  qc_types = "SPL",
  include_qualifier = FALSE,
  filter_data = TRUE
)
```

## 23. Sharing the MRMhubExperiment dataset

Finally, the whole `myexp` object can be serialized to an `RDS` file and shared.
`RDS` files open in any R session (even without the mrmhub package installed)
and the saved `MRMhubExperiment` can be reloaded for further processing,
replotting, or inspection with mrmhub.

```{r save-rds}
path <- tempfile(fileext = ".rds")
saveRDS(myexp, file = path, compress = TRUE)
my_saved_exp <- readRDS(file = path)
print(my_saved_exp)
```

## Next steps

- [Drift and batch correction](tutorial-04-drift-correction.html): correction methods and diagnostics in depth
- [RunScatter and PCA QC exploration](tutorial-05-run-scatter.html): QC visualisation in depth
- [Interference correction](tutorial-11-interference-correction.html): correcting isotopic and isobaric overlap
- [External calibration & QC](tutorial-06-external-calibration.html): quantify with external calibration curves
