---
title: "Introduction to R, Quarto, and MRMhub"
description: >
  Install R and an IDE, create a Quarto project, and run a complete MRMhub
  analysis in it — written for readers new to R and Quarto.
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Introduction to R, Quarto, and MRMhub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Beginner</span></p>

A MRMhub analysis is written as a document that mixes prose with the R code that produces the
results, so the report and the analysis are one file. This tutorial assumes no prior experience with
R or Quarto. It starts from an empty computer, installs the software, sets up a project, and then
runs a complete analysis of the bundled demonstration data — importing, normalizing, plotting, and
exporting — using MRMhub functions alone, without any custom R code. After working through it you
can open a fresh Quarto document and reproduce the same workflow on your own data.

## 1. Install R and an editor

Two pieces of software are needed. **R** is the language the analysis runs in; an **integrated
development environment** (IDE) is the editor you write and run it in.

- Install R from [CRAN](https://cran.r-project.org/) (version 4.1 or newer). Pick the download for
  your operating system and accept the defaults.
- Install an IDE. [RStudio](https://posit.co/download/rstudio-desktop/) is the most common choice and
  the one this tutorial follows. If you already work in VS Code, [Positron](https://positron.posit.co/)
  is an RStudio-like alternative built on the same editor.

With R and RStudio installed, open RStudio and install MRMhub from the console (the bottom-left
pane). The first line installs `pak`, a package installer; the second uses it to fetch MRMhub and its
dependencies:

```{r install, eval = FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/MRMhub")
```

If the installation reports an error, the [Installation guide](manual-00-installation.html) lists the
common causes (a missing compiler on Windows or macOS, a firewall blocking the download) and their
fixes.

## 2. Create a Quarto project

A **project** keeps everything for one analysis — data, code, and results — in a single folder that
the IDE treats as a unit. In RStudio, choose *File → New Project → New Directory → Quarto Project*,
give it a name, and create it. Quarto's [project documentation](https://quarto.org/docs/tools/rstudio/)
covers the dialog in detail.

Inside the new project, add two folders to keep raw inputs separate from generated results — the
convention MRMhub's own workflows follow:

```
my_study/
├── data/            # raw INTEGRATOR output and metadata files
├── output/          # exported CSVs, PDFs, and reports
└── analysis.qmd     # this document
```

A Quarto document (`.qmd`) is the home for the workflow: it interleaves explanatory text with code,
documenting each decision alongside the result it produces.

## 3. Add text and code

A `.qmd` file is plain text with two kinds of content. **Prose** is written in Markdown — plain
paragraphs, with `#` for headings, `**bold**`, and `-` for lists. **Code** lives in *chunks*: R code
fenced between ```` ```{r} ```` and ```` ``` ````. In RStudio, *Insert → Code Chunk* (or Ctrl/Cmd +
Alt + I) adds an empty one.

```` markdown
Some explanatory text describing the next step.

```{r}
1 + 1
```
````

RStudio offers two ways to edit a `.qmd`: the **Source** editor shows the raw Markdown, while the
**Visual** editor (the *Visual* toggle, top-left of the document) shows a formatted, word-processor-like
view. Both edit the same file. Quarto's [authoring guide](https://quarto.org/docs/get-started/hello/rstudio.html)
introduces both.

## 4. Preview and run code

There are two distinct actions. **Running a chunk** executes its code immediately and shows the
result inline: click the green ▶ arrow at the chunk's top-right, or press Ctrl/Cmd + Shift + Enter
with the cursor inside it. **Rendering** turns the whole document into a finished HTML (or PDF)
report: click **Render**, and RStudio runs every chunk in order and assembles the output. The
**Preview** that opens updates each time you render, so the report stays in step with the code.

While developing an analysis, run chunks one at a time to check each step; render at the end to
produce the shareable report.

## 5. The setup chunk

The first chunk of an analysis loads the packages and sets defaults that apply throughout. Loading
MRMhub makes its functions available; `mrmhub_enable_cli_color()` colours the console messages in the
rendered report; and `mrmhub_set_plot_defaults()` sets a house font size and resolution for every
figure, so individual plots need not repeat them.

```{r first-chunk}
library(mrmhub)
mrmhub_enable_cli_color()
mrmhub_set_plot_defaults(font_base_size = 8)
```

Figure dimensions are controlled by Quarto itself, through the document's `fig-width` and
`fig-height` options; [Quarto workflows](manual-11-quarto-workflows.html) explains how those interact
with MRMhub's plot sizing.

## 6. Load the data

Every analysis begins with a `MRMhubExperiment` — the container that holds the data and every result
computed from it (see [The MRMhubExperiment object](manual-02-data-object.html)). We create an empty
one and import the demonstration dataset bundled with the package. Using `system.file()` locates that
file wherever the package is installed, so the code runs unchanged on any machine; in your own
project you would instead give the path to a file in `data/`.

```{r load-data}
#| message: true
demo_file <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")

mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(
  mexp,
  path = demo_file,
  import_metadata = TRUE)
```

Setting `import_metadata = TRUE` reads the sample and internal-standard annotations embedded in the
INTEGRATOR file at the same time. MRMhub reads several other formats; [Importing analytical
data](manual-04-data-import.html) has a decision flowchart and the full importer reference.

## 7. Process the data

The core processing step normalizes each feature to its assigned internal standard, which corrects
for extraction and injection variation. The result is stored in the `feature_norm_intensity`
variable, leaving the raw intensities untouched.

```{r process}
#| message: true
mexp <- normalize_by_istd(mexp)
```

This is the shortest useful pipeline. A real study usually adds drift and batch correction and QC
filtering in a set order; [A basic MRMhub workflow](tutorial-02-basic-workflow.html) and [Drift and
batch correction](tutorial-04-drift-correction.html) build on the object from here.

## 8. Plot signals across the run

A run-scatter plot shows each feature's signal against injection order — a quick check for trends or
outliers. We plot the normalized intensities for two lipid classes, selected by a regular expression
on their names.

```{r runscatter}
#| fig.cap: !expr fig_cap("Normalized intensities of Cer and PC features across injection order.")
#| fig.alt: Run-scatter of normalized intensities for Cer and PC features by injection order
plot_runscatter(
  mexp,
  variable = "norm_intensity",
  include_feature_filter = "^(Cer|PC)",
  rows_page = 2, cols_page = 3)
```

To save the plot to a file instead of showing it, `plot_runscatter()` can write a multi-page PDF —
one page per grid of features — directly to `output/`:

```{r runscatter-pdf, eval = FALSE}
plot_runscatter(
  mexp,
  variable = "norm_intensity",
  include_feature_filter = "^(Cer|PC)",
  rows_page = 2, cols_page = 3,
  output_pdf = TRUE,
  path = "output/runscatter.pdf")
```

## 9. Run a PCA, inline and as a PDF

A principal-component analysis (PCA) summarises all features at once, placing each sample in a plane
so that similar samples fall close together — useful for spotting QC clustering or outliers.

```{r pca}
#| fig.cap: !expr fig_cap("PCA of the normalized intensities, coloured by QC type.")
#| fig.alt: PCA score plot of samples coloured by QC type
plot_pca(mexp, variable = "norm_intensity")
```

To keep a copy of the figure, pipe it into `save_plot()`. Because `save_plot()` returns the plot
visibly, one line both writes the PDF and still renders the figure in the report; `width` and
`height` are in millimetres:

```{r pca-save, eval = FALSE}
plot_pca(mexp, variable = "norm_intensity") |>
  save_plot("output/pca.pdf", width = 180, height = 120)
```

Giving `save_plot()` a path with no extension and `format = c("pdf", "png")` writes both formats at
once; [Visualisation functions](manual-08-visualization.html) lists the plots available at each stage.

## 10. Export results

Finally, export the results for downstream use. A single-variable CSV gives a plain table of
normalized intensities, and `save_report_xlsx()` writes a multi-sheet Excel workbook holding every
data table and QC metric in one file.

```{r export, eval = FALSE}
save_dataset_csv(
  mexp,
  path = "output/norm_intensities.csv",
  variable = "norm_intensity")

save_report_xlsx(
  mexp,
  path = "output/report.xlsx",
  filtered_variable = "norm_intensity")
```

With these chunks written, click **Render** once more: RStudio runs the whole analysis top to bottom
and produces a self-contained report, with the CSV, PDFs, and Excel workbook waiting in `output/`.

## Next steps

- [A basic MRMhub workflow](tutorial-02-basic-workflow.html): the full pipeline with quantification and QC filtering
- [Drift and batch correction](tutorial-04-drift-correction.html): correcting run-order and batch effects
- [Exploring QC: RunScatter and PCA](tutorial-05-run-scatter.html): QC visualisation in depth
- [Quarto workflows](manual-11-quarto-workflows.html): rendering, figure sizing, and report options
