---
title: "Calibration by a reference sample"
description: >
  Re-calibrate or normalise concentrations against a reference sample
  (e.g. NIST SRM1950), apply it batch-wise, and check the reference bias.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Calibration by a reference sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-prereq">Prerequisites: <a href="tutorial-03-lipidomics-workflow.html">Full workflow</a>, <a href="tutorial-06-external-calibration.html">External calibration</a></span></p>

Feature abundances in samples can also be calibrated to corresponding abundances in a specified
reference sample. MRMhub supports absolute (re-)calibration and normalization (relative calibration).

Absolute calibration of feature abundances is based on known metabolite concentrations in a
reference sample (e.g. NIST SRM1950 plasma). Normalization (relative calibration) is based on
calculating the abundance ratios of features in samples and a reference sample.

Both absolute and relative calibration are demonstrated below using NIST SRM1950 plasma samples
that were measured as part of the same analysis.

## 1. Import data and metadata

```{r import}
#| results: markup
#| message: false
library(mrmhub)
library(dplyr)

# Get example data paths
dat_file <- system.file("extdata", "S1P_MHQuant.csv", package = "mrmhub")
meta_file <- system.file("extdata", "S1P_metadata_tables.xlsx", package = "mrmhub")

# Load data and metadata
mexp <- MRMhubExperiment()
mexp <- import_data_masshunter(mexp, dat_file, import_metadata = FALSE)
mexp <- import_metadata_analyses(mexp, path = meta_file, sheet = "Analyses")
mexp <- import_metadata_features(mexp, path = meta_file, sheet = "Features")
mexp <- import_metadata_istds(mexp, path = meta_file, sheet = "ISTDs")
```

## 2. Load known concentrations of the reference sample

A table of known analyte concentrations for the NIST SRM1950 reference sample is added to the
`MRMhubExperiment` object. The S1P concentrations in this table are for illustration
only; the actual absolute S1P concentrations in NIST SRM1950 may differ significantly.

```{r load-qcconc}
#| results: false
#| message: true
mexp <- import_metadata_qcconcentrations(mexp, path = meta_file, sheet = "QCconcentrations")
```

## 3. Process the data

The analysis was performed using HILIC chromatography; the isotope interferences from
S1P 18:2;O2 M+2 and S1P 18:1;O2 M+2 must therefore be corrected. Initial quantification is then
done using the spiked-in ISTD concentration.

```{r process}
#| results: false
#| message: true
# Isotope-interference correction
mexp <- correct_custom_interferences(mexp)

# Quantify against the internal standards
mexp <- normalize_by_istd(mexp)
mexp <- quantify_by_istd(mexp)
```

## 4. Absolute calibration

Absolute re-calibration is performed with `calibrate_by_reference()`. The reference sample is set
via `reference_sample_id`. Where multiple analyses of the same reference sample are present in the
dataset, either their mean or median is used (defined via `summarize_fun`).

The calibrated concentration is calculated as:

$$
c_\text{calibrated}^\text{Analyte} = \frac{c_\text{sample}^\text{Analyte}}{c_\text{ref}^\text{Analyte}} \times c_\text{known}^\text{Analyte}
$$

```{r calibrate-absolute}
#| results: false
#| message: true
mexp_res <- calibrate_by_reference(
  data = mexp,
  variable = "conc",
  reference_sample_id = "SRM1950",
  absolute_calibration = TRUE,
  batch_wise = FALSE,
  summarize_fun = "mean",
  undefined_conc_action = "na")
```

The re-calibrated concentrations are written to the variable `conc`, overwriting any previously
calculated concentrations; the values from before calibration remain available as `conc_beforecal`.
They can be exported as usual and also appear as concentrations in the MRMhub XLSX report.

```{r export-absolute}
#| results: false
#| message: true
# Export absolute-calibrated concentrations
save_dataset_csv(mexp_res, tempfile(fileext = ".csv"), variable = "conc")

# Export non-calibrated concentrations
save_dataset_csv(mexp_res, tempfile(fileext = ".csv"), variable = "conc_beforecal")

# Create XLSX report with calibrated concentrations as filtered dataset
save_report_xlsx(mexp_res, tempfile(fileext = ".xlsx"), filtered_variable = "conc")
```

## 5. Normalization (relative calibration)

Normalization against a reference sample is performed with the same function, setting
`absolute_calibration = FALSE`. Where multiple analyses of the reference sample are
present, either their mean or median is used (defined via `summarize_fun`).

```{r normalize}
#| results: false
#| message: true
mexp_res <- calibrate_by_reference(
  data = mexp,
  variable = "conc",
  reference_sample_id = "SRM1950",
  absolute_calibration = FALSE,
  summarize_fun = "mean")
```

Unlike absolute calibration, the normalization results are stored as **ratios** in a new variable,
`[VARIABLE]_normalized`, where `[VARIABLE]` is the input variable, e.g. `conc_normalized` or
`intensity_normalized`.

The normalized values can be exported as `[VARIABLE]_normalized` with `save_dataset_csv()`. In the
MRMhub XLSX report from `save_report_xlsx()`, the unfiltered dataset with normalized concentrations
is included by default; to include them as the filtered dataset, set
`filtered_variable = "[VARIABLE]_normalized"`.

```{r export-normalized}
#| results: false
#| message: true
# Export NIST SRM1950-normalized concentrations
save_dataset_csv(mexp_res, "norm.csv", variable = "conc_normalized")

# Create XLSX report with normalized concentrations as filtered dataset
save_report_xlsx(
  mexp_res,
  path = tempfile(fileext = ".xlsx"),
  filtered_variable = "conc_normalized")
```

## 6. Batch-wise calibration

Calibration can also be applied batch-wise, in which case each batch is calibrated separately using
the reference sample in the same batch. Set `batch_wise = TRUE`; this works for both absolute and
relative calibration. It is useful to correct batches, assays, or plates using a reference material
shared across them.

```{r calibrate-batchwise}
#| results: false
#| message: true
mexp_res <- calibrate_by_reference(
  data = mexp,
  variable = "conc",
  reference_sample_id = "SRM1950",
  absolute_calibration = TRUE,
  batch_wise = TRUE,
  summarize_fun = "mean",
  undefined_conc_action = "na")

save_dataset_csv(mexp_res, tempfile(fileext = ".csv"), variable = "conc_beforecal")
```

## 7. Concentration ratio and bias

To examine the ratio between measured and expected (known) concentrations in the reference samples,
a table of concentration ratios can be generated with `store_conc_ratio = TRUE`. The ratio is

$$
R_\text{ratio}^\text{Analyte} = \frac{c_\text{measured}^\text{Analyte}}{c_\text{expected}^\text{Analyte}}
$$

```{r conc-ratio}
#| results: true
#| message: true
mexp_res <- calibrate_by_reference(
  data = mexp,
  variable = "conc",
  reference_sample_id = "SRM1950",
  absolute_calibration = TRUE,
  summarize_fun = "mean",
  undefined_conc_action = "na",
  store_conc_ratio = TRUE)

tbl_ref_bias <- mexp_res$dataset |>
  filter(sample_id == "SRM1950", is_quantifier) |>
  group_by(feature_id) |>
  summarise(bias_mean = mean(feature_conc_ratio))

gt::gt(tbl_ref_bias)
```

A ratio of 1 indicates perfect agreement between the measured and expected concentrations; values
above 1 suggest overestimation and values below 1 underestimation. The ratios can also be
visualised or analysed further to identify outliers or investigate issues in the analytical process
or calibration.

The ratio and bias of QC samples can also be obtained directly, without applying
`calibrate_by_reference()`. For illustration we compute them on the re-calibrated data above, where
all corrected feature concentrations in the reference sample are expected to have no bias (0%) and
a ratio of 1.

```{r qc-bias-after}
#| results: true
#| message: true
tbl <- get_qc_bias_variability(mexp_res, qc_types = "NIST")
gt::gt(tbl) |> gt::fmt_number(decimals = 3)
```

The bias and concentration ratios before re-calibration can be viewed from the `MRMhubExperiment`
object that had no calibration applied.

```{r qc-bias-before}
#| results: true
#| message: true
tbl <- get_qc_bias_variability(mexp, qc_types = "NIST", with_conc_ratio = TRUE)
gt::gt(tbl) |> gt::fmt_number(decimals = 3)
```

## Next steps

- [External calibration and QC](tutorial-06-external-calibration.html): full calibration-curve workflow with QC
- [Lipidomics workflow](tutorial-03-lipidomics-workflow.html): revisit the full pipeline
