---
title: "Getting started with R and Quarto notebooks"
description: >
  Install R and an IDE, create a Quarto project, learn to read and run notebook
  code, and render a report — for readers new to R and Quarto.
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Getting started with R and Quarto notebooks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span></p>

A MRMhub analysis is written as a document that mixes prose with the R code that produces the
results, so the report and the analysis are one file. This tutorial assumes no prior experience with
R or Quarto: it starts from an empty computer, installs the software, sets up the project and notebook
to work in, and shows how to read and run the code. Once your machine is set up, [Getting started with
MRMhub](tutorial-02-getting-started-mrmhub.html) runs a complete analysis in the project you create
here.

## 1. Install R and an editor

Two pieces of software are needed. **R** is the language the analysis runs in; an **integrated
development environment** (IDE) is the editor you write and run it in.

- Install R from [CRAN](https://cran.r-project.org/) (version 4.1 or newer). Pick the download for
  your operating system and accept the defaults.
- Install an IDE. [RStudio](https://posit.co/download/rstudio-desktop/) is the most common choice and
  the one this tutorial follows. If you already work in VS Code, [Positron](https://positron.posit.co/)
  is an RStudio-like alternative built on the same editor.

With R and RStudio installed, open RStudio and install MRMhub from the console (the bottom-left
pane). The first line installs `pak`, a package installer; the second uses it to fetch MRMhub and its
dependencies:

```{r install, eval = FALSE}
if (!require("pak")) install.packages("pak")
pak::pak("SLINGhub/MRMhub")
```

If the installation reports an error, the [Installation guide](manual-00-installation.html) lists the
common causes (a missing compiler on Windows or macOS, a firewall blocking the download) and their
fixes.

## 2. RStudio at a glance

When RStudio opens it shows four panes, each used at some point in an analysis:

- **Source** (top-left) — where you edit the notebook and run its code.
- **Console** (bottom-left) — runs a single line of R immediately; this is where you installed the
  package above.
- **Environment** (top-right) — lists the objects you have created, such as the `mexp` data object
  once it exists.
- **Files / Plots / Help** (bottom-right) — browses the project folder, shows figures, and displays
  function help.

The defaults are fine to start; nothing here needs configuring.

## 3. Create a Quarto project

A **project** keeps everything for one analysis — data, code, and results — in a single folder that
the IDE treats as a unit. In RStudio, choose *File → New Project → New Directory → Quarto Project*,
give it a name, and create it. Quarto's [project documentation](https://quarto.org/docs/tools/rstudio/)
covers the dialog in detail.

Inside the new project, add two folders to keep raw inputs separate from generated results — the
convention MRMhub's own workflows follow:

```
my_study/
├── data/            # raw INTEGRATOR output and metadata files
├── output/          # exported CSVs, PDFs, and reports
└── analysis.qmd     # this document
```

Paths in your code are written relative to this project folder: RStudio treats it as the working
directory, so `data/my_file.csv` points to the same place on any machine, without a full path. This
is why the `data/` and `output/` folders above are all the analysis needs to find its files.

## 4. Add text and code

A `.qmd` file has three kinds of content. At the top, a short **header** fenced by `---` sets the
title and the output format — `html`, `pdf`, or `docx`; a new project starts with a sensible HTML
header, and [Quarto workflows](manual-11-quarto-workflows.html) covers the options. Below it,
**prose** is written in Markdown — plain paragraphs, with `#` for headings, `**bold**`, and `-` for
lists. **Code** lives in *chunks*: R code fenced between ```` ```{r} ```` and ```` ``` ````. In
RStudio, *Insert → Code Chunk* (or Ctrl/Cmd + Alt + I) adds an empty one. A chunk that loads the
MRMhub package looks like this:

```` markdown
Load the package before calling any of its functions.

```{r}
library(mrmhub)
```
````

RStudio offers two ways to edit a `.qmd`: the **Source** editor shows the raw Markdown, while the
**Visual** editor (the *Visual* toggle, top-left of the document) shows a formatted, word-processor-like
view. Both edit the same file. Quarto's [authoring guide](https://quarto.org/docs/get-started/hello/rstudio.html)
introduces both.

## 5. Preview and run code

There are two distinct actions. **Running a chunk** executes its code immediately and shows the
result inline: click the green ▶ arrow at the chunk's top-right, or press Ctrl/Cmd + Shift + Enter
with the cursor inside it. **Rendering** turns the whole document into a finished HTML (or PDF)
report: click **Render**, and RStudio runs every chunk in order and assembles the output. The
**Preview** that opens updates each time you render, so the report stays in step with the code.

While developing an analysis, run chunks one at a time to check each step; render at the end to
produce the shareable report.

Try it now: add the chunk above and run it. Loading MRMhub without error confirms the installation
worked.

```{r load-mrmhub}
library(mrmhub)
```

## 6. Reading the code you'll run

The next tutorial uses MRMhub functions and no custom R, but three pieces of R syntax recur
throughout. Recognising them is enough to follow along:

- **Assignment** (`<-`) stores a result under a name. `mexp <- MRMhubExperiment()` creates a data
  object and names it `mexp`; writing `mexp` again later refers back to it.
- **Function calls with named arguments.** In `import_data_mrmhub(mexp, path = demo_file)`, the
  function acts on `mexp`, and `path =` names which input the file is. Most MRMhub functions take the
  data object first and options as named arguments.
- **The pipe** (`|>`) passes a result straight into the next function, so
  `plot_pca(mexp) |> save_plot("pca.pdf")` plots and then saves in one line, with no temporary name.

Each processing step follows the same shape — take the data object, return an updated one, and store
it back under the same name: `mexp <- normalize_by_istd(mexp)`.

## 7. Getting help

To see what a function does and which arguments it takes, type `?` followed by its name in the
Console:

```{r help, eval = FALSE}
?plot_pca
```

Its help page opens in the **Help** pane, listing every argument with an explanation and, usually,
examples. The same pages for every function are in the [function reference](../reference/index.html),
and the [MRMhub overview](manual-01-key-concepts.html) introduces the concepts they build on.

With the software installed, the project set up, and the syntax demystified, you are ready to run a
real analysis. Continue with [Getting started with MRMhub](tutorial-02-getting-started-mrmhub.html),
which imports the demo data and takes it through to an exported report.

## Next steps

- [Getting started with MRMhub](tutorial-02-getting-started-mrmhub.html): run a complete analysis in the project you just created
- [Quarto workflows](manual-11-quarto-workflows.html): rendering, figure sizing, and report options
- [Installation guide](manual-00-installation.html): detailed setup and troubleshooting
