---
title: "Interference Correction"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4
description: |
  Annotate mrm_pattern, derive isotopic interference relationships automatically
  with calc_isotopic_interferences(), inspect them, and apply the subtraction
  with correct_isotopic_interferences(). Declared and single-pair manual
  corrections are also covered.
vignette: >
  %\VignetteIndexEntry{Interference Correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      fig.width = 7, fig.height = 4, out.width = "100%")
```

<p class="page-kind page-kind-tutorial">Tutorial</p>

In class-based targeted assays, the natural-abundance heavy isotopologues of one species can overlap
the transition of a species two mass units heavier, inflating its measured area. MRMhub derives these
isotopic (M+2) interference relationships automatically from a per-feature `mrm_pattern` annotation
and removes them with a contribution-based subtraction ported from the LICAR method
[@gao2021LICARApplicationIsotopic]. Correction is worth considering whenever a blank shows non-zero
signal in a feature that should be empty, a species and its M+2 shoulder correlate strongly across
injections, or the theoretical M+2 contribution from an adjacent same-class species exceeds a few
percent of the target signal. It runs on raw feature intensities (`feature_intensity`) and applies to
*every* sample — unlike drift and batch correction, which are fitted on QC samples only.

For MRM data the correction must be *fragment-based*: a heavy isotope's contribution depends on
whether it lands on the retained product ion or the neutral loss. The whole-molecule `level = "MS1"`
path is therefore for genuine full-scan data only, **never a fallback for MRM** data that happens to
lack a product m/z. The underlying concepts (front/back overlaps, the co-elution requirement) are
covered in the [Isotopic interference correction](manual-12-interference-correction.html) manual page.

::: {.callout .callout-note}
**Time** ~15 min &nbsp;·&nbsp; **Level** Advanced &nbsp;·&nbsp; **Prerequisites** [Basic workflow](tutorial-02-basic-workflow.html)
:::

## 1. Preparing for isotopic correction

Automatic derivation needs one annotation: the `mrm_pattern` column on the `Features` metadata sheet,
giving each feature its class + MRM pattern. The label encodes both the lipid class and the
product-ion origin, so `calc_isotopic_interferences()` can build the correct fragment formula. The
precursor and product m/z the derivation also needs — and the optional `polarity` that narrows the
pattern choices — can come from the imported data (INTEGRATOR) *or* the metadata; only `mrm_pattern`
must be annotated by hand. A `Features` sheet ready for correction looks like this — precursor/product
m/z and `polarity` supplied by the import, with `mrm_pattern` added as the final column:

```{r, eval = TRUE, echo = FALSE}
tibble::tribble(
  ~feature_id, ~feature_class, ~precursor_mz, ~product_mz, ~polarity, ~mrm_pattern,
  "PC 34:2", "PC", 758.6, 184.1, "Pos", "PC (Pos) Pro=184.1",
  "PC 34:1", "PC", 760.6, 184.1, "Pos", "PC (Pos) Pro=184.1",
  "PC 34:0", "PC", 762.6, 184.1, "Pos", "PC (Pos) Pro=184.1",
  "Cer 18:1;O2/16:0", "Cer", 538.5, 264.3, "Pos", "Cer (Pos) SphB-2H2O",
  "PC 34:1 [FA 16:0]", "PC", 804.6, 255.2, "Neg", "PC (Neg, FA) FA"
) |>
  knitr::kable(caption = "Example `Features` metadata (`mrm_pattern` is the added annotation).")
```

The valid labels are listed by [`licar_pattern_choices()`](../reference/licar_pattern_choices.html):

```{r}
library(mrmhub)
licar_pattern_choices("Head Group")
```

The metadata template (`save_metadata_templates()`) provides `mrm_pattern` as a filtered dropdown on
the `Features` sheet, with an optional `polarity` column that narrows the choices and colour warnings
when a species name does not match the chosen pattern's class. On import, MRMhub validates the column:
an unknown label is an error, and a name/pattern class mismatch or a sum-only name under a
chain-resolved (FA/LCB) pattern is a warning.

```{r}
mexp <- readRDS("results/mexp_processed.rds")
```

## 2. Derive interference relationships

[`calc_isotopic_interferences()`](../reference/calc_isotopic_interferences.html) discovers the M+2
overlaps and stores them in the `annot_interferences` slot. Choose the derivation `level` to match the
acquisition:

```{r}
# Class-based LC-MRM: fragment-level front/back correction (needs precursor + product m/z).
mexp <- calc_isotopic_interferences(mexp, level = "MRM")

# Genuine MS1 / full-scan: whole-molecule M+2 (needs precursor m/z only).
# mexp <- calc_isotopic_interferences(mexp, level = "MS1")
```

Derived factors are pinned to `enviPat` 2.8 [@loos2015NontargetEnviPat] and are deterministic, so
re-importing the annotation and re-deriving reproduces identical edges. Raise `min_contribution` to
drop negligible or borderline pairs, and see [co-elution filtering](#co-elution-filtering-experimental)
for the experimental option to keep only chromatographically co-eluting pairs.

## 3. Inspect the derived relationships

The two-stage API is a built-in *preview*: review the derived edges before subtracting anything. Each
row of `annot_interferences` is one overlap, with its contribution factor `K` and an `overlap_type`
(`m2_front`, `m2_back`, or `ms1_m2`):

```{r}
mexp@annot_interferences
```

[`summarize_interferences()`](../reference/summarize_interferences.html) rolls this up — how many
features are affected, split by source (auto-derived vs. declared) and overlap type, with the
contribution-factor range:

```{r}
summarize_interferences(mexp)
```

A victim carrying both a `m2_front` and a `m2_back` row is corrected against two interferers. The
`source` column distinguishes auto-derived (`"auto"`) from declared (`"manual"`) edges.

## 4. Apply the correction

[`correct_isotopic_interferences()`](../reference/correct_isotopic_interferences.html) subtracts the
derived edges from the raw intensities in one pass:

$$
\text{Value}_\text{corrected} = \text{Value}_\text{target} -
   \sum_i K_i \cdot \text{Value}_{\text{interfering}_i}
$$

```{r}
mexp <- correct_isotopic_interferences(mexp)
```

If nothing has been derived yet the function stops with a reminder to run
`calc_isotopic_interferences()` first. Declared interferences — in-source fragments, co-eluting
isobars, or other overlaps you annotate yourself in the feature metadata — are applied with the
sibling [`correct_custom_interferences()`](../reference/correct_custom_interferences.html).

Upstream features in a chain (A ← B ← C) are corrected first via a topological ordering; circular
dependencies are detected and aborted with an informative error. For auto-derived edges, a
subtraction is skipped when the running-corrected interferer has already fallen to zero, and corrected
values are clamped at zero (LICAR parity). The original raw values are preserved in
`feature_intensity_orig`. Apply interference correction *before* ISTD normalisation, drift, and batch
correction; running it later resets those steps (with a warning) because they must use the corrected
raw signal.

## 5. Manual correction of a single pair

For a one-off correction, or to validate a factor before trusting the automatic derivation, use
[`correct_interference_manual()`](../reference/correct_interference_manual.html). `variable` here is the
actual `dataset` column name (`feature_intensity`).

```{r}
mexp <- correct_interference_manual(
  mexp,
  variable                  = "feature_intensity",
  feature                   = "PC 32:0",
  interfering_feature       = "SM 36:1 M+3",
  interference_contribution = 0.0107,
  neg_to_na                 = FALSE,
  updated_feature_id        = NA
)
```

Setting `updated_feature_id` renames the corrected feature so raw and corrected channels can coexist
under different IDs. Manual factors can be sourced from a theoretical isotopologue calculation
(`enviPat`), an empirical pure-standard injection (which also captures Q1/Q3 transmission and
in-source effects), or published class-level interference tables — verified empirically when moving a
method between instruments.

## 6. Verifying the correction

```{r, eval = TRUE, include = FALSE, message = FALSE, warning = FALSE}
# Build a corrected object for the QC plots below from the bundled multi-class
# LICAR HILIC demo (PC, SM, Cer, dhCer in positive mode plus PC/PE via the
# negative FA fragment), mirroring sections 2-4: import the long analysis data
# (carrying precursor/product m/z and polarity) together with the mrm_pattern
# feature sheet, auto-derive the M+2 edges, and subtract them. The derivation
# needs enviPat 2.8 and rgoslin.
library(mrmhub)
demo_csv <- system.file("extdata", "licar_hilic_demo.csv", package = "mrmhub")
demo_features <- system.file(
  "extdata",
  "licar_hilic_demo_features.csv",
  package = "mrmhub"
)

mexp <- MRMhubExperiment() |>
  import_data_csv_long(path = demo_csv, import_metadata = TRUE, silent = TRUE) |>
  import_metadata_features(path = demo_features, ignore_warnings = TRUE) |>
  calc_isotopic_interferences(level = "MRM") |>
  correct_isotopic_interferences()
```

```{r}
d <- get_analyticaldata(mexp, annotated = TRUE) |>
  dplyr::filter(feature_id == "PC 32:0") |>
  dplyr::select(analysis_id, qc_type,
                intensity_before = feature_intensity_orig,
                intensity_after  = feature_intensity) |>
  dplyr::mutate(pct_change = 100 * (intensity_after - intensity_before) / intensity_before)

summary(d$pct_change)
```

For blanks (`SBLK`/`PBLK`), residual signal after correction should approach zero. A non-zero blank
median often indicates an underestimated contribution factor.
[`plot_qc_interference_impact()`](../reference/plot_qc_interference_impact.html) shows how many
features were affected by which magnitude of correction (percent of signal removed) in the study
samples:

```{r, eval = TRUE, message = FALSE, warning = FALSE}
plot_qc_interference_impact(mexp, qc_types = "SPL")
```

For a per-feature view, [`plot_qc_interferences()`](../reference/plot_qc_interferences.html) shows each
corrected feature's residual signal as a percent of its uncorrected value, split by QC type. With many
features corrected, `min_correction_pct` focuses the plot on the substantially-affected ones (here,
those with at least 40 % of their signal removed); `include_feature_filter` / `exclude_feature_filter`
narrow it to specific features or classes. When blanks are present, a blank approaching zero alongside
study samples (`SPL`) retaining most of their signal is the expected pattern:

```{r, eval = TRUE, message = FALSE, warning = FALSE}
plot_qc_interferences(mexp, qc_types = "SPL", min_correction_pct = 40)
```

## Co-elution filtering (experimental)

An isotopologue elutes at its monoisotopic apex, so an interferer's M+2 lands in a victim's integrated
area only if the two co-elute. The experimental `check_coelution` option keeps an m/z-matched edge
only when the interferer's apex falls inside the victim's integration window, dropping
chromatographically resolved pairs:

```{r}
mexp <- calc_isotopic_interferences(mexp, level = "MRM", check_coelution = TRUE)
```

It is **off by default** while the gate is validated. When enabled, inspect `annot_interferences`
carefully — resolved pairs it drops (and, when retention data are missing, pairs it cannot verify) are
reported in the console.

## Recommendations

- Apply interference correction on raw `feature_intensity` *before* normalisation, drift, and batch
  correction.
- Prefer `level = "MRM"` for MRM data and supply the product m/z; reserve `level = "MS1"` for genuine
  full-scan measurements.
- Inspect `annot_interferences` (or `summarize_interferences()`) before applying.
- Keep the `mrm_pattern` annotation with the project: it reproduces the derived factors under the
  `enviPat` 2.8 pin.
- After correction, verify that blank residual signal approaches zero.

## Next steps

- [Isotopic interference correction](manual-12-interference-correction.html) — the conceptual reference
- [Drift and Batch Correction](tutorial-04-drift-correction.html) — apply after interference correction
- [Basic MRMhub Workflow](tutorial-02-basic-workflow.html) — full processing pipeline
- [The MRMhubExperiment Data Object](manual-02-data-object.html#feature-variables) — how `_orig` postfixes preserve raw values

## References
