---
title: "RunScatter Plot"
output: 
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 3
vignette: >
  %\VignetteIndexEntry{RunScatter Plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`plot_runscatter()` visualises feature signals across the analysis sequence, helping to identify
trends, detect outliers, and assess analytical performance. This tutorial demonstrates the main
parameters using a single feature (`CE 18:1`) from the built-in lipidomics dataset.

## Setup

```{r setup}
#| results: false  
#| message: false
library(mrmhub)

mexp <- lipidomics_dataset
mexp <- normalize_by_istd(mexp)
mexp <- calc_qc_metrics(mexp)
```

## Basic plot

A minimal call plots all QC and sample types for one feature.

```{r}
#| fig.alt: >
#|   Basic RunScatter plot of CE 18:1
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                rows_page = 1, cols_page = 1)
```

## Selecting QC types

Use `qc_types` to display only specific sample types.

```{r}
#| fig.alt: >
#|   RunScatter plot showing BQC and SPL only
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                qc_types = c("BQC", "SPL"),
                rows_page = 1, cols_page = 1)
```

## Batch display

Set `show_batches = TRUE` and `batch_zebra_stripe = TRUE` to highlight batch boundaries with
alternating shaded areas.

```{r}
#| fig.alt: >
#|   RunScatter plot with batch zebra stripes
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                show_batches = TRUE,
                batch_zebra_stripe = TRUE,
                rows_page = 1, cols_page = 1)
```

## Reference lines

Show mean ± k × SD reference lines with `show_reference_lines = TRUE`. Use `reference_sd_shade`
to display the range as a shaded band instead of lines.

```{r}
#| fig.alt: >
#|   RunScatter plot with mean ± 2 SD reference band
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                qc_types = c("BQC", "SPL"),
                show_reference_lines = TRUE,
                ref_qc_types = "BQC",
                reference_k_sd = 2,
                reference_sd_shade = TRUE,
                rows_page = 1, cols_page = 1)
```

## Outlier capping

Use `cap_outliers = TRUE` to cap extreme values based on MAD fences. This is useful when outliers
obscure the trends of QC or study samples.

```{r}
#| fig.alt: >
#|   RunScatter plot with outlier capping
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                cap_outliers = TRUE,
                cap_sample_k_mad = 3,
                rows_page = 1, cols_page = 1)
```

## Log scale

Set `log_scale = TRUE` to apply a log10 transformation to the y-axis. Zero or negative values are
replaced with the minimum positive value divided by 5.

```{r}
#| fig.alt: >
#|   RunScatter plot with log10 y-axis
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                log_scale = TRUE,
                rows_page = 1, cols_page = 1)
```

## Removing gaps in the analysis sequence

Gaps in the x-axis can arise in two ways:

1. **Filtered QC types** — when only a subset of QC types is selected, the
   unselected positions leave gaps. Use `collapse_excluded = TRUE` to close
   them.
2. **Unannotated analyses** — when some runs in the analytical sequence have no
   corresponding entry in `@dataset` (e.g. solvent blanks or system
   suitability injections that were not imported), the analysis order contains
   discontinuities. Use `remove_gaps = TRUE` to collapse these and mark the
   former gap positions with vertical indicator lines.

Both options can be combined.

### Collapsing excluded QC types

```{r}
#| fig.alt: >
#|   RunScatter plot with collapsed gaps from excluded QC types
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                qc_types = c("LTR"),
                collapse_excluded = TRUE,
                rows_page = 1, cols_page = 1)
```

### Removing gaps from unannotated analyses

To demonstrate this feature, we first simulate a dataset with unannotated analyses by excluding a
block of runs from batch 2. This creates a contiguous gap in the analysis order, as would occur
when solvent blanks or system suitability injections are not imported.

```{r}
#| results: false 
#| message: false
mexp_gaps <- exclude_analyses(
  mexp,
  analyses = c(
    paste0("Longit_batch2_", 1:45),
    "Longit_batch2_PQC 12",
    "Longit_batch2_PQC 13",
    "Longit_batch2_PQC 14",
    "Longit_batch2_PQC 15",
    "Longit_batch2_TQC13",
    "Longit_batch2_TQC14",
    "Longit_batch2_TQC15"
  ),
  clear_existing = TRUE
)
mexp_gaps <- normalize_by_istd(mexp_gaps)
mexp_gaps <- calc_qc_metrics(mexp_gaps)
```

```{r}
#| fig.alt: >
#|   RunScatter plot with gap indicator lines
#| results: false 
#| message: false
plot_runscatter(mexp_gaps, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                remove_gaps = TRUE,
                rows_page = 1, cols_page = 1)
```

### Combining both

When filtering to a single QC type *and* the sequence contains unannotated
runs, combine `collapse_excluded` and `remove_gaps` to produce a compact,
fully annotated plot.

```{r}
#| fig.alt: >
#|   RunScatter plot with collapsed excluded types and gap indicators
#| results: false 
#| message: false
plot_runscatter(mexp_gaps, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                qc_types = c("BQC"),
                collapse_excluded = TRUE,
                remove_gaps = TRUE,
                rows_page = 1, cols_page = 1)
```

## Label wrapping

When feature names are long, strip labels can overflow. Set `label_wrap = TRUE` to wrap labels
across multiple lines, controlled by `label_wrap_width`. Here we show three features to illustrate
the effect on strip text.

```{r}
#| fig.alt: >
#|   RunScatter plot with wrapped strip labels
#| results: false 
#| message: false
plot_runscatter(mexp, variable = "intensity",
                include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
                rows_page = 1, cols_page = 3,
                specific_page = 1,
                label_wrap = TRUE,
                label_wrap_width = 8)
```

## Trend curves

Trend curves visualise the fitted signal used during drift or batch correction. They require a
prior correction step — `show_trend = TRUE` will error if no drift or batch correction has been
applied. To plot the values *before* the last correction, append `_before` to the variable name
(e.g. `variable = "intensity_before"`); for the original uncorrected values use `_raw`
(e.g. `variable = "intensity_raw"`).

See the [Drift Correction](tutorial-04-drift-correction.html) tutorial for a worked example.

## Next Steps

- [Drift Correction](tutorial-04-drift-correction.html) — apply signal drift correction
- [Batch Correction](tutorial-06-batch-correction.html) — apply inter-batch correction
