---
title: "Quantification with external calibration"
description: >
  Fit external calibration curves for a targeted assay, quantify the samples,
  and check the result against QC samples with known concentrations.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Quantification with external calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Intermediate</span> <span class="page-prereq">Prerequisites: <a href="tutorial-02-basic-workflow.html">Basic workflow</a></span></p>

External calibration quantifies each analyte from a calibration curve measured alongside the
samples, as used in clinical chemistry, toxicology and environmental analysis. This tutorial fits
one curve per analyte from a dilution series of calibrators, quantifies the samples, and checks the
result against QC samples with assigned target concentrations. The example is a 15-analyte serum
steroid panel measured by LC-MRM-MS, with calibrants (Cal A–F), low and high QCs (LQC, HQC), and
external quality-assessment samples in place of study samples.

The target concentrations for the calibrators and QCs live in the `QCconcentrations` metadata,
which requires `sample_id` in the analysis metadata and `analyte_id` in the feature metadata.

::: {.callout .callout-note}
The full, reproducible report for this dataset — including drift-range flags, exports and the
underlying data — is published as
[Dataset 4](https://slinghub.github.io/MRMhub-workflows/Dataset4.html) in the MRMhub-workflows
supplement.
:::

## 1. Import data and metadata

We load the INTEGRATOR peak areas into a fresh `MRMhubExperiment`, then attach the metadata
workbook that defines the analytes, their internal standards, the sample annotations, and the
calibration and target concentrations.

```{r import}
#| message: true
library(mrmhub)

mexp <- MRMhubExperiment(title = "Steroid Assay")
mexp <- import_data_mrmhub(
  mexp,
  path = "datasets/Dataset4_MRMhub-INTEGRATOR_ASSAY.csv",
  import_metadata = TRUE)
mexp <- import_metadata_msorganiser(
  mexp,
  path = "datasets/Dataset4_Metadata.xlsx",
  excl_unmatched_analyses = TRUE,
  ignore_warnings = TRUE)
```

## 2. Fit the calibration curves

`normalize_by_istd()` divides each analyte's peak area by that of its internal standard, writing the
ratio to `feature_norm_intensity`. `calc_calibration_results()` then fits one curve per analyte from
the calibrator levels — here a quadratic model with `1/x` weighting for all analytes. The model and
weighting can also be set per analyte in the feature metadata.

```{r calibrate}
#| message: true
mexp <- normalize_by_istd(mexp)
mexp <- calc_calibration_results(
  mexp,
  fit_overwrite = TRUE,
  fit_model = "quadratic",
  fit_weighting = "1/x")
```

```{r calibration-curves}
#| fig.cap: !expr fig_cap("Quadratic, 1/x-weighted calibration curves for the 15 steroid analytes. Curves are solid within the calibrator range and dotted where extrapolated.")
#| fig.alt: Calibration curves for the 15 steroid analytes, plotted on a log scale
#| fig.width: 9
#| fig.height: 7
#| out.width: "100%"
plot_calibrationcurves(
  mexp,
  fit_overwrite = TRUE,
  fit_model = "quadratic",
  fit_weighting = "1/x",
  include_istd = FALSE,
  include_qualifier = FALSE,
  rows_page = 4, cols_page = 4,
  show_progress = FALSE)
```

`get_calibration_metrics()` summarises each fit: the R² for linearity and the limits of detection
and quantification (`lod`, `loq`), in the calibrated concentration unit.

```{r cal-metrics}
get_calibration_metrics(
  mexp,
  include_qualifier = FALSE, summary_table = TRUE)
```

## 3. Quantify the samples

Once the curves are satisfactory, `quantify_by_calibration()` inverts each fitted curve to convert
the normalized intensities into absolute concentrations, written to `feature_conc`. Using
`ignore_failed_calibration = TRUE` skips any analyte whose curve failed to fit rather than aborting
the whole run.

```{r quantify}
#| message: true
mexp <- quantify_by_calibration(
  mexp,
  fit_overwrite = FALSE,
  include_qualifier = FALSE,
  ignore_failed_calibration = TRUE,
  fit_model = "quadratic",
  fit_weighting = "1/x")
```

## 4. Check QC bias and variability

The final check compares the measured QC concentrations against their known targets.
`get_qc_bias_variability()` reports, per analyte and QC level, the mean measured concentration, the
**bias** (percent deviation from target — accuracy) and the intra-batch **%CV** (precision).

```{r qc-check}
get_qc_bias_variability(
  mexp,
  qc_types = c("LQC", "HQC"), summary_table = TRUE)
```

## 5. Export the concentrations

`save_dataset_csv()` writes a flat table of concentrations, one row per sample. `add_qctype = TRUE`
keeps the QC-type column so calibrators, QCs and samples stay distinguishable. For a multi-sheet
workbook that also bundles the calibration and QC metrics, use `save_report_xlsx()`.

```{r export}
save_dataset_csv(
  mexp,
  path = "steroid_conc.csv",
  variable = "conc",
  add_qctype = TRUE)
```

## Next steps

- [Calibration by a reference sample](tutorial-07-calibration-reference.html): an alternative when no calibration series is available
- [Custom QC report](recipe-02-custom-qc-report.html): build a formatted QC report from processed data
- [Visualisation functions](manual-08-visualization.html): the plotting reference, including calibration and QC plots
