---
title: "Batch-effect correction"
output: 
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Batch-effect correction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  markdown: 
    wrap: 100
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

MRMhub currently supports median centering-based batch effect correction.

The data must be provided via a MRMhubExperiment object, whereby raw data that was imported or
processed data can be corrected, such as `intensity` or `conc` values.

::: {.callout .callout-note}
**Time:** ~10 min &nbsp;|&nbsp; **Level:** Intermediate &nbsp;|&nbsp; **Prerequisites:** [Basic workflow](tutorial-02-basic-workflow.html)
:::

## Import data

This tutorial imports pre-calculated concentration values from a CSV file. The file must
contain a column with batch information (`batch_id`); see `import_data_csv_wide()` for details.

```{r setup}
#| results: false  
#| message: false
library(mrmhub)

myexp <- mrmhub::MRMhubExperiment()

myexp <- import_data_csv_wide(
  myexp,
  path = "simdata-u1000-sd100_7batches.csv",
  variable_name = "conc",
  import_metadata = TRUE)
```

## Apply batch-centering

The concentration values of each batch are centered on a reference QC type, here the samples
(`ref_qc_types = "SPL"`). The `correct_scale` parameter controls whether the batch-to-batch
differences in variance (scale) are also corrected. Here it is set to `FALSE`, so only the median
(location) of each batch is aligned; variance scaling is disabled.

```{r}
#| results: false  
#| code-overflow: 'wrap'

myexp <- correct_batch_centering(
  data = myexp, 
  variable = "conc",
  ref_qc_types = "SPL", 
  correct_scale = FALSE
)

```

The data before and after batch correction are compared below. The batches are now
aligned. Note: if the samples or other quality control types do not follow the reference samples,
they may not be corrected appropriately.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false
plot_runscatter(myexp, variable = "conc_before", rows_page = 1, cols_page = 1)
plot_runscatter(myexp, variable = "conc", rows_page = 1, cols_page = 1)
```

## Batch-centering with variance scaling

Although the batches are aligned, the spread (variance) of the data points
varies considerably between the batches. This can be corrected by scaling the variance via
`correct_scale = TRUE`. In the plot below, the variance of the data points is now
fairly consistent across the batches.

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: true
myexp <- mrmhub::correct_batch_centering(
  myexp, 
  ref_qc_types = "SPL", 
  variable = "conc",
  correct_scale = TRUE
)
```

```{r}
#| fig.alt: >
#|   RunScatter plots before and after batch correction
#| results: false 
#| message: false

plot_runscatter(myexp, variable = "conc", rows_page = 1, cols_page = 1)

```

## Export batch-corrected data

The corrected data can be processed further with `MRMhub` functions, or exported.

```{r}
save_dataset_csv(
  myexp, 
  path = "batch-corrected-conc-data.csv", 
  variable = "conc", 
  filter_data = FALSE
  )
```

## Next Steps

- [Drift Correction](tutorial-04-drift-correction.html) — correct signal drift within a batch
- [Calibration by Reference Sample](tutorial-07-calibration-reference.html) — normalise to a reference
- [External Calibration & QC](recipe-01-ext-calibration-qc.html) — quantify with calibration curves
