---
title: "Your First Analysis"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
description: |
  A 5-minute walkthrough: import demo data, inspect, normalize, and export.
vignette: >
  %\VignetteIndexEntry{Your First Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", out.width = "100%",
  message = FALSE, warning = FALSE, eval = FALSE
)
```

This walkthrough produces a normalized dataset in under 5 minutes using
bundled demo data; no external files are needed.

::: {.callout .callout-note}
**Time:** ~5 min &nbsp;|&nbsp; **Level:** Beginner &nbsp;|&nbsp; **Prerequisites:** MRMhub installed (`check_setup()` passes)
:::

## The complete workflow in one script

The complete workflow on the bundled demo data is shown below. The printed summary and plot are
produced by this code.

```{r whole-thing, eval=TRUE, fig.height=4}
library(mrmhub)

# 1. Import the bundled demo dataset (produced by INTEGRATOR)
demo_file <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")
mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(mexp, path = demo_file, import_metadata = TRUE)
mexp

# 2. Normalise each feature by its internal standard
mexp <- normalize_by_istd(mexp)

# 3. Inspect the normalised signal across the analytical run
plot_runscatter(mexp, variable = "norm_intensity")
```

To save the processed data to an Excel report (or a tidy CSV), add one line:

```{r eval=FALSE}
save_report_xlsx(mexp, path = "my_first_results.xlsx")
```

::: {.callout .callout-tip}
**Interactive alternative.** `run_walkthrough()` opens a point-and-click application that validates
the data, generates the equivalent workflow script, and displays the results. The code-first
workflow above remains the reproducible path.
:::

## Step by step

The rest of this page explains each step of the script above.

### Step 1: Load MRMhub

```{r}
library(mrmhub)
```

### Step 2: Import Demo Data

MRMhub ships with a demo dataset produced by INTEGRATOR.
Import it with `import_data_mrmhub()`:

```{r}
demo_file <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")
mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(mexp, path = demo_file, import_metadata = TRUE)
mexp
```

The result is an `MRMhubExperiment` object, a structured container
holding the peak area data, sample annotations, and feature metadata.

### Step 3: Explore the Data

```{r}
# Dimensions: samples × features
get_analysis_count(mexp)
get_feature_count(mexp)

# First few sample annotations
head(mexp@annot_analyses)

# Feature list
get_featurelist(mexp)
```

### Step 4: Normalize by Internal Standard

Apply ISTD normalization to correct for extraction and injection variability:

```{r}
mexp <- normalize_by_istd(mexp)
```

::: {.callout .callout-tip}
After this step, `mexp@is_istd_normalized` is `TRUE`.
The original data is preserved in `dataset_orig`.
:::

### Step 5: Visualise

Check the run sequence for signal drift:

```{r, fig.height=4}
plot_runscatter(mexp, variable = "norm_intensity")
```

### Step 6: Export

Export the processed data to Excel or CSV:

```{r}
# Excel report (multiple sheets)
save_report_xlsx(mexp, path = "my_first_results.xlsx")

# Or as tidy CSV
save_dataset_csv(mexp, path = "my_first_results.csv")
```

## Next Steps

- [Key Concepts](manual-00-key-concepts.html) — understand the data model
- [Lipidomics Workflow](tutorial-03-lipidomics-workflow.html) — full lipidomics workflow
- [Drift Correction](tutorial-04-drift-correction.html) — drift & batch correction
- [External Calibration](recipe-01-ext-calibration-qc.html) — quantitative assay with calibration
