---
title: "Preparing and importing data"
description: >
  Import analytical data from the supported platforms and bring in the sample
  and feature metadata that the MRMhub workflow depends on.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 2
vignette: >
  %\VignetteIndexEntry{Preparing and importing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
source("_common.R")
set.seed(1041)
options(dplyr.print_max = 10)

if (fs::file_exists("metadata_template.xlsx")) {
  fs::file_delete("metadata_template.xlsx")
}
if (fs::file_exists("metadata_msorganiser_template.xlsx")) {
  fs::file_delete("metadata_msorganiser_template.xlsx")
}
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-prereq">Prerequisites: <a href="tutorial-02-getting-started-mrmhub.html">Getting started with MRMhub</a></span></p>

Every analysis begins with the analytical data: the pre-processed measurements
from the instrument, such as peak areas. That alone is already enough for some
basic exploration, for example plotting signal intensities against injection
order. Most processing, though, also draws on metadata: the information
describing the samples and features behind those measurements. Some metadata
travels with the data file itself (INTEGRATOR and MassHunter results in
particular can carry it, pulled in with the importer's `import_metadata`
argument) while the rest is imported from separate sources. This tutorial
covers both, from a raw result file to a fully annotated `MRMhubExperiment`
ready for processing.

## 1. Importing analytical data

Analytical data, such as pre-processed measurement data (e.g. peak areas), can be
imported from several software platforms. MRMhub reads results from INTEGRATOR,
its bundled peak picker (`import_data_mrmhub()`), from Agilent MassHunter
Quantitative Analysis (`import_data_masshunter()`), from Skyline
(`import_data_skyline()`), and from generic wide- or long-format CSV files
(`import_data_csv_wide()` and `import_data_csv_long()`).

Do not manually edit these files before importing them. Hand edits can corrupt
the file or introduce silent errors in the data; MRMhub is built around
reproducible, automated workflows, and editing the source data works against
that.

We start from an empty object and read an INTEGRATOR result file. The imported
measurements land in the `dataset_orig` slot and are copied to the working
`dataset`; with `import_metadata = TRUE`, the sample- and feature-level metadata
embedded in the file (such as acquisition time stamps and precursor/product
m/z) is brought across as well.

```{r import-mrmhub}
#| message: true
library(mrmhub)

mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(
  mexp,
  path = "datasets/sPerfect_MRMhub.tsv",
  import_metadata = TRUE)
```

A plain wide-format CSV of peak areas is imported the same way, into a fresh
object. Here `variable_name` names the quantity stored in the table's cells, so
MRMhub knows which measurement it is reading.

```{r import-csv}
mexp2 <- MRMhubExperiment()
mexp2 <- import_data_csv_wide(
  mexp2,
  path = "data/plain_wide_dataset.csv",
  variable_name = "area",
  import_metadata = TRUE)
```

See [Importing analytical data](manual-04-data-import.html) for a decision guide
and full documentation of every importer.

## 2. Metadata

Metadata (the analysis metadata that describes and annotates the analytical
data at the sample and feature level) is essential to the MRMhub workflow. Some
of it can be read straight from the imported result file, but more commonly the
required metadata comes from other sources and has to be imported separately.
Which tables are needed depends on the intended processing; see
[Data and metadata in MRMhub](manual-02-data-object.html) for an overview of the
metadata categories.

When the data file already carries embedded analysis metadata (an INTEGRATOR
result, for instance, can hold `qc_type`, `batch_id` and `analysis_order`
columns), `import_metadata_from_data()` lifts those columns into the annotation
tables in a single call:

```{r import-metadata-from-data, eval = FALSE}
mexp <- import_metadata_from_data(mexp)
```

This is the fast path for data that already contains its analysis metadata.
Internal standards, calibration concentrations and other annotations still come
from separate sources, prepared and imported as shown below.

Integrity of the data and metadata is key to accurate, reproducible processing.
MRMhub therefore inspects imported metadata for completeness and for consistency
of the IDs used across the different tables, and reports a summary of the errors,
warnings and notes it finds after each import.

## 3. Preparing metadata

Metadata can be imported from Excel sheets, CSV files, or R data frames.
Preparing it is often a manual step, collecting information from various sources.
To support this, MRMhub ships Excel templates that carry the expected column
headers and usage notes for every supported metadata type. Starting from a
template and filling it in is the recommended approach.
`save_metadata_templates()` writes the workbook (`metadata_template.xlsx` by
default) to the working directory:

```{r save-templates, eval = TRUE, message = TRUE}
save_metadata_templates()
```

## 4. Importing metadata

Metadata can be imported table by table from CSV files. Below we bring in the
analysis and feature tables. `excl_unmatched_analyses = TRUE` drops any imported
analysis that has no matching row in the metadata, and `ignore_warnings = TRUE`
lets the import proceed despite the integrity warnings reported above.

```{r import-metadata-csv}
#| message: true
mexp <- import_metadata_analyses(
  mexp,
  path = "datasets/analysis_metadata.csv",
  excl_unmatched_analyses = TRUE,
  ignore_warnings = TRUE)

mexp <- import_metadata_features(
  mexp,
  path = "datasets/feature_metadata.csv",
  ignore_warnings = TRUE)
```

Metadata can also be read from the sheets of an Excel workbook, which lets all
tables live in a single file. Here we add the internal-standard metadata from
the `ISTDs` sheet:

```{r import-metadata-xlsx}
mexp <- import_metadata_istds(
  mexp,
  path = "datasets/metadata_tables.xlsx",
  sheet = "ISTDs",
  ignore_warnings = TRUE)
```

Finally, metadata can be passed directly as an R `data.frame`, so it can be
prepared in R or pulled from a database or LIMS. Below we import the quality
control sample concentrations from a table read into R:

```{r import-metadata-df}
df_qcinfo <- readr::read_table(file = "datasets/qc_metadata.txt")
mexp <- import_metadata_qcconcentrations(mexp, table = df_qcinfo)
```

## 5. Metadata from the MSOrganiser template

Rather than importing each table on its own, all metadata can be collected in a
single MSOrganiser template (an Excel file providing a sheet for every metadata
type, with built-in validity and integrity checks) and imported in one step.
Obtain the template from <https://github.com/SLINGhub/mrmhub>, or save a blank
copy to the working directory:

```{r save-msorganiser}
save_metadata_msorganiser_template()
```

Only the tables required by the intended workflow need to be completed; the
import reads every table that has been filled in:

```{r import-msorganiser, eval = FALSE}
mexp <- import_metadata_msorganiser(
  mexp,
  path = "datasets/sPerfect_Metadata.xlsx",
  ignore_warnings = TRUE)
```

See [Importing metadata](manual-05-metadata.html) for the full set of metadata
tables, templates and validation checks.

## Next steps

- [Importing analytical data](manual-04-data-import.html): decision guide and reference for every importer
- [Importing metadata](manual-05-metadata.html): metadata tables, templates and validation
- [Common defects and fixes](manual-05-metadata.html): checking metadata for common errors
- [Lipidomics workflow](tutorial-03-lipidomics-workflow.html): a complete walkthrough from import to export
