---
title: "Exploring QC: RunScatter and PCA"
description: >
  Read run-order signal quality with RunScatter and multivariate structure with
  PCA, using the built-in lipidomics dataset.
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 3
vignette: >
  %\VignetteIndexEntry{Exploring QC: RunScatter and PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 100
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Intermediate</span> <span class="page-prereq">Prerequisites: <a href="tutorial-02-basic-workflow.html">Basic workflow</a></span></p>

This tutorial covers two complementary QC-exploration views on the built-in
lipidomics dataset. `plot_runscatter()` shows a feature's signal across the
analysis sequence, which helps to identify trends, detect outliers, and assess
analytical performance. Principal component analysis (`plot_pca()`,
`plot_pca_loading()`, `detect_outlier_pca()`) adds a multivariate view for
spotting injection outliers and residual batch effects. The RunScatter examples
all use a single feature, `TG 48:2 [-18:1]`.

## 1. Setup

We load the dataset, normalise to internal standards, and compute the QC metrics
that several of the plots below draw on.

```{r prepare}
#| results: false
library(mrmhub)

mexp <- lipidomics_dataset
mexp <- normalize_by_istd(mexp)
mexp <- calc_qc_metrics(mexp)
```

## 2. Basic plot

A minimal call plots all QC and sample types for one feature.

```{r basic}
#| fig.cap: !expr fig_cap("RunScatter of TG 48:2 [-18:1] across the analysis sequence, all sample types shown.")
#| fig.alt: Basic RunScatter plot of TG 48:2 [-18:1]
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  rows_page = 1, cols_page = 1)
```

## 3. Selecting QC types

Use `qc_types` to display only specific sample types.

```{r qctypes}
#| fig.cap: !expr fig_cap("RunScatter restricted to batch QC (BQC) and study (SPL) samples.")
#| fig.alt: RunScatter plot showing BQC and SPL only
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  qc_types = c("BQC", "SPL"),
  rows_page = 1, cols_page = 1)
```

## 4. Batch display

Set `show_batches = TRUE` and `batch_zebra_stripe = TRUE` to highlight batch boundaries with
alternating shaded areas.

```{r batches}
#| fig.cap: !expr fig_cap("RunScatter with batch boundaries marked by alternating zebra stripes.")
#| fig.alt: RunScatter plot with batch zebra stripes
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  show_batches = TRUE,
  batch_zebra_stripe = TRUE,
  rows_page = 1, cols_page = 1)
```

## 5. Reference lines

Show mean ± k × SD reference lines with `show_reference_lines = TRUE`. Use `reference_sd_shade`
to display the range as a shaded band instead of lines.

```{r reflines}
#| fig.cap: !expr fig_cap("RunScatter with a mean ± 2 SD reference band computed from the BQC samples.")
#| fig.alt: RunScatter plot with mean ± 2 SD reference band
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  qc_types = c("BQC", "SPL"),
  show_reference_lines = TRUE,
  ref_qc_types = "BQC",
  reference_k_sd = 2,
  reference_sd_shade = TRUE,
  rows_page = 1, cols_page = 1)
```

## 6. Outlier capping

Use `cap_outliers = TRUE` to cap extreme values based on MAD fences. This is useful when outliers
obscure the trends of QC or study samples.

```{r capping}
#| fig.cap: !expr fig_cap("RunScatter with extreme values capped at MAD-based fences.")
#| fig.alt: RunScatter plot with outlier capping
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  cap_outliers = TRUE,
  cap_sample_k_mad = 3,
  rows_page = 1, cols_page = 1)
```

## 7. Log scale

Set `log_scale = TRUE` to apply a log10 transformation to the y-axis. Zero or negative values are
replaced with the minimum positive value divided by 5.

```{r logscale}
#| fig.cap: !expr fig_cap("RunScatter with a log10-transformed y-axis.")
#| fig.alt: RunScatter plot with log10 y-axis
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  log_scale = TRUE,
  rows_page = 1, cols_page = 1)
```

## 8. Removing gaps in the analysis sequence

Gaps in the x-axis can arise in two ways:

1. **Filtered QC types** — when only a subset of QC types is selected, the
   unselected positions leave gaps. Use `collapse_excluded = TRUE` to close
   them.
2. **Unannotated analyses** — when some runs in the analytical sequence have no
   corresponding entry in `@dataset` (e.g. solvent blanks or system
   suitability injections that were not imported), the analysis order contains
   discontinuities. Use `remove_gaps = TRUE` to collapse these and mark the
   former gap positions with vertical indicator lines.

Both options can be combined.

### Collapsing excluded QC types

```{r collapse}
#| fig.cap: !expr fig_cap("RunScatter of a single QC type (LTR) with the excluded positions collapsed.")
#| fig.alt: RunScatter plot with collapsed gaps from excluded QC types
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  qc_types = c("LTR"),
  collapse_excluded = TRUE,
  rows_page = 1, cols_page = 1)
```

### Removing gaps from unannotated analyses

To demonstrate this feature, we first simulate a dataset with unannotated analyses by excluding a
block of runs from batch 2. This creates a contiguous gap in the analysis order, as would occur
when solvent blanks or system suitability injections are not imported.

```{r simulate-gaps}
#| results: false
mexp_gaps <- exclude_analyses(
  mexp,
  analyses = c(
    paste0("Longit_batch2_", 1:45),
    "Longit_batch2_PQC 12",
    "Longit_batch2_PQC 13",
    "Longit_batch2_PQC 14",
    "Longit_batch2_PQC 15",
    "Longit_batch2_TQC13",
    "Longit_batch2_TQC14",
    "Longit_batch2_TQC15"
  ),
  clear_existing = TRUE)
mexp_gaps <- normalize_by_istd(mexp_gaps)
mexp_gaps <- calc_qc_metrics(mexp_gaps)
```

```{r gaps}
#| fig.cap: !expr fig_cap("RunScatter with vertical indicator lines marking the collapsed unannotated runs.")
#| fig.alt: RunScatter plot with gap indicator lines
#| results: false
plot_runscatter(
  mexp_gaps,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  remove_gaps = TRUE,
  rows_page = 1, cols_page = 1)
```

### Combining both

When filtering to a single QC type *and* the sequence contains unannotated
runs, combine `collapse_excluded` and `remove_gaps` to produce a compact,
fully annotated plot.

```{r gaps-combined}
#| fig.cap: !expr fig_cap("RunScatter combining collapsed excluded types with gap indicators.")
#| fig.alt: RunScatter plot with collapsed excluded types and gap indicators
#| results: false
plot_runscatter(
  mexp_gaps,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  qc_types = c("BQC"),
  collapse_excluded = TRUE,
  remove_gaps = TRUE,
  rows_page = 1, cols_page = 1)
```

## 9. Label wrapping

When feature names are long, strip labels can overflow. Set `label_wrap = TRUE` to wrap labels
across multiple lines, controlled by `label_wrap_width`. Here we show three features to illustrate
the effect on strip text.

```{r labelwrap}
#| fig.cap: !expr fig_cap("RunScatter of three features with wrapped strip labels.")
#| fig.alt: RunScatter plot with wrapped strip labels
#| results: false
plot_runscatter(
  mexp,
  variable = "intensity",
  include_feature_filter = "TG 48\\:2 \\[\\-18:1\\]",
  label_wrap = TRUE,
  label_wrap_width = 8,
  rows_page = 1, cols_page = 3, specific_page = 1)
```

## 10. Trend curves

Trend curves visualise the fitted signal used during drift or batch correction. They require a
prior correction step — `show_trend = TRUE` will error if no drift or batch correction has been
applied. To plot the values *before* the last correction, append `_before` to the variable name
(e.g. `variable = "intensity_before"`); for the original uncorrected values use `_raw`
(e.g. `variable = "intensity_raw"`).

See the [drift and batch correction](tutorial-04-drift-correction.html) tutorial for a worked example.

## 11. Principal component analysis (PCA)

PCA is a routine multivariate check in targeted MS workflows: it helps spot injection
outliers, visualise residual batch effects, and confirm that drift or batch corrections
reduced unwanted variance. The plots below reuse the `mexp` prepared in Setup.

### Score plot by QC type

The score plot summarises sample variance in two dimensions. Biological QCs (BQC) should
cluster tightly near the centre if normalisation and any corrections succeeded, while study
samples (SPL) typically show wider biological spread. The `ellipse_variable` controls the
grouping of the confidence ellipses (`qc_type`, `batch_id`, or `"none"`); a markedly
dispersed QC cluster suggests insufficient normalisation or remaining instrument drift.

```{r pca-score-qc}
#| fig.cap: !expr fig_cap("PCA score plot coloured by QC type, with confidence ellipses per group.")
#| fig.alt: PCA score plot coloured by QC type
#| results: false
plot_pca(
  mexp,
  variable = "norm_intensity",
  qc_types = c("BQC", "SPL"),
  ellipse_variable = "qc_type")
```

### Score plot by batch

If samples separate along PC1 or PC2 by `batch_id`, batch effects persist and a
centering-based correction is likely warranted. Running the same plot before and after
`correct_batch_centering()` confirms that the batch ellipses overlap after correction.

```{r pca-score-batch}
#| fig.cap: !expr fig_cap("PCA score plot coloured by batch.")
#| fig.alt: PCA score plot coloured by batch
#| results: false
plot_pca(
  mexp,
  variable = "norm_intensity",
  qc_types = c("BQC", "SPL"),
  ellipse_variable = "batch_id")
```

### Loadings

The loading plot identifies the features driving the principal components; features at the
extremes contribute most to sample separation. A single feature dominating PC1 should be
inspected — it is often a saturated or contaminated transition rather than a biological
signal.

```{r pca-loadings}
#| fig.cap: !expr fig_cap("PCA loadings identifying the features driving each component.")
#| fig.alt: PCA loadings plot
#| results: false
plot_pca_loading(
  mexp,
  variable = "norm_intensity",
  qc_types = c("BQC", "SPL"))
```

### Detecting outlier injections

`detect_outlier_pca()` flags analyses whose score on a chosen principal component lies
outside a fence, defined as either `mean ± k·SD` (`outlier_detection = "sd"`) or
`median ± k·MAD` (`outlier_detection = "mad"`), with `fence_multiplicator` setting *k*. It
returns the `analysis_id` values that exceed the fence on the selected component, or `NULL`
if none are flagged.

```{r pca-outliers}
#| results: markup
outlier_ids <- detect_outlier_pca(
  mexp,
  variable = "intensity",
  filter_data = FALSE,
  pca_component = 1,
  qc_types = c("BQC", "SPL"),
  outlier_detection = "mad",
  fence_multiplicator = 4)

outlier_ids
```

<details>
<summary><strong>Choice of fence method</strong></summary>

The SD-based fence is sensitive to the very outliers it tries to detect: a single extreme
observation inflates the SD estimate. The MAD-based fence is robust to a few large
deviations and is generally preferable for QC screening. A typical multiplier is
`fence_multiplicator = 3` (≈99.7% under normality) or `4` for a more permissive screen.
The function evaluates one principal component at a time — re-run with `pca_component = 2`
to screen PC2.
</details>

A sample flagged by PCA is a candidate for investigation, not a verdict. Outlier patterns
frequently reflect genuine biology (e.g. a disease group or a sex difference), so an
injection should be excluded only where a documented technical cause is identified — a
failed injection, contamination, instrument fault, or sample-handling error.

### Excluding analyses and features

After visual and documented confirmation, remove the offending injections or features.
`exclude_analyses()` and `exclude_features()` flip the affected rows' `valid_analysis` /
`valid_feature` flag so downstream steps skip them, while the original data remain in
`mexp@dataset_orig`; setting `clear_existing = TRUE` replaces any previous exclusion list
rather than appending.

```{r exclude}
#| eval: false
# Review candidates before excluding
mexp@annot_analyses |>
  dplyr::filter(analysis_id %in% outlier_ids) |>
  dplyr::select(analysis_id, qc_type, batch_id, analysis_order)

# Exclude after technical confirmation
mexp <- exclude_analyses(mexp, analyses = outlier_ids, clear_existing = FALSE)

# Exclude a known-problematic feature (e.g. a saturated transition)
mexp <- exclude_features(mexp, features = c("PC 32:0"), clear_existing = FALSE)
```

### Interpretation guide

| PCA pattern | Likely cause | Action |
|-------------|--------------|--------|
| BQC samples dispersed across the score plot | Poor precision; ISTD assignment problem | Inspect `plot_normalization_qc()`; verify ISTD pairing |
| Clear separation by `batch_id` on PC1 or PC2 | Uncorrected batch effect | Apply `correct_batch_centering()` |
| Single injection isolated from all groups | Technical outlier | Investigate cause; exclude only with documented reason |
| One feature dominates loadings on PC1 | Saturation, contamination, or single-transition artefact | Inspect with `plot_runscatter()`; exclude feature if confirmed |
| BQC tight, SPL spread | Genuine biological variability | Proceed |

## Next steps

- [Drift and batch correction](tutorial-04-drift-correction.html) — apply and inspect corrections
- [Visualisation functions](manual-08-visualization.html) — the full list of plotting functions
- [Customising plots](manual-13-plot-customization.html) — sizing text, legends and points without `+ theme()`
- [Basic MRMhub workflow](tutorial-02-basic-workflow.html) — where these checks fit in the pipeline
