---
title: "Your first analysis"
description: >
  A short walkthrough on the bundled demo data: import a result file, normalize
  by internal standard, inspect the run, and export: no external files needed.
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 3
vignette: >
  %\VignetteIndexEntry{Your first analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
source("_common.R")
```

<p class="page-meta"><span class="page-kind page-kind-tutorial">Tutorial</span> <span class="page-level">Beginner</span> <span class="page-prereq">Prerequisites: <a href="manual-00-installation.html">MRMhub installed</a></span></p>

This walkthrough takes an MRMhub result from import to a normalized, exportable
dataset in a handful of lines, using demo data bundled with the package so
nothing external is needed. It is the shortest path through the pipeline; the
workflow tutorials that follow unpack each step in full.

## 1. Run the complete workflow

The block below is the whole analysis: import the demo result, normalize each
feature by its internal standard, and print the resulting object. Run it and you
have a processed dataset.

```{r pipeline}
#| message: true
library(mrmhub)

demo_file <- system.file("extdata", "MRMhub_demo.tsv", package = "mrmhub")

mexp <- MRMhubExperiment()
mexp <- import_data_mrmhub(mexp, path = demo_file, import_metadata = TRUE)
mexp <- normalize_by_istd(mexp)
mexp
```

We then look at the normalized signal across the analytical run, the first thing
to inspect for run-order drift.

```{r runscatter}
#| fig.height: 4
#| fig.cap: !expr fig_cap("Internal-standard–normalized signal of each feature plotted against injection order.")
#| fig.alt: RunScatter of ISTD-normalized feature intensities against run order
plot_runscatter(mexp, variable = "norm_intensity")
```

Printing the object returns a compact status overview: `mrmhub_status(mexp)` prints the
full report. If you prefer not to write code, `build_workflow()` opens an
interactive application that validates your data and metadata, flags pipeline
mismatches, and generates an equivalent Quarto workflow to download (see
[Build a workflow without code](tutorial-12-workflow-builder.html)); the script
above remains the reproducible path.

## 2. What the script does

The importer returns an `MRMhubExperiment`: a structured container holding the
peak-area data, the sample annotations, and the feature metadata together.
`normalize_by_istd()` divides each feature by its assigned internal standard,
correcting for extraction and injection variability. The ratio is stored in
`feature_norm_intensity`, the `is_istd_normalized` flag is set, and the raw
import is left untouched in the `dataset_orig` slot. A few helpers report what
was imported:

```{r inspect}
#| eval: false
get_analysis_count(mexp)   # number of analyses
get_feature_count(mexp)    # number of features
get_featurelist(mexp)      # the measured features
```

## 3. Export the results

Write the processed data to a multi-sheet Excel report, or to a single tidy CSV. Figures are saved
the same way, at a size you specify (in mm by default):

```{r export}
#| eval: false
save_report_xlsx(mexp, path = "my_first_results.xlsx")
save_dataset_csv(mexp, path = "my_first_results.csv")

save_plot(
  plot_pca(mexp, variable = "norm_intensity"),
  path = "my_first_pca.pdf", width = 180, height = 120
)
```

## Next steps

- [Key concepts and glossary](manual-01-key-concepts.html): understand the data model
- [Preparing and importing data](tutorial-01-prep-data.html): import your own data and metadata
- [Basic MRMhub workflow](tutorial-02-basic-workflow.html): the full end-to-end pipeline
- [Lipidomics data processing](tutorial-03-lipidomics-workflow.html): a detailed real-world workflow
- [Visualisation functions](manual-08-visualization.html): the QC plots, and saving figures at a defined size
- [Customising plots](manual-13-plot-customization.html): size text, legends and points for reports
- [Getting help from an AI assistant](manual-09-ai-assistants.html): use Claude or ChatGPT to help write and troubleshoot your workflow, accurately and safely
